### 2016-09-23 mixed-model.Rmd
preliminary look at mixed model & residuals

---
title: "R Notebook"
output: html_notebook
---

```{r, message = FALSE, warning = FALSE, results = "hide"}
devtools::load_all()
library(tidyverse)
library(ggplot2)
start_yr = 1982
end_yr = 2013
min_num_yrs = 25
last_train_year = 2003
d = get_richness_ts_env_data(start_yr, end_yr, min_num_yrs) %>% 
  filter(!is.na(observer_id))
train = d[d$year <= last_train_year, ]
test = d[d$year > last_train_year, ]
```

Partition the variance among site effects, observer effects, annual effects,
and residuals.

```{r}
library(lme4)
model = lmer(richness ~ (1|site_id) + (1|observer_id), data = train)
summary(model)
```

Predict forward:

```{r}
predictions = test %>%
  mutate(predicted = predict(model, test, allow.new.levels = TRUE)) %>% 
  mutate(diff = richness - predicted)

predictions %>% 
  group_by(year) %>% 
  summarize(mean(diff^2)) %>% 
  plot(ylab = "Mean Squared Errorr", type = "l", ylim = c(0, 50))

predictions %>% 
  filter(observer_id %in% train$observer_id) %>% 
  group_by(year) %>% 
  summarize(mean(diff^2)) %>% 
  lines(col = 2)
legend("topleft", 
       legend = c("all observations", "known observers only", "training observations"), 
       lty = c(1, 1, 2), col = c(1, 2, 1))

abline(h = var(resid(model)), lty = 2)
```

## Can we do better?

None of the predictor variables correlate with the residuals, suggesting that 
their explanatory effect is mainly *among* sites, rather than in the residuals
for individual sites.

```{r}
stopifnot(nrow(train) == length(resid(model)))
round(sort(cor(resid(model), train)[1, colnames(train) != "richness"]^2), 4)
```

There also isn't much in the way of consistent autocorrelation in the residuals
(correlation of 0.2 explains 4% of the residual variance [100 * 0.2^2], which 
isn't tremendously useful even if it is statistically significant---especially
since the residual variance is only a small portion of the total variance anyway).

```{r}
resid_matrix = data.frame(site_id = train$site_id, 
                          year = train$year, 
                          residuals = resid(model)) %>% 
  complete(site_id, year) %>% 
  arrange(site_id, year) %>% 
  magrittr::extract2("residuals") %>% 
  matrix(nrow = n_distinct(train$year))

autocorr_matrix = resid_matrix %>% 
  t() %>% 
  cor(use = "pairwise.complete.obs")

plot_upper = function(x, title){
  reshape2::melt(x * upper.tri(x)) %>% 
    ggplot(aes(x = Var1, y = Var2, fill = value)) + 
    geom_tile() +
    scale_fill_gradient2() +
    ggtitle(title)
}

plot_upper(autocorr_matrix, "autocorrelation")

summary(lm(resid_matrix[4, ] ~ t(resid_matrix[1:3, ])))

```

### 2016-09-26 mixed-model.Rmd
play with results

@@
 
-predictions %>% 
-  group_by(year) %>% 
-  summarize(mean(diff^2)) %>% 
-  plot(ylab = "Mean Squared Errorr", type = "l", ylim = c(0, 50))
+predictions = predictions %>% 
+  mutate(observer_count = c(table(train$observer_id)[as.character(observer_id)])) %>% 
+  mutate(observer_count = ifelse(is.na(observer_count), 0, observer_count)) %>% 
+  mutate(observer_class = cut(observer_count, c(-Inf,0, 2^(0:7), Inf)))
 
 predictions %>% 
-  filter(observer_id %in% train$observer_id) %>% 
-  group_by(year) %>% 
-  summarize(mean(diff^2)) %>% 
-  lines(col = 2)
-legend("topleft", 
-       legend = c("all observations", "known observers only", "training observations"), 
-       lty = c(1, 1, 2), col = c(1, 2, 1))
+  group_by(year, observer_class) %>% 
+  summarize(rmse = sqrt(mean((richness - predicted)^2))) %>% 
+  ggplot(aes(x = year, y = rmse, color = observer_class)) + 
+  geom_smooth(se = FALSE, span = 2)
 
-abline(h = var(resid(model)), lty = 2)
+#predictions %>% 
+#  filter(observer_id %in% train$observer_id) %>% 
+#  group_by(year) %>% 
+#  summarize(mean(diff^2)) %>% 
+#  lines(col = 2)
+#legend("topleft", 
+#       legend = c("all observations", "known observers only", "training observations"), 
+#       lty = c(1, 1, 2), col = c(1, 2, 1))
+
+#abline(h = var(resid(model)), lty = 2)
 ```
@@
 ```
+```{r}
+sapply(1:ncol(autocorr_matrix),
+       function(i){
+         mean(autocorr_matrix[row(autocorr_matrix) == (col(autocorr_matrix) - i)])
+       }
+)
+```
+
+```{r}
+
+```

### 2016-09-26 sdm.R
simple SDM fitting

@@ -1,0 +1,82 @@
+devtools::load_all()
+library(parallel)
+library(tidyverse)
+library(gbm)
+
+start_yr = 1982
+end_yr = 2013
+min_num_yrs = 25
+last_train_year = 2003
+
+# GBM parameters
+first_validation_year = 2000 # for temporal cross-validation
+interaction.depth = 8
+max_n_trees = 2E4
+my_formula = present ~ start_time + bio2 + bio5 + bio15 + ndvi_sum + elevs +
+  observer_effect + site_effect
+
+# Munge -------------------------------------------------------------------
+d = get_pop_ts_env_data(start_yr, end_yr, min_num_yrs) %>% 
+  add_ranefs(last_training_year = last_train_year)
+
+train = d[d$year <= last_train_year, ]
+test = d[d$year > last_train_year, ]
+
+# Training data must be arranged by year first for temporal cross-validation
+# to work!
+train_x = select(train, -species_id, -abundance) %>% 
+  distinct() %>% 
+  filter(!is.na(lat)) %>% 
+  distinct() %>% 
+  arrange(year, site_id)
+
+test_x = select(test, -species_id, -abundance) %>% 
+  distinct() %>% 
+  filter(!is.na(lat)) %>% 
+  distinct() %>% 
+  arrange(year, site_id)
+
+species_ids = unique(train$species_id)
+
+
+# Fitting/predicting ------------------------------------------------------
+fit_species = function(species_id){
+  xy = filter(train, species_id == species_id) %>% 
+    select(site_id, year) %>% 
+    mutate(present = 1) %>% 
+    right_join(train_x, c("site_id", "year")) %>% 
+    arrange(year, site_id) %>% 
+    mutate(present = ifelse(is.na(present), 0, 1))
+  
+  # Determine the number of trees to include by using the first portion of
+  # the training set to predict the remainder of the training set.
+  g = gbm(
+    my_formula,
+    data = xy,
+    distribution = "bernoulli",
+    interaction.depth = interaction.depth,
+    train.fraction = mean(xy$year < first_validation_year),
+    n.trees = max_n_trees
+  )
+  n.trees = gbm.perf(g)
+  
+  # Fit the full time series with the same number of trees
+  g_full = gbm(
+    my_formula,
+    data = xy,
+    distribution = "bernoulli",
+    interaction.depth = interaction.depth,
+    train.fraction = 1,
+    n.trees = n.trees
+  )
+  
[…]

### 2016-09-26 sdm.R
print status while fitting

@@ -41,6 +41,8 @@ species_ids = unique(train$species_id)
 
 # Fitting/predicting ------------------------------------------------------
 fit_species = function(species_id){
+  cat(strsplit(date(), " ")[[1]][4], " fitting species", species_id, "\n")
+  
   xy = filter(train, species_id == species_id) %>% 
     select(site_id, year) %>% 
     mutate(present = 1) %>% 

### 2016-09-26 R/forecast-bbs-core.R
save observers (sqlite workaround)

@@ -293,20 +293,26 @@ add_env_data <- function(bbs_data){
 #' Add observer ID's to data
 #' @param bbs_data dataframe that contains BBS site_id and year columns
 #' @importFrom dplyr %>% left_join
-add_observers=function(bbs_data){
-  observer_query='SELECT
-  bbs_weather.obsn as observer_id,
-  year,
-  (statenum*1000)+route as site_id
-  FROM
-  bbs_weather
-  WHERE
-  runtype=1 AND rpid=101'
-  
-  observer_info=db_engine(action = 'read', sql_query = observer_query)
-  
+add_observers = function(bbs_data) {
+  data_path = "data/bbs_observers.csv"
+  if (file.exists(data_path)) {
+    observer_info = read_csv(data_path)
+  } else {
+    observer_query = '
+    SELECT
+      bbs_weather.obsn as observer_id,
+      year,
+      (statenum*1000)+route as site_id
+    FROM
+      bbs_weather
+    WHERE
+      runtype=1 AND rpid=101'
+    
+    observer_info = db_engine(action = 'read', sql_query = observer_query)
+    write.csv(observer_info, file = data_path, row.names = FALSE, quote = FALSE)
+  }  
   bbs_data %>%
-    left_join(observer_info, by=c('site_id','year'))
+    left_join(observer_info, by = c('site_id','year'))
 }
 
 

### 2016-09-26 sdm.R
shebangs seem like a good idea

@@ -1,3 +1,5 @@
+#!/usr/bin/Rscript
+
 devtools::load_all()
 library(parallel)
 library(tidyverse)

### 2016-09-26 sdm.R
redirect shebang

@@ -1,4 +1,4 @@
-#!/usr/bin/Rscript
+#!/apps/R/3.2.0/bin/Rscript
 
 devtools::load_all()
 library(parallel)

### 2016-09-26 sdm.R
only use 8 cores

@@ -78,7 +78,7 @@ fit_species = function(species_id){
 }
 
 predictions = mclapply(species_ids, fit_species, 
-                       mc.cores = detectCores(logical = FALSE))
+                       mc.cores = 8)
 
 
 # Save results ------------------------------------------------------------

### 2016-09-26 sdm.R
avoid nonstandard evaluation problem

@@ -42,10 +42,10 @@ species_ids = unique(train$species_id)
 
 
 # Fitting/predicting ------------------------------------------------------
-fit_species = function(species_id){
-  cat(strsplit(date(), " ")[[1]][4], " fitting species", species_id, "\n")
+fit_species = function(sp_id){
+  cat(strsplit(date(), " ")[[1]][4], " fitting species", sp_id, "\n")
   
-  xy = filter(train, species_id == species_id) %>% 
+  xy = filter(train, species_id == sp_id) %>% 
     select(site_id, year) %>% 
     mutate(present = 1) %>% 
     right_join(train_x, c("site_id", "year")) %>% 
@@ -84,3 +84,5 @@ predictions = mclapply(species_ids, fit_species,
 # Save results ------------------------------------------------------------
 
 saveRDS(predictions, file = "gbm_predictions.rds")
+
+

### 2016-09-26 sdm.R
avoid problems with 100% or 0% presence

@@ -52,6 +52,12 @@ fit_species = function(sp_id){
     arrange(year, site_id) %>% 
     mutate(present = ifelse(is.na(present), 0, 1))
   
+  # avoid problems with 100% or 0% presence
+  if (var(xy$present[xy$year < first_validation_year]) == 0) {
+    return(NA)
+  }
+  
+  
   # Determine the number of trees to include by using the first portion of
   # the training set to predict the remainder of the training set.
   g = gbm(

### 2016-09-27 sdm.R
don't preschedule parallel models

@@ -84,7 +84,8 @@ fit_species = function(sp_id){
 }
 
 predictions = mclapply(species_ids, fit_species, 
-                       mc.cores = 8)
+                       mc.cores = 8,
+                       mc.preschedule = FALSE)
 
 
 # Save results ------------------------------------------------------------

### 2016-09-27 sdm.R
save SDM results as we go

@@ -5,6 +5,8 @@ library(parallel)
 library(tidyverse)
 library(gbm)
 
+dir.create("sdm_predictions", showWarnings = FALSE)
+
 start_yr = 1982
 end_yr = 2013
 min_num_yrs = 25
@@ -80,16 +82,14 @@ fit_species = function(sp_id){
     n.trees = n.trees
   )
   
-  predict(g_full, test_x, n.trees = n.trees, type = "response")
+  p = predict(g_full, test_x, n.trees = n.trees, type = "response")
+  
+  write.table(p, file = paste0("sdm_predictions/", sp_id, ".txt"),
+              row.names = FALSE, col.names = FALSE)
+  
+  NULL
 }
 
 predictions = mclapply(species_ids, fit_species, 
                        mc.cores = 8,
                        mc.preschedule = FALSE)
-
-
-# Save results ------------------------------------------------------------
-
-saveRDS(predictions, file = "gbm_predictions.rds")
-
-

### 2016-09-27 analysis/make_shear_map.R
remove dead folders; shear map in new 'analysis' folder

@@ -1,0 +1,40 @@
+library(ggplot2)
+library(viridis)
+library(maps)
+
+# Assumes "results" is a data frame with long, lat, and p (which determines
+# intensity and is between 0 and 1).
+# Note that any points with p==0 will not be plotted
+# Shear amount is how much to move the top to the right (so negative numbers
+# move to the left)
+make_shear_map = function(results, shear_amount = -1, map_color = "#FEFFEF"){
+  shear = matrix(c(1, shear_amount, 0, 1), nrow = 2)
+  
+  results[ , c("x", "y")] = as.matrix(results[,c("long", "lat")]) %*% shear
+  
+  # Sheared version of the US state map
+  shear_borders = function (database = "world", regions = ".", fill = NA, colour = "grey50",
+                            xlim = NULL, ylim = NULL, ...) {
+    df <- map_data(database, regions, xlim = xlim, ylim = ylim)
+    df[ , c("long", "lat")] = as.matrix(df[ , c("long", "lat")]) %*% shear
+    geom_polygon(aes_(~long, ~lat, group = ~group), data = df,
+                 fill = fill, colour = colour, ..., inherit.aes = FALSE)
+  }
+  
+  # Define the parallelogram that will represent the map area
+  xlim = range(map_data("state")$long)
+  ylim = range(map_data("state")$lat)
+  corner_matrix = cbind(x = rep(xlim, each = 2), y = c(ylim, rev(ylim)))
+  corners = corners = structure(as.data.frame(corner_matrix %*% shear), names = c("x", "y"))
+  
+  results$alpha = ifelse(results$p == 0, 0, 1)
+  
+  ggplot(results, aes(x = x, y = y, color = p)) +
+    geom_polygon(aes_(~x, ~y), data = corners, inherit.aes = FALSE, color = "black", fill = "white") +
+    shear_borders("state", fill = map_color) +
+    geom_point(size = 2, alpha = results$alpha) +
+    scale_color_viridis(guide = FALSE, option = "magma", limits = c(0, 1),
+                        direction = -1) +
+    coord_equal() +
+    theme_void()
+}

### 2016-09-27 sdm.R
drop time variable

@@ -16,7 +16,7 @@ last_train_year = 2003
 first_validation_year = 2000 # for temporal cross-validation
 interaction.depth = 8
 max_n_trees = 2E4
-my_formula = present ~ start_time + bio2 + bio5 + bio15 + ndvi_sum + elevs +
+my_formula = present ~ bio2 + bio5 + bio15 + ndvi_sum + elevs +
   observer_effect + site_effect
 
 # Munge -------------------------------------------------------------------

### 2016-09-27 manuscript/draft.Rmd
manuscript outline

# Title

Authors: Harris, Taylor and White?

## Introduction

Quality of predictions from SDMs is important:
  * helps determine policy/set priorities
  * something something climate change
  * need to know both accuracy of point estimates and how reasonable our prediction intervals are.
  * Overconfident predictions can be worse than uncertainty
SDM assumptions:
  * we've measured all the important variables (sites don't have any other important properties)
  * Something like stationarity:
    * e.g. "systems are at equilibrium now and will be at equilibrium in future"
    * or "species track predictor variables quickly" (no lags)
    * etc
  * independent/identically distributed observations across species, years
  * Space substitutes well for time
Difficult to test SDM assumptions/predictions because of time scales
  * Most scientists making predictions today will have retired by 2050
  * Use long-term data sets such as BBS to evaluate predictive skill
    * Note the large change in average temperatures since the 80's; comparable to the changes expected in the next N years
  * Test with simple baselines (e.g. random walk, future-is-like-present)
  * Focus on richness because... reasons.

## Methods

**Data**

* BBS + retriever
* Monthly climate data from interpolated weather stations
* NDVI
* Altitude

**Models**

* Mixed model for observer effort \& site effects
* Naive AR1 model (adjusted for observer effects)
* species-level GBM w/ mixed model point estimates as bonus predictors where available
* richness-level GBM w/ mixed model point estimates as bonus predictors where available

(need a side note about why we're not using a proper count distribution; underdispersion)

**Evaluation**

* (R)MSE
* Log-likelihood (simultaneously rewards good point estimates, precision, and coverage)
* Coverage at 95%
* Calibration/bias
* Evaluated each year after the last training year
* Dig around in the residuals for insights

## Results and conclusions

* Temporal richness variability doesn't track the predictors we use; sites have "inertia"
* More important to use a reasonable error model than to get the best predictors \& basis expansions
* We've posted predictions for the next N years of BBS data, using several methods
* In many ways, this is the best-case scenario for SDMs:
  * large data set, known \& consistent sampling technique over a period of decades
  * True absences
  * extended predictor set
  * birds arguably migrate quickly enough to make equlibrium \& especially space-for-time assumptions less ridiculous
  * errors for individual species can cancel out; only asking for aggregate values
* Value of stacking versus predicting the quantitity of interest directly
*

### 2016-09-27 manuscript/draft.Rmd
minor manuscript changes

@@
   * Overconfident predictions can be worse than uncertainty
+  * Quality of prediction is one metric for how much we know about ecological processes
 SDM assumptions:
@@
 * More important to use a reasonable error model than to get the best predictors \& basis expansions
-* We've posted predictions for the next N years of BBS data, using several methods
+* We've posted predictions for the next N years of BBS data, using several methods, and invite others to do so as well.
 * In many ways, this is the best-case scenario for SDMs:

### 2016-09-29 R/make_shear_map.R
move shear to ./R/

@@ -1,12 +1,15 @@
-library(ggplot2)
-library(viridis)
-library(maps)
-
-# Assumes "results" is a data frame with long, lat, and p (which determines
-# intensity and is between 0 and 1).
-# Note that any points with p==0 will not be plotted
-# Shear amount is how much to move the top to the right (so negative numbers
-# move to the left)
+#' @param results a data frame with long, lat, and p (which color determines 
+#' intensity and is between 0 and 1). Note that any points with p==0 will not 
+#' be plotted
+#' @param shear_amout how much to move the top to the right (so negative numbers
+#' move to the left)
+#' @param map_color color for landmass in the background
+#' @export
+#' @importFrom ggplot2 geom_polygon geom_point
+#' @importFrom ggplot2 ggplot aes aes_
+#' @importFrom ggplot2 map_data
+#' @importFrom ggplot2 theme_void coord_equal
+#' @importFrom viridis scale_color_viridis
 make_shear_map = function(results, shear_amount = -1, map_color = "#FEFFEF"){
   shear = matrix(c(1, shear_amount, 0, 1), nrow = 2)
   
@@ -38,3 +41,4 @@ make_shear_map = function(results, shear_amount = -1, map_color = "#FEFFEF"){
     coord_equal() +
     theme_void()
 }
+

### 2016-09-30 manuscript/draft.Rmd

@@
 
-Quality of predictions from SDMs is important:
+Quality of predictions is important:
   * helps determine policy/set priorities
@@
   * Quality of prediction is one metric for how much we know about ecological processes
-SDM assumptions:
-  * we've measured all the important variables (sites don't have any other important properties)
-  * Something like stationarity:
-    * e.g. "systems are at equilibrium now and will be at equilibrium in future"
-    * or "species track predictor variables quickly" (no lags)
-    * etc
-  * independent/identically distributed observations across species, years
-  * Space substitutes well for time
-Difficult to test SDM assumptions/predictions because of time scales
+Dominant method is stacked SDMs:
+  *
+Difficult to test model assumptions/predictions because of time scales
   * Most scientists making predictions today will have retired by 2050
@@
   * Test with simple baselines (e.g. random walk, future-is-like-present)
-  * Focus on richness because... reasons.
+  * Focus on richness/diversity because \[citations saying that it's important\]
 
@@
 * BBS + retriever
-* Monthly climate data from interpolated weather stations
-* NDVI
-* Altitude
+* Monthly climate data from interpolated weather stations (PRISM)
+* NDVI (GIMMS)
+* Altitude (SRTM 90m Digital Elevation Database via raster::getData)
 
@@
 * Value of stacking versus predicting the quantitity of interest directly
-*
+
+## misc
+
+SDM assumptions:
+  * we've measured all the important variables (sites don't have any other important properties)
+  * Something like stationarity:
+    * e.g. "systems are at equilibrium now and will be at equilibrium in future"
+    * or "species track predictor variables quickly" (no lags)
+    * etc
+  * independent/identically distributed observations across species, years
+  * Space substitutes well for time

### 2016-10-03 manuscript/draft.Rmd
some SDM stuff

@@
 Dominant method is stacked SDMs:
-  *
+  * Strong assumptions (see "misc" below), many *interesting* predictions
+  * Potential problems with confidence intervals
+  * Potential disconnect between training target and actual goals
 Difficult to test model assumptions/predictions because of time scales

### 2016-10-03 manuscript/draft.Rmd

--- 2016-09-30 manuscript/draft.Rmd
@@
   * Quality of prediction is one metric for how much we know about ecological processes
-Dominant method is stacked SDMs:
-  *
-Difficult to test model assumptions/predictions because of time scales
+Unclear how much to trust existing predictions:
+  * Strong assumptions (see "misc" below)
+  * many *interesting* predictions
+  * Potential problems with confidence intervals
+  * Potential disconnect between training target and actual goals
+  * Difficult to test model assumptions/predictions because of time scales
     * Most scientists making predictions today will have retired by 2050
+Demonstrate best practices \& their contribution to better predictions
   * Use long-term data sets such as BBS to evaluate predictive skill
   * Note the large change in average temperatures since the 80's; comparable to the changes expected in the next N years
+  * Lots of talk about how we should do it, fewer examples of doing them all together
+  * Examples of best practices
+    * not just bioclim
+    * time series
+    * spatial/site-level stuff
+    * probabilistic/error bars
+    * multiple models
     * Test with simple baselines (e.g. random walk, future-is-like-present)
   * Focus on richness/diversity because \[citations saying that it's important\]
+  * Publish forecasts at different timescales to evaluate performance/credibility
+Here we do that:
+  * forecasts for biodiversity in NAmerican birds with as many best practices as possible
+  * Discuss the influence of these improved methods \& of our error rates on our ability to understand:
+    * How to improve
+    * How much to trust the predictions
 
@@
 * richness-level GBM w/ mixed model point estimates as bonus predictors where available
+* JSDM
 
-(need a side note about why we're not using a proper count distribution; underdispersion)
+(need a side note about why we're not using a proper count distribution; underdispersion, easier access to out-of-the-box time series methods)
 
@@
   * errors for individual species can cancel out; only asking for aggregate values
+* None of the models actually work all that well, compared to baseline.
 * Value of stacking versus predicting the quantitity of interest directly
+* Maybe richness is too coarse
+  * abundance
+  * species-level
 

### 2016-10-05 manuscript/draft.Rmd

@@
 * JSDM
+* Ensembles (weighted and unweighted)
 
@@
 
+* None of the models actually work all that well, compared to baseline.
 * Temporal richness variability doesn't track the predictors we use; sites have "inertia"
+  * Cite 3 papers Ethan found on consistancy of alpha even when composition changes dramatically
+  * Good for absolute accuracy, little room for improvement over baseline
+  * Maybe richness is too coarse
+    * abundance
+    * species-level
 * More important to use a reasonable error model than to get the best predictors \& basis expansions
-* We've posted predictions for the next N years of BBS data, using several methods, and invite others to do so as well.
+* Ensembles help/don't help (depending on results)
+* Point estimate error versus coverage
 * In many ways, this is the best-case scenario for SDMs:
   * large data set, known \& consistent sampling technique over a period of decades
-  * True absences
+  * "True" absences
   * extended predictor set
@@
   * errors for individual species can cancel out; only asking for aggregate values
-* None of the models actually work all that well, compared to baseline.
 * Value of stacking versus predicting the quantitity of interest directly
-* Maybe richness is too coarse
-  * abundance
-  * species-level
+* Need to start doing this, even if we're bad at it.
+  * We've posted predictions for the next N years of BBS data, using several methods, and invite others to do so as well.
 

### 2016-10-13 manuscript/draft.Rmd
add thoughts on climate and variables

@@
   * Use long-term data sets such as BBS to evaluate predictive skill
-  * Note the large change in average temperatures since the 80's; comparable to the changes expected in the next N years
+  * Note the large change in average temperatures since the 80's; comparable to the changes expected in the next N years (change since 80's is a true trend but whether it continues in uncertain, 10.1007/978-3-319-03768-4_2, 10.1007/s00382-014-2070-3)
+, 
   * Lots of talk about how we should do it, fewer examples of doing them all together
@@
 SDM assumptions:
-  * we've measured all the important variables (sites don't have any other important properties)
+  * we've measured the predictive power of all avialable variables, other potentially important thing that aren't available are landcover, observer skill level, (probably some others)
   * Something like stationarity:
@@
   * Space substitutes well for time
+
+None of the models are true forcasts as they use observed predictor variables (except the AR1).

### 2016-10-13 manuscript/important_references.Rmd
move citation clutter to its own file

#Various factoids
---------------------
On North American climate
There has been an increase in mean temperature in North America since the 1980's (pg. 90, fig 2.7) 
Redmond, Kelly T., and John T. Abatzoglou. "Current climate and recent trends." Climate Change in North America. Springer International Publishing, 2014. 53-94.

Predictions for the next 10-20 years are uncertain (Meehl et al 2014) but models currently suggest a continued trend in warming, with potential for short term decreases (natural variability around increasing trend)(fig 3 & 4, Kumar et al 2014)
Kumar, Devashish, Evan Kodra, and Auroop R. Ganguly. "Regional and seasonal intercomparison of CMIP3 and CMIP5 climate model ensembles for temperature and precipitation." Climate dynamics 43.9-10 (2014): 2491-2518.
Meehl, Gerald A., et al. "Decadal climate prediction: an update from the trenches." Bulletin of the American Meteorological Society 95.2 (2014): 243-267.

#Forcasting
People should do a better job at forecasting.
Evans, M. R., Norris, K. J. & Benton, T. G. Predictive ecology: systems approaches. Philos. Trans. R. Soc. B Biol. Sci. 367, 163–169 (2012).
Houlahan, J. E. The priority of prediction in ecological understanding. Oikos (2016). doi:10.1111/oik.03726
Petchey, O. L. et al. The ecological forecast horizon, and examples of its uses and determinants. Ecol. Lett. 18, 597–611 (2015).

#SDM references

GBM's are one  of the best models for SDM's (Elith et al 2006) but JSDMs could potentially be better (Warton et al 2015)

Elith, J. et al. Novel methods improve prediction of species’ distributions from occurrence data. Ecography (Cop.). 29, 129–151 (2006).
Warton, David I., et al. "So many variables: joint modeling in community ecology." Trends in ecology & evolution 30.12 (2015): 766-779.

### 2016-10-13 manuscript/important_references.Rmd
formatting
one more quick one

@@
 #Various factoids
----------------------
+
 On North American climate
@@
 Meehl, Gerald A., et al. "Decadal climate prediction: an update from the trenches." Bulletin of the American Meteorological Society 95.2 (2014): 243-267.  
+-------------------------------
+Biodiversity is an important thing to consider 
+Cardinale, Bradley J., et al. "Biodiversity loss and its impact on humanity." Nature 486.7401 (2012): 59-67.
 

### 2016-10-13 manuscript/important_references.Rmd
formating
formatting

@@
 
-On North American climate
+####On North American climate  
 There has been an increase in mean temperature in North America since the 1980's (pg. 90, fig 2.7)   
@@
 Meehl, Gerald A., et al. "Decadal climate prediction: an update from the trenches." Bulletin of the American Meteorological Society 95.2 (2014): 243-267.
+
 -------------------------------
-Biodiversity is an important thing to consider 
+####Biodiversity is an important thing to consider   
 Cardinale, Bradley J., et al. "Biodiversity loss and its impact on humanity." Nature 486.7401 (2012): 59-67.
@@
 #Forcasting
-People should do a better job at forecasting.  
+####People should do a better job at forecasting.  
 Evans, M. R., Norris, K. J. & Benton, T. G. Predictive ecology: systems approaches. Philos. Trans. R. Soc. B Biol. Sci. 367, 163–169 (2012).  

### 2016-10-13 R/forecast-bbs-core.R
use species instead of _common_name columns for filtering unidentified species
the two common name columns have a lot of variety in them, the species column in the species table is more specific in identified species vs. hybrids vs. subspecies.

@@ -60,7 +60,8 @@ db_engine=function(action, db='./data/bbsforecasting.sqlite', sql_query=NULL,
 #' Removes waterbirds, shorebirds, owls, kingfishers, knightjars,
 #' dippers. These species are poorly sampled due to their aquatic or
 #' noctural nature. Also removes taxa that were either partially unidentified
-#' (e.g. "sp.") or were considered hybrids (e.g. "A x B").
+#' (e.g. "sp.") or were considered hybrids (e.g. "A x B") or were listed as more
+#' than one species (e.g. "B / B")
 #'
 #' @param df dataframe containing an species_id column
 #'
@@ -70,12 +71,13 @@ filter_species <- function(df){
   species_table = get_species_data()
 
   is_unidentified = function(names) {
-    grepl("/|unid\\.|sp\\.| or |hybrid| X | x ", names)
+    #Befor filter account for this one hybrid of 2 subspecies so it's kept
+    names[names=='auratus auratus x auratus cafer']='auratus auratus'
+    grepl('sp\\.| x |\\/', names)
   }
 
   valid_taxa = species_table %>%
-    filter(!is_unidentified(english_common_name)) %>%
-    filter(!is_unidentified(spanish_common_name)) %>%
+    filter(!is_unidentified(species)) %>%
     filter(aou > 2880) %>%
     filter(aou < 3650 | aou > 3810) %>%
     filter(aou < 3900 | aou > 3910) %>%

### 2016-10-13 R/forecast-bbs-core.R
use stringr::word() instead of regex to pull main species name
regex was not working on oddball subspeces 'Colaptes auratus auratus x auratus cafer'. word() works on all of them.

@@ -91,6 +91,7 @@ filter_species <- function(df){
 #'
 #' @importFrom dplyr "%>%" filter slice group_by summarise ungroup
 #' @importFrom magrittr extract2
+#' @importFrom stringr word
 combine_subspecies = function(df){
 
   species_table = get_species_data()
@@ -108,7 +109,7 @@ combine_subspecies = function(df){
   # Drop the third word of the subspecies name to get the species name,
   # then find the AOU code
   new_subspecies_ids = species_table %>%
-    slice(match(gsub(" [^ ]+$", "", subspecies_names),
+    slice(match(word(subspecies_names, 1,2),
                 species_table$spanish_common_name)) %>%
     extract2("aou")
 

### 2016-10-13 R/forecast-bbs-core.R
fix typo

@@ -61,7 +61,7 @@ db_engine=function(action, db='./data/bbsforecasting.sqlite', sql_query=NULL,
 #' dippers. These species are poorly sampled due to their aquatic or
 #' noctural nature. Also removes taxa that were either partially unidentified
 #' (e.g. "sp.") or were considered hybrids (e.g. "A x B") or were listed as more
-#' than one species (e.g. "B / B")
+#' than one species (e.g. "A / B")
 #'
 #' @param df dataframe containing an species_id column
 #'

### 2016-10-13 R/forecast-bbs-core.R
update comment

@@ -106,7 +106,7 @@ combine_subspecies = function(df){
     filter(spanish_common_name %in% subspecies_names) %>%
     extract2("aou")
 
-  # Drop the third word of the subspecies name to get the species name,
+  # Drop all but the first two words to get the root species name,
   # then find the AOU code
   new_subspecies_ids = species_table %>%
     slice(match(word(subspecies_names, 1,2),

### 2016-10-27 manuscript/draft.Rmd

@@
   * Use long-term data sets such as BBS to evaluate predictive skill
-  * Note the large change in average temperatures since the 80's; comparable to the changes expected in the next N years (change since 80's is a true trend but whether it continues in uncertain, 10.1007/978-3-319-03768-4_2, 10.1007/s00382-014-2070-3)
-, 
+  * Note the large change in average temperatures since the 80's; comparable to the potential changed in the next 20 years 
   * Lots of talk about how we should do it, fewer examples of doing them all together

### 2016-10-31 manuscript/draft.Rmd
update methods/models

--- 2016-10-05 manuscript/draft.Rmd
@@
   * Examples of best practices
-    * not just bioclim
-    * time series
-    * spatial/site-level stuff
+    * Not just bioclim
+    * Dynamic predictor variables (not just site-level averages over 50 years like 
+      Worldclim)
+    * time series models (temporal autocorrelation)
+    * spatial/site-level stuff (repeated measures aren't fully independent)
+    * Properly handling uncertainty:
       * probabilistic/error bars
+      * error propagation
     * multiple models
@@
 
-* Mixed model for observer effort \& site effects
-* Naive AR1 model (adjusted for observer effects)
-* species-level GBM w/ mixed model point estimates as bonus predictors where available
-* richness-level GBM w/ mixed model point estimates as bonus predictors where available
-* JSDM
-* Ensembles (weighted and unweighted)
+* Mixed model for observer effort \& site effects: used lme4, included random
+  effects observer_id to account for situations were one observer consistently
+  found more (or fewer) species than average. Observer effects are inherent in
+  large data sets taken by different observers, and are known to occur in BBS 
+  [citation]. These models also included a site-level random effect to account
+  for differences in actual richness across BBS routes. [[Maybe year effect?]].
+  No fixed effects because not interested in explaining differences among sites
+  at this point; just want to account for their existence.
+* Residuals from this model give an estimate of how richness changed over time
+  at each site, accounting for observer differences.  We then used this "corrected"
+  version of the data for our time-series models such as the AR1, etc.
+* Naive AR1 model (adjusted for observer effects). This model says that richness 
+  can diffuse up or down from the previous year's value, without any constraints.
+  As a result, the confidence intervals expand quickly over time.
+* Average (adjusted for observer effects). This model says that richness at a site
+  fluctuates around a fixed mean value, so its confidence intervals do not expand
+  as the time since last observation increases.
+* Auto-ARIMA (Hyndman 2016). Choose a differencing order, then do forward stepwise
+  model selection based on AIC.
+* Auto-ARIMA + environment.
+* SDMs: use graf model w/ mixed model estimates of site \& observer effects
+  as predictors (include error in predictors as estimated from the mixed model
+  to propagate our uncertainty forward). This is basically SDM, but with a 
+  better model (graf) and more information (monthly climate data instead of using constant values from the Worldclim data, and observer effects).
+We used Poisson-binomial confidence intervals, as discussed in Calabrese et al.
+2014.
+* richness-level GBM w/ mixed model point estimates as bonus predictors where 
+  available. This lets us predict richness directly instead of via noisy 
+  species-level estimates, but doesn't use the full data set in one model.
+  Maybe need to run multiple times to propagate uncertainty from lme4.
+* JSDM. Should get best of the previous 2. Might be able to propagate uncertainty
+  from lme4.
+* Ensembles. Averaging predictions from multiple models tends to reduce the 
+  noise in the estimates, and can thus lead to better predictions. Choosing the
+  weights for the ensemble isn't straightforward because our estimates of model 
+  error on the training set are biased in favor of models that
+  overfit. [[Another option is to spatially cross-validate on the training set]]
 
@@
 * Calibration/bias
-* Evaluated each year after the last training year
+* Evaluated each year after the last training year (2010?)
 * Dig around in the residuals for insights

### 2016-10-31 manuscript/draft.Rmd

@@
   * Use long-term data sets such as BBS to evaluate predictive skill
-  * Note the large change in average temperatures since the 80's; comparable to the changes expected in the next N years
+  * Note the large change in average temperatures since the 80's; comparable to the potential changed in the next 20 years 
   * Lots of talk about how we should do it, fewer examples of doing them all together

### 2016-11-07 manuscript/draft.Rmd

@@
   * Use long-term data sets such as BBS to evaluate predictive skill
-  * Note the large change in average temperatures since the 80's; comparable to the potential changed in the next 20 years 
+  * Note the large change in average temperatures since the 80's; comparable to 
+    the potential changed in the next 20 years 
   * Lots of talk about how we should do it, fewer examples of doing them all together
@@
 Here we do that:
-  * forecasts for biodiversity in NAmerican birds with as many best practices as possible
-  * Discuss the influence of these improved methods \& of our error rates on our ability to understand:
+  * forecasts for biodiversity in NAmerican birds with as many best practices as 
+    possible
+  * Discuss the influence of these improved methods \& of our error rates on our 
+    ability to understand:
     * How to improve
@@
 
-**Data**
+### Data
 
@@
 
-**Models**
-
-* Mixed model for observer effort \& site effects: used lme4, included random
+### Accounting for site \& observer effects
+* Observer effects are inherent in
+  large data sets taken by different observers, and are known to occur in BBS 
+  [citation].
+* Mixed model for observer effort \& site effects: used Stan, included random
   effects observer_id to account for situations were one observer consistently
-  found more (or fewer) species than average. Observer effects are inherent in
-  large data sets taken by different observers, and are known to occur in BBS 
-  [citation]. These models also included a site-level random effect to account
-  for differences in actual richness across BBS routes. [[Maybe year effect?]].
+  found more (or fewer) species than average. These models also included a 
+  site-level random effect to account for differences in actual richness across 
+  BBS routes. [[Maybe year effect?]].
   No fixed effects because not interested in explaining differences among sites
@@
   version of the data for our time-series models such as the AR1, etc.
+* Note: throughout this manuscript, we use Gaussian errors for richness rather than 
+  a proper count distribution because:
+  * underdispersion (variance/mean << 1)
+  * Easier to access out-of-the-box time series methods like auto-ARIMA.
+  * Central limit theorem, large number of species (mean ~50 spp)
+
+### Models
+
+**Baseline models**
+* The lme4 model fitted above.  This model assumes that the site-level average
+  is constant, and tries to estimate its value.  The confidence intervals don't
+  expand over time (except when observers change) because the model says nothing
+  has changed
 * Naive AR1 model (adjusted for observer effects). This model says that richness 
@@
   As a result, the confidence intervals expand quickly over time.
-* Average (adjusted for observer effects). This model says that richness at a site
-  fluctuates around a fixed mean value, so its confidence intervals do not expand
-  as the time since last observation increases.
+
+**Time series models**
 * Auto-ARIMA (Hyndman 2016). Choose a differencing order, then do forward stepwise
@@
 * Auto-ARIMA + environment.
+
+**SDMs**
 * SDMs: use graf model w/ mixed model estimates of site \& observer effects
@@
   to propagate our uncertainty forward). This is basically SDM, but with a 
-  better model (graf) and more information (monthly climate data instead of using constant values from the Worldclim data, and observer effects).
-We used Poisson-binomial confidence intervals, as discussed in Calabrese et al.
-2014.
+  better model (graf) and more information (monthly climate data instead of 
+  using constant values from the Worldclim data, and observer effects). We used 
+  Poisson-binomial confidence intervals, as discussed in Calabrese et al. 2014.
+* JSDM. Should get best of the previous 2. Might be able to propagate uncertainty
+  from lme4.
+
+**"Macroecological" model**
 * richness-level GBM w/ mixed model point estimates as bonus predictors where 
@@
   Maybe need to run multiple times to propagate uncertainty from lme4.
-* JSDM. Should get best of the previous 2. Might be able to propagate uncertainty
-  from lme4.
+
+**Ensembles**
 * Ensembles. Averaging predictions from multiple models tends to reduce the 
@@
 
-(need a side note about why we're not using a proper count distribution; underdispersion, easier access to out-of-the-box time series methods)
-
 **Evaluation**
 
+Evaluated each year after the last training year (2010?):
 * (R)MSE
@@
 * Calibration/bias
-* Evaluated each year after the last training year (2010?)
-* Dig around in the residuals for insights
+
+Looked at how error changed as the time horizon of forecasting lengthened
+* Error will generally tend to increase
+* Error might flatten out after some number of years
+* Error might increase at different rates for different methods
+  * One model might be better at short time scales \& worse at long time scales
 
@@
   * We've posted predictions for the next N years of BBS data, using several methods, and invite others to do so as well.
+* Dig around in the residuals for insights
 
@@
 None of the models are true forcasts as they use observed predictor variables (except the AR1).
+
+* Need predictions at multiple time scales
+* Prediction quality at short time scales might not be a good indicator of 
+  long-term effectiveness

### 2017-02-14 manuscript/draft.Rmd
Create box on good forecasting practices

@@
   long-term effectiveness
+
+## Box 1: Ten simple rules for making and evaluating ecological forecasts
+
+### 1. Compare multiple modeling approaches
+
+Typically ecological forecasts use one modeling approach or a small number of
+related approaches. By fitting and evaluating multiple modeling approaches we
+can learn more rapidly about the best approaches for making predictions for a
+given ecological quantity. This includes comparing process based and data-driven
+models.
+
+### 2. Use time-series data when possible
+
+Forecasts describe how systems are expected to change through time. While some
+areas of ecological forecasting focus primarily on time-series data, others
+primarily focus on using spatial models and space-for time substitutions. Using
+ecological and environmental time-series data allows the consideration of actual
+dynamics from both a process [@treddenick2016a] and error structure perspective.
+
+### 3. Pay attention to uncertainty
+
+Understanding confidence in a forecast is just as important as understanding the
+average or expected outcome. Failing to account for uncertainty can result in
+overconfidence in highly uncertain outcomes leading to poor decision making and
+erosion of confidence in ecological forecasts. Models should explicitly include
+sources of uncertainty where possible and evaluations of forecasts should assess
+the accuracy of uncertainties as well as point estimates [@dietz2017].
+
+### 4. Use predictors related to the question
+
+Many ecological forecasts use data that is readily available and easy to work
+with. While ease of use is a reasonable consideration it is also important to
+include predictor variables that are expected to relate to the ecological
+quantity being forecast and dynamic time-series of predictors instead of
+long-term averages. Investing time in identifying and acquiring better predictor
+variables may have at least as many benefits as using more sophisticated
+modeling techniques.
+
+### 5. Assess how forecast accuracy changes with time-lag
+
+In general the accuracy of forecasts decreases with the length of time into the
+future being forecast [@petch2015]. This decay in accuracy and associated
+forecast horizons should be considered when evaluating forecasts and comparing
+models.
+
+### 6. Include an observation model
+
+Ecological observations are influenced by both the underlying processes and how
+the system is sampled. When possible forecasts should model the factors
+influencing the observation of the data.
+
+### 7. Validate using hindcasting
+
+To evaluate the expected accuracy and uncertainty of forecasts assess the
+performance of these forecasts within existing time-series data.
+
+### 8. Publicly archive forecasts
+
+To allow the future evaluation of the accuracy and uncertainty of forecasts the
+forecast values and/or models should be archived so that they can be assessed
+after new data is generated [@mcgill2012]. Enough information should be provided
+to allow an unambiguous assessment of the forecast performance.
+
+### 9. Make short-term and long-term predictions
+
+In cases where long-term predictions are the primary goal, short-term should
+also be made to accommodate the time-scales of planning and management decisions
+and to allow the accuracy of the forecasts to be quickly evaluated
+[@treddenick2016a];
+
+### 10. Compare forecasts to simple baselines
+
+Compare the accuracy of the forecasts to simple baselines such as the
+time-series average to determine if the modeled forecasts are more accurate than
+the naive assumption that the world is static.

### 2017-02-14 manuscript/draft.Rmd
Draft of Introduction

@@
 
-Quality of predictions is important:
-  * helps determine policy/set priorities
-  * something something climate change
-  * need to know both accuracy of point estimates and how reasonable our prediction intervals are.
-  * Overconfident predictions can be worse than uncertainty
-  * Quality of prediction is one metric for how much we know about ecological processes
-Unclear how much to trust existing predictions:
-  * Strong assumptions (see "misc" below)
-  * many *interesting* predictions
-  * Potential problems with confidence intervals
-  * Potential disconnect between training target and actual goals
-  * Difficult to test model assumptions/predictions because of time scales
-    * Most scientists making predictions today will have retired by 2050
-Demonstrate best practices \& their contribution to better predictions
-  * Use long-term data sets such as BBS to evaluate predictive skill
-  * Note the large change in average temperatures since the 80's; comparable to 
-    the potential changed in the next 20 years 
-  * Lots of talk about how we should do it, fewer examples of doing them all together
-  * Examples of best practices
-    * Not just bioclim
-    * Dynamic predictor variables (not just site-level averages over 50 years like 
-      Worldclim)
-    * time series models (temporal autocorrelation)
-    * spatial/site-level stuff (repeated measures aren't fully independent)
-    * Properly handling uncertainty:
-      * probabilistic/error bars
-      * error propagation
-    * multiple models
-      * Test with simple baselines (e.g. random walk, future-is-like-present)
-  * Focus on richness/diversity because \[citations saying that it's important\]
-  * Publish forecasts at different timescales to evaluate performance/credibility
-Here we do that:
-  * forecasts for biodiversity in NAmerican birds with as many best practices as 
-    possible
-  * Discuss the influence of these improved methods \& of our error rates on our 
-    ability to understand:
-    * How to improve
-    * How much to trust the predictions
+Forecasting the future state of ecological systems is increasingly important for
+planning, management, and testing the performance of ecological models
+[@clark2001, @dietz2017]. One area of particular interest for forecasting is
+biodiversity since it is important for ecosystem function, central to
+conservation planning, and expected to be influenced by anthropogenic global
+change. Anticipating potential changes in biodiversity is crucial for large
+scale management and conservation, and for addressing debates regarding whether
+local scale diversity is declining and may continue to decline in response to
+anthropogenic influences [@dornelas2014].
+
+Previous efforts to predict future patterns of species richness, and diversity
+more generally, have focused primarily on building species distributions models
+to quantify the spatial relationships between the occurrence of individual
+species and the environment. Forecasts for future environmental conditions are
+then used to predict where each species will occur in the future and the
+probabilities of occurrence are summed across species to predict future species
+richness [e.g., @thomas2004]. Alternatively, models that directly relate spatial
+patterns of species richness to the environment have been developed and
+generally perform equivalently to species distribution modeling based methods
+[@algar2009, @distler2015].
+
+However, despite the emerging interest in forecasting species richness and other
+aspects of biodiversity, little is known about how effectively we can anticipate
+their dynamics. What we do know comes from a small number of hindcasting
+studies, where models are built using data on species occurrence and richness
+from the past and evaluated on their ability to predict contemporary patterns
+[e.g., @algar2009, @distler2015]. These studies are a valuable first step, but
+lack several components that are important for developing good forecasting
+models and understanding how accurately these models can predict the
+future. These components of good forecasting and evaluation (Box 1) broadly
+involve: 1) expanding the use of data to include biological and environmental
+time-series [@treddenick2016]; 2) accounting for uncertainty in observations,
+processes, model choice, and forecast evaluation [@dietz2017]; and 3) conducting
+meaningful evaluations of the forecasts by hindcasting, archiving short-term
+forecasts, and comparing forecasts to baselines to determine if the forecasts
+are more accurate than assuming the system is basically static [@Perretti2013].
+
+In this paper we attempt to forecast the species richness of breeding birds at
+hundreds of locations throughout North America while following the best
+practices in Box 1 for ecological forecasting. To do this we combine 30 years of
+time-series data on bird distributions with monthly time-series of climate data
+and satellite-based remote-sensing. We compare traditional distribution modeling
+based approaches to spatial models of species richness, time-series methods, and
+simple baselines. All forecasts model uncertainty and observation, are evaluated
+across different time lags using hindcasting, and are publicly archived to allow
+future assessment. We discuss the implications of these practices for our
+understanding of, and confidence in, the resulting forecasts, and how we can
+continue to improve on these approaches in the future.
 

### 2017-02-23 manuscript/draft.Rmd
Draft of Title
Add possible changes in model performance w/lead time

@@
-# Title
+# Moving towards large-scale data-intensive forecasting of species richness in breeding birds
 
@@
 In general the accuracy of forecasts decreases with the length of time into the
-future being forecast [@petch2015]. This decay in accuracy and associated
-forecast horizons should be considered when evaluating forecasts and comparing
-models.
+future being forecast [@petch2015]. This decay in accuracy and the potential for
+different rates of decay to result in different relative model performance at
+different lead times should be considered when evaluating forecasts and
+comparing models.
 

### 2017-03-02 manuscript/draft.Rmd
Edits to Intro/Box in response to @davharris's suggestions

@@
+---
+output:
+  pdf_document: default
+  html_document: default
+---
 # Moving towards large-scale data-intensive forecasting of species richness in breeding birds
-
+# Using best practices for forecasting biodiversity changes in breeding birds
 Authors: Harris, Taylor and White?
@@
 Forecasting the future state of ecological systems is increasingly important for
-planning, management, and testing the performance of ecological models
-[@clark2001, @dietz2017]. One area of particular interest for forecasting is
-biodiversity since it is important for ecosystem function, central to
-conservation planning, and expected to be influenced by anthropogenic global
-change. Anticipating potential changes in biodiversity is crucial for large
-scale management and conservation, and for addressing debates regarding whether
-local scale diversity is declining and may continue to decline in response to
-anthropogenic influences [@dornelas2014].
+planning, management, and evaluating how well ecological models capture the key
+ecological processes governing a system [@clark2001, @dietz2017]. One area of
+particular interest for forecasting is biodiversity since it is important for
+ecosystem function, central to conservation planning, and expected to be
+influenced by anthropogenic global change. Anticipating potential changes in
+biodiversity is crucial for large scale management and conservation, and for
+addressing debates regarding whether local scale diversity is declining and may
+continue to decline in response to anthropogenic influences [@dornelas2014].
 
@@
 aspects of biodiversity, little is known about how effectively we can anticipate
-their dynamics. What we do know comes from a small number of hindcasting
-studies, where models are built using data on species occurrence and richness
-from the past and evaluated on their ability to predict contemporary patterns
-[e.g., @algar2009, @distler2015]. These studies are a valuable first step, but
-lack several components that are important for developing good forecasting
-models and understanding how accurately these models can predict the
-future. These components of good forecasting and evaluation (Box 1) broadly
-involve: 1) expanding the use of data to include biological and environmental
-time-series [@treddenick2016]; 2) accounting for uncertainty in observations,
-processes, model choice, and forecast evaluation [@dietz2017]; and 3) conducting
-meaningful evaluations of the forecasts by hindcasting, archiving short-term
-forecasts, and comparing forecasts to baselines to determine if the forecasts
-are more accurate than assuming the system is basically static [@Perretti2013].
+their dynamics. This is due in part to the long time scales over which many
+ecological forecasts are applied and the resulting difficulty in assessing
+whether the predicted changes occurred. What we do know comes from a small
+number of hindcasting studies, where models are built using data on species
+occurrence and richness from the past and evaluated on their ability to predict
+contemporary patterns [e.g., @algar2009, @distler2015]. These studies are a
+valuable first step, but lack several components that are important for
+developing good forecasting models and understanding how accurately these models
+can predict the future. These components of good forecasting and evaluation (Box
+1) broadly involve: 1) expanding the use of data to include biological and
+environmental time-series [@treddenick2016]; 2) accounting for uncertainty in
+observations, processes, model choice, and forecast evaluation
+[@clark2001, @dietz2017]; and 3) conducting meaningful evaluations of the
+forecasts by hindcasting, archiving short-term forecasts, and comparing
+forecasts to baselines to determine if the forecasts are more accurate than
+assuming the system is basically static [@Perretti2013].
 
@@
 time-series data on bird distributions with monthly time-series of climate data
-and satellite-based remote-sensing. We compare traditional distribution modeling
-based approaches to spatial models of species richness, time-series methods, and
-simple baselines. All forecasts model uncertainty and observation, are evaluated
-across different time lags using hindcasting, and are publicly archived to allow
-future assessment. We discuss the implications of these practices for our
-understanding of, and confidence in, the resulting forecasts, and how we can
-continue to improve on these approaches in the future.
+and satellite-based remote-sensing. This 30 year time-scale has only recently
+become possible for large-scale time-series based forecasting and allows us to
+model and assess changes a decade or more into the future in the presence of
+shifts in environmental conditions on par with predicted climate change. We
+compare traditional distribution modeling based approaches to spatial models of
+species richness, time-series methods, and simple baselines. All forecasts model
+uncertainty and observation, are evaluated across different time lags using
+hindcasting, and are publicly archived to allow future assessment. We discuss
+the implications of these practices for our understanding of, and confidence in,
+the resulting forecasts, and how we can continue to improve on these approaches
+in the future.
 
@@
 given ecological quantity. This includes comparing process based and data-driven
-models.
+models and comparing the accuracy of forecasts to simple baselines to determine
+if the modeled forecasts are more accurate than the naive assumption that the
+world is static.
 
@@
 erosion of confidence in ecological forecasts. Models should explicitly include
-sources of uncertainty where possible and evaluations of forecasts should assess
-the accuracy of uncertainties as well as point estimates [@dietz2017].
+sources of uncertainty and how the propagate through the forecast where
+possible. Evaluations of forecasts should assess the accuracy of uncertainties
+as well as point estimates [@dietz2017].
 
@@
 and to allow the accuracy of the forecasts to be quickly evaluated
-[@treddenick2016a];
-
-### 10. Compare forecasts to simple baselines
-
-Compare the accuracy of the forecasts to simple baselines such as the
-time-series average to determine if the modeled forecasts are more accurate than
-the naive assumption that the world is static.
+[@treddenick2016a].

### 2017-03-10 R/get_ndvi_data.R
updates for latest gimms library updates
No longer have to deal with quality flag stuff because gimms now deals with the quality flags automatically in rasterizeGimms.

@@ -26,15 +26,22 @@ get_gimms_download_list=function(gimms_folder = './data/gimms_ndvi/'){
 }
 
 ################################################
-#Extract values from a single gimms file given a set of coordinates
+#Extract values from a single gimms file given a set of coordinates.
+#Excludes values which don't meet NDVI quality flags.
+#From the GIMMS readme:
+#FLAG = 7 (missing data)
+#FLAG = 6 (NDVI retrieved from average seasonal profile, possibly snow)
+#FLAG = 5 (NDVI retrieved from average seasonal profile)
+#FLAG = 4 (NDVI retrieved from spline interpolation, possibly snow)
+#FLAG = 3 (NDVI retrieved from spline interpolation)
+#FLAG = 2 (Good value)
+#FLAG = 1 (Good value)
 ################################################
 #' @importFrom gimms rasterizeGimms
 extract_gimms_data=function(gimms_file_path, route_locations){
   #Have to load and extract twice. Once for the actual NDVI, once for the quality flag.
-  gimmsRaster=rasterizeGimms(gimms_file_path, flag=FALSE)
-  ndvi=raster::extract(gimmsRaster, route_locations)
-  gimmsRaster=rasterizeGimms(gimms_file_path, flag=TRUE)
-  flag=raster::extract(gimmsRaster, route_locations)
+  gimmsRaster=rasterizeGimms(gimms_file_path, keep=c(1,2,3))
+  ndvi=as.numeric(raster::extract(gimmsRaster, route_locations))
 
   year=as.numeric(substr(basename(gimms_file_path), 4,5))
   month=substr(basename(gimms_file_path), 6,8)
@@ -46,14 +53,13 @@ extract_gimms_data=function(gimms_file_path, route_locations){
   #Convert 2 digit year to 4 digit year
   year=ifelse(year>50, year+1900, year+2000)
 
-  return(data.frame(year=year, month=month, day=day, ndvi=ndvi, flag=flag, site_id=route_locations@data$site_id, stringsAsFactors = FALSE))
+  return(data.frame(year=year, month=month, day=day, ndvi=ndvi, site_id=route_locations@data$site_id, stringsAsFactors = FALSE))
 }
 
 ################################################
 #Extract the NDVI time series for all bbs routes
 #from all years of gimms data
 ################################################
-
[…]

### 2017-03-10 manuscript/draft.Rmd

@@
 * Dig around in the residuals for insights
+* SDM's and time series models may not be accounting for the ecology of north american birds.
+Most of them migrate, so whether they are observed or not at the same time every year is a function of their migratory phenology.
+The migratory patterns of several species have been correlated with NDVI (Renfrew et al 2013).
 

### 2017-03-13 manuscript/draft.Rmd
Dave's edits

--- 2017-03-02 manuscript/draft.Rmd
@@
 * BBS + retriever
-* Monthly climate data from interpolated weather stations (PRISM)
+* Predictor variables
+  * Monthly climate data from interpolated weather stations (PRISM), converted
+    to BIOCLIM variables. Many of these variables were tightly correlated, so
+    we only included the 8 variables selected by Harris's (2015) BBS analysis
   * NDVI (GIMMS)
   * Altitude (SRTM 90m Digital Elevation Database via raster::getData)
+* Data summary: 
+  * Number of sites, species, unique observers
+  * central moments of richness (mean about 51, sd about 12 spp)
+  * The within-site variation was much smaller (sd about 5 spp), making the
+    site-level distributions underdispersed.  
+    * This underdispersion implies
+      that common count models such as binomial, Poisson, negative binomial 
+      won't fit well.
+    * For this reason (and because the time series methods assumed 
+      Gaussian errors), we defined model performance throughout this paper in 
+      terms of Gaussians (e.g. mean squared error), rather than with count
+      distributions.
+        * Some of the modeling approaches (SDMs and JSDMs) did use discrete
+          distributions internally, however.
 
 ### Accounting for site \& observer effects
-* Observer effects are inherent in
-  large data sets taken by different observers, and are known to occur in BBS 
-  [citation].
-* Mixed model for observer effort \& site effects: used Stan, included random
-  effects observer_id to account for situations were one observer consistently
-  found more (or fewer) species than average. These models also included a 
-  site-level random effect to account for differences in actual richness across 
-  BBS routes. [[Maybe year effect?]].
-  No fixed effects because not interested in explaining differences among sites
-  at this point; just want to account for their existence.
-* Residuals from this model give an estimate of how richness changed over time
-  at each site, accounting for observer differences.  We then used this "corrected"
+* Observer effects are inherent in large data sets taken by different 
+  observers, and are known to occur in BBS [citation].
+* While these biases are known to exist, they are difficult to quantify: even if
+  the different observers ran the same transects under identical conditions (which 
+  they don't), we would still only have an estimate of their differences, not an 
+  exact value.
+* We estimated the observer effects, along with our uncertainty about those
+  effects' size & direction, with a linear mixed model.
+* This model partitions the variance in observed richness values into site-level
+  variance, observer-level variance, and residual variance (e.g. variation within
+  a site from year to year or from morning to morning).
+  * No fixed effects in this model (e.g. environmental predictor variables because 
+    we're not interested in explaining differences among sites at this point; we 
+    just want to account for their existence while estimating observer effects.
+  * Subtracting off the estimated observer effect provides an estimate of how many
+    would have been found by a "typical" observer on that day; in this sense, it 
+    enables us to correct for observer differences. We then used this "corrected" 
     version of the data for our time-series models such as the AR1, etc.
-* Note: throughout this manuscript, we use Gaussian errors for richness rather than 
-  a proper count distribution because:
-  * underdispersion (variance/mean << 1)
-  * Easier to access out-of-the-box time series methods like auto-ARIMA.
-  * Central limit theorem, large number of species (mean ~50 spp)
+* A simple model like this one can't tell us exactly how each observer differed from
+  the others, so it is important to represent our uncertainty about these differences.
+  Here, we did so by collecting [[N]] Monte Carlo samples from the model's posterior
+  distribution, using Stan. This provided us with [[N]] plausible values for each 
+  observer, so we did not have to commit to a single value.
+* While some methods for predicting richness might be able to use the uncertain
+  estimates of the observer effects directly, we take the simpler approach of just
+  fitting each of the downstream models with each of the Monte Carlo samples. This
+  increases the amount of computational work by a factor of [[N]], but CPU time
+  wasn't our limiting resource.
 
@@
   expand over time (except when observers change) because the model says nothing
-  has changed
-* Naive AR1 model (adjusted for observer effects). This model says that richness 
-  can diffuse up or down from the previous year's value, without any constraints.
-  As a result, the confidence intervals expand quickly over time.
+  has changed.
+* Naive AR1 model (fitted to individual sites' residuals from the observer model). 
+  This model says that richness can diffuse up or down from the previous year's 
+  value, without any constraints. As a result, the confidence intervals expand 
+  quickly over time.
+* Fit once per MCMC sample, to account for uncertainty in what observer effect
+  values we should subtract off
 
@@
 * Auto-ARIMA + environment.
+* Uncertainty treated as above
+
+**"Macroecological" model**
+* Use a nonlinear model to predict richness as a function of environmental predictors
+  (BIOCLIM + elevation + NDVI).
+* Compared to SDMs, this lets us predict richness directly instead of using hundreds of
+  noisy species-level estimates, but doesn't use the full data set in one model.
+* For this part of the analysis, we used the `gbm` package for fitting boosted regression
+  trees with Gaussian errors.  
+* We included observer effects & the associated uncertainty by running this model [[N]]
+  times (once per MCMC sample).
+* We did *not* include the site-level random effects as predictors, meaning that this
+  approach implicitly assumes that two sites with identical Bioclim, elevation, and 
+  NDVI values should have identical richness distributions.
 
 **SDMs**
-* SDMs: use graf model w/ mixed model estimates of site \& observer effects
-  as predictors (include error in predictors as estimated from the mixed model
-  to propagate our uncertainty forward). This is basically SDM, but with a 
-  better model (graf) and more information (monthly climate data instead of 
-  using constant values from the Worldclim data, and observer effects). We used 
-  Poisson-binomial confidence intervals, as discussed in Calabrese et al. 2014.
-* JSDM. Should get best of the previous 2. Might be able to propagate uncertainty
+* Probably the most common approach?
+* Predict each species' occurrence probabilities individually, then add up the predictions
+* [[Something about not using thresholds that throw away information]]
+* Taking this approach in the face of uncertainty about observer effects would either 
+  require a customized model (cite "neighborly advice" paper), or fitting [[N]] models
+  times the number of species.
+* To make it feasible to fit this large number of models, we
+  used a custom wrapper around the randomForest package. Random forests are flexible
+  models that are built by combining many independent models ("trees") fit to different 
+  versions of the training data, which made it straightforward to adapt them to our 
+  situation: we simply fit one tree using each of the Monte Carlo estimates of our 
+  observer effects and averaged the predictions of the resulting forest.
+
+**JSDM** 
+* New approach that tries to get the , cite examples.
+* Should get best of the previous 2. Might be able to propagate uncertainty
   from lme4.
-
-**"Macroecological" model**
-* richness-level GBM w/ mixed model point estimates as bonus predictors where 
-  available. This lets us predict richness directly instead of via noisy 
-  species-level estimates, but doesn't use the full data set in one model.
-  Maybe need to run multiple times to propagate uncertainty from lme4.
 
@@
 * (R)MSE 
-* Log-likelihood (simultaneously rewards good point estimates, precision, and coverage)
+  * How far, on average, are the models' predictions from the true value?
+  * [[RMSE has better units than MSE (species versus squared species), but MSE is
+    nice because it's additive (i.e. MSE of 2 is twice as big as MSE of 1).]]
+* Gaussian Log-likelihood (simultaneously rewards good point estimates, precision, 
+  and coverage)
 * Coverage at 95%
@@
 [@treddenick2016a].
+
+### 10. [[?]]
+
+Sites' biotic and abiotic environments differ in more ways than the 
+few ones you happened to measure. As a result, some sites will 
+consistently have more species than a regression model (or sum of 
+regression models) would predict, while others willl consistently have 
+fewer. Depending on the spatial scale of these unmeasured variables, 
+independent site-level random effects could be sufficient for dealing 
+with this, or spatially-autocorrelated random effects might be needed.

### 2017-03-14 manuscript/important_references.Rmd

--- 2016-10-13 manuscript/important_references.Rmd
@@
-#Various factoids
----------------------
-On North American climate
+## Various factoids
+
+#### On North American climate  
 There has been an increase in mean temperature in North America since the 1980's (pg. 90, fig 2.7)   
@@
 
-#Forcasting
-People should do a better job at forecasting.
+-------------------------------
+#### Biodiversity is an important thing to consider   
+Cardinale, Bradley J., et al. "Biodiversity loss and its impact on humanity." Nature 486.7401 (2012): 59-67.
+
+## Forcasting
+#### People should do a better job at forecasting.  
 Evans, M. R., Norris, K. J. & Benton, T. G. Predictive ecology: systems approaches. Philos. Trans. R. Soc. B Biol. Sci. 367, 163–169 (2012).  
@@
 
-#SDM references
+## SDM references
 
@@
 Warton, David I., et al. "So many variables: joint modeling in community ecology." Trends in ecology & evolution 30.12 (2015): 766-779.
+
+## Bird ecology
+Renfrew, R.
+B., Kim, D., Perlut, N., Smith, J., Fox, J.
+and Marra, P.
+P.
+(2013), Phenological matching across hemispheres in a long-distance migratory bird.
+Diversity Distrib., 19: 1008–1019.
+doi:10.1111/ddi.12080
+Stephen Mayor, a postdoc with Rob Guralnick, also has unpublished work looking at migration/ndvi patterns for all north american birds.

### 2017-03-21 manuscript/draft.Rmd

@@
     we only included the 8 variables selected by Harris's (2015) BBS analysis
+  * Weather/climate and NDVI were each calculated separately for each year,
+    rather than averaged over the full multi-decade period (as Worldclim does)
   * NDVI (GIMMS)
@@
           distributions internally, however.
+* Splitting the data temporally
+  * Our data set began in `r settings$start_yr`, which was the first year that 
+    NDVI data was available from GIMMS
+  * All the models were trained on the data that was collected through 
+    `r settings$last_train_year`
+  * Sites were included in the analysis if there were at least 
+    `r settings$min_num_yrs` years of data available during this period.
+  * The models were tested/evaluated on the data from 
+    `r settings$last_train_year + 1` `r settings$end_yr`.
 
@@
 **Time series models**
+* These models only looked at one site at a time and looked for patterns in 
+  univariate time series.
 * Auto-ARIMA (Hyndman 2016). Choose a differencing order, then do forward stepwise
-  model selection based on AIC.
-* Auto-ARIMA + environment.
+  model selection based on AICc.
+    * Autoregressive: current state depends on previous state(s)
+    * Moving average: current observation error depends on previous observation error
+    * Integrated/differencing: is analysis performed on raw data, on sequential 
+      differences of the raw data, on sequential differences of sequential differences,
+      etc.
+    * default settings, except seasonal = FALSE
+* Auto-ARIMA + site-level environmental variables over time.
 * Uncertainty treated as above
@@
   (BIOCLIM + elevation + NDVI).
+* We chose the `gbm` package for fitting boosted regression trees as our example of
+  a nonlinear "macroecological model".
+    * Works by fitting thousands of small models sequentially, and then adding up
+      their predictions
+    * the individual small models were regression trees with [[settings]].
 * Compared to SDMs, this lets us predict richness directly instead of using hundreds of
@@
 * Probably the most common approach?
+* These models say that species' occurrence probabilities depend only on the values of
+  the environmental variables at the time and place of the transect, not on the history
+  of the environment or the history of the birds at that site
 * Predict each species' occurrence probabilities individually, then add up the predictions
@@
   times the number of species.
-* To make it feasible to fit this large number of models, we
-  used a custom wrapper around the randomForest package. Random forests are flexible
-  models that are built by combining many independent models ("trees") fit to different 
-  versions of the training data, which made it straightforward to adapt them to our 
-  situation: we simply fit one tree using each of the Monte Carlo estimates of our 
-  observer effects and averaged the predictions of the resulting forest.
+* To make it feasible to fit this large number of models, we used a custom wrapper 
+  around the randomForest package. Random forests are flexible models that are built 
+  by combining many independent models ("trees") fit to different versions of the training
+  data, which made it straightforward to adapt them to our situation: we simply fit one 
+  tree using each of the Monte Carlo estimates of our observer effects and averaged the 
+  predictions of the resulting forest.
+* For the uncertainty, we used Gaussian approximations, based on the means and variances
+  of the sums of independent Bernoulli random variables (cite Calabrese et al.).
 
 **JSDM** 
-* New approach that tries to get the , cite examples.
-* Should get best of the previous 2. Might be able to propagate uncertainty
-  from lme4.
+* Joint species distribution model (JSDM) is a new approach that makes predictions
+  about the full community composition, rather than modeling it species-by-species.
+* In stacked single-species SDMs, the species-level occurrences are treated as 
+  independent coin-flips whose proability depends only on the environment.
+* JSDMs remove this independence assumption, and explicitly account for the
+  possibility that a site will be much more (or less) suitable for birds in 
+  general than one would expect based on a few environmental measurements alone.
+* As a result, JSDMs do a better job of representing our uncertainty about 
+  richness, while stacked SDMs underestimate it (Harris 2015).
+* We used the `mistnet` package (Harris 2015) because it is the only JSDM that
+  describes species' environmental associations with nonlinear functions.
+* The mistnet package doesn't "know" about time series or spatial autocorrelation,
+  so it did not share information among the repeated runs of the same transect.
 
@@
 
-Evaluated each year after the last training year (2010?):
+* Evaluated each year between  `r settings$last_train_year` and `r settings$end_yr`
+* Metrics:
   * (R)MSE 
@@
       nice because it's additive (i.e. MSE of 2 is twice as big as MSE of 1).]]
+  * Calibration/bias
+    * In general, or in specific situations, is the model consistently predicting
+      that richness will be too high or too low?
+    * Can also report this value in terms of RMSE or MSE
+  * Coverage of 95% confidence intervals
   * Gaussian Log-likelihood (simultaneously rewards good point estimates, precision, 
     and coverage)
-* Coverage at 95%
-* Calibration/bias
 

### 2017-03-30 manuscript/draft.Rmd

@@
 Authors: Harris, Taylor and White?
+
+`r settings = yaml::yaml.load_file("settings.yaml")`
 
@@
 
-### Models
+### Site-level models
 
-**Baseline models**
-* The lme4 model fitted above.  This model assumes that the site-level average
-  is constant, and tries to estimate its value.  The confidence intervals don't
-  expand over time (except when observers change) because the model says nothing
-  has changed.
-* Naive AR1 model (fitted to individual sites' residuals from the observer model). 
-  This model says that richness can diffuse up or down from the previous year's 
-  value, without any constraints. As a result, the confidence intervals expand 
-  quickly over time.
-* Fit once per MCMC sample, to account for uncertainty in what observer effect
-  values we should subtract off
+[[note that these all use a Gaussian error distribution]]
 
-**Time series models**
-* These models only looked at one site at a time and looked for patterns in 
-  univariate time series.
-* Auto-ARIMA (Hyndman 2016). Choose a differencing order, then do forward stepwise
-  model selection based on AICc.
-    * Autoregressive: current state depends on previous state(s)
-    * Moving average: current observation error depends on previous observation error
-    * Integrated/differencing: is analysis performed on raw data, on sequential 
-      differences of the raw data, on sequential differences of sequential differences,
-      etc.
-    * default settings, except seasonal = FALSE
-* Auto-ARIMA + site-level environmental variables over time.
-* Uncertainty treated as above
+**Baseline models.** The two simplest models we fit treated site-level richness
+observations either as uncorrelated noise around a site-level constant (the 
+"average" model) or as an autoregressive model with a single year of history 
+(the "naive" model). The predictions from these two models are qualitatively 
+different: predictions from the "average" model are centered on the average 
+richness and the confidence intervals are narrow and constant-width. The
+"naive" model, in contrast, predicts that future observations will be similar to
+the final observed value (e.g. the value observed in `$r settings$last_train_year`), 
+and the confidence intervals expand rapidly as the predictions extend farther into 
+the future.  Each of these model types was fit separately to each BBS route in two
+ways: first, we fit the model to the raw richness values, without accounting for
+possible differences among observers at the site.  Then, we fit the models to the
+residuals from our observer model (i.e. fitting the model to an estimate of how
+many species a typical observer would have found, rather than to the raw data). 
+In the latter case, we averaged across [[N]] models that were fit separately to 
+the [[N]] Monte Carlo estimates of the observer effects, to account for our
+uncertainty in the true values of those effects.
 
-**"Macroecological" model**
-* Use a nonlinear model to predict richness as a function of environmental predictors
-  (BIOCLIM + elevation + NDVI).
-* We chose the `gbm` package for fitting boosted regression trees as our example of
-  a nonlinear "macroecological model".
-    * Works by fitting thousands of small models sequentially, and then adding up
-      their predictions
-    * the individual small models were regression trees with [[settings]].
-* Compared to SDMs, this lets us predict richness directly instead of using hundreds of
-  noisy species-level estimates, but doesn't use the full data set in one model.
-* For this part of the analysis, we used the `gbm` package for fitting boosted regression
-  trees with Gaussian errors.  
-* We included observer effects & the associated uncertainty by running this model [[N]]
-  times (once per MCMC sample).
-* We did *not* include the site-level random effects as predictors, meaning that this
-  approach implicitly assumes that two sites with identical Bioclim, elevation, and 
-  NDVI values should have identical richness distributions.
+**Time series models.** Like the two baseline models, these models only looked at 
+one site at a time, but they could include additional complexity. As above, these
+models were trained both with the raw richness time series and with Monte Carlo
+estimates of the observer-corrected time series. The  Auto-ARIMA models (based on 
+Hyndman's 2016 `auto.arima` function) could include an autoregressive component 
+(as in the "naive" model, but with the possibility of longer-term dependencies
+in the underlying process), a moving average component (where the noise can have 
+serial autocorrelation) and an integration/differencing component (so that the 
+analysis could be performed on sequential differences of the raw data, rather than
+on the observed time series itself). The `auto.arima` function chooses whether to 
+include each of these components, and how many terms to include for each one, using
+AICc, as described in Hyndman (2016). We used the default settings for this function,
+apart from setting `seasonal = FALSE` because our time series does not include a 
+seasonal component.  [[Note: should probably record somewhere what order model
+the auto.arima function tends to choose.  Eyeballing the results, seems to be most 
+similar to the naive model, but would be good to check.]] We also fit an 
+"Auto-ARIMA + environment" model, in which each site-level time series could 
+also be predicted by annual changes in our environmental variables at that site.
 
-**SDMs**
+### Continental-scale models
+
+So far, each model has been fit to one site at a time, with no information 
+(apart from observer effects) shared among sites. When continental-scale data 
+is available, ecologists often use correlative models to predict richness 
+changes over time from environmental data. Here, we tested three variants of 
+this approach. Here, we did *not* include the site-level random effects as 
+predictors, meaning that this approach implicitly assumes that two sites 
+with identical Bioclim, elevation, and NDVI values should have identical 
+richness distributions. As above, we included observer effects & the 
+associated uncertainty by running this model [[N]] times (once per MCMC 
+sample).
+
+**"Macroecological" model.** The most straightforward way to predict richness is 
+to fit a richness model. We chose the `gbm` package for fitting boosted 
+regression trees as our example of such a model. Boosted regression trees is a 
+common approach in the species distribution modeling literature [refs] and 
+works by fitting thousands of small tree-structured models sequentially, with
+each tree optimized to reduce the error of its predecessors. For the reasons 
+discussed above, this model was optimized using a Gaussian likelihood. [[other 
+gbm settings go here.]] Compared to SDMs, this lets us predict richness directly 
+instead of using hundreds of noisy species-level estimates, but it also discards 
+any potentially-useful information regarding species turnover.
+
+**SDMs.**
 * Probably the most common approach?
@@
   around the randomForest package. Random forests are flexible models that are built 
-  by combining many independent models ("trees") fit to different versions of the training
+  by combining many independent sub-models ("trees") fit to different versions of the training
   data, which made it straightforward to adapt them to our situation: we simply fit one 
@@
 
-**Ensembles**
+### Ensembles
 * Ensembles. Averaging predictions from multiple models tends to reduce the 
@@
 
-**Evaluation**
+### Model evaluation
 
 * Evaluated each year between  `r settings$last_train_year` and `r settings$end_yr`
+* When a model was run multiple times (e.g. once per MCMC sample for the observer
+  estimates), we used the mean of those runs' point estimates as our final point 
+  estimate and we calculated the uncertainty using the law of total variance (i.e.
+  the average of the model runs' variance, plus the variance in the point estimates).
 * Metrics:
@@
   * Coverage of 95% confidence intervals
-  * Gaussian Log-likelihood (simultaneously rewards good point estimates, precision, 
+  * Gaussian deviance (simultaneously rewards good point estimates, precision, 
     and coverage)

### 2017-03-30 analysis/cv-mistnet.R
cross-validate additional arguments, add leaky relu & time constraint

@@ -5,6 +5,11 @@ start = args[[1]]
 end = args[[2]]
 N = args[[3]]
 
+discrete_log_runif = function(N, a, b){
+  # Sample from a discrete log-uniform distribution between a and b
+  floor(exp(runif(N, log(a), log(b + 1))))
+}
+
 # check for mistnet version: this will fail if mistnet version is too old.
 # (Otherwise, bogus arguments will get absorbed into `...` without warning)
 invisible(mistnet::adam.updater()$annealing_rate)
@@ -14,11 +19,18 @@ seed = .Random.seed
 
 # Deterministically set hyperparameter list by setting the seed
 set.seed(1)
-full_arglist = list(a_0 = rlnorm(N, log(.01), .25),
+mistnet_arglist = list(n.minibatch = discrete_log_runif(N, 10, 100),
+                       latent_dim =  discrete_log_runif(N, 5, 20),
+                       n.importance.samples = discrete_log_runif(N, 10, 50),
+                       N1 = discrete_log_runif(N, 20, 100),
+                       N2 = discrete_log_runif(N, 5, 50))
+
+updater_arglist = list(a_0 = rlnorm(N, log(.01), 1),
                     annealing_rate = rlnorm(N, -10, 1),
                     b1 = rbeta(N, 7, 1),
                     b2 = rbeta(N, 100, 1),
                     e = rlnorm(N, -18, 2))
+
 # Re-set the seed to its previous value
 set.seed(seed)
 
@@ -29,5 +41,7 @@ for (i in start:end) {
   # effects.
   fit_mistnet(iter = i, 
               use_obs_model = FALSE, 
-              updater_arglist = purrr::map(full_arglist, i), CV = TRUE)
+              mistnet_arglist = purrr::map(mistnet_arglist, i),
+              updater_arglist = purrr::map(updater_arglist, i), 
+              CV = TRUE)
 }

### 2017-03-30 mistnet.R

@@ -2,7 +2,7 @@ library(Heraclitus)
 library(tidyverse)
 library(mistnet)
 library(progress)
-devtools::load_all()
+devtools::load_all() 
 if (!file.exists("observer_model.rds")) {
   fit_observer_model()
 }
@@ -14,14 +14,9 @@ bbs = get_bbs_data() %>%
 
 # Mistnet settings based on the second-best model in Appendix D of 
 # the mistnet paper
-n.minibatch = 79L
-latent_dim = 10L
-n.importance.samples = 28L
-N1 = 37L
-N2 = 15L
 
 # Other mistnet settings
-n_opt_iterations = 1E5
+n_hours = 4
 iteration_size = 10
 n_prediction_samples = 500
 
@@ -29,6 +24,7 @@ n_prediction_samples = 500
 # a_0 and annealing_rate are hyperparameters of the `adam` optimizer
 fit_mistnet = function(iter,
                        use_obs_model, 
+                       mistnet_arglist,
                        updater_arglist,
                        CV = FALSE){
   
@@ -81,13 +77,13 @@ fit_mistnet = function(iter,
     y = y[data$in_train, ],
     layer.definitions = list(
       defineLayer(
-        nonlinearity = rectify.nonlinearity(),
-        size = N1,
+        nonlinearity = leaky.rectify.nonlinearity(),
+        size = mistnet_arglist$N1,
         prior = gaussian.prior(mean = 0, sd = .5)
       ),
       defineLayer(
-        nonlinearity = linear.nonlinearity(),
-        size = N2,
+        nonlinearity = leaky.rectify.nonlinearity(),
+        size = mistnet_arglist$N2,
         prior = gaussian.prior(mean = 0, sd = .5)
       ),
       defineLayer(
@@ -99,28 +95,25 @@ fit_mistnet = function(iter,
     loss = bernoulliRegLoss(a = 1 + 1E-6, b = 1 + 1E-6),
     updater = purrr::invoke(adam.updater$new, updater_arglist),  
     sampler = gaussian.sampler(ncol = latent_dim, sd = 1),
-    n.importance.samples = n.importance.samples,
-    n.minibatch = n.minibatch,
+    n.importance.samples = mistnet_arglist$n.importance.samples,
+    n.minibatch = mistnet_arglist$n.minibatch,
     training.iterations = 0,
     initialize.biases = TRUE,
     initialize.weights = TRUE
   )
[…]

### 2017-03-30 mistnet.R
scope

@@ -94,7 +94,7 @@ fit_mistnet = function(iter,
     ),
     loss = bernoulliRegLoss(a = 1 + 1E-6, b = 1 + 1E-6),
     updater = purrr::invoke(adam.updater$new, updater_arglist),  
-    sampler = gaussian.sampler(ncol = latent_dim, sd = 1),
+    sampler = gaussian.sampler(ncol = mistnet_arglist$latent_dim, sd = 1),
     n.importance.samples = mistnet_arglist$n.importance.samples,
     n.minibatch = mistnet_arglist$n.minibatch,
     training.iterations = 0,

### 2017-03-30 analysis/cv-mistnet.R
int != float

@@ -7,7 +7,9 @@ N = args[[3]]
 
 discrete_log_runif = function(N, a, b){
   # Sample from a discrete log-uniform distribution between a and b
-  floor(exp(runif(N, log(a), log(b + 1))))
+  # Note that as.integer rounds down, so the upper limit needs to be
+  # incremented by 1.
+  as.integer(exp(runif(N, log(a), log(b + 1))))
 }
 
 # check for mistnet version: this will fail if mistnet version is too old.

### 2017-03-30 R/post-observer.R
save a list of ARIMA coefficients

@@ -31,8 +31,8 @@ make_forecast = function(x, fun_name, use_obs_model, settings, ...){
                          seasonal = FALSE)
   }
   
-  fcst = fun(y = x[[response_variable]], ...) %>% 
-    forecast::forecast(h = h, level = level)
+  model = fun(y = x[[response_variable]], ...)
+  fcst = forecast::forecast(model, h = h, level = level)
   
   if (any(is.na(fcst$mean))) {
     warning("NA in predictions")
@@ -42,7 +42,8 @@ make_forecast = function(x, fun_name, use_obs_model, settings, ...){
   # Distance between `upper` and `lower` is 2 sd, so divide by 2
   data_frame(year = seq(settings$last_train_year + 1, settings$end_yr), 
          mean = c(fcst$mean), sd = c(fcst$upper - fcst$lower) / 2, 
-         model = fun_name, use_obs_model = use_obs_model)
+         model = fun_name, use_obs_model = use_obs_model,
+         coef_names = list(names(coef(model))))
 }
 
 make_all_forecasts = function(x, fun_name, use_obs_model, 

### 2017-03-30 R/post-observer.R
coef_names choked when zero coefs were passed to it

@@ -39,11 +39,17 @@ make_forecast = function(x, fun_name, use_obs_model, settings, ...){
     return(list(NA))
   }
   
+  coef_names = if (length(coef(model)) > 0) {
+    list(names(coef(model)))
+  } else {
+    list(NA)
+  }
+  
   # Distance between `upper` and `lower` is 2 sd, so divide by 2
   data_frame(year = seq(settings$last_train_year + 1, settings$end_yr), 
          mean = c(fcst$mean), sd = c(fcst$upper - fcst$lower) / 2, 
          model = fun_name, use_obs_model = use_obs_model,
-         coef_names = list(names(coef(model))))
+         coef_names = coef_names)
 }
 
 make_all_forecasts = function(x, fun_name, use_obs_model, 

### 2017-05-10 manuscript/draft.Rmd
Draft of Discussion

@@
 
-## Results and conclusions
+## Results
 
-* None of the models actually work all that well, compared to baseline.
-* Temporal richness variability doesn't track the predictors we use; sites have "inertia"
-  * Cite 3 papers Ethan found on consistancy of alpha even when composition changes dramatically
-  * Good for absolute accuracy, little room for improvement over baseline
-  * Maybe richness is too coarse
-    * abundance
-    * species-level
-* More important to use a reasonable error model than to get the best predictors \& basis expansions
-* Ensembles help/don't help (depending on results)
-* Point estimate error versus coverage
-* In many ways, this is the best-case scenario for SDMs:
-  * large data set, known \& consistent sampling technique over a period of decades
-  * "True" absences
-  * extended predictor set
-  * birds arguably migrate quickly enough to make equlibrium \& especially space-for-time assumptions less ridiculous
-  * errors for individual species can cancel out; only asking for aggregate values
-* Value of stacking versus predicting the quantitity of interest directly
-* Need to start doing this, even if we're bad at it.
-  * We've posted predictions for the next N years of BBS data, using several methods, and invite others to do so as well.
-* Dig around in the residuals for insights
-* SDM's and time series models may not be accounting for the ecology of north american birds.
-Most of them migrate, so whether they are observed or not at the same time every year is a function of their migratory phenology.
-The migratory patterns of several species have been correlated with NDVI (Renfrew et al 2013).
+## Discussion
 
-## misc
+Forecasting is an emerging imperative in ecology and as such the field needs to
+develop and follow good practices for conducting and evaluating ecological
+forecasts. We have laid out a number of these practices (Box 1) and attempted to
+implement them in a single study that builds and evaluates forecasts of
+biodiversity in the form of species richness. The results of this effort are
+both promising and humbling. On an absolute basis, when comparing forecasts
+across sites, many different approaches to forecasting produce reasonable
+forecasts. If a site is predicted to have a high number of species in the
+future, relative to other sites, it generally does. However, within a site the
+forecasts perform relatively poorly. If a site is predicted to have a higher
+richness next year than it did last year there is little confidence that this
+will be the case. This results in simple site-level time-series and baseline
+models like the long-term average and the naive model being some of the best
+performing overall models for future biodiversity.
 
-SDM assumptions:
-  * we've measured the predictive power of all avialable variables, other potentially important thing that aren't available are landcover, observer skill level, (probably some others)
-  * Something like stationarity:
-    * e.g. "systems are at equilibrium now and will be at equilibrium in future"
-    * or "species track predictor variables quickly" (no lags)
-    * etc
-  * independent/identically distributed observations across species, years
-  * Space substitutes well for time
+The most commonly used method to forecast the future state of biodiversity,
+fitting a species distribution model separately to each species and then summing
+the probability of presence across those models, provides the worst overall
+forecasts of all methods evaluated. Most forecasting based on this approach is
+never evaluated and when it is it is typically only assessed based on cross-site
+rather than within site approaches. As a result the forecasts may appear to
+perform well even when change through time is not accurately predicted. This
+issue has also recently been recently highlighted as an issue for using SDMs to
+predict the future locations of individual species [plos one 2012 paper]. This
+result is particularly sobering because this method forms the foundation for
+major claims regarding the predicted loss of biodiversity to climate change
+[2004 science/nature papers]. The poor performance of SDM-based forecasts
+highlights the crucial importance of comparing multiple modeling approaches when
+conducting ecological forecasts, and in particular the value of comparing
+results to simple baselines to avoid over-interpreting the accuracy of
+ecological forecasts. Disciplines that have more mature forecasting cultures
+often do this by reporting "forecast skill", which is the improvement in the
+forecast relative to a simple baseline. We recommend adopting this approach in
+future ecological forecasting research. It also highlights the need for using
+hindcasting and time-series data for assessing the effectiveness of forecasting
+methods, since this allows comparisons to baselines and assessment of the
+predicted dynamics at each site.
 
-None of the models are true forcasts as they use observed predictor variables (except the AR1).
+The most common alternative to summed SDMs is modeling richness directly using
+environmental variables and using the resulting models to make forecasts [].
+Two comparisons of the the SDM and direct richness modeling approaches reported
+that the methods yielded equivalent results for forecasting diversity. While our
+results generally support this for point estimates of richness they also show
+that modeling richness directly results in much better estimation of the
+uncertainty of the forecast and therefore is a better overall approach to
+forecasting richness. A similar result is seen when comparing joint species
+distribution models (JSDMs) to stacked single species distribution models. For
+point estimates the joint distribution models are roughly equivalent to stacked
+SDMs, but the JSDMs provide much better estimates of uncertainty. In fact, JSDMs
+and modeling richness directly provide some of the best estimates of uncertainty
+across all modeling approaches. This highlights the importance evaluating models
+based on their uncertainty as well as their mean estimates. Not doing so is
+particularly problematic in this case because the summed species distribution
+models are overly confident meaning that using them would lead to believing that
+richness would be restricted to a much narrower range than would actually
+occur.
 
-* Need predictions at multiple time scales
-* Prediction quality at short time scales might not be a good indicator of 
-  long-term effectiveness
+The accurate estimates of uncertainty by JSDMs and richness models result in
+these models approaching the performance of time-series models and baselines at
+the longest time-scales assessed when using measures that consider both point
+estimates and uncertainty. This convergence at longer time scales highlights the
+importance of considering how forecasting method performance changes with the
+distance into the future being predicted. The value of this approach, and
+associated forecast horizons, has been raised for knowing how far into the
+future a model can be effectively used for forecasting [@petchy2015], but our
+results suggest a broader value to this approach for considering the potential
+importance of different models and processes for making forecasts at different
+scales. [[Should make sure Petchy et al. don't make this point]] Our results
+show that for forecasts 1-3 years in the future, baselines and time-series based
+approaches outperform all other methods using integrative metrics. At these time
+scales environmental changes are relatively small and changes in the biota may
+lag behind changes in the environment. However, as the timescale of the forecast
+increases to a decade the JSDMs and environmental richness models get closer to
+the accuracy of the time-series methods and even surpass some of those
+methods. As the forecast distance increases the amount of environmental change
+is expected to increase and the system will have more time to respond,
+potentially leading to shifts towards improved relative performance of the
+models incorporating environmental and ecological information. If this shift
+with forecast distance in the relative importance of different processes and
+models continues, this could lead to JSDMs and environmental richness models
+outperforming time-series based approaches at sufficiently long time
+scales. These results suggest that assessment of how forecast performance
+changes with time lag is not only important for determining how far in the
+future to forecast, but also for determining models and processes are most
+relevant for making forecasts at different time scales.
+
+It is also possible that models that include species' relationships to their
+environments or direct environmental constraints on richness may continue to
+produce forecasts that are worse than simply assuming the systems are
+static. This would be expected to occur is the systems are in fact not changing
+their richness over the relevant time scales and therefore simple models of no
+directional change are appropriate. Recent suggestions that local scale richness
+in some systems is not changing directionally at multi-decadal scales supports
+this possibility [@brown2001; @ernest2001; @dornelas201X; @theplantone]. This
+lack of change may be expected even in the presence of substantial changes in
+environmental conditions and species composition at a site due to the
+replacement of species from the regional pool [@brown2001; @ernest2001]. On
+average the Breeding Bird Survey sites used in this study show little change in
+richness ([[Add summary stat here]]) [see also @lasorte2007]. This absence of
+rapid change is beneficial for the absolute accuracy of forecasts across sites,
+because if a past years richness is known it is easy to estimate a future
+richness, and explains why stacked SDMs performs relatively well at this task
+despite failing to capture meaningful dynamics. However it makes it difficult to
+improve forecasts relative to simple baselines, since those baselines are close
+to representing what is actually occurring in the system. This suggests that
+simple time-series models and baselines should be actively considered for
+forecasts of richness and other stable aspects of biodiversity and also suggests
+that future efforts to understand and forecast biodiversity should also focus on
+composition since it is expected to be more dynamic
+[@ernest2001; @dornelas201X].
+
+In addition to consideration of the different process models used for
+forecasting it is important to consider the observation models. When working
+with any ecological dataset there are imperfections in sampling that have the
+potential to influence results. With large scale survey and citizen science
+datasets like the Breeding Bird Survey these issues are potentially magnified by
+the large number of different observers and major differences in habitat and
+species. We included an observation model for one of the two major observation
+issues known for the Breeding Bird Survey, differences among
+observers. Accounting for differences in observers substantially reduced the
+error in point estimates in all models and also improved the coverage of the
+confidence intervals. In addition, it resulted in changes in which models
+performed best, most notably reducing the performance of the naive model for
+point estimates. This suggests that the naive model performed well in part
+because it was capable of accommodating rapid shifts in estimated richness
+introduced by changes in the observer. These kinds of rapid changes were
+difficult for the other time-series models to accommodate and so the ARIMA and
+average models improved substantially once this source of observation error was
+addressed. This demonstrates that properly modeling observation error can be
+important for reducing uncertainty in forecast but can also lead to changes in
+the best methods for forecasting. We did not address differences in detection
+probability across species and sites since there is no clear way to address this
+issue without making strong assumptions about the data, but this would be a
+valuable addition to future forecasting models.
+
+Future efforts will also need to begin to address the additional uncertainty
+that comes from error in forecasting the environmental conditions themselves. In
+this and other hindcasting studies the environmental conditions for the "future"
+are known because the data has already been observed. In real forecasts the
+environmental conditions themselves have to be forecast and those forecasts have
+uncertainty and bias. This will cause ecological forecasts that use
+environmental data to be more uncertain. It is important to correctly
+incorporate this uncertainty in the predictor variables into forecasting models
+[@clark2001; @dietz2017]. Difficulty in forecasting future environmental
+conditions at small scales will present continued challenges for models
+incorporating these conditions and this may result in a continued advantage to
+simple time-series based approaches.
+
+The science of forecasting biodiversity, and ecology more broadly, remains in
+its infancy and it is important to consider the general inability the
+forecasting methods to improve on simple baselines in that context. When weather
+forecasting first started the forecasts were likewise worse than simple
+baselines []. One of the things that helped weather forecasts improve was large
+numbers of forecasts were made in public, which allowed different approaches to
+forecasting to be regularly assessed and improved
+[@mcgill2012 blog post; history of weather forecasting book]. This suggests that
+it is important for ecologists to start regularly making and evaluating true
+ecological forecasts, even if they perform poorly, and to make these forecasts
+publicly available for assessment. These forecasts should include both
+short-term predictions, which can be assessed quickly, and mid to long-term
+forecasts to help assess long time-scale processes and determine how far into
+the future we can successfully forecast [@tredenick ???, @dietz2017]. Forecasts
+for the next XX years for the models in this paper are openly archived on Dryad
+(XXXXXXXXXXX) so that we and others can assess how well they perform and we plan
+to evaluate these forecasts and report the results as each new year of BBS data
+becomes available. Weather forecasting has continually improved throughout its
+history [@bauer2015] in part due to making and evaluating public forecasts
+[@mcgill2012] and we hope this will help ecology do the same.
+
+Making successful ecological forecasts will be challenging. Ecological systems
+are complex, fundamental theory is less refined than simpler physical/chemical
+systems, and we currently lack the kinds of truly "big" data that produce
+effective forecasts through machine learning. Despite this we believe that
+progress can be made if we build an active forecasting culture in ecology that
+builds and assesses forecasts in ways that will allow us to improve the
+effectiveness of ecological forecasts most rapidly (Box 1). This includes
+expanding the scope of the ecological and environmental data we work with,
+paying attention to uncertainty in both model building and forecast evaluation,
+and rigorously assessing forecasts using a combination of hindcasting, archived
+forecasts, and comparisons to simple baselines.
 

### 2017-05-11 manuscript/draft.Rmd
first go at data

--- 2017-03-30 manuscript/draft.Rmd
@@
     `r settings$last_train_year + 1` `r settings$end_yr`.
+
+  Bird species richness was compiled from the Breeding Bird Survey (BBS) obtained using the Data Retriever software [@pardieck2017, @morris203].
+The BBS data was initially filtered to exclude all aquatic species, unidentified species, and hybrids.
+All data from routes that did not meet BBS acceptable criteria were excluded.
+Only routes within the continental U.S.A were used so as to match the climatic dataset.
+  Nineteen bioclimatic variables were derived from mean monthly temperature and precipitation from the 4km resolution PRISM dataset, allowing us to have yearly values for each variable [@prism].
+This was subset to the eight variables identified by Harris [-@harris2015] to be non-correlated.
+For each BBS route the mean value for each variable within a 40km radius, equal to the length of a BBS route, was used to account for spatial heterogeneity.
+Satellite derived NDVI was from the NDIV3g dataset and was available from the time period 1981-2015 [@pinzon2014].
+Elevation was from the SRTM 90m elevation dataset obtained using the R package raster [@hijmans2014, @jarvis2008].
+  All models described below were fit using data from the years 1982-2003, and evaluated using data from the years 2004-2015.
+The year 1982 corresponds to the first complete year of NDVI observations.
+Only BBS routes with at least 25 yearly observations in the years 1982-2015 were used.
 

### 2017-05-16 manuscript/draft.Rmd
edits and suggestions from ethan

--- 2017-03-30 manuscript/draft.Rmd
@@
     `r settings$last_train_year + 1` `r settings$end_yr`.
+
+  Bird species richness was compiled from the Breeding Bird Survey (BBS) [@pardieck2017] obtained using the Data Retriever software [@morris203].
+The BBS data was initially filtered to exclude all nocturnal, cepuscular, and aquatic species (since these species are not well sampled by BBS methods), as well as unidentified species, and hybrids.
+All data from routes that did not meet BBS acceptable criteria were excluded.
+Only routes within the continental U.S.A were used so as to match the climatic dataset.
+  We used the 4 km resolution PRISM data [@prism] to calculate eight bioclimatic variables identified as relevant to bird distributions [@harris2015] for each year in the dataset: mean diurnal range, isothermality, max temperature of the warmest month, mean temperature of the wettest quarter, mean temperature of the driest quarter, precipitation seasonality, precipitation of the wettest quarter, precipitation of the warmest quarter.
+Satellite derived NDVI with an 8 km resolution was from the NDIV3g dataset [@pinzon2014] and was available from the time period 1981-2013.
+Elevation was from the SRTM 90m elevation dataset [@jarvis2008] obtained using the R package raster [@hijmans2014].
+BBS route starting point coordinates were used to extract all environmental data.
+Climatic variables and NDVI were aggregated to the average values within a 40 km radius of the start point, equal to the length of a BBS route.
+NDVI was also subset to average summer (May, June, July) and winter (December, January, Feburary) values.
+  Data from 1982 (the first complete year of NDVI data) to 2003 were used for training models and data from 2004 to 2013 was used for testing.
+Only BBS routes with at least 25 yearly observations in the years 1982-2015 were used.
 

### 2017-05-17 R/get_prism_data.R
upscale PRISM to 40km cells
Also give the bbs locations a CRS since raster::extract() doesn't do it automatically anymore.

@@ -66,7 +66,8 @@ get_prism_data=function(){
     locations <- dplyr::select(bbs_data, site_id, long, lat) %>%
       distinct()
     coordinates(locations) <- c("long", "lat")
-
+    raster::crs(locations) <- '+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,0'
+    
     #Check to see if all the raw data in the years specified are downloaded,
     #download everything again if not. (Getting only what is needed, say if a
     #previous download failed, might be overly complicated)
@@ -76,6 +77,9 @@ get_prism_data=function(){
 
     #Load the prism data and extract using the bbs locations.
     prism_stacked <- prism_stack(ls_prism_data())
+    #Aggregate to 40km. Original cells average ~4.5 across N. America,
+    #so aggregation of 9x9 creates mostly 40km cells. 
+    prism_stacked = raster::aggregate(prism_stacked, fact=9)
     save_provenance(prism_stacked)
     extracted <- raster::extract(prism_stacked, locations)
     prism_bbs_data <- data.frame(site_id = locations$site_id, coordinates(locations), extracted)

### 2017-05-17 R/get_prism_data.R
offset data to better match bbs observations

@@ -130,12 +130,16 @@ process_bioclim_data=function(){
   #Get the prism data.
   prism_bbs_data=get_prism_data()
 
+  #Offset the year by 6 months so that the window for calculating bioclim variables
+  #will be July 1 - July 31. See https://github.com/weecology/bbs-forecasting/issues/114
+  prism_bbs_data$year = with(prism_bbs_data, ifelse(month %in% 6:12, year+1, year))
+  
   #Spread out the climate variables ppt, tmean, etc into columns
   prism_bbs_data = prism_bbs_data %>%
     spread(clim_var, value)
 
   #Process the quarter ones first.
-  quarter_info=data.frame(month=1:12, quarter=c(1,1,1,2,2,2,3,3,3,4,4,4))
+  quarter_info=data.frame(month=1:12, quarter=c(3,3,3,4,4,4,1,1,1,2,2,2))
   bioclim_quarter_data= prism_bbs_data %>%
     left_join(quarter_info, by='month') %>%
     group_by(site_id, year, quarter) %>%

### 2017-05-18 R/post-observer.R
don't drop coef names

@@ -76,7 +76,8 @@ make_all_forecasts = function(x, fun_name, use_obs_model,
     out = mutate(out, mean = mean + observer_effect)
   }
   
-  select(out, site_id, year, mean, sd, iteration, richness, model, use_obs_model)
+  select(out, site_id, year, mean, sd, iteration, richness, model, 
+         use_obs_model, coef_names)
 }
 
 make_gbm_predictions = function(x, use_obs_model){

### 2017-05-18 figures/plots.R
add plotting script

@@ -1,0 +1,168 @@
+devtools::load_all()
+library(tidyverse)
+library(cowplot)
+settings = yaml::yaml.load_file("settings.yaml")
+
+forecast_results = readRDS("forecast.rds") %>% 
+  filter(!is.na(richness))
+gbm_results = bind_rows(readRDS("gbm_TRUE.rds"),
+                        readRDS("gbm_FALSE.rds"))
+mistnet_results = lapply(dir("mistnet_output/", full.names = TRUE,
+                             pattern = "TRUE|FALSE"),
+                         readRDS) 
+
+
+mean(map_dbl(mistnet_results, ~mean(.x$mean - .x$richness)))
+
+
+average_results = bind_rows(readRDS("avg_TRUE.rds"), readRDS("avg_FALSE.rds"))
+
+my_var = function(x){
+  ifelse(length(x) == 1, 0, var(x))
+}
+
+
+rf_results = bind_rows(readRDS("rf_predictions/all_FALSE.rds"),
+                       readRDS("rf_predictions/all_TRUE.rds"))
+
+bound = bind_rows(forecast_results, gbm_results, average_results, 
+                  rf_results, mistnet_results) %>% 
+  group_by(site_id, year, richness, model, use_obs_model) %>% 
+  summarize(sd = sqrt(mean(sd^2 + my_var(mean))), mean = mean(mean)) %>% 
+  mutate(diff = richness - mean, z = diff / sd,
+         p = pnorm(richness, mean, sd),
+         deviance = -2 * dnorm(richness, mean, sd, log = TRUE))
+
+
+for_violins = bound %>% 
+  group_by(site_id, year, use_obs_model, model) %>% 
+  summarize(mse = Metrics::mse(richness, mean), mean_deviance = mean(deviance)) %>% 
+  arrange(use_obs_model) %>% 
+  group_by(site_id, year, model) %>% 
+  summarize(n = n(), diff_mse = mse[2] - mse[1], diff_deviance = mean_deviance[2] - mean_deviance[1])
+
+for_violins %>% 
+  ggplot(aes(x = model, y = diff_mse)) + 
+  geom_violin(fill = 2, size = 0, alpha = .5, adjust = 2) +
+  geom_hline(yintercept = 0, color = alpha(1, .25)) +
+  stat_summary(fun.data = "mean_cl_boot", colour = "darkred", geom = "point") +
[…]

### 2017-05-18 R/forecast-bbs-core.R
add imports, update namespace

@@ -270,7 +270,7 @@ get_richness_ts_env_data <- function(start_yr, end_yr, min_num_yrs){
 #' Replace species-level information with richness values, eliminating redundant rows
 #' @param df a data frame, such as produced by get_bbs_data or get_pop_ts_env_data
 #' @export
-#' @importFrom dplyr n
+#' @importFrom dplyr n group_by mutate select ungroup distinct right_join
 collapse_to_richness = function(df){
 
   # Code below assumes that NA abundance always means no observations for the

### 2017-05-18 R/get_prism_data.R
correct offset details

@@ -131,8 +131,8 @@ process_bioclim_data=function(){
   prism_bbs_data=get_prism_data()
 
   #Offset the year by 6 months so that the window for calculating bioclim variables
-  #will be July 1 - July 31. See https://github.com/weecology/bbs-forecasting/issues/114
-  prism_bbs_data$year = with(prism_bbs_data, ifelse(month %in% 6:12, year+1, year))
+  #will be July 1 - June 30. See https://github.com/weecology/bbs-forecasting/issues/114
+  prism_bbs_data$year = with(prism_bbs_data, ifelse(month %in% 7:12, year+1, year))
   
   #Spread out the climate variables ppt, tmean, etc into columns
   prism_bbs_data = prism_bbs_data %>%

### 2017-05-18 manuscript/draft.Rmd
preliminary results section

@@
 
+The mixed model (which formed the basis of our observer-corrected "average"
+predictions) found that 70% of the variance in richness in the training set
+could be explained by differences among sites, and 21% could be explained by
+differences among observers. The remaining 9% represents residual variation,
+where a given observer might report a different number of species on different
+runs of the same transect. We found remarkably little autocorrelation in these
+residuals ([[about 0.04, +/- something]]), suggesting richness is close to
+stationary on these time scales, once observer differences are accounted for.
+
+The substantial uncertainty associated with differences among observers caused
+reliable improvements in both mean-squared-error and deviance when observer
+effects were controlled for (Figure 4). This improvement was smallest for the
+"Naive" model, as discussed below.
+
+In general, each model's error increased more-or-less linearly with time,
+although the slopes and initial error rates varied across models (Figure 3). The
+site-level models (i.e. the baselines and the ARIMA methods) tended to do very
+well (Figure 3), especially in terms of RMSE and especially on short time
+horizons. For most combinations of year and metric, the best-performing model
+was one of the two baselines. Usually, the "naive" model had the lowest error
+when observer effects were not explicitly accounted for, and the "average" model
+had the lowest error when they were. We discuss the reasons for this difference
+in the next section.
+
+Our two error metrics produced qualitatively similar results in most cases, with
+the notable exception of the stacked SDMs (Figure 3A and 3B). Deviance, unlike
+RMSE, depends on the full predictive distribution of a model, including the
+model's estimated uncertainty. As noted elsewhere [@harris2015], these models
+are generally far too confident, with observed richness values frequently
+falling well outside their confidence limits (Figure 3C). As a result, the
+stacked SDMs' deviance was dramatically higher than the next-worst model even
+though its RMSE was not unusual. On the other hand, the richness-GBM model and
+especially the mistnet JSDM had reasonably well-calibrated uncertainty
+estimates, even in the presence of uncorrected observer biases (Figure 3C). This
+prevented their deviance from rising as quickly as the site-level models when
+the observer population changed (Figure 3B).
+
+In general, the models produced confidence intervals that were too narrow,
+indicating overconfident predictions. Of these, the random forest-based SDM
+stack was the most overconfident, as discussed above. As one might expect,
+correcting for observer differences improved the coverage of most models' 95%
+intervals. Interestingly, this effect was very small for mistnet, which seems to
+indicate that its latent variables were already accounting for observer-based
+uncertainty. The "naive" model was the only model whose confidence intervals
+were too wide (Figure 3), which can be attributed to the rapid rate at which
+these intervals expand as the time lag increases (Figure 1).
+
 ## Discussion

### 2017-05-18 manuscript/draft.Rmd
numbers for mistnet coverage

@@
 correcting for observer differences improved the coverage of most models' 95%
-intervals. Interestingly, this effect was very small for mistnet, which seems to
-indicate that its latent variables were already accounting for observer-based
-uncertainty. The "naive" model was the only model whose confidence intervals
-were too wide (Figure 3), which can be attributed to the rapid rate at which
+intervals. Interestingly, this effect was very small for mistnet (its
+coverage only improved from 92% to 93%), which seems to indicate that its
+latent variables were already accounting for observer-related uncertainty. 
+The "naive" model was the only model whose confidence intervals were too wide
+(Figure 3), which can be attributed to the rapid rate at which
 these intervals expand as the time lag increases (Figure 1).

### 2017-05-19 manuscript/draft.Rmd
some discussion edits

--- 2017-05-17 manuscript/draft.Rmd
@@
 forecasts. If a site is predicted to have a high number of species in the
-future, relative to other sites, it generally does. However, within a site the
-forecasts perform relatively poorly. If a site is predicted to have a higher
-richness next year than it did last year there is little confidence that this
-will be the case. This results in simple site-level time-series and baseline
-models like the long-term average and the naive model being some of the best
-performing overall models for future biodiversity.
+future, relative to other sites, it generally does. However, the forecasts
+perform relatively poorly at predicting variation in richness over time for any
+given site. If a site is predicted to have a higher richness next year than it
+did last year there is little confidence that this will be the case. This
+results in simple site-level time-series, which are best predicted by simple
+baselines like the long-term average and the "naive" model.
 
-The most commonly used method to forecast the future state of biodiversity,
-fitting a species distribution model separately to each species and then summing
-the probability of presence across those models, provides the worst overall
-forecasts of all methods evaluated. Most forecasting based on this approach is
-never evaluated and when it is it is typically only assessed based on cross-site
-rather than within site approaches. As a result the forecasts may appear to
-perform well even when change through time is not accurately predicted. This
-issue has also recently been recently highlighted as an issue for using SDMs to
-predict the future locations of individual species [plos one 2012 paper]. This
-result is particularly sobering because this method forms the foundation for
-major claims regarding the predicted loss of biodiversity to climate change
-[2004 science/nature papers]. The poor performance of SDM-based forecasts
-highlights the crucial importance of comparing multiple modeling approaches when
-conducting ecological forecasts, and in particular the value of comparing
-results to simple baselines to avoid over-interpreting the accuracy of
-ecological forecasts. Disciplines that have more mature forecasting cultures
-often do this by reporting "forecast skill", which is the improvement in the
-forecast relative to a simple baseline. We recommend adopting this approach in
-future ecological forecasting research. It also highlights the need for using
-hindcasting and time-series data for assessing the effectiveness of forecasting
-methods, since this allows comparisons to baselines and assessment of the
-predicted dynamics at each site.
+While stacked species distribution models are the most commonly used approach
+for forecasting the future state of biodiversity [[cite?]], this approach
+provided the worst overall forecasts of all methods evaluated. In fact, the most
+common version of the stacking approach (based on rounding species' occurrence
+probabilities up or down before adding them up) would have led to even worse
+results than what we presented [Pellissier et al. 2013's "probabilistic
+approach" paper, Calabese et al. 2014]. In general, SDM-based forecasts are
+rarely evaluated, particularly in terms of repeated observations at the same
+site, and especially with long time horizons. Since site-level richness is
+relatively stable, the model evaluations that show high accuracy in predicting
+differences among sites may not indicate much about the model's ability to
+predict changes over time. This issue has also recently been recently
+highlighted as an issue for using SDMs to predict the future locations of
+individual species [plos one 2012 paper]. This result is particularly sobering
+because this approach forms the foundation for major claims regarding the
+predicted loss of biodiversity to climate change [2004 science/nature papers].
+The poor performance of SDM-based forecasts highlights the crucial importance of
+comparing multiple modeling approaches when conducting ecological forecasts, and
+in particular the value of comparing results to simple baselines to avoid
+over-interpreting the accuracy of ecological forecasts. Disciplines that have
+more mature forecasting cultures often do this by reporting "forecast skill",
+which is the improvement in the forecast relative to a simple baseline. We
+recommend adopting this approach in future ecological forecasting research. It
+also highlights the need for using hindcasting and time-series data for
+assessing the effectiveness of forecasting methods, since this allows
+comparisons to baselines and assessment of the predicted dynamics at each site.
 
 The most common alternative to summed SDMs is modeling richness directly using
-environmental variables and using the resulting models to make forecasts [].
-Two comparisons of the the SDM and direct richness modeling approaches reported
-that the methods yielded equivalent results for forecasting diversity. While our
-results generally support this for point estimates of richness they also show
-that modeling richness directly results in much better estimation of the
-uncertainty of the forecast and therefore is a better overall approach to
-forecasting richness. A similar result is seen when comparing joint species
-distribution models (JSDMs) to stacked single species distribution models. For
-point estimates the joint distribution models are roughly equivalent to stacked
-SDMs, but the JSDMs provide much better estimates of uncertainty. In fact, JSDMs
-and modeling richness directly provide some of the best estimates of uncertainty
-across all modeling approaches. This highlights the importance evaluating models
-based on their uncertainty as well as their mean estimates. Not doing so is
-particularly problematic in this case because the summed species distribution
-models are overly confident meaning that using them would lead to believing that
-richness would be restricted to a much narrower range than would actually
-occur.
+environmental variables and using the resulting models to make forecasts []. Two
+comparisons of the the SDM and direct richness modeling approaches reported that
+the methods yielded equivalent results for forecasting diversity (some other
+comparisons that found poor SDM performance were [[tainted]] by the rounding
+error alluded to above and are not considered here). While our results generally
+support rough equivalence between stacked SDMs and direct richness modeling for
+point estimates, they also show that modeling richness directly
+results in much better estimation of the uncertainty of the forecast and
+therefore is a better overall approach to forecasting richness. A similar result
+is seen when comparing joint species distribution models (JSDMs) to stacked
+single species distribution models. For point estimates the joint distribution
+models are roughly equivalent to stacked SDMs, but the JSDMs provide much better
+estimates of uncertainty. In fact, JSDMs and modeling richness directly provide
+some of the best estimates of uncertainty across all modeling approaches. This
+highlights the importance evaluating models based on their uncertainty as well
+as their mean estimates. Not doing so is particularly problematic in this case
+because the summed species distribution models are overly confident meaning that
+using them would lead to believing that richness would be restricted to a much
+narrower range than would actually occur.
 
-The accurate estimates of uncertainty by JSDMs and richness models result in
-these models approaching the performance of time-series models and baselines at
-the longest time-scales assessed when using measures that consider both point
-estimates and uncertainty. This convergence at longer time scales highlights the
-importance of considering how forecasting method performance changes with the
-distance into the future being predicted. The value of this approach, and
-associated forecast horizons, has been raised for knowing how far into the
-future a model can be effectively used for forecasting [@petchy2015], but our
-results suggest a broader value to this approach for considering the potential
-importance of different models and processes for making forecasts at different
-scales. [[Should make sure Petchy et al. don't make this point]] Our results
-show that for forecasts 1-3 years in the future, baselines and time-series based
-approaches outperform all other methods using integrative metrics. At these time
-scales environmental changes are relatively small and changes in the biota may
-lag behind changes in the environment. However, as the timescale of the forecast
-increases to a decade the JSDMs and environmental richness models get closer to
-the accuracy of the time-series methods and even surpass some of those
-methods. As the forecast distance increases the amount of environmental change
-is expected to increase and the system will have more time to respond,
-potentially leading to shifts towards improved relative performance of the
-models incorporating environmental and ecological information. If this shift
-with forecast distance in the relative importance of different processes and
-models continues, this could lead to JSDMs and environmental richness models
-outperforming time-series based approaches at sufficiently long time
-scales. These results suggest that assessment of how forecast performance
-changes with time lag is not only important for determining how far in the
-future to forecast, but also for determining models and processes are most
-relevant for making forecasts at different time scales.
+For error metrics that evaluate both models' point estimates and their
+uncertainty, the joint SDM and the nonlinear richness model began to outperform
+the time-series baselines at the longest time-scales assessed. This convergence
+at longer time scales highlights the importance of considering how forecasting
+method performance changes with the distance into the future being predicted.
+The value of this approach, and associated forecast horizons, has been raised
+for knowing how far into the future a model can be effectively used for
+forecasting [@petchy2015], but our results suggest a broader value to this
+approach for considering the potential importance of different models and
+processes for making forecasts at different scales. [[Should make sure Petchy et
+al. don't make this point]] Our results show that for forecasts 1-3 years in the
+future, baselines and time-series based approaches outperform all other methods
+using integrative metrics. At these time scales, environmental changes are
+relatively small and changes in the biota may lag behind changes in the
+environment. However, as the timescale of the forecast increases to a decade the
+JSDMs and environmental richness models get closer to the accuracy of the
+time-series methods and even surpass some of those methods. As the forecast
+distance increases the amount of environmental change is expected to increase
+and the system will have more time to respond, potentially leading to shifts
+towards improved relative performance of the models incorporating environmental
+and ecological information. If this shift with forecast distance in the relative
+importance of different processes and models continues, this could lead to JSDMs
+and environmental richness models outperforming time-series based approaches at
+sufficiently long time scales. These results suggest that assessment of how
+forecast performance changes with time lag is not only important for determining
+how far in the future to forecast, but also for determining models and processes
+are most relevant for making forecasts at different time scales.
 

### 2017-05-19 R/forecast-bbs-core.R
offset NDVI to match BBS observations

@@ -193,7 +193,11 @@ get_env_data <- function(){
   bioclim_data <- get_bioclim_data()
   elev_data <- get_elev_data()
   ndvi_data_raw <- get_bbs_gimms_ndvi()
-
+  
+  #Offset the NDVI year by 6 months so that the window for will be July 1 - June 30. 
+  #See https://github.com/weecology/bbs-forecasting/issues/114
+  ndvi_data_raw$year = with(ndvi_data_raw, ifelse(month %in% c('jul','aug','sep','oct','nov','dec'), year+1, year))
+  
   ndvi_data_summer <- ndvi_data_raw %>%
     filter(!is.na(ndvi), month %in% c('may', 'jun', 'jul'), year > 1981) %>%
     group_by(site_id, year) %>%

### 2017-06-09 manuscript/draft.Rmd

--- 2017-05-18 manuscript/draft.Rmd
@@
 
-The mixed model (which formed the basis of our observer-corrected "average"
-predictions) found that 70% of the variance in richness in the training set
-could be explained by differences among sites, and 21% could be explained by
-differences among observers. The remaining 9% represents residual variation,
-where a given observer might report a different number of species on different
-runs of the same transect. We found remarkably little autocorrelation in these
-residuals ([[about 0.04, +/- something]]), suggesting richness is close to
-stationary on these time scales, once observer differences are accounted for.
-
-The substantial uncertainty associated with differences among observers caused
-reliable improvements in both mean-squared-error and deviance when observer
-effects were controlled for (Figure 4). This improvement was smallest for the
-"Naive" model, as discussed below.
-
-In general, each model's error increased more-or-less linearly with time,
-although the slopes and initial error rates varied across models (Figure 3). The
-site-level models (i.e. the baselines and the ARIMA methods) tended to do very
-well (Figure 3), especially in terms of RMSE and especially on short time
-horizons. For most combinations of year and metric, the best-performing model
-was one of the two baselines. Usually, the "naive" model had the lowest error
-when observer effects were not explicitly accounted for, and the "average" model
-had the lowest error when they were. We discuss the reasons for this difference
-in the next section.
-
-Our two error metrics produced qualitatively similar results in most cases, with
-the notable exception of the stacked SDMs (Figure 3A and 3B). Deviance, unlike
-RMSE, depends on the full predictive distribution of a model, including the
-model's estimated uncertainty. As noted elsewhere [@harris2015], these models
-are generally far too confident, with observed richness values frequently
-falling well outside their confidence limits (Figure 3C). As a result, the
-stacked SDMs' deviance was dramatically higher than the next-worst model even
-though its RMSE was not unusual. On the other hand, the richness-GBM model and
-especially the mistnet JSDM had reasonably well-calibrated uncertainty
-estimates, even in the presence of uncorrected observer biases (Figure 3C). This
-prevented their deviance from rising as quickly as the site-level models when
-the observer population changed (Figure 3B).
-
-In general, the models produced confidence intervals that were too narrow,
-indicating overconfident predictions. Of these, the random forest-based SDM
-stack was the most overconfident, as discussed above. As one might expect,
-correcting for observer differences improved the coverage of most models' 95%
-intervals. Interestingly, this effect was very small for mistnet, which seems to
-indicate that its latent variables were already accounting for observer-based
-uncertainty. The "naive" model was the only model whose confidence intervals
-were too wide (Figure 3), which can be attributed to the rapid rate at which
-these intervals expand as the time lag increases (Figure 1).
-
 ## Discussion
@@
 
+Here we have made forecasts using yearly timesteps fifty (?) years into the
+future. Many ecological forecasts are temporally aggregated to 5-30 year
+timesteps and projected up to a century into the future. These are commonly
+beyond the career or even lifespan of the researchers, making reasonable model
+assessment impossible. Currently assessments at these large scales can be only
+be made with a small number of opportune datasets. For examples hindcasts have
+been used to asses models of species richness [@alger2009, distler2015] and
+distribution [@rappacciuolo2012, @moran-ordonez2017, @araujo2005,
+@eskildsen2013] using data aggregated across similar time scales. More
+evaluation studies like these are needed but there is a paucity of adequate long
+term data available. Short term forecasts will allow for relatively rapid
+validation of models by researchers without needing to wait for enough data to
+accumulate. This is not to say they should replace large scale long term
+forecasts. The relationships of temporal scaling is currently understudied, but
+it is possible they are similar to spatial scale dynamics [@mcgill2010,
+@levin1992]. For example drivers of species richness differ at different spatial
+scales [@rahbek2001, @hurlbert2003]. Thus short term forecasts should be
+expected to add value to end users and inform, not replace, long term larger
+scaled forecasts.
+
 Future efforts will also need to begin to address the additional uncertainty

### 2017-06-13 R/forecast-bbs-core.R
filter years by % training data available (fixes #123)

@@ -70,7 +70,7 @@ filter_species <- function(df){
   species_table = get_species_data()
 
   is_unidentified = function(names) {
-    #Befor filter account for this one hybrid of 2 subspecies so it's kept
+    #Before filtering, account for this one hybrid of 2 subspecies so it's kept
     names[names=='auratus auratus x auratus cafer']='auratus auratus'
     grepl('sp\\.| x |\\/', names)
   }
@@ -224,27 +224,30 @@ get_env_data <- function(){
 
 #' Get BBS population time-series data with environmental variables
 #'
-#' Selects sites with data spanning 1982 through 2013 containing at least 25
-#' samples during that period.
+#' Selects data from the subset of sites that were visited in at least 
+#' `min_year_percentage` percent of years in the training set, and from
+#' all years between `start_yr` to `end_yr`.
 #'
-#' Attaches associated environmental data
+#' Also attaches associated environmental data
 #'
 #' @param start_yr num first year of time-series
 #' @param end_yr num last year of time-series
-#' @param min_num_yrs num minimum number of years of data between start_yr & end_yr
+#' @param last_train_year num last year of the time-series' training portion
+#' @param min_year_percentage num percent of years between of data between 
+#'         start_yr & last_train_year
 #'
 #' @return dataframe with site_id, lat, long, year, species_id, and abundance
 #' @export
 #' @importFrom dplyr "%>%" filter select
-get_pop_ts_env_data <- function(start_yr, end_yr, min_num_yrs){
+get_pop_ts_env_data <- function(start_yr, end_yr, last_train_year, 
+                                min_year_percentage){
   pop_ts_env_data = get_bbs_data() %>%
-    filter_ts(start_yr, end_yr, min_num_yrs) %>%
+    filter_ts(start_yr, end_yr, last_train_year, min_year_percentage) %>%
     complete(site_id, year) %>% 
     add_env_data() %>%
     filter(!is.na(bio1), !is.na(ndvi_sum), !is.na(elevs)) %>%
-    group_by(site_id) %>%
     #Filter min_num_years again after accounting for missing environmental data
-    filter(length(unique(year)) >= min_num_yrs) %>%
+    filter_ts(start_yr, end_yr, last_train_year, min_year_percentage) %>%
     ungroup() %>%
     add_observers()
 
@@ -258,14 +261,17 @@ get_pop_ts_env_data <- function(start_yr, end_yr, min_num_yrs){
 #'
 #' @param start_yr num first year of time-series
[…]

### 2017-06-13 R/forecast-bbs-core.R
fix minor problems Ethan noted

@@ -233,7 +233,7 @@ get_env_data <- function(){
 #' @param start_yr num first year of time-series
 #' @param end_yr num last year of time-series
 #' @param last_train_year num last year of the time-series' training portion
-#' @param min_year_percentage num percent of years between of data between 
+#' @param min_year_percentage num percent of years of non-missing data between 
 #'         start_yr & last_train_year
 #'
 #' @return dataframe with site_id, lat, long, year, species_id, and abundance
@@ -351,7 +351,6 @@ add_observers = function(bbs_data) {
 filter_ts <- function(bbs_data, start_yr, end_yr, last_train_year, 
                       min_year_percentage){
   min_num_years = (last_train_year - start_yr + 1) * min_year_percentage / 100
-  browser()
   sites_to_keep = bbs_data %>%
     filter(year >= start_yr, year <= last_train_year) %>%
     group_by(site_id) %>%

### 2017-06-19 R/forecast-bbs-core.R
typos

@@ -235,7 +235,7 @@ get_env_data <- function(timeframe = 'past'){
   #If projecting forward, use NDVI averages for future NDVI values, and use
   #CMIP5 data for bioclim values
   if(timeframe == 'future') {
-    ndvi_long_term_averges = ndvi_data %>%
+    ndvi_long_term_averages = ndvi_data %>%
       filter(year %in% 2000:2013) %>%
       gather(season,value, -site_id, -year) %>%
       group_by(site_id, season) %>%
@@ -245,7 +245,7 @@ get_env_data <- function(timeframe = 'past'){
     all_sites = unique(ndvi_data$site_id)
     
     ndvi_data = expand.grid(site_id = all_sites, year=2014:2050) %>%
-      left_join(ndvi_long_term_averges, by='site_id') %>%
+      left_join(ndvi_long_term_averages, by='site_id') %>%
       spread(season, value)
     
     bioclim_data = get_bioclim_data(source='cmip5') %>%

### 2017-06-20 R/post-observer.R
Change namespace of calls to by_slice; delete vestigial file

@@ -64,9 +64,7 @@ make_all_forecasts = function(x, fun_name, use_obs_model,
     forecast_data = filter(forecast_data, iteration == 1)
   }
   
-  # TODO: by_slice will be deprecated. See if I can use mutate_all instead?
-  # Dropping x_richness$sd before joining so it can be replaced by forecast sd.
-  out = by_slice(forecast_data, make_forecast, fun_name = fun_name, 
+  out = purrrlyr::by_slice(forecast_data, make_forecast, fun_name = fun_name, 
                  use_obs_model = use_obs_model, settings = settings, ...,
                  .collate = "row") %>%
     left_join(select(x_richness, -sd), c("site_id", "year", "iteration"))

### 2017-06-20 analysis/rf.R

@@ -1,131 +1,0 @@
-library(rpart)
-library(tidyverse)
-devtools::load_all()
-
-settings = yaml::yaml.load_file("settings.yaml")
-iters = 1:200 # During interactive use/testing, only include a few MCMC samples
-
-
-obs_model = readRDS("observer_model.rds")
-x_richness = obs_model$data %>% 
-  mutate(intercept = c(obs_model$intercept[iteration]), 
-         sd = c(obs_model$sigma[iteration]),
-         expected_richness = richness - observer_effect) %>% 
-  filter(iteration %in% iters)
-
-bbs = get_bbs_data() %>% 
-  select(site_id, year, species_id) %>% 
-  mutate(presence = 1)
-
-
-sdm_tree = function(spid, iter, obs_model){
-  d = x_richness %>% 
-    filter(iteration == iter) %>% 
-    left_join(filter(bbs, species_id == spid), c("year", "site_id")) %>% 
-    mutate(presence = ifelse(is.na(presence), 0, 1))
-  
-  vars = c(settings$vars, "presence", if (obs_model) {"observer_effect"})
-  
-  boot_rows = sample.int(sum(d$in_train), replace = TRUE)
-  non_boot_rows = seq(1, sum(d$in_train)) %>% discard(~.x %in% boot_rows)
-  
-  train = d %>% filter(in_train) %>% .[boot_rows, vars] # Bootstrap resample
-  oob = d %>% filter(in_train) %>% .[non_boot_rows, vars] # Out of bag
-  test = d %>% filter(!in_train)
-  
-  if (var(train$presence) == 0) {
-    # if presence is constant, return that constant
-    p = rep(train$presence[1], nrow(test))
-  } else{
-    t = rpart(presence ~ ., data = train[boot_rows, ], method = "class",
-              maxcompete = 0, maxsurrogate = 0, cp = 0.00001, minbucket = 25,
-              xval = 0)
-    p = predict(t, test)[,2]
-  }
-  
-  data_frame(
-    iteration = iter,
-    species_id = spid,
-    site_id = test$site_id,
-    year = test$year,
-    obs_model = obs_model,
-    p = p
-  )
-}
-
-
-sdm = function(iters, spid, obs_model) {
-  map(iters, ~sdm_tree(spid = spid, iter = .x, obs_model = TRUE)) %>% 
-    bind_rows() %>% 
[…]

### 2017-06-20 R/forecast-bbs-core.R
minor pipeline fixes

@@ -77,19 +77,18 @@ filter_species <- function(df){
 
   valid_taxa = species_table %>%
     filter(!is_unidentified(species)) %>%
-    filter(AOU > 2880) %>%
-    filter(AOU < 3650 | AOU > 3810) %>%
-    filter(AOU < 3900 | AOU > 3910) %>%
-    filter(AOU < 4160 | AOU > 4210) %>%
-    filter(AOU != 7010)
+    filter(aou > 2880) %>%
+    filter(aou < 3650 | aou > 3810) %>%
+    filter(aou < 3900 | aou > 3910) %>%
+    filter(aou < 4160 | aou > 4210) %>%
+    filter(aou != 7010)
 
-  filter(df, species_id %in% valid_taxa$AOU)
+  filter(df, species_id %in% valid_taxa$aou)
 }
 
 #' Combine subspecies into their common species
 #'
-#' @importFrom dplyr "%>%" filter slice group_by summarise ungroup
-#' @importFrom magrittr extract2
+#' @importFrom dplyr "%>%" filter slice group_by summarise ungroup pull
 #' @importFrom stringr word
 combine_subspecies = function(df){
 
@@ -97,20 +96,20 @@ combine_subspecies = function(df){
 
   # Subspecies have two spaces separated by non-spaces
   subspecies_names = species_table %>%
-    filter(species_table$AOU %in% unique(df$species_id)) %>%
-    magrittr::extract2("spanish_common_name") %>%
+    filter(aou %in% unique(df$species_id)) %>%
+    pull(spanish_common_name) %>%
     grep(" [^ ]+ ", ., value = TRUE)
 
   subspecies_ids = species_table %>%
     filter(spanish_common_name %in% subspecies_names) %>%
-    extract2("AOU")
+    pull(aou)
 
   # Drop all but the first two words to get the root species name,
   # then find the AOU code
   new_subspecies_ids = species_table %>%
     slice(match(word(subspecies_names, 1,2),
                 species_table$spanish_common_name)) %>%
-    extract2("AOU")
+    pull(aou)
 
   # replace the full subspecies names with species-level names
   for (i in seq_along(subspecies_ids)) {

### 2017-06-20 manuscript/draft.Rmd

@@
 given site. If a site is predicted to have a higher richness next year than it
-did last year there is little confidence that this will be the case. This
-results in simple site-level time-series, which are best predicted by simple
-baselines like the long-term average and the "naive" model.
+did last year there is little confidence that this will be the case. As a
+result, the baseline models, like the long-term average and the "naive" model,
+along with site-level time-series models, provided some of the best forecasts
+for future biodiversity.
 
-While stacked species distribution models are the most commonly used approach
-for forecasting the future state of biodiversity [[cite?]], this approach
-provided the worst overall forecasts of all methods evaluated. In fact, the most
-common version of the stacking approach (based on rounding species' occurrence
-probabilities up or down before adding them up) would have led to even worse
-results than what we presented [Pellissier et al. 2013's "probabilistic
-approach" paper, Calabese et al. 2014]. In general, SDM-based forecasts are
-rarely evaluated, particularly in terms of repeated observations at the same
-site, and especially with long time horizons. Since site-level richness is
-relatively stable, the model evaluations that show high accuracy in predicting
-differences among sites may not indicate much about the model's ability to
-predict changes over time. This issue has also recently been recently
-highlighted as an issue for using SDMs to predict the future locations of
-individual species [plos one 2012 paper]. This result is particularly sobering
-because this approach forms the foundation for major claims regarding the
-predicted loss of biodiversity to climate change [2004 science/nature papers].
-The poor performance of SDM-based forecasts highlights the crucial importance of
-comparing multiple modeling approaches when conducting ecological forecasts, and
-in particular the value of comparing results to simple baselines to avoid
-over-interpreting the accuracy of ecological forecasts. Disciplines that have
-more mature forecasting cultures often do this by reporting "forecast skill",
-which is the improvement in the forecast relative to a simple baseline. We
-recommend adopting this approach in future ecological forecasting research. It
-also highlights the need for using hindcasting and time-series data for
-assessing the effectiveness of forecasting methods, since this allows
-comparisons to baselines and assessment of the predicted dynamics at each site.
+The most commonly used method to forecast future biodiversity, stacked SDMS
+[[cite?]], provided the worst overall forecasts of all methods evaluated. In
+general, SDM-based forecasts are rarely evaluated, particularly in terms of
+repeated observations at the same site, and especially with long time horizons.
+Since site-level richness is relatively stable, the model evaluations that show
+high accuracy in predicting differences among sites may not indicate much about
+the model's ability to predict changes over time. This issue has also recently
+been recently highlighted as an issue for using SDMs to predict the future
+locations of individual species [plos one 2012 paper]. This result is
+particularly sobering because this approach forms the foundation for major
+claims regarding the predicted loss of biodiversity to climate change [2004
+science/nature papers]. The poor performance of SDM-based forecasts highlights
+the crucial importance of comparing multiple modeling approaches when conducting
+ecological forecasts, and in particular the value of comparing results to simple
+baselines to avoid over-interpreting the accuracy of ecological forecasts.
+Disciplines that have more mature forecasting cultures often do this by
+reporting "forecast skill", which is the improvement in the forecast relative to
+a simple baseline. We recommend adopting this approach in future ecological
+forecasting research. It also highlights the need for using hindcasting and
+time-series data for assessing the effectiveness of forecasting methods, since
+this allows comparisons to baselines and assessment of the predicted dynamics at
+each site.
 
@@
 comparisons of the the SDM and direct richness modeling approaches reported that
-the methods yielded equivalent results for forecasting diversity (some other
-comparisons that found poor SDM performance were [[tainted]] by the rounding
-error alluded to above and are not considered here). While our results generally
-support rough equivalence between stacked SDMs and direct richness modeling for
-point estimates, they also show that modeling richness directly
-results in much better estimation of the uncertainty of the forecast and
-therefore is a better overall approach to forecasting richness. A similar result
-is seen when comparing joint species distribution models (JSDMs) to stacked
-single species distribution models. For point estimates the joint distribution
-models are roughly equivalent to stacked SDMs, but the JSDMs provide much better
-estimates of uncertainty. In fact, JSDMs and modeling richness directly provide
-some of the best estimates of uncertainty across all modeling approaches. This
-highlights the importance evaluating models based on their uncertainty as well
-as their mean estimates. Not doing so is particularly problematic in this case
-because the summed species distribution models are overly confident meaning that
-using them would lead to believing that richness would be restricted to a much
-narrower range than would actually occur.
+the methods yielded equivalent results for forecasting diversity. While our
+results generally support rough equivalence between stacked SDMs and direct
+richness modeling for point estimates, they also show that modeling richness
+directly results in much better estimation of the uncertainty of the forecast
+and therefore is a better overall approach to forecasting richness. A similar
+result is seen when comparing joint species distribution models (JSDMs) to
+stacked single species distribution models. For point estimates the joint
+distribution models are roughly equivalent to stacked SDMs, but the JSDMs
+provide much better estimates of uncertainty. In fact, JSDMs and modeling
+richness directly provide some of the best estimates of uncertainty across all
+modeling approaches. This highlights the importance evaluating models based on
+their uncertainty as well as their mean estimates. Not doing so is particularly
+problematic in this case because the summed species distribution models are
+overly confident meaning that using them would lead to believing that richness
+would be restricted to a much narrower range than would actually occur.
 

### 2017-06-20 mistnet.R
fix subtle non-standard evaluation issue

@@ -33,7 +33,8 @@ fit_mistnet = function(iter,
   # Collect data for this iteration, then discard the rest to save
   # memory
   obs_model = readRDS("observer_model.rds")
-  data = filter(obs_model$data, iteration == iter) %>% 
+  data = obs_model$data %>% 
+    filter(iteration == iter) %>% 
     left_join(bbs, c("site_id", "year"))
   rm(obs_model)
   gc(TRUE)

### 2017-07-05 manuscript/draft.Rmd

@@
 ## Results
+
+The mixed model (which formed the basis of our observer-corrected "average"
+predictions) found that 70% of the variance in richness in the training set
+could be explained by differences among sites, and 21% could be explained by
+differences among observers. The remaining 9% represents residual variation,
+where a given observer might report a different number of species on different
+runs of the same transect. We found remarkably little autocorrelation in these
+residuals ([[about 0.04, +/- something]]), suggesting richness is close to
+stationary on these time scales, once observer differences are accounted for.
+
+The substantial uncertainty associated with differences among observers caused
+reliable improvements in both mean-squared-error and deviance when observer
+effects were controlled for (Figure 4). This improvement was smallest for the
+"Naive" model, as discussed below.
+
+In general, each model's error increased more-or-less linearly with time,
+although the slopes and initial error rates varied across models (Figure 3). The
+site-level models (i.e. the baselines and the ARIMA methods) tended to do very
+well (Figure 3), especially in terms of RMSE and especially on short time
+horizons. For most combinations of year and metric, the best-performing model
+was one of the two baselines. Usually, the "naive" model had the lowest error
+when observer effects were not explicitly accounted for, and the "average" model
+had the lowest error when they were. We discuss the reasons for this difference
+in the next section.
+
+Our two error metrics produced qualitatively similar results in most cases, with
+the notable exception of the stacked SDMs (Figure 3A and 3B). Deviance, unlike
+RMSE, depends on the full predictive distribution of a model, including the
+model's estimated uncertainty. As noted elsewhere [@harris2015], these models
+are generally far too confident, with observed richness values frequently
+falling well outside their confidence limits (Figure 3C). As a result, the
+stacked SDMs' deviance was dramatically higher than the next-worst model even
+though its RMSE was not unusual. On the other hand, the richness-GBM model and
+especially the mistnet JSDM had reasonably well-calibrated uncertainty
+estimates, even in the presence of uncorrected observer biases (Figure 3C). This
+prevented their deviance from rising as quickly as the site-level models when
+the observer population changed (Figure 3B).
+
+In general, the models produced confidence intervals that were too narrow,
+indicating overconfident predictions. Of these, the random forest-based SDM
+stack was the most overconfident, as discussed above. As one might expect,
+correcting for observer differences improved the coverage of most models' 95%
+intervals. Interestingly, this effect was very small for mistnet (its
+coverage only improved from 92% to 93%), which seems to indicate that its
+latent variables were already accounting for observer-related uncertainty. 
+The "naive" model was the only model whose confidence intervals were too wide
+(Figure 3), which can be attributed to the rapid rate at which
+these intervals expand as the time lag increases (Figure 1).
 

### 2017-07-06 R/post-observer.R
Add future predictions for GBM & RF

@@ -1,3 +1,27 @@
+iter_rep = function(df, .f, ...) {
+  # Get all distinct non-future iteration numbers
+  iters = df %>% 
+    ungroup() %>% 
+    filter(!is_future) %>% 
+    distinct(iteration) %>% 
+    pull(iteration)
+  
+  parallel::mclapply(
+    iters,
+    function(i){
+      # Run the function on the target iteration, retaining "future" rows for
+      # prediction. Finally, mark everything (including "future" predictions)
+      # with the iteration number.
+      df %>% 
+        filter(iteration == !!i | is_future) %>% 
+        .f(...) %>% 
+        mutate(iteration = !!i)
+    },
+    mc.cores = 8,
+    mc.preschedule = FALSE
+  )
+}
+
 #' @importFrom forecast Arima auto.arima
 make_forecast = function(x, fun_name, use_obs_model, settings, ...){
   # `Forecast` functions want NAs for missing years, & want the years in order
@@ -107,8 +131,8 @@ make_gbm_predictions = function(x, use_obs_model){
   
   cbind(test, mean = mean, model = "richness_gbm",  use_obs_model = use_obs_model, 
         stringsAsFactors = FALSE) %>% 
-    select(site_id, year, mean, richness, model, use_obs_model) %>% 
-    mutate(sd = sd)
+    select(site_id, year, mean, richness, model, use_obs_model, is_future) %>% 
+    mutate(sd = !!sd)
 }
 
 

### 2017-07-06 R/sdm-rf.R

@@ -18,9 +18,10 @@ one_rf_tree = function(bbs, vars, sp_id, iter, use_obs_model, x_richness,
     ntree = 1
   )  
   
-  test = filter(d, !in_train, iteration == iter)
+  test = filter(d, !in_train, iteration == iter | is_future)
   test$mean = predict(rf, test, type = "prob")[,2]
   test$use_obs_model = use_obs_model
+  test$iteration = iter
   
   select(test, site_id, year, species_id, mean, richness, use_obs_model, iteration)
 }

### 2017-07-06 analysis/fit-predict.R

@@ -24,7 +24,8 @@ bioclim_to_discard = colnames(obs_model$data) %>%
 x_richness = obs_model$data %>% 
   mutate(intercept = c(obs_model$intercept[iteration]), 
          sd = c(obs_model$sigma[iteration]),
-         expected_richness = richness - observer_effect) %>% 
+         expected_richness = richness - observer_effect,
+         is_future = FALSE) %>% 
   select(-ndvi_ann, -lat, -long, -site_index, -one_of(bioclim_to_discard))
 
 # Reclaim memory
@@ -32,7 +33,12 @@ rm(obs_model)
 gc()
 
 future = get_env_data(timeframe = "future") %>% 
-  filter(site_id %in% !!x_richness$site_id)
+  filter(site_id %in% !!x_richness$site_id) %>% 
+  select(which(colnames(.) %in% !!colnames(x_richness))) %>% 
+  mutate(richness = 0, iteration = 0,
+         observer_effect = 0, is_future = TRUE)
+
+x_richness = bind_rows(x_richness, future)
 
 # Fit & save models --------------------------------------------------------
 
@@ -89,18 +95,14 @@ expand.grid(fun_name = c("naive", "auto.arima"),
   saveRDS(file = "forecast.rds")
 
 # GBM richness with observer effects:
-x_richness %>%
-  group_by(iteration) %>%
-  purrrlyr::by_slice(make_gbm_predictions, use_obs_model = TRUE, .collate = "rows") %>% 
+iter_rep(x_richness, make_gbm_predictions, use_obs_model = TRUE) %>% 
+  bind_rows() %>% 
   saveRDS(file = "gbm_TRUE.rds")
 
 # GBM richness without observer effects:
-# Don't need to group/by_slice because only fitting one iteration
-x_richness %>%
-  filter(iteration == 1) %>%
-  make_gbm_predictions(use_obs_model = FALSE) %>% 
-  saveRDS(file = "gbm_FALSE.rds")
-
+iter_rep(x_richness, make_gbm_predictions, use_obs_model = FALSE) %>% 
+  bind_rows() %>% 
+  saveRDS(file = "gbm_TRUE.rds")
 
 # fit random forest SDMs -------------------------------------------------------
 

### 2017-07-06 R/post-observer.R
commit the "forecast" changes

@@ -35,7 +35,7 @@ make_forecast = function(x, fun_name, use_obs_model, settings, ...){
     response_variable = "richness"
   }
   
-  h = settings$end_yr - settings$last_train_year
+  h = settings$future_year - settings$last_train_year
   
   # Set the `level` so that `upper` and `lower` are 2 sd apart.
   level = pnorm(0.5)

### 2017-07-06 manuscript/draft.Rmd
Edit and add refs to Introduction

@@
 ---
+title: Using best practices for forecasting biodiversity changes in breeding birds
+author:
+- David J. Harris
+- Shawn D. Taylor
+- Ethan P. White
 output:
@@
 ---
-# Moving towards large-scale data-intensive forecasting of species richness in breeding birds
-# Using best practices for forecasting biodiversity changes in breeding birds
-Authors: Harris, Taylor and White?
 
@@
 planning, management, and evaluating how well ecological models capture the key
-ecological processes governing a system [@clark2001, @dietz2017]. One area of
-particular interest for forecasting is biodiversity since it is important for
-ecosystem function, central to conservation planning, and expected to be
-influenced by anthropogenic global change. Anticipating potential changes in
-biodiversity is crucial for large scale management and conservation, and for
-addressing debates regarding whether local scale diversity is declining and may
-continue to decline in response to anthropogenic influences [@dornelas2014].
+ecological processes governing a system [@clark2001; @houlahan2017;
+@dietz2017]. One area of particular interest for forecasting is biodiversity
+since it is important for ecosystem function, central to conservation planning,
+and expected to be influenced by anthropogenic global change [@cardinale2012;
+@diaz2015; @tilman2017]. Anticipating potential changes in biodiversity is
+crucial for large scale management and conservation, and for addressing debates
+regarding whether local scale diversity is declining and may continue to decline
+in response to anthropogenic influences [@diaz2015; @dornelas2014].
 
@@
 to quantify the spatial relationships between the occurrence of individual
-species and the environment. Forecasts for future environmental conditions are
-then used to predict where each species will occur in the future and the
-probabilities of occurrence are summed across species to predict future species
-richness [e.g., @thomas2004]. Alternatively, models that directly relate spatial
-patterns of species richness to the environment have been developed and
-generally perform equivalently to species distribution modeling based methods
-[@algar2009, @distler2015].
+species and the environment [@thomas2004; @urban2015]. Forecasts
+for future environmental conditions are then used to predict where each species
+will occur in the future and the probabilities of occurrence are summed across
+species to predict future species richness [e.g., @thuiller2005]. Alternatively,
+models that directly relate spatial patterns of species richness to the
+environment have been developed and generally perform equivalently to species
+distribution modeling based methods [@algar2009; @distler2015].
 
@@
 ecological forecasts are applied and the resulting difficulty in assessing
-whether the predicted changes occurred. What we do know comes from a small
-number of hindcasting studies, where models are built using data on species
-occurrence and richness from the past and evaluated on their ability to predict
-contemporary patterns [e.g., @algar2009, @distler2015]. These studies are a
-valuable first step, but lack several components that are important for
+whether the predicted changes occurred [@dietzinreview]. What we do know comes
+from a small number of hindcasting studies, where models are built using data on
+species occurrence and richness from the past and evaluated on their ability to
+predict contemporary patterns [e.g., @algar2009; @distler2015]. These studies
+are a valuable first step, but lack several components that are important for
 developing good forecasting models and understanding how accurately these models
@@
 time-series data on bird distributions with monthly time-series of climate data
-and satellite-based remote-sensing. This 30 year time-scale has only recently
-become possible for large-scale time-series based forecasting and allows us to
-model and assess changes a decade or more into the future in the presence of
-shifts in environmental conditions on par with predicted climate change. We
-compare traditional distribution modeling based approaches to spatial models of
-species richness, time-series methods, and simple baselines. All forecasts model
-uncertainty and observation, are evaluated across different time lags using
-hindcasting, and are publicly archived to allow future assessment. We discuss
-the implications of these practices for our understanding of, and confidence in,
-the resulting forecasts, and how we can continue to improve on these approaches
-in the future.
+and satellite-based remote-sensing. This 30 year annual time-scale has only
+recently become possible for large-scale time-series based forecasting and
+allows us to model and assess changes a decade or more into the future in the
+presence of shifts in environmental conditions on par with predicted climate
+change. We compare traditional distribution modeling based approaches to spatial
+models of species richness, time-series methods, and simple baselines. All
+forecasts model uncertainty and observation, are evaluated across different time
+lags using hindcasting, and are publicly archived to allow future assessment. We
+discuss the implications of these practices for our understanding of, and
+confidence in, the resulting forecasts, and how we can continue to build on
+these approaches improve ecological forecasting in the future.
 

### 2017-07-06 R/post-observer.R
Fix forecast models, drop past years from "future

@@ -70,7 +70,8 @@ make_forecast = function(x, fun_name, use_obs_model, settings, ...){
   }
   
   # Distance between `upper` and `lower` is 2 sd, so divide by 2
-  data_frame(year = seq(settings$last_train_year + 1, settings$end_yr), 
+  tibble::data_frame(year = seq(settings$last_train_year + 1, 
+                                settings$future_year), 
          mean = c(fcst$mean), sd = c(fcst$upper - fcst$lower) / 2, 
          model = fun_name, use_obs_model = use_obs_model,
          coef_names = coef_names)
@@ -88,10 +89,15 @@ make_all_forecasts = function(x, fun_name, use_obs_model,
     forecast_data = filter(forecast_data, iteration == 1)
   }
   
+  # Joining by iteration creates NAs for the "future" data, which are all
+  # listed as "iteration 0". Filling those in with `mutate`, keeping in mind
+  # that the "future" observer effect is defined as zero.
   out = purrrlyr::by_slice(forecast_data, make_forecast, fun_name = fun_name, 
                  use_obs_model = use_obs_model, settings = settings, ...,
                  .collate = "row") %>%
-    left_join(select(x_richness, -sd), c("site_id", "year", "iteration"))
+    left_join(select(x_richness, -sd), c("site_id", "year", "iteration")) %>% 
+    mutate(is_future = ifelse(is.na(is_future), TRUE, FALSE),
+           observer_effect = ifelse(is_future, observer_effect, 0))
   
   if (use_obs_model) {
     # Observer effect was subtraced out in make_forcast. Add it back in here.
@@ -99,7 +105,7 @@ make_all_forecasts = function(x, fun_name, use_obs_model,
   }
   
   select(out, site_id, year, mean, sd, iteration, richness, model, 
-         use_obs_model, coef_names)
+         use_obs_model, coef_names, is_future)
 }
 
 make_gbm_predictions = function(x, use_obs_model){

### 2017-07-06 analysis/fit-predict.R

@@ -33,7 +33,8 @@ rm(obs_model)
 gc()
 
 future = get_env_data(timeframe = "future") %>% 
-  filter(site_id %in% !!x_richness$site_id) %>% 
+  filter(site_id %in% !!x_richness$site_id,
+         year > !!settings$end_yr) %>% 
   select(which(colnames(.) %in% !!colnames(x_richness))) %>% 
   mutate(richness = 0, iteration = 0,
          observer_effect = 0, is_future = TRUE)
@@ -82,11 +83,11 @@ x_richness %>%
 expand.grid(fun_name = c("naive", "auto.arima"), 
             use_obs_model = c(TRUE, FALSE),
             stringsAsFactors = FALSE) %>% 
-  transpose() %>% 
+  purrr::transpose() %>% 
   parallel::mclapply(
     function(grid_row){
       do.call(make_all_forecasts, 
-             c(x = list(x_richness), settings = list(settings), grid_row))
+              c(x = list(x_richness), settings = list(settings), grid_row))
     },
     mc.cores = 8, 
     mc.preschedule = FALSE
@@ -132,3 +133,5 @@ rf_sdm_no_obs = rf_predict_richness(bbs = bbs, x_richness = x_richness,
                                     settings = settings, use_obs_model = FALSE,
                                     mc.cores = 8) %>% 
   saveRDS(file = "rf_predictions/all_FALSE.rds")
+
+

### 2017-07-06 R/sdm-rf.R
fix RF

@@ -11,7 +11,7 @@ one_rf_tree = function(bbs, vars, sp_id, iter, use_obs_model, x_richness,
   
   my_formula = as.formula(paste("present ~", 
                                 paste(vars, collapse = "+")))
-   
+  
   rf = randomForest(
     my_formula, 
     filter(d, in_train),
@@ -23,12 +23,16 @@ one_rf_tree = function(bbs, vars, sp_id, iter, use_obs_model, x_richness,
   test$use_obs_model = use_obs_model
   test$iteration = iter
   
-  select(test, site_id, year, species_id, mean, richness, use_obs_model, iteration)
+  select(test, site_id, year, species_id, mean, richness, use_obs_model, 
+         iteration, is_future)
 }
 
 rf_predict_species = function(sp_id, bbs, settings, x_richness, use_obs_model){
   vars = c(settings$vars, if (use_obs_model) {"observer_effect"})
-  iters = sort(unique(x_richness$iteration))
+  iters = x_richness %>% 
+    filter(iteration > 0) %>% 
+    pull(iteration) %>% 
+    unique()
   results = lapply(iters, one_rf_tree,
                    bbs = bbs, sp_id = sp_id, use_obs_model = use_obs_model, 
                    vars = vars,

### 2017-07-07 manuscript/draft.Rmd
Edits to Methods and Results sections
* Expand methods into full paragraph draft and add refs
* Add recommend results changes from #118
* General editorial work on results

@@
 
-* BBS + retriever
-* Predictor variables
-  * Monthly climate data from interpolated weather stations (PRISM), converted
-    to BIOCLIM variables. Many of these variables were tightly correlated, so
-    we only included the 8 variables selected by Harris's (2015) BBS analysis
-  * Weather/climate and NDVI were each calculated separately for each year,
-    rather than averaged over the full multi-decade period (as Worldclim does)
-  * NDVI (GIMMS)
-  * Altitude (SRTM 90m Digital Elevation Database via raster::getData)
-* Data summary: 
-  * Number of sites, species, unique observers
-  * central moments of richness (mean about 51, sd about 12 spp)
-  * The within-site variation was much smaller (sd about 5 spp), making the
-    site-level distributions underdispersed.  
-    * This underdispersion implies
-      that common count models such as binomial, Poisson, negative binomial 
-      won't fit well.
-    * For this reason (and because the time series methods assumed 
-      Gaussian errors), we defined model performance throughout this paper in 
-      terms of Gaussians (e.g. mean squared error), rather than with count
-      distributions.
-        * Some of the modeling approaches (SDMs and JSDMs) did use discrete
-          distributions internally, however.
-* Splitting the data temporally
-  * Our data set began in `r settings$start_yr`, which was the first year that 
-    NDVI data was available from GIMMS
-  * All the models were trained on the data that was collected through 
-    `r settings$last_train_year`
-  * Sites were included in the analysis if there were at least 
-    `r settings$min_num_yrs` years of data available during this period.
-  * The models were tested/evaluated on the data from 
-    `r settings$last_train_year + 1` `r settings$end_yr`.
+Bird species richness was obtained from the Breeding Bird Survey (BBS)
+[@pardieck2017] using the Data Retriever software [@morris2013] and
+rdataretriever R package [@mcglinn2017]. The BBS data was filtered to exclude
+all nocturnal, cepuscular, and aquatic species
+[since these species are not well sampled by BBS methods; @hurlbert2005], as
+well as unidentified species, and hybrids. All data from routes that did not
+meet BBS acceptable criteria were excluded. Only routes within the continental
+United States were used so as to match the climate dataset. The resulting
+dataset included [[N]] species and [[N]] sites. Among sites richness varied from
+[[N]] to [[N]] with an average richness of [[N]] species. Within-site richness
+variation (among years) was small (~5 species).
 
-  Bird species richness was compiled from the Breeding Bird Survey (BBS) [@pardieck2017] obtained using the Data Retriever software [@morris203].
-The BBS data was initially filtered to exclude all nocturnal, cepuscular, and aquatic species (since these species are not well sampled by BBS methods), as well as unidentified species, and hybrids.
-All data from routes that did not meet BBS acceptable criteria were excluded.
-Only routes within the continental U.S.A were used so as to match the climatic dataset.
-  We used the 4 km resolution PRISM data [@prism] to calculate eight bioclimatic variables identified as relevant to bird distributions [@harris2015] for each year in the dataset: mean diurnal range, isothermality, max temperature of the warmest month, mean temperature of the wettest quarter, mean temperature of the driest quarter, precipitation seasonality, precipitation of the wettest quarter, precipitation of the warmest quarter.
-Satellite derived NDVI with an 8 km resolution was from the NDIV3g dataset [@pinzon2014] and was available from the time period 1981-2013.
-Elevation was from the SRTM 90m elevation dataset [@jarvis2008] obtained using the R package raster [@hijmans2014].
-BBS route starting point coordinates were used to extract all environmental data.
-Climatic variables and NDVI were aggregated to the average values within a 40 km radius of the start point, equal to the length of a BBS route.
-NDVI was also subset to average summer (May, June, July) and winter (December, January, Feburary) values.
-  Data from 1982 (the first complete year of NDVI data) to 2003 were used for training models and data from 2004 to 2013 was used for testing.
-Only BBS routes with at least 25 yearly observations in the years 1982-2015 were used.
+Environmental data included a combination of elevation, bioclimatic variables
+and a remotely sensed vegetation index (the normalized difference vegetation
+index; NDVI), all of which are know to influence richness and distribution in
+the BBS data. We used the 4 km resolution PRISM data [@prism] to calculate eight
+bioclimatic variables identified as relevant to bird distributions [@harris2015]
+for each year in the dataset: mean diurnal range, isothermality, max temperature
+of the warmest month, mean temperature of the wettest quarter, mean temperature
+of the driest quarter, precipitation seasonality, precipitation of the wettest
+quarter, precipitation of the warmest quarter. Satellite derived NDVI, a primary
+correlated of richness in BBS data [@hurlbert2002], with an 8 km resolution was
+obtained from the NDIV3g dataset [@pinzon2014] and was available from the time
+period 1981-2013. Average summer (May, June, July) and winter (December,
+January, Feburary) NDVI values were used as predictors. Elevation was from the
+SRTM 90m elevation dataset [@jarvis2008] obtained using the R package raster
+[@hijmans2016]. All environmental variables were aggregated to the average
+values within a 40 km radius (the length of a BBS route) of the starting point
+of each BBS route.
 
-### Accounting for site \& observer effects
-* Observer effects are inherent in large data sets taken by different 
-  observers, and are known to occur in BBS [citation].
-* While these biases are known to exist, they are difficult to quantify: even if
-  the different observers ran the same transects under identical conditions (which 
-  they don't), we would still only have an estimate of their differences, not an 
-  exact value.
-* We estimated the observer effects, along with our uncertainty about those
-  effects' size & direction, with a linear mixed model.
-* This model partitions the variance in observed richness values into site-level
-  variance, observer-level variance, and residual variance (e.g. variation within
-  a site from year to year or from morning to morning).
-  * No fixed effects in this model (e.g. environmental predictor variables because 
-    we're not interested in explaining differences among sites at this point; we 
-    just want to account for their existence while estimating observer effects.
-  * Subtracting off the estimated observer effect provides an estimate of how many
-    would have been found by a "typical" observer on that day; in this sense, it 
-    enables us to correct for observer differences. We then used this "corrected" 
-    version of the data for our time-series models such as the AR1, etc.
-* A simple model like this one can't tell us exactly how each observer differed from
-  the others, so it is important to represent our uncertainty about these differences.
-  Here, we did so by collecting [[N]] Monte Carlo samples from the model's posterior
-  distribution, using Stan. This provided us with [[N]] plausible values for each 
-  observer, so we did not have to commit to a single value.
-* While some methods for predicting richness might be able to use the uncertain
-  estimates of the observer effects directly, we take the simpler approach of just
-  fitting each of the downstream models with each of the Monte Carlo samples. This
-  increases the amount of computational work by a factor of [[N]], but CPU time
-  wasn't our limiting resource.
+Data from `r settings$start_yr` (the first complete year of NDVI data) to `r
+settings$last_train_year` were used for training models and data from `r
+settings$last_train_year + 1` to `r settings$end_yr` was used for testing. Only
+BBS routes with at least `r settings$min_num_yrs` annual observations in the
+years 1982-2015 were used.
+
+### Accounting for observer effects
+
+Observer effects are inherent in large data sets taken by different observers,
+and are known to occur in BBS [@sauer1994]. We estimated the observer effects,
+and associated uncertainty about those effects, with a linear mixed model with
+observer as a random effect. Because observer and site are strongly related (an
+observer tends to repeatedly sample the same site), site was also included as a
+random effect to ensure that inferred deviations were actually observer-related
+(as opposed to being related to the sites that a given observer happened to
+see). The resulting model partitions the variance in observed richness values
+into site-level variance, observer-level variance, and residual variance
+(e.g. variation within a site from year to year), and can therefore also be used
+directly as the "average" baseline model (see below). The estimated observer
+effects can be subtracted from the richness values for a particular observer to
+provide an estimate of how many species would have been found by a "typical"
+observer. To incorporate uncertainty in these "corrected" richness values into
+the forecasting models we collected [[N]] Monte Carlo samples from the model's
+posterior distribution, using Stan, and fit each of the downstream models with
+each of the Monte Carlo samples.
 
@@
 
-[[note that these all use a Gaussian error distribution]]
+We fit three sets of models that were fit to each site independently. These
+models were fit to each BBS route both on the raw richness values and on the
+residuals from our observer model. When correcting for observer effects, we
+averaged across [[N]] models that were fit separately to the [[N]] Monte Carlo
+estimates of the observer effects, to account for our uncertainty in the true
+values of those effects. All of these models use a Gaussian error distribution.
 
-**Baseline models.** The two simplest models we fit treated site-level richness
+**Baseline models.** We used two simple models as baselines for comparison to
+the fits of more complex models. These baselines treated site-level richness
 observations either as uncorrelated noise around a site-level constant (the
 "average" model) or as an autoregressive model with a single year of history
-(the "naive" model). The predictions from these two models are qualitatively 
-different: predictions from the "average" model are centered on the average 
-richness and the confidence intervals are narrow and constant-width. The
-"naive" model, in contrast, predicts that future observations will be similar to
-the final observed value (e.g. the value observed in `$r settings$last_train_year`), 
-and the confidence intervals expand rapidly as the predictions extend farther into 
-the future.  Each of these model types was fit separately to each BBS route in two
-ways: first, we fit the model to the raw richness values, without accounting for
-possible differences among observers at the site.  Then, we fit the models to the
-residuals from our observer model (i.e. fitting the model to an estimate of how
-many species a typical observer would have found, rather than to the raw data). 
-In the latter case, we averaged across [[N]] models that were fit separately to 
-the [[N]] Monte Carlo estimates of the observer effects, to account for our
-uncertainty in the true values of those effects.
+(the "naive" model) [@hyndman2014]. Predictions from the "average" model are
+centered on the average richness and the confidence intervals are narrow and
+constant-width. The "naive" model, in contrast, predicts that future
+observations will be similar to the final observed value (e.g. the value
+observed in `$r settings$last_train_year`), and the confidence intervals expand
+rapidly as the predictions extend farther into the future.
 
-**Time series models.** Like the two baseline models, these models only looked at 
-one site at a time, but they could include additional complexity. As above, these
-models were trained both with the raw richness time series and with Monte Carlo
-estimates of the observer-corrected time series. The  Auto-ARIMA models (based on 
-Hyndman's 2016 `auto.arima` function) could include an autoregressive component 
-(as in the "naive" model, but with the possibility of longer-term dependencies
-in the underlying process), a moving average component (where the noise can have 
-serial autocorrelation) and an integration/differencing component (so that the 
-analysis could be performed on sequential differences of the raw data, rather than
-on the observed time series itself). The `auto.arima` function chooses whether to 
-include each of these components, and how many terms to include for each one, using
-AICc, as described in Hyndman (2016). We used the default settings for this function,
-apart from setting `seasonal = FALSE` because our time series does not include a 
-seasonal component.  [[Note: should probably record somewhere what order model
-the auto.arima function tends to choose.  Eyeballing the results, seems to be most 
-similar to the naive model, but would be good to check.]] We also fit an 
-"Auto-ARIMA + environment" model, in which each site-level time series could 
-also be predicted by annual changes in our environmental variables at that site.
+**Time series models.** We used Auto-ARIMA models [based on the `auto.arima`
+function in @hyndman2016] to represent a array of different time-series modeling
+approaches. These models can include an autoregressive component (as in the
+"naive" model, but with the possibility of longer-term dependencies in the
+underlying process), a moving average component (where the noise can have serial
+autocorrelation) and an integration/differencing component (so that the analysis
+could be performed on sequential differences of the raw data, accommodating more
+complex patterns including trends). The `auto.arima` function chooses whether to
+include each of these components, and how many terms to include for each one,
+using AICc [@hyndman2016]. Since there is no seasonable component to the BBS
+time-series we did not include a season component in these models. Otherwise we
+used the default settings for this function. [[Note: should probably record
+somewhere what order model the auto.arima function tends to choose.  Eyeballing
+the results, seems to be most similar to the naive model, but would be good to
+check.]] We also fit an "Auto-ARIMA + environment" model, in which each
+site-level time series could also be predicted by annual changes in our
+environmental variables at that site.
 
@@
 
-So far, each model has been fit to one site at a time, with no information 
-(apart from observer effects) shared among sites. When continental-scale data 
-is available, ecologists often use correlative models to predict richness 
-changes over time from environmental data. Here, we tested three variants of 
-this approach. Here, we did *not* include the site-level random effects as 
-predictors, meaning that this approach implicitly assumes that two sites 
-with identical Bioclim, elevation, and NDVI values should have identical 
-richness distributions. As above, we included observer effects & the 
-associated uncertainty by running this model [[N]] times (once per MCMC 
-sample).
+In contrast to the time-series models, most attempts to predict species richness
+focusing on using correlative models to predict richness changes over time from
+environmental data. We tested three common variants of this approach - direct
+modeling of species richness, stacking individual species distribution models,
+and joint species distribution models. Following the standard approach,
+site-level random effects were not included in these models as predictors,
+meaning that this approach implicitly assumes that two sites with identical
+Bioclim, elevation, and NDVI values should have identical richness
+distributions. As above, we included observer effects & the associated
+uncertainty by running these models [[N]] times (once per MCMC sample).
 
-**"Macroecological" model.** The most straightforward way to predict richness is 
-to fit a richness model. We chose the `gbm` package for fitting boosted 
-regression trees as our example of such a model. Boosted regression trees is a 
-common approach in the species distribution modeling literature [refs] and 
-works by fitting thousands of small tree-structured models sequentially, with
-each tree optimized to reduce the error of its predecessors. For the reasons 
-discussed above, this model was optimized using a Gaussian likelihood. [[other 
-gbm settings go here.]] Compared to SDMs, this lets us predict richness directly 
-instead of using hundreds of noisy species-level estimates, but it also discards 
-any potentially-useful information regarding species turnover.
+**"Macroecological" model.** We used a generalized boosted regression tree model
+using the `gbm` package [@ridgeway2017] to directly model species richness as a function of
+environmental variables. Boosted regression trees are a form of tree based
+modeling that work by fitting thousands of small tree-structured models
+sequentially, with each tree optimized to reduce the error of its
+predecessors. They are flexible models that are considered well suited for
+prediction [@elith2008]. This model was optimized using a Gaussian
+likelihood. [[other gbm settings go here.]]
 
-**SDMs.**
-* Probably the most common approach?
-* These models say that species' occurrence probabilities depend only on the values of
-  the environmental variables at the time and place of the transect, not on the history
-  of the environment or the history of the birds at that site
-* Predict each species' occurrence probabilities individually, then add up the predictions
-* [[Something about not using thresholds that throw away information]]
-* Taking this approach in the face of uncertainty about observer effects would either 
-  require a customized model (cite "neighborly advice" paper), or fitting [[N]] models
-  times the number of species.
-* To make it feasible to fit this large number of models, we used a custom wrapper 
-  around the randomForest package. Random forests are flexible models that are built 
-  by combining many independent sub-models ("trees") fit to different versions of the training
-  data, which made it straightforward to adapt them to our situation: we simply fit one 
-  tree using each of the Monte Carlo estimates of our observer effects and averaged the 
-  predictions of the resulting forest.
-* For the uncertainty, we used Gaussian approximations, based on the means and variances
-  of the sums of independent Bernoulli random variables (cite Calabrese et al.).
+**Species Distribution Models.** Species distribution models (SDMs) model
+individual species' occurrence probabilities using environmental
+variables. Species level models are used to predict richness by summing the
+predicted probability of occupancy across all species at a site. We fit SDMs
+using random forests, a common approach in the species distribution modeling
+literature [[refs; would be good to provide a source that says these are a good
+choice]]. Predicted probabilities of occurrence were summed directly to predict
+the number of species. This avoids known problems with the use of thresholds for
+determining whether or not a species will be present at a site [[refs]]. When
+incorporating observer effects we avoided fitting [[N x S]] models ([[N]] MCMC
+samples * [[S]] species) by fitting one regression tree using each of the Monte
+Carlo estimates of our observer effects and averaging the predictions of the
+resulting forest [[citation?]]. We used Gaussian approximations for uncertainty,
+based on the means and variances of the sums of independent Bernoulli random
+variables [[cite Calabrese et al.]].
 
-**JSDM** 
-* Joint species distribution model (JSDM) is a new approach that makes predictions
-  about the full community composition, rather than modeling it species-by-species.
-* In stacked single-species SDMs, the species-level occurrences are treated as 
-  independent coin-flips whose proability depends only on the environment.
-* JSDMs remove this independence assumption, and explicitly account for the
-  possibility that a site will be much more (or less) suitable for birds in 
-  general than one would expect based on a few environmental measurements alone.
-* As a result, JSDMs do a better job of representing our uncertainty about 
-  richness, while stacked SDMs underestimate it (Harris 2015).
-* We used the `mistnet` package (Harris 2015) because it is the only JSDM that
-  describes species' environmental associations with nonlinear functions.
-* The mistnet package doesn't "know" about time series or spatial autocorrelation,
-  so it did not share information among the repeated runs of the same transect.
-
-### Ensembles
-* Ensembles. Averaging predictions from multiple models tends to reduce the 
-  noise in the estimates, and can thus lead to better predictions. Choosing the
-  weights for the ensemble isn't straightforward because our estimates of model 
-  error on the training set are biased in favor of models that
-  overfit. [[Another option is to spatially cross-validate on the training set]]
+**Joint Species Distribution Models.** Joint species distribution models (JSDMs)
+are a new approach that makes predictions about the full composition of a
+community instead of modeling each species independently. JSDMs remove the
+assumed independence among species and explicitly account for the possibility
+that a site will be much more (or less) suitable for birds in general, or
+particular groups of birds, than one would expect based on the available
+environmental measurements alone. As a result, JSDMs do a better job of
+representing our uncertainty about richness, while stacked SDMs underestimate it
+[@harris2015; [[other refs, e.g., Clark]]]. We used the `mistnet` package
+[@harris2015] because it is the only JSDM that describes species' environmental
+associations with nonlinear functions.
 
@@
 
-* Evaluated each year between  `r settings$last_train_year` and `r settings$end_yr`
-* When a model was run multiple times (e.g. once per MCMC sample for the observer
-  estimates), we used the mean of those runs' point estimates as our final point 
-  estimate and we calculated the uncertainty using the law of total variance (i.e.
-  the average of the model runs' variance, plus the variance in the point estimates).
-* Metrics:
-  * (R)MSE 
-    * How far, on average, are the models' predictions from the true value?
-    * [[RMSE has better units than MSE (species versus squared species), but MSE is
-      nice because it's additive (i.e. MSE of 2 is twice as big as MSE of 1).]]
-  * Calibration/bias
-    * In general, or in specific situations, is the model consistently predicting
-      that richness will be too high or too low?
-    * Can also report this value in terms of RMSE or MSE
-  * Coverage of 95% confidence intervals
-  * Gaussian deviance (simultaneously rewards good point estimates, precision, 
-    and coverage)
+We evaluated the performance of each model for forecasting using the data for
+each year between `r settings$last_train_year` and `r settings$end_yr`. We used
+three metrics for evaluating performance: 1) root-mean-square error (RMSE) to
+determine how far, on average, the models' predictions were from the observed
+value; 2) the 95% prediction interval coverage to determine how well the models
+did at predicting uncertainty; and 3) deviance as an integrative measure of fit
+incorporating good point estimates, precision, and coverage. When a model was
+run multiple times for the purpose of correcting for observer effects, we used
+the mean of those runs' point estimates as our final point estimate and we
+calculated the uncertainty using the law of total variance (i.e.  the average of
+the model runs' variance, plus the variance in the point estimates). In addition
+to looking at forecast performance in general we evaluated how performance
+changed as the time horizon of forecasting increased by plotting performance
+metrics against the forecast horizon.
 
-Looked at how error changed as the time horizon of forecasting lengthened
-* Error will generally tend to increase
-* Error might flatten out after some number of years
-* Error might increase at different rates for different methods
-  * One model might be better at short time scales \& worse at long time scales
+Variance in species richness within sites was lower than predicted by several
+common count models (i.e. richness was underdispersed), so these count models
+would have difficulty fitting the data. Since richness also had a relatively
+large mean ([[N]]) and all models already produce continuous richness estimates
+(due to summing probabilities in the (J)SDM case), we defined model performance
+for all models in terms of continuous Gaussian errors instead of using discrete
+count distributions.
 
@@
 
-The mixed model (which formed the basis of our observer-corrected "average"
-predictions) found that 70% of the variance in richness in the training set
-could be explained by differences among sites, and 21% could be explained by
-differences among observers. The remaining 9% represents residual variation,
-where a given observer might report a different number of species on different
-runs of the same transect. We found remarkably little autocorrelation in these
-residuals ([[about 0.04, +/- something]]), suggesting richness is close to
-stationary on these time scales, once observer differences are accounted for.
+The site-observer model found that 70% of the variance in richness in the
+training set could be explained by differences among sites, and 21% could be
+explained by differences among observers. The remaining 9% represents residual
+variation, where a given observer might report a different number of species in
+different years. There was little temporal autocorrelation in these residuals
+([[about 0.04, +/- something]]), suggesting that richness is close to stationary
+on these time scales once observer differences are accounted for.
 
-The substantial uncertainty associated with differences among observers caused
-reliable improvements in both mean-squared-error and deviance when observer
-effects were controlled for (Figure 4). This improvement was smallest for the
-"Naive" model, as discussed below.
+When comparing forecasts for richness across sites all methods performed well
+(Figure 2; R^2 XX-XX%). However stacked and joint SDMs and the macroecological
+model both failed to successfully forecast the highest richness sites resulting
+in a noteable clustering of predicted values near ~60 species and the poorest
+performance in the cross-site comparison (R^2 XX-XX%).
 
-In general, each model's error increased more-or-less linearly with time,
-although the slopes and initial error rates varied across models (Figure 3). The
-site-level models (i.e. the baselines and the ARIMA methods) tended to do very
-well (Figure 3), especially in terms of RMSE and especially on short time
-horizons. For most combinations of year and metric, the best-performing model
-was one of the two baselines. Usually, the "naive" model had the lowest error
-when observer effects were not explicitly accounted for, and the "average" model
-had the lowest error when they were. We discuss the reasons for this difference
-in the next section.
-
-Our two error metrics produced qualitatively similar results in most cases, with
-the notable exception of the stacked SDMs (Figure 3A and 3B). Deviance, unlike
-RMSE, depends on the full predictive distribution of a model, including the
-model's estimated uncertainty. As noted elsewhere [@harris2015], these models
-are generally far too confident, with observed richness values frequently
-falling well outside their confidence limits (Figure 3C). As a result, the
-stacked SDMs' deviance was dramatically higher than the next-worst model even
-though its RMSE was not unusual. On the other hand, the richness-GBM model and
-especially the mistnet JSDM had reasonably well-calibrated uncertainty
-estimates, even in the presence of uncorrected observer biases (Figure 3C). This
-prevented their deviance from rising as quickly as the site-level models when
-the observer population changed (Figure 3B).
+For most combinations of metric and forecast time lag, the best-performing model
+for forecasting within site variability was either the "average" baseline model
+or the ARIMA model (Figure 3). In general, each model's error increased
+more-or-less linearly with time, although the slopes and initial error rates
+varied across models (Figure 3). The site-level models (i.e. the baselines and
+the ARIMA methods) tended to do very well (Figure 3) on short time horizons,
+especially in terms of RMSE.
 
@@
 indicating overconfident predictions. Of these, the random forest-based SDM
-stack was the most overconfident, as discussed above. As one might expect,
-correcting for observer differences improved the coverage of most models' 95%
-intervals. Interestingly, this effect was very small for mistnet (its
-coverage only improved from 92% to 93%), which seems to indicate that its
-latent variables were already accounting for observer-related uncertainty. 
-The "naive" model was the only model whose confidence intervals were too wide
-(Figure 3), which can be attributed to the rapid rate at which
-these intervals expand as the time lag increases (Figure 1).
+stack was the most overconfident. As noted elsewhere [@harris2015], these models
+are generally far too confident, with observed richness values frequently
+falling well outside their confidence limits (Figure 3C). The "naive" model was
+the only model whose confidence intervals were too wide (Figure 3), which can be
+attributed to the rapid rate at which these intervals expand as the time lag
+increases (Figure 1). The JSDM and macroecological model has the best coverage.
+
+RMSE and deviance produced qualitatively similar results in most cases, with the
+notable exception of the stacked SDMs (Figure 3A and 3B). Deviance, unlike RMSE,
+depends on the full predictive distribution of a model, including the model's
+estimated uncertainty. As a result, the stacked SDMs' deviance was notably
+higher than the next-worst model even though its RMSE was not unusual. On the
+other hand, the richness-GBM model and especially the mistnet JSDM had
+reasonably well-calibrated uncertainty estimates, even in the presence of
+uncorrected observer biases (Figure 3C). This prevented their deviance from
+rising as quickly as the site-level models when the observer population changed
+and resulted in better comparative performance for these models using deviance
+(Figure 3B).
+
+Accounting for the uncertainty associated with differences among observers
+resulted in improvements to all metrics of fit (Figure 4). The one exception was
+RMSE of the "Naive" model, which remained unchanged. The improvements in point
+estimates primarily resulted from large improvements in a small number of
+forecasts, presumably ones where a large difference between observers resulted
+in a large shift in predicted richness values that the models could not
+otherwise accommodate. Most forecasts exhibited only small changes in
+RMSE. However these changes were sufficient to reorder which models performed
+best overall. For example, the "naive" model generally had the lowest average
+error when observer effects were not explicitly accounted for, switching to the
+average and ARIMA models once observer differences were corrected. Improvements
+in coverage were notable for all models except the naive and JSDM, meaning that
+incorporating observer effects resulted in improved quantification of
+uncertainty. The lack of change in the mistnet JSDM seems to indicate that its
+latent variables were already accounting for observer-related uncertainty. In
+combination changes in point estimates and coverage resulted in changes in the
+best models based on deviance at different time scales. The macroecological
+model and the JSDM appeared to narrowly outperform the average and ARIMA models
+when the forecast horizon was greater than five years when not correcting for
+observers, but after the correction the average and ARIMA models maintained a
+small average advantage out to at least ten years.
 

### 2017-07-11 R/observer_model.R
Revised crack at 2050 predictions

@@ -8,12 +8,11 @@
 fit_observer_model = function(stan_file = "mixed.stan", seed = 1,
                               output_file = "observer_model.rds", 
                               chains = 2, cores = 2, iter = 2000,
-                              thin = 4, ...){
+                              thin = 4, settings...){
   library(rstan) # prevent annoying NAMESPACE issue described in
   # https://github.com/stan-dev/rstan/issues/353
   
   set.seed(seed)
-  settings = yaml::yaml.load_file("settings.yaml")
   
   # If we don't have any observed richness, we can't use the row, so omit
   # anything with is.na(richness)

### 2017-07-11 R/post-observer.R

@@ -1,5 +1,6 @@
 #' @importFrom forecast Arima auto.arima
-make_forecast = function(x, fun_name, use_obs_model, settings, ...){
+make_forecast = function(x, fun_name, use_obs_model, settings, 
+                         future_observer_effects, ...){
   # `Forecast` functions want NAs for missing years, & want the years in order
   x = complete(x, year = settings$start_yr:settings$last_train_year) %>% 
     arrange(year)
@@ -45,15 +46,18 @@ make_forecast = function(x, fun_name, use_obs_model, settings, ...){
     list(NA)
   }
   
+  i = unique(x$iteration)
+  
   # Distance between `upper` and `lower` is 2 sd, so divide by 2
   data_frame(year = seq(settings$last_train_year + 1, settings$end_yr), 
          mean = c(fcst$mean), sd = c(fcst$upper - fcst$lower) / 2, 
          model = fun_name, use_obs_model = use_obs_model,
-         coef_names = coef_names)
+         coef_names = coef_names, 
+         observer_effect = future_observer_effects[[i]])
 }
 
 make_all_forecasts = function(x, fun_name, use_obs_model, 
-                              settings, ...){
+                              settings, future_observer_effects, ...){
   forecast_data = x %>% 
     filter(year <= settings$last_train_year) %>% 
     group_by(site_id, iteration)
@@ -78,9 +82,23 @@ make_all_forecasts = function(x, fun_name, use_obs_model,
          use_obs_model, coef_names)
 }
 
-make_gbm_predictions = function(x, use_obs_model){
+make_test_set = function(x, future, future_observer_effects, settings){
+  if (settings$timeframe == "future") {
+    i = unique(x$iteration)
+    test = mutate(future, 
+                  observer_effect = !!future_observer_effects[[i]],
+                  expected_richness = richness - observer_effect)
+  } else{
+    test = filter(x, year > !!settings$last_train_year)
+  }
+  
+  test
+}
+
+make_gbm_predictions = function(x, use_obs_model, settings, future,
+                                future_observer_effects) {
   train = filter(x, year <= settings$last_train_year)
-  test = filter(x, year > settings$last_train_year)
+  test = make_test_set(x, future, future_observer_effects)
   
   if (use_obs_model) {
     train$y = train$expected_richness
@@ -88,7 +106,9 @@ make_gbm_predictions = function(x, use_obs_model){
[…]

### 2017-07-11 analysis/fit-predict.R

@@ -1,15 +1,24 @@
-# TODO: ensure that sd is correct for all models
-# TODO: better names for columns/functions
-# TODO: Check namespaces
-
 library(tidyverse)
 devtools::load_all()
-settings = yaml::yaml.load_file("settings.yaml")
+timeframe = "train_32"
 
-if (!file.exists("observer_model.rds")) {
-  fit_observer_model()
+
+# Ingest the `settings` and put the timeframe of interest at the top level
+# of the list.
+settings = yaml::yaml.load_file("settings.yaml")
+settings = c(settings, settings$timeframes[[timeframe]])
+settings$timeframes = NULL
+
+
+prepend_timeframe = function(x) {
+  paste0("results", "/", timeframe, "/", x)
 }
-obs_model = readRDS("observer_model.rds")
+dir.create(prepend_timeframe(NULL), showWarnings = FALSE)
+
+if (!file.exists(prepend_timeframe("observer_model.rds"))) {
+  fit_observer_model(settings = settings)
+}
+obs_model = readRDS(prepend_timeframe("observer_model.rds"))
 
 # Discard unnecessary data to save memory
 bioclim_to_discard = colnames(obs_model$data) %>% 
@@ -24,6 +33,23 @@ x_richness = obs_model$data %>%
          expected_richness = richness - observer_effect) %>% 
   select(-ndvi_ann, -lat, -long, -site_index, -one_of(bioclim_to_discard))
 
+
+if (settings$timeframe == "future") {
+  future = get_env_data(timeframe = "future") %>% 
+    filter(site_id %in% !!x_richness$site_id,
+           year > !!settings$last_train_year) %>% 
+    select(which(colnames(.) %in% !!colnames(x_richness)))
+  
+  # Randomly-sampled observers for each `iteration`. Not used for the
+  # "average" model, which will just use observer_sigma directly without
+  # simulations.
+  future_observer_effects = obs_model$observer_sigma %>% 
+    purrr::map(~rnorm(nrow(future), 0, .x))
+} else {
+  future = NULL
+  future_observer_effects = NULL
+}
+
 # Reclaim memory
 rm(obs_model)
 gc()
@@ -31,13 +57,25 @@ gc()
 
 # Fit & save models --------------------------------------------------------
 
-# "Average" model with observer effects
+# For "future" predictions of the "average" model, we'll need to know how 
[…]

### 2017-07-11 R/sdm-rf.R
rf

@@ -1,6 +1,7 @@
 #' @importFrom randomForest randomForest
 one_rf_tree = function(bbs, vars, sp_id, iter, use_obs_model, x_richness,
-                       last_train_year){
+                       last_train_year, future, future_observer_effects,
+                       settings){
   d = bbs %>% 
     select(site_id, year, species_id, abundance) %>% 
     filter(species_id == sp_id) %>% 
@@ -10,31 +11,34 @@ one_rf_tree = function(bbs, vars, sp_id, iter, use_obs_model, x_richness,
     mutate(present = factor(ifelse(is.na(abundance), 0, 1)))
   
   my_formula = as.formula(paste("present ~", 
-                                paste(vars, collapse = "+")))
-   
+                                paste(vars, collapse = " + ")))
+  
   rf = randomForest(
     my_formula, 
     filter(d, in_train),
     ntree = 1
   )  
   
-  test = filter(d, !in_train, iteration == iter)
+  test = make_test_set(d, future, future_observer_effects, settings)
   test$mean = predict(rf, test, type = "prob")[,2]
   test$use_obs_model = use_obs_model
   
   select(test, site_id, year, species_id, mean, richness, use_obs_model, iteration)
 }
 
-rf_predict_species = function(sp_id, bbs, settings, x_richness, use_obs_model){
+rf_predict_species = function(sp_id, bbs, settings, x_richness, use_obs_model,
+                              future, future_observer_effects){
   vars = c(settings$vars, if (use_obs_model) {"observer_effect"})
   iters = sort(unique(x_richness$iteration))
   results = lapply(iters, one_rf_tree,
                    bbs = bbs, sp_id = sp_id, use_obs_model = use_obs_model, 
                    vars = vars,
                    x_richness = x_richness,
-                   last_train_year = settings$last_train_year) %>% 
+                   last_train_year = settings$last_train_year,
+                   future, future_observer_effects,
+                   settings) %>% 
     bind_rows()
-
+  
   path = paste0("rf_predictions/sp_", sp_id, "_", use_obs_model, ".csv.gz")
   dir.create("rf_predictions", showWarnings = FALSE)
   write.csv(results, file = gzfile(path), row.names = FALSE)
@@ -44,16 +48,18 @@ rf_predict_species = function(sp_id, bbs, settings, x_richness, use_obs_model){
     summarize(mean = mean(mean))
 }
 
[…]

### 2017-07-11 analysis/fit-predict.R

@@ -1,4 +1,5 @@
 library(tidyverse)
+library(purrrlyr) # for by_slice
 devtools::load_all()
 timeframe = "train_32"
 
@@ -143,10 +144,12 @@ dir.create("rf_predictions", showWarnings = FALSE)
 
 gc() # Minimize memory detritus before forking
 
-rf_sdm_obs = rf_predict_richness(bbs = bbs, x_richness = x_richness, 
-                                 settings = settings, use_obs_model = TRUE) %>% 
+rf_predict_richness(bbs = bbs, x_richness = x_richness, 
+                    settings = settings, use_obs_model = TRUE,
+                    future, future_observer_effects) %>% 
   saveRDS(file = prepend_timeframe("rf_predictions/all_TRUE.rds"))
 
-rf_sdm_no_obs = rf_predict_richness(bbs = bbs, x_richness = x_richness, 
-                                    settings = settings, use_obs_model = FALSE) %>% 
+rf_predict_richness(bbs = bbs, x_richness = x_richness, 
+                    settings = settings, use_obs_model = FALSE,
+                    future, future_observer_effects) %>% 
   saveRDS(file = prepend_timeframe("rf_predictions/all_FALSE.rds"))

### 2017-07-11 R/get_prism_data.R
fix na issue
This caused a lot of unneeded NA's in the final cmip bioclim dataset. See #133

@@ -106,6 +106,11 @@ get_prism_data=function(){
 extract_cmip_data=function(cmip_filename, clim_var){
   cmip_raster = raster::stack(cmip_filename)
 
+  #Missing values from water bodies are set as NaN but need to be NA
+  #for the aggregation step, otherwise missing values propagate along
+  #coasts. 
+  cmip_raster[is.nan(cmip_raster)] = NA
+  
   #Aggregate the 12km cell size close to 40km
   cmip_raster = raster::aggregate(cmip_raster, fact=3, fun=mean)
 

### 2017-07-11 manuscript/draft.Rmd

@@
 Forecasting is an emerging imperative in ecology and as such the field needs to
-develop and follow good practices for conducting and evaluating ecological
-forecasts. We have laid out a number of these practices (Box 1) and attempted to
-implement them in a single study that builds and evaluates forecasts of
-biodiversity in the form of species richness. The results of this effort are
-both promising and humbling. On an absolute basis, when comparing forecasts
-across sites, many different approaches to forecasting produce reasonable
-forecasts. If a site is predicted to have a high number of species in the
-future, relative to other sites, it generally does. However, the forecasts
-perform relatively poorly at predicting variation in richness over time for any
-given site. If a site is predicted to have a higher richness next year than it
-did last year there is little confidence that this will be the case. As a
-result, the baseline models, like the long-term average and the "naive" model,
-along with site-level time-series models, provided some of the best forecasts
-for future biodiversity.
+develop and follow best practices for conducting and evaluating ecological
+forecasts. We have laid out ideas for a number of these practices (Box 1) and
+attempted to implement them in a single study that builds and evaluates
+forecasts of biodiversity in the form of species richness. The results of this
+effort are both promising and humbling. On an absolute basis, when comparing
+forecasts across sites, many different approaches to forecasting produce
+reasonable forecasts. If a site is predicted to have a high number of species in
+the future, relative to other sites, it generally does. However, the forecasts
+perform relatively poorly at predicting variation in richness over time at a
+location. If a site is predicted to have a higher richness next year than it did
+last year there is little confidence that this will be the case. As a result,
+the baseline models, like the long-term average, along with site-level
+time-series models, provided some of the best forecasts for future biodiversity.
 
 The most commonly used method to forecast future biodiversity, stacked SDMS
-[[cite?]], provided the worst overall forecasts of all methods evaluated. In
+[@urban2015], provided the worst overall forecasts of all methods evaluated. In
 general, SDM-based forecasts are rarely evaluated, particularly in terms of
-repeated observations at the same site, and especially with long time horizons.
-Since site-level richness is relatively stable, the model evaluations that show
-high accuracy in predicting differences among sites may not indicate much about
-the model's ability to predict changes over time. This issue has also recently
-been recently highlighted as an issue for using SDMs to predict the future
-locations of individual species [plos one 2012 paper]. This result is
-particularly sobering because this approach forms the foundation for major
-claims regarding the predicted loss of biodiversity to climate change [2004
-science/nature papers]. The poor performance of SDM-based forecasts highlights
-the crucial importance of comparing multiple modeling approaches when conducting
-ecological forecasts, and in particular the value of comparing results to simple
-baselines to avoid over-interpreting the accuracy of ecological forecasts.
-Disciplines that have more mature forecasting cultures often do this by
-reporting "forecast skill", which is the improvement in the forecast relative to
-a simple baseline. We recommend adopting this approach in future ecological
+repeated observations at the same site. Since site-level richness is relatively
+stable, model evaluations that show high accuracy in predicting differences
+among sites may not indicate much about a model's ability to predict changes
+over time. This issue has also recently been recently highlighted as an issue
+for using SDMs to predict the future locations of individual species
+[@rapacciuolo2012]. This result is particularly sobering because this approach
+forms the main foundation for estimates of the predicted loss of biodiversity to
+climate change [@thomas2004; @urban2015]. The poor performance of SDM-based
+forecasts highlights the crucial importance of comparing multiple modeling
+approaches when conducting ecological forecasts, and in particular the value of
+comparing results to simple baselines to avoid over-interpreting the information
+present in these forecasts (Box 1).  Disciplines that have more mature forecasting
+cultures often do this by reporting "forecast skill", which is the improvement
+in the forecast relative to a simple baseline [[weather citation]]. We recommend
+following Ye et al. [@ye2015] and adopting this approach in future ecological
 forecasting research. It also highlights the need for using hindcasting and
 time-series data for assessing the effectiveness of forecasting methods, since
-this allows comparisons to baselines and assessment of the predicted dynamics at
-each site.
+this allows comparisons to baselines and assessment of predicted dynamics within
+sites or regions.
 
-The most common alternative to summed SDMs is modeling richness directly using
-environmental variables and using the resulting models to make forecasts []. Two
-comparisons of the the SDM and direct richness modeling approaches reported that
-the methods yielded equivalent results for forecasting diversity. While our
-results generally support rough equivalence between stacked SDMs and direct
-richness modeling for point estimates, they also show that modeling richness
-directly results in much better estimation of the uncertainty of the forecast
-and therefore is a better overall approach to forecasting richness. A similar
-result is seen when comparing joint species distribution models (JSDMs) to
-stacked single species distribution models. For point estimates the joint
-distribution models are roughly equivalent to stacked SDMs, but the JSDMs
-provide much better estimates of uncertainty. In fact, JSDMs and modeling
-richness directly provide some of the best estimates of uncertainty across all
-modeling approaches. This highlights the importance evaluating models based on
-their uncertainty as well as their mean estimates. Not doing so is particularly
-problematic in this case because the summed species distribution models are
-overly confident meaning that using them would lead to believing that richness
-would be restricted to a much narrower range than would actually occur.
+One of the main alternatives to summed SDMs is modeling richness directly using
+environmental variables and using the resulting models to make forecasts
+[@algar2009; @distler2015]. Two comparisons of the the SDM and direct richness
+modeling approaches reported that the methods yielded equivalent results for
+forecasting diversity. While our results generally support rough equivalence
+between stacked SDMs and direct richness modeling for point estimates, they also
+show that modeling richness directly results in much better estimation of the
+uncertainty of the forecast and therefore is a better overall approach to
+forecasting richness. A similar result is seen when comparing joint species
+distribution models (JSDMs) to stacked single species distribution models. For
+point estimates the joint distribution models are roughly equivalent to stacked
+SDMs, but the JSDMs provide much better estimates of uncertainty [@harris2015;
+others?]. In fact, JSDMs and modeling richness directly provide some of the best
+estimates of uncertainty across all modeling approaches. This highlights the
+importance of evaluating models based on their uncertainty as well as their mean
+estimates (Box 1). Not doing so is particularly problematic in this case because the
+summed species distribution models are overly confident meaning that using them
+would lead to believing that richness would be restricted to a much narrower
+range than would actually occur.
 
-For error metrics that evaluate both models' point estimates and their
-uncertainty, the joint SDM and the nonlinear richness model began to outperform
-the time-series baselines at the longest time-scales assessed. This convergence
-at longer time scales highlights the importance of considering how forecasting
-method performance changes with the distance into the future being predicted.
-The value of this approach, and associated forecast horizons, has been raised
-for knowing how far into the future a model can be effectively used for
-forecasting [@petchy2015], but our results suggest a broader value to this
-approach for considering the potential importance of different models and
+For the error metric that evaluates both models' point estimates and their
+uncertainty simultaneously, the JSDM and the nonlinear richness model converge
+towards the time-series baselines at the longest time-scales assessed. This
+convergence at longer time scales highlights the importance of considering how
+forecasting method performance changes with the distance into the future being
+predicted (Box 1). The value of this approach, and associated forecast horizons,
+has been raised for knowing how far into the future a model can be effectively
+used for forecasting [@petchy2015], but our results suggest a broader value to
+this approach for considering the potential importance of different models and
 processes for making forecasts at different scales. [[Should make sure Petchy et
@@
 produce forecasts that are worse than simply assuming the systems are
-static. This would be expected to occur is the systems are in fact not changing
-their richness over the relevant time scales and therefore simple models of no
-directional change are appropriate. Recent suggestions that local scale richness
-in some systems is not changing directionally at multi-decadal scales supports
-this possibility [@brown2001; @ernest2001; @dornelas201X; @theplantone]. This
-lack of change may be expected even in the presence of substantial changes in
-environmental conditions and species composition at a site due to the
-replacement of species from the regional pool [@brown2001; @ernest2001]. On
-average the Breeding Bird Survey sites used in this study show little change in
-richness ([[Add summary stat here]]) [see also @lasorte2007]. This absence of
-rapid change is beneficial for the absolute accuracy of forecasts across sites,
-because if a past years richness is known it is easy to estimate a future
-richness, and explains why stacked SDMs performs relatively well at this task
-despite failing to capture meaningful dynamics. However it makes it difficult to
-improve forecasts relative to simple baselines, since those baselines are close
-to representing what is actually occurring in the system. This suggests that
-simple time-series models and baselines should be actively considered for
-forecasts of richness and other stable aspects of biodiversity and also suggests
-that future efforts to understand and forecast biodiversity should also focus on
-composition since it is expected to be more dynamic
-[@ernest2001; @dornelas201X].
+static. This would be expected to occur if richness in these systems is not
+changing over the relevant time scales and therefore simple models of no
+directional change are most appropriate. Recent suggestions that local scale
+richness in some systems is not changing directionally at multi-decadal scales
+supports this possibility [@brown2001; @ernest2001; @dornelas2014;
+@vellend2013]. This lack of change may be expected even in the presence of
+substantial changes in environmental conditions and species composition at a
+site due to the replacement of species from the regional pool [@brown2001;
+@ernest2001]. On average the Breeding Bird Survey sites used in this study show
+little change in richness ([[Add summary stat here]]) [see also
+@lasorte2005]. This absence of rapid change is beneficial for the absolute
+accuracy of forecasts across sites, because if a past year's richness is known
+it is easy to estimate a future richness, and explains why stacked SDMs perform
+relatively well at this task despite failing to capture meaningful
+dynamics. However it makes it difficult to improve forecasts relative to simple
+baselines, since those baselines are close to representing what is actually
+occurring in the system. This suggests that simple time-series models and
+baselines should be actively considered for forecasts of richness and other
+stable aspects of biodiversity and also suggests that future efforts to
+understand and forecast biodiversity should also focus on composition since it
+is expected to be more dynamic [@ernest2001; @dornelas2014].
 
@@
 the large number of different observers and major differences in habitat and
-species. We included an observation model for one of the two major observation
-issues known for the Breeding Bird Survey, differences among
-observers. Accounting for differences in observers substantially reduced the
-error in point estimates in all models and also improved the coverage of the
-confidence intervals. In addition, it resulted in changes in which models
-performed best, most notably reducing the performance of the naive model for
-point estimates. This suggests that the naive model performed well in part
-because it was capable of accommodating rapid shifts in estimated richness
-introduced by changes in the observer. These kinds of rapid changes were
-difficult for the other time-series models to accommodate and so the ARIMA and
-average models improved substantially once this source of observation error was
-addressed. This demonstrates that properly modeling observation error can be
-important for reducing uncertainty in forecast but can also lead to changes in
-the best methods for forecasting. We did not address differences in detection
-probability across species and sites since there is no clear way to address this
-issue without making strong assumptions about the data, but this would be a
-valuable addition to future forecasting models.
-
-Here we have made forecasts using yearly timesteps fifty (?) years into the
-future. Many ecological forecasts are temporally aggregated to 5-30 year
-timesteps and projected up to a century into the future. These are commonly
-beyond the career or even lifespan of the researchers, making reasonable model
-assessment impossible. Currently assessments at these large scales can be only
-be made with a small number of opportune datasets. For examples hindcasts have
-been used to asses models of species richness [@alger2009, distler2015] and
-distribution [@rappacciuolo2012, @moran-ordonez2017, @araujo2005,
-@eskildsen2013] using data aggregated across similar time scales. More
-evaluation studies like these are needed but there is a paucity of adequate long
-term data available. Short term forecasts will allow for relatively rapid
-validation of models by researchers without needing to wait for enough data to
-accumulate. This is not to say they should replace large scale long term
-forecasts. The relationships of temporal scaling is currently understudied, but
-it is possible they are similar to spatial scale dynamics [@mcgill2010,
-@levin1992]. For example drivers of species richness differ at different spatial
-scales [@rahbek2001, @hurlbert2003]. Thus short term forecasts should be
-expected to add value to end users and inform, not replace, long term larger
-scaled forecasts.
+species [@sauer1994]. We included an observation model for one of the two major
+observation issues known for the Breeding Bird Survey, differences among
+observers. Accounting for differences in observers reduced the average error in
+point estimates and also improved the coverage of the confidence intervals. In
+addition, it resulted in changes in which models performed best, most notably
+reducing the relative performance of the naive model for point estimates. This
+suggests that the naive model performed well in part because it was capable of
+accommodating rapid shifts in estimated richness introduced by changes in the
+observer. These kinds of rapid changes were difficult for the other time-series
+models to accommodate and so the average performance of the ARIMA and average
+models improved once this source of observation error was addressed. This
+demonstrates that properly modeling observation error can be important for
+properly estimating and reducing uncertainty in forecasts and can also lead to
+changes in the best methods for forecasting. We did not address differences in
+detection probability across species and sites [@boulinier1998] since there is
+no clear way to address this issue without making strong assumptions about the
+data (i.e., assuming there is no biological variation in stops along a route),
+but this would be a valuable addition to future forecasting models.
 
@@
 uncertainty and bias. This will cause ecological forecasts that use
-environmental data to be more uncertain. It is important to correctly
-incorporate this uncertainty in the predictor variables into forecasting models
-[@clark2001; @dietz2017]. Difficulty in forecasting future environmental
-conditions at small scales will present continued challenges for models
-incorporating these conditions and this may result in a continued advantage to
-simple time-series based approaches.
+environmental data to be more uncertain than observed in current hindcasting
+efforts. It is important to correctly incorporate this uncertainty in the
+predictor variables into forecasting models [@clark2001; @dietz2017]. Difficulty
+in forecasting future environmental conditions at small scales will present
+continued challenges for models incorporating these conditions and this may
+result in a continued advantage to simple time-series based approaches.
+
+Another challenge for ecological forecasting is the timescales of typical
+forecasts. We evaluated forecasts using yearly timesteps up to a decade into the
+future. In contrast, many ecological forecasts are temporally aggregated to 5-30
+year timesteps and projected up to a century into the future. These are commonly
+beyond the career or even lifespan of the researchers, making reasonable model
+assessment impossible. Currently assessments at these long time-scales can be
+only be made with a small number of opportune datasets. For examples hindcasts
+have been used to asses models of species richness [@alger2009, distler2015] and
+distribution [@rappacciuolo2012, @moran-ordonez2017, @araujo2005,
+@eskildsen2013] using data aggregated across similar time scales. More
+evaluation studies like these are needed but there is a paucity of adequate long
+term data available. Short term forecasts will allow for relatively rapid
+validation of models by researchers without needing to wait for enough data to
+accumulate [tredennick2016; @dietze2016]. This does not mean that they should
+replace large scale long-term forecasts. As discussed above, drivers of species
+richness are likely to differ at different temporal scales [@rosenzweig1995,
+@white2004, @white2007]. Thus short-term forecasts should be expected to add
+value to end users and inform, not replace, long-term forecasts. Hopefully by
+evaluating trends in the error shorter term forecasts as done here it will be
+possible to also use the information from short-term forecasting to improve
+longer-term estimates.
 
@@
 forecasting first started the forecasts were likewise worse than simple
-baselines []. One of the things that helped weather forecasts improve was large
-numbers of forecasts were made in public, which allowed different approaches to
-forecasting to be regularly assessed and improved
-[@mcgill2012 blog post; history of weather forecasting book]. This suggests that
-it is important for ecologists to start regularly making and evaluating true
-ecological forecasts, even if they perform poorly, and to make these forecasts
-publicly available for assessment. These forecasts should include both
-short-term predictions, which can be assessed quickly, and mid to long-term
-forecasts to help assess long time-scale processes and determine how far into
-the future we can successfully forecast [@tredenick ???, @dietz2017]. Forecasts
-for the next XX years for the models in this paper are openly archived on Dryad
-(XXXXXXXXXXX) so that we and others can assess how well they perform and we plan
-to evaluate these forecasts and report the results as each new year of BBS data
-becomes available. Weather forecasting has continually improved throughout its
-history [@bauer2015] in part due to making and evaluating public forecasts
-[@mcgill2012] and we hope this will help ecology do the same.
+baselines [@mcgill2012]. One of the things that helped weather forecasts improve
+was large numbers of forecasts were made in public, which allowed different
+approaches to forecasting to be regularly assessed and improved [@mcgill2012;
+@silver2012]. This suggests that it is important for ecologists to start
+regularly making and evaluating true ecological forecasts, even if they perform
+poorly, and to make these forecasts publicly available for assessment. These
+forecasts should include both short-term predictions, which can be assessed
+quickly, and mid to long-term forecasts to help assess long time-scale processes
+and determine how far into the future we can successfully forecast
+[@tredenick2016; @dietz2017]. Forecasts based on the models in this paper from
+now until 2050 are openly archived on Dryad so that we and others can assess how
+well they perform and we plan to evaluate these forecasts and report the results
+as each new year of BBS data becomes available. Weather forecasting has
+continually improved throughout its history [@bauer2015] in part due to making
+and evaluating public forecasts [@mcgill2012] and we hope this will help ecology
+do the same.
 
@@
 are complex, fundamental theory is less refined than simpler physical/chemical
-systems, and we currently lack the kinds of truly "big" data that produce
-effective forecasts through machine learning. Despite this we believe that
-progress can be made if we build an active forecasting culture in ecology that
-builds and assesses forecasts in ways that will allow us to improve the
-effectiveness of ecological forecasts most rapidly (Box 1). This includes
-expanding the scope of the ecological and environmental data we work with,
-paying attention to uncertainty in both model building and forecast evaluation,
-and rigorously assessing forecasts using a combination of hindcasting, archived
-forecasts, and comparisons to simple baselines.
+systems, and we currently lack the large amounts of data that produce effective
+forecasts through machine learning. Despite this we believe that progress can be
+made if we build an active forecasting culture in ecology that builds and
+assesses forecasts in ways that will allow us to improve the effectiveness of
+ecological forecasts most rapidly (Box 1). This includes expanding the scope of
+the ecological and environmental data we work with, paying attention to
+uncertainty in both model building and forecast evaluation, and rigorously
+assessing forecasts using a combination of hindcasting, archived forecasts, and
+comparisons to simple baselines.
 
@@
 if the modeled forecasts are more accurate than the naive assumption that the
-world is static.
+world is static [@ye2015].
 
@@
 ecological and environmental time-series data allows the consideration of actual
-dynamics from both a process [@treddenick2016a] and error structure perspective.
+dynamics from both a process and error structure perspective [@treddenick2016a].
 
@@
 
-Understanding confidence in a forecast is just as important as understanding the
-average or expected outcome. Failing to account for uncertainty can result in
-overconfidence in highly uncertain outcomes leading to poor decision making and
-erosion of confidence in ecological forecasts. Models should explicitly include
-sources of uncertainty and how the propagate through the forecast where
-possible. Evaluations of forecasts should assess the accuracy of uncertainties
-as well as point estimates [@dietz2017].
+Understanding uncertainty in a forecast is just as important as understanding
+the average or expected outcome. Failing to account for uncertainty can result
+in overconfidence in highly uncertain outcomes leading to poor decision making
+and erosion of confidence in ecological forecasts. Models should explicitly
+include sources of uncertainty and propagate them through the forecast where
+possible [@clark2001; @dietz2017]. Evaluations of forecasts should assess the
+accuracy of uncertainties as well as point estimates [@dietz2017].
 
@@
 variables may have at least as many benefits as using more sophisticated
-modeling techniques.
+modeling techniques. [[Need a citation or two here]]
 
-### 5. Assess how forecast accuracy changes with time-lag
+### 5. Address unknown or unmeasured predictors
+
+Ecological systems are complex and many biotic and abiotic aspects of the
+environment are not regularly measured. As a result, some sites may deviate in
+consistent ways from model predictions. Unknown or unmeasured predictors can be
+incorporated in models using site-level random effects (potentially spatially
+autocorrelated) or by using latent variables that can identify unmeasured
+gradients [@harris2015].
+
+### 6. Assess how forecast accuracy changes with time-lag
 
@@
 
-### 6. Include an observation model
+### 7. Include an observation model
 
@@
 the system is sampled. When possible forecasts should model the factors
-influencing the observation of the data.
+influencing the observation of the data. [[citation?]]
 
-### 7. Validate using hindcasting
+### 8. Validate using hindcasting
 
 To evaluate the expected accuracy and uncertainty of forecasts assess the
-performance of these forecasts within existing time-series data.
+performance of these forecasts within existing time-series data. [[citaton?]]
 
-### 8. Publicly archive forecasts
+### 9. Publicly archive forecasts
 
@@
 forecast values and/or models should be archived so that they can be assessed
-after new data is generated [@mcgill2012]. Enough information should be provided
-to allow an unambiguous assessment of the forecast performance.
+after new data is generated [@mcgill2012; @silver2012]. Enough information
+should be provided to allow an unambiguous assessment of the forecast
+performance.
 
-### 9. Make short-term and long-term predictions
+### 10. Make short-term and long-term predictions
 
@@
 and to allow the accuracy of the forecasts to be quickly evaluated
-[@treddenick2016a].
-
-### 10. [[?]]
-
-Sites' biotic and abiotic environments differ in more ways than the 
-few ones you happened to measure. As a result, some sites will 
-consistently have more species than a regression model (or sum of 
-regression models) would predict, while others willl consistently have 
-fewer. Depending on the spatial scale of these unmeasured variables, 
-independent site-level random effects could be sufficient for dealing 
-with this, or spatially-autocorrelated random effects might be needed.
+[@treddenick2016].

### 2017-07-11 manuscript/draft.Rmd
don't discuss excluded sites after counting the sites

@@
 rdataretriever R package [@mcglinn2017]. The BBS data was filtered to exclude
-all nocturnal, cepuscular, and aquatic species
-[since these species are not well sampled by BBS methods; @hurlbert2005], as
-well as unidentified species, and hybrids. All data from routes that did not
-meet BBS acceptable criteria were excluded. Only routes within the continental
-United States were used so as to match the climate dataset. The resulting
-dataset included [[N]] species and [[N]] sites. Among sites richness varied from
-[[N]] to [[N]] with an average richness of [[N]] species. Within-site richness
-variation (among years) was small (~5 species).
+all nocturnal, cepuscular, and aquatic species [since these species are not well
+sampled by BBS methods; @hurlbert2005], as well as unidentified species, and
+hybrids. Due to limitations of the environmental data (below), we did not
+include any data from outside the continental United States. Observed richness
+from 1982 (the first complete year of NDVI data) to 2003 were used for training
+the models, and data from 2004 to 2013 was used for testing their performance.
+Only BBS routes that were sampled during 70% of the years in the training period
+were used (i.e., routes with at least 16 annual observations). All data from
+routes that did not meet BBS acceptable criteria were excluded, as were routes
+that had not been visited during at least 70% of the years in our model-training
+window. The resulting dataset included [[N]] species and [[N]] sites. Among
+sites,  richness varied from [[N]] to [[N]] with an average richness of [[N]]
+species. Within-site richness variation (among years) was small (standard
+deviation of about 5 species, after controlling for observer effects).
 
@@
 and a remotely sensed vegetation index (the normalized difference vegetation
-index; NDVI), all of which are know to influence richness and distribution in
+index; NDVI), all of which are known to influence richness and distribution in
 the BBS data. We used the 4 km resolution PRISM data [@prism] to calculate eight
@@
 of the driest quarter, precipitation seasonality, precipitation of the wettest
-quarter, precipitation of the warmest quarter. Satellite derived NDVI, a primary
-correlated of richness in BBS data [@hurlbert2002], with an 8 km resolution was
+quarter, precipitation of the warmest quarter. Satellite-derived NDVI, a primary
+correlate of richness in BBS data [@hurlbert2002], with an 8 km resolution was
 obtained from the NDIV3g dataset [@pinzon2014] and was available from the time
@@
 of each BBS route.
-
-Data from `r settings$start_yr` (the first complete year of NDVI data) to `r
-settings$last_train_year` were used for training models and data from `r
-settings$last_train_year + 1` to `r settings$end_yr` was used for testing. Only
-BBS routes with at least `r settings$min_num_yrs` annual observations in the
-years 1982-2015 were used.
 

### 2017-07-11 manuscript/draft.Rmd
update model descriptions

@@
 Observer effects are inherent in large data sets taken by different observers,
-and are known to occur in BBS [@sauer1994]. We estimated the observer effects,
-and associated uncertainty about those effects, with a linear mixed model with
-observer as a random effect. Because observer and site are strongly related (an
-observer tends to repeatedly sample the same site), site was also included as a
-random effect to ensure that inferred deviations were actually observer-related
-(as opposed to being related to the sites that a given observer happened to
-see). The resulting model partitions the variance in observed richness values
-into site-level variance, observer-level variance, and residual variance
-(e.g. variation within a site from year to year), and can therefore also be used
-directly as the "average" baseline model (see below). The estimated observer
-effects can be subtracted from the richness values for a particular observer to
-provide an estimate of how many species would have been found by a "typical"
-observer. To incorporate uncertainty in these "corrected" richness values into
-the forecasting models we collected [[N]] Monte Carlo samples from the model's
-posterior distribution, using Stan, and fit each of the downstream models with
-each of the Monte Carlo samples.
+and are known to occur in BBS [@sauer1994]. We estimated the observer effects
+(and associated uncertainty about those effects) with a linear mixed model with
+observer as a random effect built in the Stan probabilistic programming language
+[@stan] and its R interface [@rstan]. Because observer and site are strongly
+related (an observer tends to repeatedly sample the same site), site was also
+included as a random effect to ensure that inferred deviations were actually
+observer-related (as opposed to being related to the sites that a given observer
+happened to see). The resulting model partitions the variance in observed
+richness values into site-level variance, observer-level variance, and residual
+variance (e.g. variation within a site from year to year), and can therefore
+also be used directly as the "average" baseline model (see below). The estimated
+observer effects can be subtracted from the richness values for a particular
+observer to provide an estimate of how many species would have been found by a
+"typical" observer. To incorporate uncertainty in these "corrected" richness
+values into the forecasting models we collected 500 Monte Carlo samples from the
+model's posterior distribution, and fit each of the downstream models with each
+of the Monte Carlo samples.
 
@@
 residuals from our observer model. When correcting for observer effects, we
-averaged across [[N]] models that were fit separately to the [[N]] Monte Carlo
+averaged across 500 models that were fit separately to the 500 Monte Carlo
 estimates of the observer effects, to account for our uncertainty in the true
@@
 (the "naive" model) [@hyndman2014]. Predictions from the "average" model are
-centered on the average richness and the confidence intervals are narrow and
-constant-width. The "naive" model, in contrast, predicts that future
-observations will be similar to the final observed value (e.g. the value
-observed in `$r settings$last_train_year`), and the confidence intervals expand
-rapidly as the predictions extend farther into the future.
+centered on the average richness observed during training, and the confidence
+intervals are narrow and constant-width. The "naive" model, in contrast,
+predicts that future observations will be similar to the final observed value
+(e.g. the value observed in `$r settings$last_train_year`), and the confidence
+intervals expand rapidly as the predictions extend farther into the future.
 
@@
 include each of these components, and how many terms to include for each one,
-using AICc [@hyndman2016]. Since there is no seasonable component to the BBS
-time-series we did not include a season component in these models. Otherwise we
+using AICc [@hyndman2016]. Since there is no seasonal component to the BBS
+time-series, we did not include a season component in these models. Otherwise we
 used the default settings for this function. [[Note: should probably record
@@
 the results, seems to be most similar to the naive model, but would be good to
-check.]] We also fit an "Auto-ARIMA + environment" model, in which each
-site-level time series could also be predicted by annual changes in our
-environmental variables at that site.
+check.]]
 
@@
 modeling of species richness, stacking individual species distribution models,
-and joint species distribution models. Following the standard approach,
+and joint species distribution models (JSDMs). Following the standard approach,
 site-level random effects were not included in these models as predictors,
@@
 distributions. As above, we included observer effects & the associated
-uncertainty by running these models [[N]] times (once per MCMC sample).
+uncertainty by running these models 500 times (once per MCMC sample).
 
-**"Macroecological" model.** We used a generalized boosted regression tree model
-using the `gbm` package [@ridgeway2017] to directly model species richness as a function of
-environmental variables. Boosted regression trees are a form of tree based
+**"Macroecological" model.** We used a boosted regression tree model using the
+`gbm` package [@ridgeway2017] to directly model species richness as a function
+of environmental variables. Boosted regression trees are a form of tree based
 modeling that work by fitting thousands of small tree-structured models
-sequentially, with each tree optimized to reduce the error of its
-predecessors. They are flexible models that are considered well suited for
-prediction [@elith2008]. This model was optimized using a Gaussian
-likelihood. [[other gbm settings go here.]]
+sequentially, with each tree optimized to reduce the error of its predecessors.
+They are flexible models that are considered well suited for prediction
+[@elith2008]. This model was optimized using a Gaussian likelihood, with a 
+maximum interaction depth of 5, shrinkage of 0.015, and up to 10,000 trees.
+The number of trees was selected using the "out of bag" estimator, and 
+[[summarize how many trees were typically chosen here, for both observer-based
+and observer-free models]].
 
-**Species Distribution Models.** Species distribution models (SDMs) model
-individual species' occurrence probabilities using environmental
-variables. Species level models are used to predict richness by summing the
-predicted probability of occupancy across all species at a site. We fit SDMs
-using random forests, a common approach in the species distribution modeling
-literature [[refs; would be good to provide a source that says these are a good
-choice]]. Predicted probabilities of occurrence were summed directly to predict
-the number of species. This avoids known problems with the use of thresholds for
-determining whether or not a species will be present at a site [[refs]]. When
-incorporating observer effects we avoided fitting [[N x S]] models ([[N]] MCMC
-samples * [[S]] species) by fitting one regression tree using each of the Monte
-Carlo estimates of our observer effects and averaging the predictions of the
-resulting forest [[citation?]]. We used Gaussian approximations for uncertainty,
-based on the means and variances of the sums of independent Bernoulli random
-variables [[cite Calabrese et al.]].
+**Species Distribution Models.** Species distribution models (SDMs) predict
+individual species' occurrence probabilities using environmental variables.
+Species-level models are used to predict richness by summing the predicted
+probability of occupancy across all species at a site; following [[Calabrese et
+al.]], we calculated the uncertainty in our richness estimate by treating
+richness as a sum over independent Bernoulli random variables (one per species).
+This avoids known problems with the use of thresholds for determining whether or
+not a species will be present at a site [[refs]]. We represented SDMs in this
+analysis using random forests, a common approach in the species distribution
+modeling literature based on fitting hundreds of independent regression trees to
+randomly-perturbed versions of the data [[refs; would be good to provide a
+source that says these are a good choice]]. Each of the 500 trees in our
+species-level random forests used a different Monte Carlo estimate of the
+observer effects as a predictor variable.
 
@@
 are a new approach that makes predictions about the full composition of a
-community instead of modeling each species independently. JSDMs remove the
-assumed independence among species and explicitly account for the possibility
-that a site will be much more (or less) suitable for birds in general, or
-particular groups of birds, than one would expect based on the available
-environmental measurements alone. As a result, JSDMs do a better job of
-representing our uncertainty about richness, while stacked SDMs underestimate it
-[@harris2015; [[other refs, e.g., Clark]]]. We used the `mistnet` package
+community instead of modeling each species independently as above. JSDMs remove
+the assumed independence among species and explicitly account for the
+possibility that a site will be much more (or less) suitable for birds in
+general, or particular groups of birds, than one would expect based on the
+available environmental measurements alone. As a result, JSDMs do a better job
+of representing our uncertainty about richness, while stacked SDMs underestimate
+it [@harris2015; [[other refs, e.g., Clark]]]. We used the `mistnet` package
 [@harris2015] because it is the only JSDM that describes species' environmental

### 2017-07-11 manuscript/draft.Rmd
evaluation & results

@@
 
-**Species Distribution Models.** Species distribution models (SDMs) predict
+**Species Distribution Model.** Species distribution models (SDMs) predict
 individual species' occurrence probabilities using environmental variables.
@@
 
+Variance in species richness within sites was lower than predicted by several
+common count models, such as the Poisson or binomial (i.e. richness was
+underdispersed), so these count models would have had difficulty fitting the
+data. Since richness had a relatively large mean ([[N]]) and all models already
+produce continuous richness estimates (due to summing probabilities in the
+(J)SDM case), we defined model performance for all models in terms of continuous
+Gaussian errors instead of using discrete count distributions. When a model was
+run multiple times for the purpose of correcting for observer effects, we used
+the mean of those runs' point estimates as our final point estimate and we
+calculated the uncertainty using the law of total variance (i.e.  the average of
+the model runs' variance, plus the variance in the point estimates).
+
 We evaluated the performance of each model for forecasting using the data for
@@
 value; 2) the 95% prediction interval coverage to determine how well the models
-did at predicting uncertainty; and 3) deviance as an integrative measure of fit
-incorporating good point estimates, precision, and coverage. When a model was
-run multiple times for the purpose of correcting for observer effects, we used
-the mean of those runs' point estimates as our final point estimate and we
-calculated the uncertainty using the law of total variance (i.e.  the average of
-the model runs' variance, plus the variance in the point estimates). In addition
-to looking at forecast performance in general we evaluated how performance
-changed as the time horizon of forecasting increased by plotting performance
-metrics against the forecast horizon.
-
-Variance in species richness within sites was lower than predicted by several
-common count models (i.e. richness was underdispersed), so these count models
-would have difficulty fitting the data. Since richness also had a relatively
-large mean ([[N]]) and all models already produce continuous richness estimates
-(due to summing probabilities in the (J)SDM case), we defined model performance
-for all models in terms of continuous Gaussian errors instead of using discrete
-count distributions.
+predicted the range of possible outcomes; and 3) deviance (i.e. negative 2 times
+the Gaussian log-likelihood) as an integrative measure of fit incorporating good
+point estimates, precision, and coverage.  In addition to looking at forecast
+performance in general we evaluated how performance changed as the time horizon
+of forecasting increased by plotting performance metrics against the forecast
+horizon.
 
@@
 model both failed to successfully forecast the highest richness sites resulting
-in a noteable clustering of predicted values near ~60 species and the poorest
+in a notable clustering of predicted values near ~60 species and the poorest
 performance in the cross-site comparison (R^2 XX-XX%).
@@
 For most combinations of metric and forecast time lag, the best-performing model
-for forecasting within site variability was either the "average" baseline model
-or the ARIMA model (Figure 3). In general, each model's error increased
-more-or-less linearly with time, although the slopes and initial error rates
-varied across models (Figure 3). The site-level models (i.e. the baselines and
-the ARIMA methods) tended to do very well (Figure 3) on short time horizons,
-especially in terms of RMSE.
+for forecasting within-site variability was either the "average" baseline model
+or the ARIMA model [[Not sure what the bit about within-site variability means
+here. Also, didn't Naive usually win when observer effects weren't corrected
+for?]] (Figure 3). In general, each model's error increased more-or-less
+linearly with time, although the slopes and initial error rates varied across
+models (Figure 3). The site-level models (i.e. the baselines and the ARIMA
+methods) tended to do very well (Figure 3) on short time horizons, especially in
+terms of RMSE.
 
@@
 indicating overconfident predictions. Of these, the random forest-based SDM
-stack was the most overconfident. As noted elsewhere [@harris2015], these models
-are generally far too confident, with observed richness values frequently
-falling well outside their confidence limits (Figure 3C). The "naive" model was
-the only model whose confidence intervals were too wide (Figure 3), which can be
-attributed to the rapid rate at which these intervals expand as the time lag
-increases (Figure 1). The JSDM and macroecological model has the best coverage.
+stack was the most overconfident, with observed richness values frequently
+falling well outside their confidence limits (Figure 3C). As noted elsewhere
+[@harris2015], this overconfidence is a product of the assumption in stacked
+SDMs that errors in the species-level predictions are independent. The JSDM and
+macroecological model has the best coverage, though each model still
+underestimated the possible variation in richness. The "naive" model
+was the only model whose confidence intervals were too wide (Figure 3), which
+can be attributed to the rapid rate at which these intervals expand as the time
+lag increases (Figure 1).
 
@@
 higher than the next-worst model even though its RMSE was not unusual. On the
-other hand, the richness-GBM model and especially the mistnet JSDM had
-reasonably well-calibrated uncertainty estimates, even in the presence of
-uncorrected observer biases (Figure 3C). This prevented their deviance from
-rising as quickly as the site-level models when the observer population changed
-and resulted in better comparative performance for these models using deviance
-(Figure 3B).
+other hand, the GBM-based "macroecological" model and especially the mistnet
+JSDM had reasonably well-calibrated uncertainty estimates, even in the presence
+of uncorrected observer biases (Figure 3C). This prevented their deviance from
+rising as quickly as the site-level models as the observer population evolved
+and resulted in better comparative performance for these models in terms of
+deviance (Figure 3B).
 
@@
 in a large shift in predicted richness values that the models could not
-otherwise accommodate. Most forecasts exhibited only small changes in
-RMSE. However these changes were sufficient to reorder which models performed
-best overall. For example, the "naive" model generally had the lowest average
+otherwise accommodate. Most forecasts exhibited only small changes in RMSE.
+However these changes were sufficient to change the rank ordering of model
+performance. For example, the "naive" model generally had the lowest average
 error when observer effects were not explicitly accounted for, switching to the
-average and ARIMA models once observer differences were corrected. Improvements
-in coverage were notable for all models except the naive and JSDM, meaning that
+"average" and ARIMA model once observer differences were corrected. Improvements
+in coverage were notable for all models except Naive and JSDM, meaning that
 incorporating observer effects resulted in improved quantification of
 uncertainty. The lack of change in the mistnet JSDM seems to indicate that its
-latent variables were already accounting for observer-related uncertainty. In
-combination changes in point estimates and coverage resulted in changes in the
-best models based on deviance at different time scales. The macroecological
-model and the JSDM appeared to narrowly outperform the average and ARIMA models
-when the forecast horizon was greater than five years when not correcting for
-observers, but after the correction the average and ARIMA models maintained a
-small average advantage out to at least ten years.
+latent variables were already accounting for observer-related uncertainty. These
+differences meant that, in terms of deviance, neither the JSDM nor the
+macroecological model model benefitted as much from the as the site-level models
+did from the observer corrections. As a result, the average and ARIMA models
+maintained a small average advantage over the JSDM in terms of deviance at
+longer time scales, rather than falling behind after about five years.
 

### 2017-07-12 manuscript/draft.Rmd
minor discussion edits

--- 2017-07-11 manuscript/draft.Rmd
@@
 rdataretriever R package [@mcglinn2017]. The BBS data was filtered to exclude
-all nocturnal, cepuscular, and aquatic species [since these species are not well
-sampled by BBS methods; @hurlbert2005], as well as unidentified species, and
-hybrids. Due to limitations of the environmental data (below), we did not
-include any data from outside the continental United States. Observed richness
-from 1982 (the first complete year of NDVI data) to 2003 were used for training
-the models, and data from 2004 to 2013 was used for testing their performance.
-Only BBS routes that were sampled during 70% of the years in the training period
-were used (i.e., routes with at least 16 annual observations). All data from
-routes that did not meet BBS acceptable criteria were excluded, as were routes
-that had not been visited during at least 70% of the years in our model-training
-window. The resulting dataset included [[N]] species and [[N]] sites. Among
-sites,  richness varied from [[N]] to [[N]] with an average richness of [[N]]
-species. Within-site richness variation (among years) was small (standard
-deviation of about 5 species, after controlling for observer effects).
+all nocturnal, cepuscular, and aquatic species
+[since these species are not well sampled by BBS methods; @hurlbert2005], as
+well as unidentified species, and hybrids. All data from routes that did not
+meet BBS acceptable criteria were excluded. Only routes within the continental
+United States were used so as to match the climate dataset. The resulting
+dataset included [[N]] species and [[N]] sites. Among sites richness varied from
+[[N]] to [[N]] with an average richness of [[N]] species. Within-site richness
+variation (among years) was small (~5 species).
 
@@
 and a remotely sensed vegetation index (the normalized difference vegetation
-index; NDVI), all of which are known to influence richness and distribution in
+index; NDVI), all of which are know to influence richness and distribution in
 the BBS data. We used the 4 km resolution PRISM data [@prism] to calculate eight
@@
 of the driest quarter, precipitation seasonality, precipitation of the wettest
-quarter, precipitation of the warmest quarter. Satellite-derived NDVI, a primary
-correlate of richness in BBS data [@hurlbert2002], with an 8 km resolution was
+quarter, precipitation of the warmest quarter. Satellite derived NDVI, a primary
+correlated of richness in BBS data [@hurlbert2002], with an 8 km resolution was
 obtained from the NDIV3g dataset [@pinzon2014] and was available from the time
@@
 of each BBS route.
+
+Data from `r settings$start_yr` (the first complete year of NDVI data) to `r
+settings$last_train_year` were used for training models and data from `r
+settings$last_train_year + 1` to `r settings$end_yr` was used for testing. Only
+BBS routes with at least `r settings$min_num_yrs` annual observations in the
+years 1982-2015 were used.
 
@@
 cultures often do this by reporting "forecast skill", which is the improvement
-in the forecast relative to a simple baseline [[weather citation]]. We recommend
+in the forecast relative to a simple baseline [@jolliffe2003]. We recommend
 following Ye et al. [@ye2015] and adopting this approach in future ecological
@@
 beyond the career or even lifespan of the researchers, making reasonable model
-assessment impossible. Currently assessments at these long time-scales can be
+assessment impossible. Currently assessments at these long time-scales can
 only be made with a small number of opportune datasets. For examples hindcasts
-have been used to asses models of species richness [@alger2009, distler2015] and
-distribution [@rappacciuolo2012, @moran-ordonez2017, @araujo2005,
+have been used to asses models of species richness [@alger2009, @distler2015]
+and distribution [@rappacciuolo2012, @moran-ordonez2017, @araujo2005,
 @eskildsen2013] using data aggregated across similar time scales. More
@@
 validation of models by researchers without needing to wait for enough data to
-accumulate [tredennick2016; @dietze2016]. This does not mean that they should
+accumulate [@tredennick2016; @dietze2016]. This does not mean that they should
 replace large scale long-term forecasts. As discussed above, drivers of species
@@
 value to end users and inform, not replace, long-term forecasts. Hopefully by
-evaluating trends in the error shorter term forecasts as done here it will be
+evaluating trends in the shorter term forecasts as done here it will be
 possible to also use the information from short-term forecasting to improve
@@
 The science of forecasting biodiversity, and ecology more broadly, remains in
-its infancy and it is important to consider the general inability the
+its infancy and it is important to consider the general inability of
 forecasting methods to improve on simple baselines in that context. When weather

### 2017-07-12 manuscript/draft.Rmd
re-order the results

--- 2017-07-11 manuscript/draft.Rmd
@@
 sites,  richness varied from [[N]] to [[N]] with an average richness of [[N]]
-species. Within-site richness variation (among years) was small (standard
-deviation of about 5 species, after controlling for observer effects).
+species.
 
@@
 variation, where a given observer might report a different number of species in
-different years. There was little temporal autocorrelation in these residuals
-([[about 0.04, +/- something]]), suggesting that richness is close to stationary
-on these time scales once observer differences are accounted for.
+different years. In the training set, the residuals had a standard deviation of
+about [[+/- 5]] species. After correcting for observer differences, there was
+little temporal autocorrelation in these residuals ([[about 0.04, +/-
+something]]), suggesting that richness is approximately stationary on these time
+scales.
 
 When comparing forecasts for richness across sites all methods performed well
-(Figure 2; R^2 XX-XX%). However stacked and joint SDMs and the macroecological
-model both failed to successfully forecast the highest richness sites resulting
-in a notable clustering of predicted values near ~60 species and the poorest
-performance in the cross-site comparison (R^2 XX-XX%).
+(Figure 2; R^2 XX-XX%). However SDMs (both stacked and joint) and the
+macroecological model all failed to successfully forecast the highest-richness
+sites resulting in a notable clustering of predicted values near ~60 species and
+the poorest performance in the cross-site comparison (R^2 XX-XX%).
 
-For most combinations of metric and forecast time lag, the best-performing model
-for forecasting within-site variability was either the "average" baseline model
-or the ARIMA model [[Not sure what the bit about within-site variability means
-here. Also, didn't Naive usually win when observer effects weren't corrected
-for?]] (Figure 3). In general, each model's error increased more-or-less
-linearly with time, although the slopes and initial error rates varied across
-models (Figure 3). The site-level models (i.e. the baselines and the ARIMA
-methods) tended to do very well (Figure 3) on short time horizons, especially in
-terms of RMSE.
+In general, each model's error increased more-or-less linearly with time,
+although the slopes and initial error rates varied across models (Figure 3). The
+site-level models (i.e. the baselines and the ARIMA methods) tended to do very
+well (Figure 3), especially on short time horizons, and especially in terms of
+RMSE. Of these, the "Naive" model performed especially well at sites with
+uncorrected observer turnover (since its predictions were best-calibrated to
+observers that were active in later years).
 
-In general, the models produced confidence intervals that were too narrow,
-indicating overconfident predictions. Of these, the random forest-based SDM
-stack was the most overconfident, with observed richness values frequently
-falling well outside their confidence limits (Figure 3C). As noted elsewhere
-[@harris2015], this overconfidence is a product of the assumption in stacked
-SDMs that errors in the species-level predictions are independent. The JSDM and
-macroecological model has the best coverage, though each model still
-underestimated the possible variation in richness. The "naive" model
-was the only model whose confidence intervals were too wide (Figure 3), which
-can be attributed to the rapid rate at which these intervals expand as the time
-lag increases (Figure 1).
-
-RMSE and deviance produced qualitatively similar results in most cases, with the
-notable exception of the stacked SDMs (Figure 3A and 3B). Deviance, unlike RMSE,
-depends on the full predictive distribution of a model, including the model's
-estimated uncertainty. As a result, the stacked SDMs' deviance was notably
-higher than the next-worst model even though its RMSE was not unusual. On the
-other hand, the GBM-based "macroecological" model and especially the mistnet
-JSDM had reasonably well-calibrated uncertainty estimates, even in the presence
-of uncorrected observer biases (Figure 3C). This prevented their deviance from
-rising as quickly as the site-level models as the observer population evolved
-and resulted in better comparative performance for these models in terms of
-deviance (Figure 3B).
+Most models produced confidence intervals that were too narrow, indicating
+overconfident predictions. The random forest-based SDM stack was the most
+overconfident model, with observed richness values frequently falling well
+outside its confidence limits (Figure 3C). This stacked SDM's narrow
+predictive distribution also caused it to have notably higher deviance (Figure
+3B) than the next-worst model, even though its point estimates were not
+unusually bad in terms of RMSE (3A), and the other models' relative performance
+was not nearly as sensitive to the choice of error metric. As discussed
+elsewhere [@harris2015], this overconfidence is a product of the assumption in
+stacked SDMs that errors in the species-level predictions are independent.
+Because they avoided this assumption, the GBM-based "macroecological" model and
+especially the mistnet JSDM had reasonably well-calibrated uncertainty
+estimates, even in the presence of uncorrected observer biases (Figure 3C). This
+calibration improved both models' deviance relative to the baseline models,
+especially as the observer population evolved (Figure 3B). The "naive" model was
+the only model whose confidence intervals were too wide (Figure 3), which can be
+attributed to the rapid rate at which these intervals expand as the time lag
+increases (Figure 1).
 
 Accounting for the uncertainty associated with differences among observers
-resulted in improvements to all metrics of fit (Figure 4). The one exception was
-RMSE of the "Naive" model, which remained unchanged. The improvements in point
-estimates primarily resulted from large improvements in a small number of
-forecasts, presumably ones where a large difference between observers resulted
-in a large shift in predicted richness values that the models could not
-otherwise accommodate. Most forecasts exhibited only small changes in RMSE.
-However these changes were sufficient to change the rank ordering of model
-performance. For example, the "naive" model generally had the lowest average
-error when observer effects were not explicitly accounted for, switching to the
-"average" and ARIMA model once observer differences were corrected. Improvements
-in coverage were notable for all models except Naive and JSDM, meaning that
-incorporating observer effects resulted in improved quantification of
-uncertainty. The lack of change in the mistnet JSDM seems to indicate that its
-latent variables were already accounting for observer-related uncertainty. These
-differences meant that, in terms of deviance, neither the JSDM nor the
-macroecological model model benefitted as much from the as the site-level models
-did from the observer corrections. As a result, the average and ARIMA models
-maintained a small average advantage over the JSDM in terms of deviance at
-longer time scales, rather than falling behind after about five years.
+resulted in improvements to nearly all metrics of fit (Figure 4). The
+improvements in point estimates (as measured by RMSE) primarily resulted from
+large improvements in a small number of forecasts where a large difference
+between observers resulted in a large shift in predicted richness values that
+the models could not otherwise accommodate. The "Average" and ARIMA models
+generally showed the largest improvement here, and the Naive model showed the
+smallest (since it was already very responsive to observer turnover). As a
+result, the "Average" and "ARIMA" models tended to perform the best, once
+observer differences were accounted for. Improvements in coverage were notable
+for all models except Naive and JSDM, meaning that unmodeled observer effects
+had been partly responsible for the poor quantification of uncertainty. As a
+result, the average and ARIMA models maintained a small average advantage over
+the JSDM in terms of deviance at longer time scales, rather than falling behind
+after about five years as in Figure 3B.
 

### 2017-07-12 manuscript/draft.Rmd
re-arrange Data section

@@
 
-Bird species richness was obtained from the Breeding Bird Survey (BBS)
-[@pardieck2017] using the Data Retriever software [@morris2013] and
+Bird species richness was obtained from the North American Breeding Bird Survey
+(BBS) [@pardieck2017] using the Data Retriever software [@morris2013] and
 rdataretriever R package [@mcglinn2017]. The BBS data was filtered to exclude
@@
 sampled by BBS methods; @hurlbert2005], as well as unidentified species, and
-hybrids. Due to limitations of the environmental data (below), we did not
-include any data from outside the continental United States. Observed richness
-from 1982 (the first complete year of NDVI data) to 2003 were used for training
-the models, and data from 2004 to 2013 was used for testing their performance.
-Only BBS routes that were sampled during 70% of the years in the training period
-were used (i.e., routes with at least 16 annual observations). All data from
-routes that did not meet BBS acceptable criteria were excluded, as were routes
-that had not been visited during at least 70% of the years in our model-training
-window. The resulting dataset included [[N]] species and [[N]] sites. Among
-sites,  richness varied from [[N]] to [[N]] with an average richness of [[N]]
-species.
+hybrids. All data from routes that did not meet BBS acceptable criteria were
+also excluded.
 
@@
 of each BBS route.
+
+We used observed richness values from 1982 (the first complete year of NDVI
+data) to 2003 for training the models, and data from 2004 to 2013 was used for
+testing their performance. We only used BBS routes from the continental United
+States (i.e. routes where PRISM data was available), and restricted the analysis
+to routes that were sampled during 70% of the years in the training period were
+used (i.e., routes with at least 16 annual observations). The resulting dataset
+included [[N]] transect runs of [[N]] unique sites, and included [[N]] species.
+Site-level richness varied from [[N]] to [[N]] with an average richness of [[N]]
+species.
 

### 2017-07-12 R/observer_model.R
minor fixes

@@ -8,7 +8,7 @@
 fit_observer_model = function(stan_file = "mixed.stan", seed = 1,
                               output_file = "observer_model.rds", 
                               chains = 2, cores = 2, iter = 2000,
-                              thin = 4, settings...){
+                              thin = 4, settings, ...){
   library(rstan) # prevent annoying NAMESPACE issue described in
   # https://github.com/stan-dev/rstan/issues/353
   

### 2017-07-12 R/sdm-rf.R

@@ -35,8 +35,9 @@ rf_predict_species = function(sp_id, bbs, settings, x_richness, use_obs_model,
                    vars = vars,
                    x_richness = x_richness,
                    last_train_year = settings$last_train_year,
-                   future, future_observer_effects,
-                   settings) %>% 
+                   future = future, 
+                   future_observer_effects = future_observer_effects,
+                   settings = settings) %>% 
     bind_rows()
   
   path = paste0("rf_predictions/sp_", sp_id, "_", use_obs_model, ".csv.gz")
@@ -57,7 +58,8 @@ rf_predict_richness = function(bbs, x_richness, settings, use_obs_model,
       rf_predict_species(sp_id, bbs = bbs, x_richness = x_richness, 
                          settings = settings, 
                          use_obs_model = use_obs_model,
-                         future, future_observer_effects)
+                         future = future, 
+                         future_observer_effects = future_observer_effects)
     },
     mc.cores = 8,
     mc.preschedule = FALSE

### 2017-07-12 analysis/fit-predict.R

@@ -47,8 +47,8 @@ if (settings$timeframe == "future") {
   future_observer_effects = obs_model$observer_sigma %>% 
     purrr::map(~rnorm(nrow(future), 0, .x))
 } else {
-  future = NULL
-  future_observer_effects = NULL
+    future = NULL
+    future_observer_effects = NULL
 }
 
 # Reclaim memory
@@ -101,7 +101,7 @@ expand.grid(fun_name = c("naive", "auto.arima"),
   parallel::mclapply(
     function(grid_row){
       do.call(make_all_forecasts, 
-             c(x = list(x_richness), settings = list(settings), grid_row))
+              c(x = list(x_richness), settings = list(settings), grid_row))
     },
     mc.cores = 8, 
     mc.preschedule = FALSE

### 2017-07-13 R/observer_model.R
fix stupid bugs

@@ -5,10 +5,11 @@
 #' 
 #' @return \code{NULL}. Saves a list of relevant values in `output_file`
 #'   and the `stanfit` object to disk.
-fit_observer_model = function(stan_file = "mixed.stan", seed = 1,
+fit_observer_model = function(settings, output_dir,
+                              stan_file = "mixed.stan", seed = 1,
                               output_file = "observer_model.rds", 
                               chains = 2, cores = 2, iter = 2000,
-                              thin = 4, settings, ...){
+                              thin = 4, ...){
   library(rstan) # prevent annoying NAMESPACE issue described in
   # https://github.com/stan-dev/rstan/issues/353
   

### 2017-07-13 R/post-observer.R

@@ -1,6 +1,5 @@
 #' @importFrom forecast Arima auto.arima
-make_forecast = function(x, fun_name, use_obs_model, settings, 
-                         future_observer_effects, ...){
+make_forecast = function(x, fun_name, use_obs_model, settings, ...){
   # `Forecast` functions want NAs for missing years, & want the years in order
   x = complete(x, year = settings$start_yr:settings$last_train_year) %>% 
     arrange(year)
@@ -46,18 +45,15 @@ make_forecast = function(x, fun_name, use_obs_model, settings,
     list(NA)
   }
   
-  i = unique(x$iteration)
-  
   # Distance between `upper` and `lower` is 2 sd, so divide by 2
   data_frame(year = seq(settings$last_train_year + 1, settings$end_yr), 
          mean = c(fcst$mean), sd = c(fcst$upper - fcst$lower) / 2, 
          model = fun_name, use_obs_model = use_obs_model,
-         coef_names = coef_names, 
-         observer_effect = future_observer_effects[[i]])
+         coef_names = coef_names)
 }
 
 make_all_forecasts = function(x, fun_name, use_obs_model, 
-                              settings, future_observer_effects, ...){
+                              settings, observer_sigmas, ...){
   forecast_data = x %>% 
     filter(year <= settings$last_train_year) %>% 
     group_by(site_id, iteration)
@@ -74,6 +70,11 @@ make_all_forecasts = function(x, fun_name, use_obs_model,
     left_join(select(x_richness, -sd), c("site_id", "year", "iteration"))
   
   if (use_obs_model) {
+    if (settings$timeframe == "future") {
+      # Calculate random observer effects
+      out$observer_effect = rnorm(nrow(out), 
+                                  sd = observer_sigmas[out$iteration])
+    }
     # Observer effect was subtraced out in make_forcast. Add it back in here.
     out = mutate(out, mean = mean + observer_effect)
   }
@@ -82,12 +83,12 @@ make_all_forecasts = function(x, fun_name, use_obs_model,
          use_obs_model, coef_names)
 }
 
-make_test_set = function(x, future, future_observer_effects, settings){
+make_test_set = function(x, future, observer_sigmas, settings){
   if (settings$timeframe == "future") {
-    i = unique(x$iteration)
+    obs_sd = unique(x$observer_sigma)
     test = mutate(future, 
-                  observer_effect = !!future_observer_effects[[i]],
-                  expected_richness = richness - observer_effect)
[…]

### 2017-07-13 analysis/fit-predict.R

@@ -17,7 +17,7 @@ prepend_timeframe = function(x) {
 dir.create(prepend_timeframe(NULL), showWarnings = FALSE)
 
 if (!file.exists(prepend_timeframe("observer_model.rds"))) {
-  fit_observer_model(settings = settings)
+  fit_observer_model(settings = settings, output_dir = prepend_timeframe(""))
 }
 obs_model = readRDS(prepend_timeframe("observer_model.rds"))
 
@@ -41,14 +41,15 @@ if (settings$timeframe == "future") {
            year > !!settings$last_train_year) %>% 
     select(which(colnames(.) %in% !!colnames(x_richness)))
   
-  # Randomly-sampled observers for each `iteration`. Not used for the
-  # "average" model, which will just use observer_sigma directly without
-  # simulations.
-  future_observer_effects = obs_model$observer_sigma %>% 
-    purrr::map(~rnorm(nrow(future), 0, .x))
+  # Standard deviation of "future" observers for each iteration, add as a 
+  # column in x_richness
+  observer_sigmas = obs_model$observer_sigma
+  x_richness = enframe(observer_sigmas, name = "iteration", 
+                       value = "observer_sigma") %>% 
+    right_join(x_richness, "iteration")
 } else {
     future = NULL
-    future_observer_effects = NULL
+    observer_sigmas = NULL
 }
 
 # Reclaim memory
@@ -57,15 +58,6 @@ gc()
 
 
 # Fit & save models --------------------------------------------------------
-
-# For "future" predictions of the "average" model, we'll need to know how 
-# much uncertainty to use at the observer level. Otherwise, this is already
-# taken care of elsewhere.
-observer_sigma = if (settings$timeframe == "future") {
-  sqrt(mean(obs_model$observer_sigma^2))
-} else{
-  0
-}
 
 # "Average" model with observer effects. Observer_sigma is only nonzero in
 # the "future", where we haven't already simulated the observer-related 
@@ -98,13 +90,15 @@ expand.grid(fun_name = c("naive", "auto.arima"),
             use_obs_model = c(TRUE, FALSE),
             stringsAsFactors = FALSE) %>% 
   transpose() %>% 
-  parallel::mclapply(
+  lapply(
     function(grid_row){
-      do.call(make_all_forecasts, 
-              c(x = list(x_richness), settings = list(settings), grid_row))
-    },
-    mc.cores = 8, 
-    mc.preschedule = FALSE
+      make_all_forecasts(x = x_richness, settings = settings, 
[…]

### 2017-07-13 R/sdm-rf.R
fix RF

@@ -1,6 +1,6 @@
 #' @importFrom randomForest randomForest
 one_rf_tree = function(bbs, vars, sp_id, iter, use_obs_model, x_richness,
-                       last_train_year, future, future_observer_effects,
+                       last_train_year, future, observer_sigmas,
                        settings){
   d = bbs %>% 
     select(site_id, year, species_id, abundance) %>% 
@@ -19,7 +19,8 @@ one_rf_tree = function(bbs, vars, sp_id, iter, use_obs_model, x_richness,
     ntree = 1
   )  
   
-  test = make_test_set(d, future, future_observer_effects, settings)
+  test = cbind(make_test_set(d, future, observer_sigmas, settings),
+               species_id = sp_id, iteration = unique(d$iteration))
   test$mean = predict(rf, test, type = "prob")[,2]
   test$use_obs_model = use_obs_model
   
@@ -27,7 +28,7 @@ one_rf_tree = function(bbs, vars, sp_id, iter, use_obs_model, x_richness,
 }
 
 rf_predict_species = function(sp_id, bbs, settings, x_richness, use_obs_model,
-                              future, future_observer_effects){
+                              future, observer_sigmas){
   vars = c(settings$vars, if (use_obs_model) {"observer_effect"})
   iters = sort(unique(x_richness$iteration))
   results = lapply(iters, one_rf_tree,
@@ -36,7 +37,7 @@ rf_predict_species = function(sp_id, bbs, settings, x_richness, use_obs_model,
                    x_richness = x_richness,
                    last_train_year = settings$last_train_year,
                    future = future, 
-                   future_observer_effects = future_observer_effects,
+                   observer_sigmas = observer_sigmas,
                    settings = settings) %>% 
     bind_rows()
   
@@ -50,16 +51,16 @@ rf_predict_species = function(sp_id, bbs, settings, x_richness, use_obs_model,
 }
 
 rf_predict_richness = function(bbs, x_richness, settings, use_obs_model,
-                               future, future_observer_effects) {
+                               future, observer_sigmas) {
   
-  out = parallel::mclapply(
+  out = lapply(
     unique(bbs$species_id), 
     function(sp_id){
       rf_predict_species(sp_id, bbs = bbs, x_richness = x_richness, 
                          settings = settings, 
                          use_obs_model = use_obs_model,
                          future = future, 
-                         future_observer_effects = future_observer_effects)
+                         observer_sigmas = observer_sigmas)
     },
     mc.cores = 8,
     mc.preschedule = FALSE

### 2017-07-13 analysis/fit-predict.R

@@ -90,15 +90,15 @@ expand.grid(fun_name = c("naive", "auto.arima"),
             use_obs_model = c(TRUE, FALSE),
             stringsAsFactors = FALSE) %>% 
   transpose() %>% 
-  lapply(
+  mclapply(
     function(grid_row){
       make_all_forecasts(x = x_richness, settings = settings, 
                          observer_sigmas = observer_sigmas,
                          fun_name = grid_row$fun_name, 
                          use_obs_model = grid_row$use_obs_model)
-    }#,
-    #mc.cores = 8, 
-    #mc.preschedule = FALSE
+    },
+    mc.cores = 8, 
+    mc.preschedule = FALSE
   ) %>% 
   bind_rows() %>% 
   saveRDS(file = prepend_timeframe("forecast.rds"))

### 2017-07-13 R/sdm-rf.R
parallel

@@ -53,7 +53,7 @@ rf_predict_species = function(sp_id, bbs, settings, x_richness, use_obs_model,
 rf_predict_richness = function(bbs, x_richness, settings, use_obs_model,
                                future, observer_sigmas) {
   
-  out = lapply(
+  out = parallel::mclapply(
     unique(bbs$species_id), 
     function(sp_id){
       rf_predict_species(sp_id, bbs = bbs, x_richness = x_richness, 

### 2017-07-13 analysis/fit-predict.R
fix the "average" model

@@ -59,28 +59,50 @@ gc()
 
 # Fit & save models --------------------------------------------------------
 
-# "Average" model with observer effects. Observer_sigma is only nonzero in
-# the "future", where we haven't already simulated the observer-related 
-# uncertainty
-x_richness %>% 
-  filter(!in_train) %>% 
-  mutate(mean = intercept + observer_effect + site_effect,
-         model = "average", use_obs_model = TRUE,
-         sd = sqrt(sd^2 + !!observer_sigma^2)) %>% 
-  select(site_id, year, mean, sd, iteration, richness, model, use_obs_model) %>% 
-  saveRDS(file = prepend_timeframe("avg_TRUE.rds"))
 
-# "Average" model without observer effects
-# Use site-level means and sds from the training set as test-set predictions
-x_richness %>% 
-  filter(in_train) %>% 
-  group_by(site_id) %>% 
-  summarize(mean = mean(richness), sd = sd(richness), model = "average", 
-            use_obs_model = FALSE) %>% 
-  left_join(select(x_richness, -sd), "site_id") %>% 
-  filter(!in_train) %>% 
-  select(site_id, year, mean, sd, richness, model, use_obs_model) %>% 
-  saveRDS(file = prepend_timeframe("avg_FALSE.rds"))
+
+if (settings$timeframe == "future") {
+  year_grid = expand.grid(iteration = unique(x_richness$iteration), 
+                          year = years_to_use)
+  
+  # Calculate expected richness for "typical" observer, add in noise for
+  # unknown observers. Then repeat for all year/iteration combinations
+  x_richness %>% 
+    mutate(mean = intercept + site_effect,
+           sd = sqrt(sd^2 + observer_sigma^2)) %>% 
+    distinct(iteration, site_id, mean, sd) %>% 
+    left_join(year_grid, "iteration") %>% 
+    saveRDS(file = prepend_timeframe("avg_TRUE.rds"))
+  
+  # When there's no observer effect, just calculate the empirical mean and sd 
+  # for each site, then replicate across years.
+  avg_false = x_richness %>% 
+    filter(in_train) %>% 
+    group_by(site_id) %>% 
+    summarize(mean = mean(richness), sd = sd(richness))
+  map_df(years_to_use, function(x) cbind(avg_false, year = x)) %>% 
+    saveRDS(file = prepend_timeframe("avg_false.rds"))
+} else {
+  # "Average" model with observer effects
[…]

### 2017-07-13 analysis/fit-predict.R
filenames

@@ -81,7 +81,7 @@ if (settings$timeframe == "future") {
     group_by(site_id) %>% 
     summarize(mean = mean(richness), sd = sd(richness))
   map_df(years_to_use, function(x) cbind(avg_false, year = x)) %>% 
-    saveRDS(file = prepend_timeframe("avg_false.rds"))
+    saveRDS(file = prepend_timeframe("avg_FALSE.rds"))
 } else {
   # "Average" model with observer effects
   x_richness %>% 
@@ -89,7 +89,7 @@ if (settings$timeframe == "future") {
     mutate(mean = intercept + observer_effect + site_effect,
            model = "average", use_obs_model = TRUE) %>% 
     select(site_id, year, mean, sd, iteration, richness, model, use_obs_model) %>% 
-    saveRDS(file = "avg_TRUE.rds")
+    saveRDS(file = prepend_timeframe("avg_TRUE.rds"))
   
   # "Average" model without observer effects
   # Use site-level means and sds from the training set as test-set predictions
@@ -101,7 +101,7 @@ if (settings$timeframe == "future") {
     left_join(select(x_richness, -sd), "site_id") %>% 
     filter(!in_train) %>% 
     select(site_id, year, mean, sd, richness, model, use_obs_model) %>% 
-    saveRDS(file = "avg_FALSE.rds")
+    saveRDS(file = prepend_timeframe("avg_FALSE.rds"))
 }
 
 

### 2017-07-13 manuscript/draft.Rmd
Add R and R package citations

@@
 States (i.e. routes where PRISM data was available), and restricted the analysis
-to routes that were sampled during 70% of the years in the training period were
-used (i.e., routes with at least 16 annual observations). The resulting dataset
+to routes that were sampled during 70% of the years in the training period
+(i.e., routes with at least 16 annual observations). The resulting dataset
 included [[N]] transect runs of [[N]] unique sites, and included [[N]] species.
@@
 
+All analyses were conducted using R [@R]. Primary R packages used in the
+analysis included dplyr [@dplyr], tidyr [@tidyr], gimms [@gimms], sp
+[@sp1; @sp2], raster [@raster], prism [@prism], rdataretriever
+[@rdataretriever], forecasts [@forecast1; @forecast2], git2r [@git2r], ggplot
+[@ggplot2], viridius [@viridis], rstan [@rstan], yaml [@yaml], purrr [@purrr],
+gbm [@gbm], randomForest [@randomForest], purrrlyr [@purrrlyr]. Code to fully
+reproduce this analysis is available on GitHub
+(https://github.com/weecology/bbs-forecasting) and archived on Zenodo ().
+
 ## Results
@@
 
-In general, each model's error increased more-or-less linearly with time,
-although the slopes and initial error rates varied across models (Figure 3). The
-site-level models (i.e. the baselines and the ARIMA methods) tended to do very
-well (Figure 3), especially on short time horizons, and especially in terms of
-RMSE. Of these, the "Naive" model performed especially well at sites with
-uncorrected observer turnover (since its predictions were best-calibrated to
-observers that were active in later years).
+While the models generally did well in absolute terms (Figure 2), they tended
+not to outperform the "average" baseline from the generalized linear mixed model
+(Figure 3A). [[Stats go here]].  The auto-ARIMA was generally the best-performing
+non-baseline model, but in many cases, the best-fitting ARIMA model chosen by
+the model selection procedure was no different from the "average" model.
 
 Most models produced confidence intervals that were too narrow, indicating
-overconfident predictions. The random forest-based SDM stack was the most
-overconfident model, with observed richness values frequently falling well
-outside its confidence limits (Figure 3C). This stacked SDM's narrow
-predictive distribution also caused it to have notably higher deviance (Figure
-3B) than the next-worst model, even though its point estimates were not
-unusually bad in terms of RMSE (3A), and the other models' relative performance
-was not nearly as sensitive to the choice of error metric. As discussed
-elsewhere [@harris2015], this overconfidence is a product of the assumption in
-stacked SDMs that errors in the species-level predictions are independent.
-Because they avoided this assumption, the GBM-based "macroecological" model and
-especially the mistnet JSDM had reasonably well-calibrated uncertainty
-estimates, even in the presence of uncorrected observer biases (Figure 3C). This
-calibration improved both models' deviance relative to the baseline models,
-especially as the observer population evolved (Figure 3B). The "naive" model was
-the only model whose confidence intervals were too wide (Figure 3), which can be
-attributed to the rapid rate at which these intervals expand as the time lag
-increases (Figure 1).
+overconfident predictions (Figure 3B). The random forest-based SDM stack was the
+most overconfident model, with only XX% of observations falling inside its 95%
+confidence intervals. This stacked SDM's narrow predictive distribution also
+caused it to have notably higher deviance (Figure 3C) than the next-worst model,
+even though its point estimates were not unusually bad in terms of RMSE (3A). As
+discussed elsewhere [@harris2015], this overconfidence is a product of the
+assumption in stacked SDMs that errors in the species-level predictions are
+independent. Because they avoided this assumption, the GBM-based
+"macroecological" model and especially the mistnet JSDM had reasonably
+well-calibrated uncertainty estimates (Figure 3B); as a result, their relative
+performance was higher in terms of deviance than in terms of RMSE. The "naive"
+model was the only model whose confidence intervals were too wide (Figure 3B),
+which can be attributed to the rapid rate at which these intervals expand as the
+time lag increases (Figure 1).
 
-Accounting for the uncertainty associated with differences among observers
-resulted in improvements to nearly all metrics of fit (Figure 4). The
-improvements in point estimates (as measured by RMSE) primarily resulted from
-large improvements in a small number of forecasts where a large difference
-between observers resulted in a large shift in predicted richness values that
-the models could not otherwise accommodate. The "Average" and ARIMA models
-generally showed the largest improvement here, and the Naive model showed the
-smallest (since it was already very responsive to observer turnover). As a
-result, the "Average" and "ARIMA" models tended to perform the best, once
-observer differences were accounted for. Improvements in coverage were notable
-for all models except Naive and JSDM, meaning that unmodeled observer effects
-had been partly responsible for the poor quantification of uncertainty. As a
-result, the average and ARIMA models maintained a small average advantage over
-the JSDM in terms of deviance at longer time scales, rather than falling behind
-after about five years as in Figure 3B.
+In general, each model's error increased more-or-less linearly with time,
+although the slopes and initial error rates varied across models (Figure 4). The
+mistnet JSDM and GBM-based macroecological model's deviance tended to rise less
+rapidly over time than the baseline models' deviance did; for the ten-year
+forecasts, these environment-based models outperformed the Naive model and 
+nearly matched the performance of the Average model.
+
+When the differences among observers were not explicitly accounted for, nearly
+all metrics of fit worsened (Figure 5). The increased error resulted from a
+small number of forecasts where observer turnover caused a large shift in the
+reported richness values. The Naive model was less sensitive to these shifts,
+because it largely ignored the richness values reported by observers that had
+retired by the end of the training period (Figure 1). The Average model, which
+gave equal weight to observations from the whole training period, showed a
+larger decline in performance -- especially in terms of coverage. The mistnet
+JSDM was notable here because the coverage of its prediction intervals retained
+very good coverage in the presence of this observer-based noise, which we
+attribute to its ability to model this variation with its latent variables.
 

### 2017-07-13 manuscript/draft.Rmd
Add Acknowledgments section

@@
 
+## Acknowledgments
+
+This research was supported by the Gordon and Betty Moore Foundation's
+Data-Driven Discovery Initiative through Grant GBMF4563 to E.P. White. We thank
+the developers and providers of the data and software that made this research
+possible including: the PRISM Climate Group at Oregon State University, the
+staff at USGS and volunteer citizen scientists associated with the North
+American Breeding Bird Survey, NASA, the World Climate Research Programme's
+Working Group on Coupled Modelling and its working groups, the
+U.S. Department of Energy's Program for Climate Model Diagnosis and
+Intercomparison, and the Global Organization for Earth System Science Portals.
+
 ## Box 1: Ten simple rules for making and evaluating ecological forecasts

### 2017-07-13 manuscript/draft.Rmd

--- 2017-07-13 manuscript/draft.Rmd
@@
 horizon.
-
-All analyses were conducted using R [@R]. Primary R packages used in the
-analysis included dplyr [@dplyr], tidyr [@tidyr], gimms [@gimms], sp
-[@sp1; @sp2], raster [@raster], prism [@prism], rdataretriever
-[@rdataretriever], forecasts [@forecast1; @forecast2], git2r [@git2r], ggplot
-[@ggplot2], viridius [@viridis], rstan [@rstan], yaml [@yaml], purrr [@purrr],
-gbm [@gbm], randomForest [@randomForest], purrrlyr [@purrrlyr]. Code to fully
-reproduce this analysis is available on GitHub
-(https://github.com/weecology/bbs-forecasting) and archived on Zenodo ().
 

### 2017-07-13 analysis/fit-predict.R
actually create the directory (#146)

@@ -14,7 +14,7 @@ settings$timeframes = NULL
 prepend_timeframe = function(x) {
   paste0("results", "/", timeframe, "/", x)
 }
-dir.create(prepend_timeframe(NULL), showWarnings = FALSE)
+dir.create(prepend_timeframe(NULL), showWarnings = FALSE, recursive = TRUE)
 
 if (!file.exists(prepend_timeframe("observer_model.rds"))) {
   fit_observer_model(settings = settings, output_dir = prepend_timeframe(""))

### 2017-07-13 manuscript/draft.Rmd
Make the manuscript knitable
* Fix reference formatting issues
* Fix location of settings.yml
* Fix typos in in line refs
* Adding missing refs

--- 2017-07-13 manuscript/draft.Rmd
@@
 - Ethan P. White
+bibliography: refs.bib
 output:
@@
 
-`r settings = yaml::yaml.load_file("settings.yaml")`
+`r settings = yaml::yaml.load_file("../settings.yaml")`
 
@@
 ecological processes governing a system [@clark2001; @houlahan2017;
-@dietz2017]. One area of particular interest for forecasting is biodiversity
+@dietze2017]. One area of particular interest for forecasting is biodiversity
 since it is important for ecosystem function, central to conservation planning,
@@
 ecological forecasts are applied and the resulting difficulty in assessing
-whether the predicted changes occurred [@dietzinreview]. What we do know comes
+whether the predicted changes occurred [@dietze2016]. What we do know comes
 from a small number of hindcasting studies, where models are built using data on
@@
 1) broadly involve: 1) expanding the use of data to include biological and
-environmental time-series [@treddenick2016]; 2) accounting for uncertainty in
+environmental time-series [@tredennick2016]; 2) accounting for uncertainty in
 observations, processes, model choice, and forecast evaluation
-[@clark2001, @dietz2017]; and 3) conducting meaningful evaluations of the
+[@clark2001, @dietze2017]; and 3) conducting meaningful evaluations of the
 forecasts by hindcasting, archiving short-term forecasts, and comparing
 forecasts to baselines to determine if the forecasts are more accurate than
-assuming the system is basically static [@Perretti2013].
+assuming the system is basically static [@perretti2013].
 
@@
 **Time series models.** We used Auto-ARIMA models [based on the `auto.arima`
-function in @hyndman2016] to represent a array of different time-series modeling
+function in @forecast1] to represent a array of different time-series modeling
 approaches. These models can include an autoregressive component (as in the
@@
 include each of these components, and how many terms to include for each one,
-using AICc [@hyndman2016]. Since there is no seasonal component to the BBS
+using AICc [@forecast1]. Since there is no seasonal component to the BBS
 time-series, we did not include a season component in these models. Otherwise we
@@
 has been raised for knowing how far into the future a model can be effectively
-used for forecasting [@petchy2015], but our results suggest a broader value to
+used for forecasting [@petchey2015], but our results suggest a broader value to
 this approach for considering the potential importance of different models and
-processes for making forecasts at different scales. [[Should make sure Petchy et
+processes for making forecasts at different scales. [[Should make sure Petchey et
 al. don't make this point]] Our results show that for forecasts 1-3 years in the
@@
 efforts. It is important to correctly incorporate this uncertainty in the
-predictor variables into forecasting models [@clark2001; @dietz2017]. Difficulty
+predictor variables into forecasting models [@clark2001; @dietze2017]. Difficulty
 in forecasting future environmental conditions at small scales will present
@@
 only be made with a small number of opportune datasets. For examples hindcasts
-have been used to asses models of species richness [@alger2009, @distler2015]
-and distribution [@rappacciuolo2012, @moran-ordonez2017, @araujo2005,
+have been used to assess models of species richness [@algar2009, @distler2015]
+and distribution [@rapacciuolo2012, @moran-ordonez2017, @araujo2005,
 @eskildsen2013] using data aggregated across similar time scales. More
@@
 and determine how far into the future we can successfully forecast
-[@tredenick2016; @dietz2017]. Forecasts based on the models in this paper from
+[@tredennick2016; @dietze2017]. Forecasts based on the models in this paper from
 now until 2050 are openly archived on Dryad so that we and others can assess how
@@
 ecological and environmental time-series data allows the consideration of actual
-dynamics from both a process and error structure perspective [@treddenick2016a].
+dynamics from both a process and error structure perspective [@tredennick2016].
 
@@
 include sources of uncertainty and propagate them through the forecast where
-possible [@clark2001; @dietz2017]. Evaluations of forecasts should assess the
-accuracy of uncertainties as well as point estimates [@dietz2017].
+possible [@clark2001; @dietze2017]. Evaluations of forecasts should assess the
+accuracy of uncertainties as well as point estimates [@dietze2017].
 
@@
 In general the accuracy of forecasts decreases with the length of time into the
-future being forecast [@petch2015]. This decay in accuracy and the potential for
+future being forecast [@petchey2015]. This decay in accuracy and the potential for
 different rates of decay to result in different relative model performance at
@@
 and to allow the accuracy of the forecasts to be quickly evaluated
-[@treddenick2016].
+[@tredennick2016].

### 2017-07-13 manuscript/draft.Rmd
add long term forecast data methods

--- 2017-07-13 manuscript/draft.Rmd
@@
 of each BBS route.
+
+We made long term forecasts from 2014-2050 using CMIP5 multi-model ensemble
+dataset as the source for climatic variables. Precipitation and temperature from
+37 downscaled model runs [@brekke2013, see Table S1] using the RCP6.0 scenario
+were averaged together to create a single ensemble used to calculate the
+bioclimatic variables for North America. For NDVI we used the per-site average
+values from 2000-2013 as a naive forecast. For observer effects (see below) each
+site was set to have zero observer bias.
 
@@
 horizon.
+
+All analyses were conducted using R [@R]. Primary R packages used in the
+analysis included dplyr [@dplyr], tidyr [@tidyr], gimms [@gimms], sp
+[@sp1; @sp2], raster [@raster], prism [@prism], rdataretriever
+[@rdataretriever], forecasts [@forecast1; @forecast2], git2r [@git2r], ggplot
+[@ggplot2], viridius [@viridis], rstan [@rstan], yaml [@yaml], purrr [@purrr],
+gbm [@gbm], randomForest [@randomForest], purrrlyr [@purrrlyr]. Code to fully
+reproduce this analysis is available on GitHub
+(https://github.com/weecology/bbs-forecasting) and archived on Zenodo ().
 

### 2017-07-14 analysis/fit-predict.R
fix directory again

@@ -17,7 +17,7 @@ prepend_timeframe = function(x) {
 dir.create(prepend_timeframe(NULL), showWarnings = FALSE, recursive = TRUE)
 
 if (!file.exists(prepend_timeframe("observer_model.rds"))) {
-  fit_observer_model(settings = settings, output_dir = prepend_timeframe(""))
+  fit_observer_model(settings = settings, output_file = prepend_timeframe("observer_model.rds"))
 }
 obs_model = readRDS(prepend_timeframe("observer_model.rds"))
 

### 2017-07-17 analysis/fit-predict.R
fix stupid bugs

@@ -131,7 +131,7 @@ x_richness %>%
   purrrlyr::by_slice(make_gbm_predictions, use_obs_model = TRUE,
                      settings = settings, future = future,
                      observer_sigmas = observer_sigmas) %>% 
-  bind_rows()
+  bind_rows() %>% 
   saveRDS(file = prepend_timeframe("gbm_TRUE.rds"))
 
 # GBM richness without observer effects:
@@ -140,8 +140,8 @@ x_richness %>%
     purrrlyr::by_slice(make_gbm_predictions, use_obs_model = FALSE,
                        settings = settings, future = future,
                        observer_sigmas = NULL) %>% 
-    bind_rows()
-  saveRDS(file = prepend_timeframe("gbm_FALSE.rds"))
+    bind_rows() %>% 
+    saveRDS(file = prepend_timeframe("gbm_FALSE.rds"))
 
 # fit random forest SDMs -------------------------------------------------------
 

### 2017-07-17 mistnet.R

@@ -5,7 +5,7 @@ library(progress)
 devtools::load_all() 
 
 
-timeframe = "train_32"
+timeframe = "train_22"
 prepend_timeframe = function(x) {
   paste0("results", "/", timeframe, "/", x)
 }
@@ -196,7 +196,7 @@ fit_mistnet = function(iter,
            log_lik = apply(ll, 1, mistnet:::logMeanExp))
   
   # Save the predictions
-  dir.create("mistnet_output", showWarnings = FALSE)
+  dir.create(prepend_timeframe("mistnet_output"), showWarnings = FALSE)
   saveRDS(out, file = prepend_timeframe(paste0("mistnet_output/", "iteration_", 
                                                iter, "_", 
                                                ifelse(CV, "CV", use_obs_model), 

### 2017-07-17 analysis/fit-predict.R
run gbm in parallel

@@ -1,5 +1,6 @@
 library(tidyverse)
 library(purrrlyr) # for by_slice
+library(parallel)
 devtools::load_all()
 timeframe = "train_32"
 
@@ -127,21 +128,29 @@ expand.grid(fun_name = c("naive", "auto.arima"),
 
 # GBM richness with observer effects:
 x_richness %>%
-  group_by(iteration) %>% 
-  purrrlyr::by_slice(make_gbm_predictions, use_obs_model = TRUE,
-                     settings = settings, future = future,
-                     observer_sigmas = observer_sigmas) %>% 
+  split(., .$iteration) %>% 
+  mclapply(make_gbm_predictions, 
+           use_obs_model = TRUE,
+           settings = settings, 
+           future = future,
+           observer_sigmas = NULL,
+           mc.cores = 8,
+           mc.preschedule = FALSE) %>% 
   bind_rows() %>% 
   saveRDS(file = prepend_timeframe("gbm_TRUE.rds"))
 
 # GBM richness without observer effects:
-  x_richness %>%
-    group_by(iteration) %>% 
-    purrrlyr::by_slice(make_gbm_predictions, use_obs_model = FALSE,
-                       settings = settings, future = future,
-                       observer_sigmas = NULL) %>% 
-    bind_rows() %>% 
-    saveRDS(file = prepend_timeframe("gbm_FALSE.rds"))
+x_richness %>%
+  split(., .$iteration) %>% 
+  mclapply(make_gbm_predictions, 
+           use_obs_model = FALSE,
+           settings = settings, 
+           future = future,
+           observer_sigmas = NULL,
+           mc.cores = 8,
+           mc.preschedule = FALSE) %>% 
+  bind_rows() %>% 
+  saveRDS(file = prepend_timeframe("gbm_FALSE.rds"))
 
 # fit random forest SDMs -------------------------------------------------------
 

### 2017-07-17 figures/plots.R
update plotting script

@@ -3,11 +3,18 @@ library(tidyverse)
 library(cowplot)
 settings = yaml::yaml.load_file("settings.yaml")
 
-forecast_results = readRDS("forecast.rds") %>% 
+timeframe = "train_22"
+prepend_timeframe = function(x) {
+  paste0("results", "/", timeframe, "/", x)
+}
+
+
+forecast_results = readRDS(prepend_timeframe("forecast.rds")) %>% 
   filter(!is.na(richness))
 gbm_results = bind_rows(readRDS("gbm_TRUE.rds"),
                         readRDS("gbm_FALSE.rds"))
-mistnet_results = lapply(dir("mistnet_output/", full.names = TRUE,
+mistnet_results = lapply(dir(prepend_timeframe("mistnet_output/"), 
+                             full.names = TRUE,
                              pattern = "TRUE|FALSE"),
                          readRDS) 
 
@@ -15,15 +22,16 @@ mistnet_results = lapply(dir("mistnet_output/", full.names = TRUE,
 mean(map_dbl(mistnet_results, ~mean(.x$mean - .x$richness)))
 
 
-average_results = bind_rows(readRDS("avg_TRUE.rds"), readRDS("avg_FALSE.rds"))
+average_results = bind_rows(readRDS(prepend_timeframe("avg_TRUE.rds")), 
+                            readRDS(prepend_timeframe("avg_FALSE.rds")))
 
 my_var = function(x){
   ifelse(length(x) == 1, 0, var(x))
 }
 
 
-rf_results = bind_rows(readRDS("rf_predictions/all_FALSE.rds"),
-                       readRDS("rf_predictions/all_TRUE.rds"))
+rf_results = bind_rows(readRDS(prepend_timeframe("rf_predictions/all_FALSE.rds")),
+                       readRDS(prepend_timeframe("rf_predictions/all_TRUE.rds")))
 
 bound = bind_rows(forecast_results, gbm_results, average_results, 
                   rf_results, mistnet_results) %>% 
@@ -32,6 +40,16 @@ bound = bind_rows(forecast_results, gbm_results, average_results,
   mutate(diff = richness - mean, z = diff / sd,
          p = pnorm(richness, mean, sd),
          deviance = -2 * dnorm(richness, mean, sd, log = TRUE))
+
+bound %>% 
+  filter(model == "average") %>% 
+  left_join(filter(bound, model != "average"), 
[…]

### 2017-07-20 manuscript/draft.Rmd
Add formatting and line numbers to paper

--- 2017-07-13 manuscript/draft.Rmd
@@
 bibliography: refs.bib
+geometry: margin=1in
+fontsize: 12pt
+header-includes:
+    - \usepackage{times}
+    - \usepackage{setspace}
+    - \doublespacing
+    - \usepackage{lineno}
+    - \linenumbers
 output:
@@
 [@tredennick2016].
+
+## References

### 2017-07-20 figures/plots.R
plot styling

@@ -209,21 +209,18 @@ observer_colors = RColorBrewer::brewer.pal(4, "PuOr")
 # before 2004.
 ggplot(time_series_data, aes(x = year)) +
   geom_ribbon(aes(ymin = mean - 1.96 * sd, ymax = mean + 1.96 * sd),
-              fill = "gray90") +
+              fill = "gray80") +
   geom_ribbon(aes(ymin = mean - 1 * sd, ymax = mean + 1 * sd),
               fill = "gray60") +
-  geom_line(aes(y = mean), size = 2, color = "gray30") +
+  geom_line(aes(y = mean), size = 1.5, color = "gray30") +
   facet_grid(use_obs_model ~ model) +
   geom_line(aes(y = richness, alpha = as.numeric(year < min(bound$year - 1)))) + 
   geom_point(aes(y = richness, fill = factor(observer_id)),
              size = 3.5, shape = 21) +
   scale_alpha(range = c(0, 1), guide = FALSE) + 
-  coord_cartesian(xlim = c(1981, 2014), 
-                  ylim = c(40, 65), expand = FALSE) +
+  coord_cartesian(ylim = c(41, 65), expand = TRUE) +
   ylab("Richness") +
   scale_fill_manual(values = c("black", observer_colors), guide = FALSE) +
-  theme(
-    panel.border = element_rect(color = alpha(1, .5), size = 1, linetype = 1),
-    panel.grid.major = theme_light()$panel.grid.major,
-    panel.grid.minor = theme_light()$panel.grid.minor)
+  theme(panel.grid.major = theme_light()$panel.grid.major,
+        panel.grid.minor = theme_light()$panel.grid.minor)
 

### 2017-07-20 R/sdm-rf.R
stupid directory structure

@@ -28,7 +28,7 @@ one_rf_tree = function(bbs, vars, sp_id, iter, use_obs_model, x_richness,
 }
 
 rf_predict_species = function(sp_id, bbs, settings, x_richness, use_obs_model,
-                              future, observer_sigmas){
+                              future, observer_sigmas, path){
   vars = c(settings$vars, if (use_obs_model) {"observer_effect"})
   iters = sort(unique(x_richness$iteration))
   results = lapply(iters, one_rf_tree,
@@ -41,8 +41,9 @@ rf_predict_species = function(sp_id, bbs, settings, x_richness, use_obs_model,
                    settings = settings) %>% 
     bind_rows()
   
-  path = paste0("rf_predictions/sp_", sp_id, "_", use_obs_model, ".csv.gz")
-  dir.create("rf_predictions", showWarnings = FALSE)
+  path = paste0(path, 
+                "rf_predictions/sp_", sp_id, "_", use_obs_model, 
+                ".csv.gz")
   write.csv(results, file = gzfile(path), row.names = FALSE)
   
   results %>% 
@@ -51,7 +52,7 @@ rf_predict_species = function(sp_id, bbs, settings, x_richness, use_obs_model,
 }
 
 rf_predict_richness = function(bbs, x_richness, settings, use_obs_model,
-                               future, observer_sigmas) {
+                               future, observer_sigmas, path) {
   
   out = parallel::mclapply(
     unique(bbs$species_id), 
@@ -60,7 +61,7 @@ rf_predict_richness = function(bbs, x_richness, settings, use_obs_model,
                          settings = settings, 
                          use_obs_model = use_obs_model,
                          future = future, 
-                         observer_sigmas = observer_sigmas)
+                         observer_sigmas = observer_sigmas, path)
     },
     mc.cores = 8,
     mc.preschedule = FALSE

### 2017-07-20 analysis/fit-predict.R

@@ -2,7 +2,7 @@ library(tidyverse)
 library(purrrlyr) # for by_slice
 library(parallel)
 devtools::load_all()
-timeframe = "train_32"
+timeframe = "train_22"
 
 
 # Ingest the `settings` and put the timeframe of interest at the top level
@@ -170,15 +170,17 @@ bbs = bbs %>%
 dir.create("rf_predictions", showWarnings = FALSE)
 
 gc() # Minimize memory detritus before forking
-
+dir.create(prepend_timeframe("rf_predictions"))
 rf_predict_richness(bbs = bbs, x_richness = x_richness, 
                     settings = settings, use_obs_model = TRUE,
                     future = future, 
-                    observer_sigmas = observer_sigmas) %>% 
+                    observer_sigmas = observer_sigmas,
+                    path = prepend_timeframe("")) %>% 
   saveRDS(file = prepend_timeframe("rf_predictions/all_TRUE.rds"))
 
 rf_predict_richness(bbs = bbs, x_richness = x_richness, 
                     settings = settings, use_obs_model = FALSE,
                     future = future, 
-                    observer_sigmas = observer_sigmas) %>% 
+                    observer_sigmas = observer_sigmas,
+                    path = prepend_timeframe("")) %>% 
   saveRDS(file = prepend_timeframe("rf_predictions/all_FALSE.rds"))

### 2017-07-20 R/sdm-rf.R
fix RF path & duplicate columns

@@ -19,8 +19,8 @@ one_rf_tree = function(bbs, vars, sp_id, iter, use_obs_model, x_richness,
     ntree = 1
   )  
   
-  test = cbind(make_test_set(d, future, observer_sigmas, settings),
-               species_id = sp_id, iteration = unique(d$iteration))
+  test = make_test_set(d, future, observer_sigmas, settings) %>% 
+    mutate(species_id = sp_id)
   test$mean = predict(rf, test, type = "prob")[,2]
   test$use_obs_model = use_obs_model
   
@@ -28,7 +28,7 @@ one_rf_tree = function(bbs, vars, sp_id, iter, use_obs_model, x_richness,
 }
 
 rf_predict_species = function(sp_id, bbs, settings, x_richness, use_obs_model,
-                              future, observer_sigmas, path){
+                              future, observer_sigmas, rf_dir){
   vars = c(settings$vars, if (use_obs_model) {"observer_effect"})
   iters = sort(unique(x_richness$iteration))
   results = lapply(iters, one_rf_tree,
@@ -41,10 +41,8 @@ rf_predict_species = function(sp_id, bbs, settings, x_richness, use_obs_model,
                    settings = settings) %>% 
     bind_rows()
   
-  path = paste0(path, 
-                "rf_predictions/sp_", sp_id, "_", use_obs_model, 
-                ".csv.gz")
-  write.csv(results, file = gzfile(path), row.names = FALSE)
+  filename = paste0(rf_dir, "sp_", sp_id, "_", use_obs_model, ".csv.gz")
+  write.csv(results, file = gzfile(filename), row.names = FALSE)
   
   results %>% 
     group_by(site_id, year, species_id, richness, use_obs_model) %>% 
@@ -52,7 +50,7 @@ rf_predict_species = function(sp_id, bbs, settings, x_richness, use_obs_model,
 }
 
 rf_predict_richness = function(bbs, x_richness, settings, use_obs_model,
-                               future, observer_sigmas, path) {
+                               future, observer_sigmas, rf_dir) {
   
   out = parallel::mclapply(
     unique(bbs$species_id), 

### 2017-07-20 analysis/fit-predict.R

@@ -170,12 +170,13 @@ bbs = bbs %>%
 dir.create("rf_predictions", showWarnings = FALSE)
 
 gc() # Minimize memory detritus before forking
-dir.create(prepend_timeframe("rf_predictions"))
+rf_dir = prepend_timeframe("rf_predictions")
+dir.create(rf_dir)
 rf_predict_richness(bbs = bbs, x_richness = x_richness, 
                     settings = settings, use_obs_model = TRUE,
                     future = future, 
                     observer_sigmas = observer_sigmas,
-                    path = prepend_timeframe("")) %>% 
+                    rf_dir = rf_dir) %>% 
   saveRDS(file = prepend_timeframe("rf_predictions/all_TRUE.rds"))
 
 rf_predict_richness(bbs = bbs, x_richness = x_richness, 

### 2017-07-20 figures/plots.R
more figures

@@ -44,18 +44,18 @@ bound = bind_rows(forecast_results, gbm_results, average_results,
 
 # scatterplot -------------------------------------------------------------
 
-R2s = filter(bound, use_obs_model, year %in% c(2004, 2009, 2013)) %>% 
+R2s = filter(bound, use_obs_model, year %in% c(2004, 2013)) %>% 
   mutate(x = min(mean), y = max(richness)) %>% 
   group_by(model, x, y, year) %>% 
   summarize(R2 = paste("R^2: ", 
                        format(1 - var(mean - richness) / var(richness),
                               digits = 2, nsmall = 2)))
             
-ggplot(data = filter(bound, use_obs_model, year %in% c(2004, 2009, 2013)), 
+ggplot(data = filter(bound, use_obs_model, year %in% c(2004, 2013)), 
        aes(x = mean, y = richness)) +
   geom_hex() + 
   geom_abline(intercept = 0, slope = 1, alpha = .5) +
-  facet_grid(model ~ year) +
+  facet_grid(year ~ model) +
   coord_equal() + 
   scale_fill_continuous(low = "gray90", high = "navy") +
   geom_text(inherit.aes = FALSE,
@@ -184,10 +184,12 @@ observer_uncertainties %>%
 #sample_site_id = 82006
 #sample_site_id = 88005
 #sample_site_id = 2022
-sample_site_id = 2024
-sample_site_id
+#sample_site_id = 91012
+#sample_site_id = 2024
+sample_site_id = 82003
+sample_site_id = sample(unique(bound$site_id), 1)
 
-models = c("average", "naive", "mistnet")
+models = c("average", "naive", "auto.arima")
 grid = crossing(model = models,
                 use_obs_model = c(TRUE, FALSE),
                 year = unique(obs_model$data$year))
@@ -218,9 +220,27 @@ ggplot(time_series_data, aes(x = year)) +
   geom_point(aes(y = richness, fill = factor(observer_id)),
              size = 3.5, shape = 21) +
   scale_alpha(range = c(0, 1), guide = FALSE) + 
-  coord_cartesian(ylim = c(41, 65), expand = TRUE) +
+  coord_cartesian(ylim = c(35, 70), expand = TRUE) +
   ylab("Richness") +
   scale_fill_manual(values = c("black", observer_colors), guide = FALSE) +
   theme(panel.grid.major = theme_light()$panel.grid.major,
[…]

### 2017-07-20 manuscript/draft.Rmd

@@
 of each BBS route.
+
+We made long term forecasts from 2014-2050 using CMIP5 multi-model ensemble
+dataset as the source for climatic variables. Precipitation and temperature from
+37 downscaled model runs [@brekke2013, see Table S1] using the RCP6.0 scenario
+were averaged together to create a single ensemble used to calculate the
+bioclimatic variables for North America. For NDVI we used the per-site average
+values from 2000-2013 as a simple forecast. For observer effects (see below) each
+site was set to have zero observer bias.
 

### 2017-07-20 manuscript/draft.Rmd

@@
 `r settings = yaml::yaml.load_file("../settings.yaml")`
+
+## Abstract
 
@@
 time-series, we did not include a season component in these models. Otherwise we
-used the default settings for this function. [[Note: should probably record
-somewhere what order model the auto.arima function tends to choose.  Eyeballing
-the results, seems to be most similar to the naive model, but would be good to
-check.]]
+used the default settings for this function.
 
@@
 not to outperform the "average" baseline from the generalized linear mixed model
-(Figure 3A). [[Stats go here]].  The auto-ARIMA was generally the best-performing
-non-baseline model, but in many cases, the best-fitting ARIMA model chosen by
-the model selection procedure was no different from the "average" model.
+(Figure 3A). [[Stats go here]].  The auto-ARIMA was generally the
+best-performing non-baseline model, but in many cases (67% of the time), the
+best-fitting ARIMA model chosen by the model selection procedure had no
+autoregressive, drift, or moving average terms, making it similar to the
+"Average" model.
 

### 2017-07-20 figures/plots.R
plotz

@@ -44,27 +44,45 @@ bound = bind_rows(forecast_results, gbm_results, average_results,
 
 # scatterplot -------------------------------------------------------------
 
+ts_models = c("average", "naive", "auto.arima")
+env_models = c("richness_gbm", "naive", "mistnet")
+models = c(ts_models, env_models)
+warning("no rf_sdm")
+
 R2s = filter(bound, use_obs_model, year %in% c(2004, 2013)) %>% 
   mutate(x = min(mean), y = max(richness)) %>% 
   group_by(model, x, y, year) %>% 
   summarize(R2 = paste("R^2: ", 
                        format(1 - var(mean - richness) / var(richness),
                               digits = 2, nsmall = 2)))
-            
-ggplot(data = filter(bound, use_obs_model, year %in% c(2004, 2013)), 
-       aes(x = mean, y = richness)) +
-  geom_hex() + 
-  geom_abline(intercept = 0, slope = 1, alpha = .5) +
-  facet_grid(year ~ model) +
-  coord_equal() + 
-  scale_fill_continuous(low = "gray90", high = "navy") +
-  geom_text(inherit.aes = FALSE,
-            data = R2s,
-            aes(x = x, y = y, label = R2),
-            hjust = 0, vjust = 1,
-            size = 3) +
-  theme_light()
 
+make_scatterplots = function(models){
+  x_range = range(bound$mean)
+  ggplot(data = filter(bound, use_obs_model, year %in% c(2004, 2013),
+                       model %in% models), 
+         aes(x = mean, y = richness)) +
+    geom_hex() + 
+    geom_abline(intercept = 0, slope = 1, alpha = .5) +
+    facet_grid(year ~ model) +
+    coord_equal() + 
+    scale_fill_continuous(low = "gray90", high = "navy",
+                          limits = c(1, 30)) +
+    geom_text(inherit.aes = FALSE,
+              data = filter(R2s, model %in% models),
+              aes(x = x, y = y, label = R2),
+              hjust = 0, vjust = 1,
+              size = 3) +
+    theme_light(base_size = 14, ) +
+    xlim(x_range)
+}
+
+plot_grid(
+  make_scatterplots(ts_models),
+  make_scatterplots(env_models),
+  align = "h",
+  nrow = 2,
+  labels = c("A. Time-series models", "B. Environmental models"))
[…]

### 2017-07-27 figures/plots.R
Save manuscript numbers in ms folder

@@ -293,53 +293,56 @@ ggsave("figures/observer_violins.png", width = 4, height = 5)
 
 
 # Numerical odds and ends for manuscript ----------------------------------
-
-N_runs = obs_model$data %>% 
-  distinct(site_id, year) %>% 
-  nrow()
-N_sites = obs_model$data %>% 
-  distinct(site_id) %>% 
-  nrow()
-richness_summary = obs_model$data %>% 
-  distinct(site_id, year, richness) %>% 
-  summarize(mean = round(mean(richness)),
-            min = min(richness),
-            max = max(richness))
-
-
-n_gbm = gbm_results %>% 
-  filter(site_id == 2001, year == 2004) %>% 
-  group_by(use_obs_model) %>% 
-  summarize(means = mean(n.trees)) %>% 
-  pull(means) %>% 
-  signif(2)
-
-rf_coverage_pct = bound %>% 
-  filter(model == "rf_sdm", use_obs_model) %>% 
-  summarize(round(100 * mean(p > .025 & p < .975))) %>% 
-  pull()
-
-all_R2s = R2s %>% 
-  pull(R2) %>% 
-  range() %>% 
-  format(digits = 2) %>% 
-  paste(collapse = "-")
-
-env_R2s = R2s %>% 
-  filter(model %in% env_models) %>% 
-  pull(R2) %>% 
-  range() %>% 
-  format(digits = 2) %>% 
-  paste(collapse = "-")
-
+bbs_data = get_pop_ts_env_data(
+  start_yr = settings$start_yr, 
+  end_yr = settings$timeframes[[timeframe]]$end_yr, 
+  last_train_year = settings$timeframes[[timeframe]]$last_train_year, 
+  min_year_percentage = settings$min_year_percentage)
 
 variance_components = data_frame(site = obs_model$site_sigma^2, 
                                  observer = obs_model$observer_sigma^2,
                                  residual = obs_model$sigma^2)
 variance_frac = colMeans(variance_components / rowSums(variance_components))
 
-variance_percent = as.list(round(variance_frac * 100))
-
+ms_numbers = list(
+  N_species =  bbs_data %>% 
+    distinct(species_id) %>% 
+    nrow(),
+  N_runs = obs_model$data %>% 
+    distinct(site_id, year) %>% 
+    nrow(),
+  N_sites = obs_model$data %>% 
+    distinct(site_id) %>% 
+    nrow(),
[…]

### 2017-07-27 manuscript/draft.Rmd

@@
 ## Abstract
+
+Developing forecasts for biodiversity is important for conservation, management,
+and evaluating how well current models characterize natural systems. While there
+are an increasing number of studies forecasting biodiversity, there is little
+information available on how well these forecasts work because these studies
+fail to follow good practice for making and evaluating forecasts. Most diversity
+forecasts do not conduct meaningful evaluations of the forecasts, fail to
+account for uncertainty, and do not use time-series data that captures the
+actual dynamics being studied. Using hindcasting we made and evaluated XXXX
+[number of years * number of models * number of sites] forecasts for breeding
+bird species richness at XXX sites throughout North America following best
+practices for forecasting. Forecasts were made for up to a decade into the
+future. Our results show that while all of the models we evaluated produce
+reasonable forecasts, none of them outperforms the simple baseline of long-term
+average richness. This means that at present our best forecast for future
+richness is simply to assume that it is unchanging. This calls into question the
+efficacy of current modeling approaches. Most of the best practices implemented
+in this study directly influence the outcome of the forecasts and result in
+different conclusions from those made in previous studies. This emphasizes the
+importance of following good practice in making ecological forecasts and of
+starting to actively make more forecasts in such a way that they can be quickly
+evaluated and improved on.
 

### 2017-07-27 manuscript/draft.Rmd
manuscript stuff

@@
 bibliography: refs.bib
+csl: csl.csl
 geometry: margin=1in
@@
 `r settings = yaml::yaml.load_file("../settings.yaml")`
+`r timeframe = settings$timeframes$train_22`
+`r numbers = yaml::yaml.load_file("numbers.yaml")`
 
@@
 Previous efforts to predict future patterns of species richness, and diversity
-more generally, have focused primarily on building species distributions models
-to quantify the spatial relationships between the occurrence of individual
-species and the environment [@thomas2004; @urban2015]. Forecasts
-for future environmental conditions are then used to predict where each species
-will occur in the future and the probabilities of occurrence are summed across
-species to predict future species richness [e.g., @thuiller2005]. Alternatively,
-models that directly relate spatial patterns of species richness to the
-environment have been developed and generally perform equivalently to species
-distribution modeling based methods [@algar2009; @distler2015].
+more generally, have focused primarily on building species distributions models.
+In general, these models describe individual species' occurrence patterns across
+sites as functions of the site's environment [@thomas2004; @urban2015]. By
+plugging forecasts of future environmental conditions into these functions,
+ecologists can make predictions about where each species will occur in the
+future. These species-level predictions can then be added up ("stacked") to
+generate predicted richness values [e.g. @calabrese2014]. Alternatively, models
+that directly relate spatial patterns of species richness to the environment
+have been developed and generally perform equivalently to species distribution
+modeling based methods [@algar2009; @distler2015]. This approach is sometimes
+referred to as "macroecological" modeling, because it models the larger-scale
+pattern (richness) more directly [@distler2015].
 
-However, despite the emerging interest in forecasting species richness and other
-aspects of biodiversity, little is known about how effectively we can anticipate
-their dynamics. This is due in part to the long time scales over which many
-ecological forecasts are applied and the resulting difficulty in assessing
-whether the predicted changes occurred [@dietze2016]. What we do know comes
-from a small number of hindcasting studies, where models are built using data on
-species occurrence and richness from the past and evaluated on their ability to
-predict contemporary patterns [e.g., @algar2009; @distler2015]. These studies
-are a valuable first step, but lack several components that are important for
+Despite the emerging interest in forecasting species richness and other aspects
+of biodiversity, little is known about how effectively we can anticipate their
+dynamics. This is due in part to the long time scales over which many ecological
+forecasts are applied and the resulting difficulty in assessing whether the
+predicted changes occurred [@dietze2016]. What we do know comes from a small
+number of hindcasting studies, where models are built using data on species
+occurrence and richness from the past and evaluated on their ability to predict
+contemporary patterns [e.g., @algar2009; @distler2015]. These studies are a
+valuable first step, but lack several components that are important for
 developing good forecasting models and understanding how accurately these models
@@
 environmental time-series [@tredennick2016]; 2) accounting for uncertainty in
-observations, processes, model choice, and forecast evaluation
-[@clark2001, @dietze2017]; and 3) conducting meaningful evaluations of the
-forecasts by hindcasting, archiving short-term forecasts, and comparing
-forecasts to baselines to determine if the forecasts are more accurate than
+observations, processes, model choice, and forecast evaluation [@clark2001,
+@dietze2017]; and 3) conducting meaningful evaluations of the forecasts by
+hindcasting, archiving short-term forecasts, and comparing forecasts to
+baselines to determine whether the forecasts are more accurate than simply
 assuming the system is basically static [@perretti2013].
 
-In this paper we attempt to forecast the species richness of breeding birds at
-hundreds of locations throughout North America while following the best
-practices in Box 1 for ecological forecasting. To do this we combine 30 years of
-time-series data on bird distributions with monthly time-series of climate data
-and satellite-based remote-sensing. This 30 year annual time-scale has only
-recently become possible for large-scale time-series based forecasting and
-allows us to model and assess changes a decade or more into the future in the
-presence of shifts in environmental conditions on par with predicted climate
-change. We compare traditional distribution modeling based approaches to spatial
-models of species richness, time-series methods, and simple baselines. All
-forecasts model uncertainty and observation, are evaluated across different time
-lags using hindcasting, and are publicly archived to allow future assessment. We
-discuss the implications of these practices for our understanding of, and
-confidence in, the resulting forecasts, and how we can continue to build on
-these approaches improve ecological forecasting in the future.
+In this paper, we attempt to forecast the species richness of breeding birds at
+over 1,000 of locations throughout North America while following the best
+practices in Box 1 for ecological forecasting. To do this, we combine 30 years
+of time-series data on bird distributions from annual surveys with monthly
+time-series of climate data and satellite-based remote-sensing. This 30 year
+time-scale has only recently become possible for large-scale time-series based
+forecasting and allows us to model and assess changes a decade or more into the
+future in the presence of shifts in environmental conditions on par with
+predicted climate change. We compare traditional distribution modeling based
+approaches to spatial models of species richness, time-series methods, and
+simple baselines. All forecasts model uncertainty and observation, are evaluated
+across different time lags using hindcasting, and are publicly archived to allow
+future assessment. We discuss the implications of these practices for our
+understanding of, and confidence in, the resulting forecasts, and how we can
+continue to build on these approaches improve ecological forecasting in the
+future.
 
@@
 (i.e., routes with at least 16 annual observations). The resulting dataset
-included [[N]] transect runs of [[N]] unique sites, and included [[N]] species.
-Site-level richness varied from [[N]] to [[N]] with an average richness of [[N]]
-species.
+included `r numbers$N_runs` transect runs of `r numbers$N_sites` unique sites,
+and included `r numbers$N_species` species. Site-level richness varied from `r
+numbers$richness_summary$min` to `r numbers$richness_summary$max` with an
+average richness of `r numbers$richness_summary$mean` species.
 
@@
 
-Observer effects are inherent in large data sets taken by different observers,
-and are known to occur in BBS [@sauer1994]. We estimated the observer effects
-(and associated uncertainty about those effects) with a linear mixed model with
-observer as a random effect built in the Stan probabilistic programming language
-[@stan] and its R interface [@rstan]. Because observer and site are strongly
-related (an observer tends to repeatedly sample the same site), site was also
-included as a random effect to ensure that inferred deviations were actually
-observer-related (as opposed to being related to the sites that a given observer
-happened to see). The resulting model partitions the variance in observed
-richness values into site-level variance, observer-level variance, and residual
-variance (e.g. variation within a site from year to year), and can therefore
-also be used directly as the "average" baseline model (see below). The estimated
-observer effects can be subtracted from the richness values for a particular
-observer to provide an estimate of how many species would have been found by a
-"typical" observer. To incorporate uncertainty in these "corrected" richness
-values into the forecasting models we collected 500 Monte Carlo samples from the
-model's posterior distribution, and fit each of the downstream models with each
-of the Monte Carlo samples.
+Observer effects are inherent in large data sets collected by different
+observers, and are known to occur in BBS [@sauer1994]. We estimated the observer
+effects (and associated uncertainty about those effects) with a linear mixed
+model with observer as a random effect built in the Stan probabilistic
+programming language [@stan] and its R interface [@rstan]. Because observer and
+site are strongly related (observers tend to repeatedly sample the same site),
+site was also included as a random effect to ensure that inferred deviations
+were actually observer-related (as opposed to being related to the sites that a
+given observer happened to see). The resulting model partitions the variance in
+observed richness values into site-level variance, observer-level variance, and
+residual variance (e.g. variation within a site from year to year). The
+site-level estimates can also be used directly as the "average" baseline model
+(see below). The estimated observer effects can be subtracted from the richness
+values for a particular observer to provide an estimate of how many species
+would have been found by a "typical" observer. To incorporate uncertainty in
+these "corrected" richness values into the forecasting models we collected 500
+Monte Carlo samples from the model's posterior distribution, and fit each of the
+downstream models with each of the Monte Carlo samples.
 
@@
 
-We fit three sets of models that were fit to each site independently. These
-models were fit to each BBS route both on the raw richness values and on the
-residuals from our observer model. When correcting for observer effects, we
-averaged across 500 models that were fit separately to the 500 Monte Carlo
-estimates of the observer effects, to account for our uncertainty in the true
-values of those effects. All of these models use a Gaussian error distribution.
+We fit three sets of models that were fit to each site independently, with no
+environmental information. These models were fit to each BBS route both on the
+raw richness values and on the residuals from our observer model. When
+correcting for observer effects, we averaged across 500 models that were fit
+separately to the 500 Monte Carlo estimates of the observer effects, to account
+for our uncertainty in the true values of those effects. All of these models use
+a Gaussian error distribution.
 
@@
 predicts that future observations will be similar to the final observed value
-(e.g. the value observed in `$r settings$last_train_year`), and the confidence
+(e.g. the value observed in `r timeframe$last_train_year`), and the confidence
 intervals expand rapidly as the predictions extend farther into the future.
@@
 complex patterns including trends). The `auto.arima` function chooses whether to
-include each of these components, and how many terms to include for each one,
+include each of these components (and how many terms to include for each one)
 using AICc [@forecast1]. Since there is no seasonal component to the BBS
@@
 In contrast to the time-series models, most attempts to predict species richness
-focusing on using correlative models to predict richness changes over time from
+focus on using correlative models to predict richness changes over time from
 environmental data. We tested three common variants of this approach - direct
@@
 [@elith2008]. This model was optimized using a Gaussian likelihood, with a
-maximum interaction depth of 5, shrinkage of 0.015, and up to 10,000 trees.
-The number of trees was selected using the "out of bag" estimator, and
-[[summarize how many trees were typically chosen here, for both observer-based
-and observer-free models]].
+maximum interaction depth of 5, shrinkage of 0.015, and up to 10,000 trees. The
+number of trees used for prediction was selected using the "out of bag"
+estimator; this number averaged [[N]] for the non-observer data and [[N]] for
+the observer-corrected data.
 
@@
 Species-level models are used to predict richness by summing the predicted
-probability of occupancy across all species at a site; following [[Calabrese et
-al.]], we calculated the uncertainty in our richness estimate by treating
-richness as a sum over independent Bernoulli random variables (one per species).
-This avoids known problems with the use of thresholds for determining whether or
-not a species will be present at a site [[refs]]. We represented SDMs in this
-analysis using random forests, a common approach in the species distribution
-modeling literature based on fitting hundreds of independent regression trees to
-randomly-perturbed versions of the data [[refs; would be good to provide a
-source that says these are a good choice]]. Each of the 500 trees in our
-species-level random forests used a different Monte Carlo estimate of the
-observer effects as a predictor variable.
+probability of occupancy across all species at a site. This avoids known
+problems with the use of thresholds for determining whether or not a species
+will be present at a site [@calabrese2014; pellissier2013]. Following
+@calabrese2014, we calculated the uncertainty in our richness estimate by
+treating richness as a sum over independent Bernoulli random variables:
+$\sigma^2_{richness} = \sum_i{p_i (1 - p_i)}$, where $i$ indexes species. By
+itself, this approach is known to underestimate the true community-level
+uncertainty because it ignores the uncertainty in the species-level probabilites
+[@calabrese2014]. To mitigate this problem, we used an ensemble of 500 estimates
+for each of the species-level probabilities instead of just one, propagating the
+uncertainty forward. We obtained these estimates using using random forests, a
+common approach in the species distribution modeling literature based on fitting
+hundreds of independent regression trees to randomly-perturbed versions of the
+data [@cutler2007; @caruana2008]. When correcting for observer effects, each of
+the 500 trees in our species-level random forests used a different Monte Carlo
+estimate of the observer effects as a predictor variable.
 
@@
 common count models, such as the Poisson or binomial (i.e. richness was
-underdispersed), so these count models would have had difficulty fitting the
-data. Since richness had a relatively large mean ([[N]]) and all models already
-produce continuous richness estimates (due to summing probabilities in the
-(J)SDM case), we defined model performance for all models in terms of continuous
-Gaussian errors instead of using discrete count distributions. When a model was
-run multiple times for the purpose of correcting for observer effects, we used
-the mean of those runs' point estimates as our final point estimate and we
-calculated the uncertainty using the law of total variance (i.e.  the average of
-the model runs' variance, plus the variance in the point estimates).
+underdispersed for individual sites), so these count models would have had
+difficulty fitting the data. Since richness had a relatively large mean ([[N]])
+and all models already produce continuous richness estimates (due to summing
+probabilities in the (J)SDM case), we defined model performance for all models
+in terms of continuous Gaussian errors instead of using discrete count
+distributions. When a model was run multiple times for the purpose of correcting
+for observer effects, we used the mean of those runs' point estimates as our
+final point estimate and we calculated the uncertainty using the law of total
+variance (i.e.  the average of the model runs' variance, plus the variance in
+the point estimates).
 
 We evaluated the performance of each model for forecasting using the data for
-each year between `r settings$last_train_year` and `r settings$end_yr`. We used
+each year between `r timeframe$last_train_year` and `r timeframe$end_yr`. We used
 three metrics for evaluating performance: 1) root-mean-square error (RMSE) to
@@
 All analyses were conducted using R [@R]. Primary R packages used in the
-analysis included dplyr [@dplyr], tidyr [@tidyr], gimms [@gimms], sp
-[@sp1; @sp2], raster [@raster], prism [@prism], rdataretriever
-[@rdataretriever], forecasts [@forecast1; @forecast2], git2r [@git2r], ggplot
-[@ggplot2], viridius [@viridis], rstan [@rstan], yaml [@yaml], purrr [@purrr],
-gbm [@gbm], randomForest [@randomForest], purrrlyr [@purrrlyr]. Code to fully
-reproduce this analysis is available on GitHub
-(https://github.com/weecology/bbs-forecasting) and archived on Zenodo ().
+analysis included dplyr [@dplyr], tidyr [@tidyr], gimms [@gimms], sp [@sp1;
+@sp2], raster [@raster], prism [@prism], rdataretriever [@rdataretriever],
+forecast [@forecast1; @forecast2], git2r [@git2r], ggplot [@ggplot2], mistnet
+[@harris2015], viridis [@viridis], rstan [@rstan], yaml [@yaml], purrr [@purrr],
+gbm [@gbm], randomForest [@randomForest]. Code to fully reproduce this analysis
+is available on GitHub (https://github.com/weecology/bbs-forecasting) and
+archived on Zenodo ().
 

### 2017-07-27 manuscript/draft.Rmd
edits

@@
 account for uncertainty, and do not use time-series data that captures the
-actual dynamics being studied. Using hindcasting we made and evaluated XXXX
-[number of years * number of models * number of sites] forecasts for breeding
-bird species richness at XXX sites throughout North America following best
-practices for forecasting. Forecasts were made for up to a decade into the
-future. Our results show that while all of the models we evaluated produce
-reasonable forecasts, none of them outperforms the simple baseline of long-term
-average richness. This means that at present our best forecast for future
-richness is simply to assume that it is unchanging. This calls into question the
-efficacy of current modeling approaches. Most of the best practices implemented
-in this study directly influence the outcome of the forecasts and result in
-different conclusions from those made in previous studies. This emphasizes the
-importance of following good practice in making ecological forecasts and of
-starting to actively make more forecasts in such a way that they can be quickly
-evaluated and improved on.
+actual dynamics being studied. Using hindcasting we made and evaluated `r
+numbers$n_forecasts` forecasts for breeding bird species richness at `r
+numbers$N_predict_sites` sites throughout the continental United States,
+following best practices for forecasting. Forecasts were made for up to a decade
+after the final training observations. Our results show that while all of the
+models we evaluated produce reasonable forecasts, none of them consistently
+outperforms the simple baseline of long-term average richness. This means that
+at present, our best forecast for future richness is simply to assume that it is
+unchanging. This calls into question the efficacy of current modeling
+approaches. Most of the best practices implemented in this study directly
+influence the outcome of the forecasts and result in different conclusions from
+those made in previous studies. This emphasizes the importance of following good
+practice in making ecological forecasts and of starting to actively make more
+forecasts in such a way that they can be quickly evaluated and improved on.
 
@@
 When comparing forecasts for richness across sites all methods performed well
-(Figure 2; R^2 XX-XX%). However SDMs (both stacked and joint) and the
+(Figure 2; all $R^2 > 0.5$). However SDMs (both stacked and joint) and the
 macroecological model all failed to successfully forecast the highest-richness
 sites resulting in a notable clustering of predicted values near ~60 species and
-the poorest performance in the cross-site comparison (R^2 XX-XX%).
+the poorest performance ($R^2$ `r numbers$env_R2s`, versus XX-XX for the
+within-site methods).
 
-While the models generally did well in absolute terms (Figure 2), they tended
-not to outperform the "average" baseline from the generalized linear mixed model
+While the models generally did well in absolute terms (Figure 2), none tended to
+outperform the "average" baseline from the generalized linear mixed model
 (Figure 3A). [[Stats go here]].  The auto-ARIMA was generally the
@@
 overconfident predictions (Figure 3B). The random forest-based SDM stack was the
-most overconfident model, with only XX% of observations falling inside its 95%
-confidence intervals. This stacked SDM's narrow predictive distribution also
-caused it to have notably higher deviance (Figure 3C) than the next-worst model,
-even though its point estimates were not unusually bad in terms of RMSE (3A). As
-discussed elsewhere [@harris2015], this overconfidence is a product of the
-assumption in stacked SDMs that errors in the species-level predictions are
-independent. Because they avoided this assumption, the GBM-based
-"macroecological" model and especially the mistnet JSDM had reasonably
-well-calibrated uncertainty estimates (Figure 3B); as a result, their relative
-performance was higher in terms of deviance than in terms of RMSE. The "naive"
-model was the only model whose confidence intervals were too wide (Figure 3B),
-which can be attributed to the rapid rate at which these intervals expand as the
-time lag increases (Figure 1).
+most overconfident model, with only `r numbers$rf_coverage_pct`% of observations
+falling inside its 95% confidence intervals. This stacked SDM's narrow
+predictive distribution also caused it to have notably higher deviance (Figure
+3C) than the next-worst model, even though its point estimates were not
+unusually bad in terms of RMSE (3A). As discussed elsewhere [@harris2015], this
+overconfidence is a product of the assumption in stacked SDMs that errors in the
+species-level predictions are independent. Because they avoided this assumption,
+the GBM-based "macroecological" model and especially the mistnet JSDM had
+reasonably well-calibrated uncertainty estimates (Figure 3B); as a result, their
+relative performance was higher in terms of deviance than in terms of RMSE. The
+"naive" model was the only model whose confidence intervals were too wide
+(Figure 3B), which can be attributed to the rapid rate at which these intervals
+expand as the time lag increases (Figure 1).
 

### 2017-07-27 figures/plots.R
more numbers

@@ -314,6 +314,15 @@ ms_numbers = list(
   N_sites = obs_model$data %>% 
     distinct(site_id) %>% 
     nrow(),
+  N_predict_sites = bound %>% 
+    distinct(site_id) %>% 
+    nrow(),
+  N_predictions = bound %>% 
+    distinct(site_id, year) %>% 
+    nrow(),
+  N_obs = obs_model$data %>% 
+    distinct(observer_id) %>% 
+    nrow(),
   richness_summary = obs_model$data %>% 
     distinct(site_id, year, richness) %>% 
     summarize(mean = round(mean(richness)),
@@ -340,7 +349,13 @@ ms_numbers = list(
     range() %>% 
     format(digits = 2) %>% 
     paste(collapse = "-"),
-  variance_percent = as.list(round(variance_frac * 100))
+  variance_percent = as.list(round(variance_frac * 100)),
+  pct_is_average = forecast_results %>% 
+    group_by(use_obs_model) %>% 
+    filter(model == "auto.arima") %>% 
+    summarize(is_avg = mean(100 * map_lgl(coef_names, ~all(.x == "intercept")))) %>% 
+    pull(is_avg) %>% 
+    round()
 )  
 cat(yaml::as.yaml(ms_numbers), file = "manuscript/numbers.yaml")
 

### 2017-07-28 figures/plots.R
improve predictions plots

@@ -187,64 +187,86 @@ ggsave("figures/observer_uncertainty.png", width = 3.75, height = 4.5)
 
 # Time series -------------------------------------------------------------
 
-#sample_site_id = sample(unique(bound$site_id), 1)
-#sample_site_id = 82006
-#sample_site_id = 88005
-#sample_site_id = 2022
-#sample_site_id = 91012
-#sample_site_id = 2024
-sample_site_id = 82003
 
-grid = crossing(model = c(ts_models, env_models),
-                use_obs_model = c(TRUE, FALSE),
-                year = unique(obs_model$data$year))
+# sample_site_id = 88005
+# sample_site_id = 91012
+# sample_site_id = 2024
+# sample_site_id = 91062
+sample_site_id = 38032
+sample_site_id = 72035
 
-time_series_data = obs_model$data %>% 
-  filter(site_id == sample_site_id, iteration == 1) %>% 
-  select(site_id, year, richness, observer_id, richness) %>% 
-  left_join(grid, by = "year") %>% 
-  left_join(select(bound, -richness), by = c("year", "model", "use_obs_model",
-                                             "site_id")) %>% 
-  mutate(model = forcats::fct_relevel(model, unique(models)),
-         observer_id = ifelse(use_obs_model, observer_id, 0)) %>% 
-  select(site_id, year, model, use_obs_model, mean, sd, richness, observer_id)
-
-
-observer_colors = RColorBrewer::brewer.pal(4, "PuOr")
-
-# The warning about missing values is just saying that there are no predictions
-# before 2004.
-make_ts_plots = function(models){
+make_ts_plots = function(models, ylim, use_obs_model, sample_site_id, 
+                         main = ""){
+  grid = crossing(model = c(ts_models, env_models),
+                  use_obs_model = unique(c(use_obs_model, FALSE)),
+                  year = unique(obs_model$data$year))
+  
+  time_series_data = obs_model$data %>% 
+    filter(site_id == sample_site_id, iteration == 1) %>% 
+    select(site_id, year, richness, observer_id, richness) %>% 
+    left_join(grid, by = "year") %>% 
+    left_join(select(bound, -richness), by = c("year", "model", "use_obs_model",
+                                               "site_id")) %>% 
+    mutate(model = forcats::fct_relevel(model, unique(models)),
[…]

### 2017-07-28 figures/alternative_timeplot.R
remove alternative timeplot detritus

@@ -1,53 +1,0 @@
-i = i + 1
-sample_site_id = rev(unique(bound$site_id))[i]
-#sample_site_id = 14198
-#sample_site_id = 92017
-#sample_site_id = 91062
-#sample_site_id = 91034
-sample_site_id = 91022
-
-grid = crossing(model = c(ts_models, env_models),
-                use_obs_model = c(TRUE, FALSE),
-                year = unique(obs_model$data$year))
-
-time_series_data = obs_model$data %>% 
-  filter(site_id == sample_site_id, iteration == 1) %>% 
-  select(site_id, year, richness, observer_id, richness) %>% 
-  left_join(grid, by = "year") %>% 
-  left_join(select(bound, -richness), by = c("year", "model", "use_obs_model",
-                                             "site_id")) %>% 
-  mutate(model = forcats::fct_relevel(model, unique(models)),
-         observer_id = ifelse(use_obs_model, observer_id, 0)) %>% 
-  select(site_id, year, model, use_obs_model, mean, sd, richness, observer_id)
-
-
-observer_colors = RColorBrewer::brewer.pal(4, "PuOr")
-
-# time_series_data$observer_id = factor(time_series_data$observer_id)
-# levels(time_series_data$observer_id) = c("0", "A", "B", "C")
-ggplot(time_series_data, aes(x = year)) +
-  geom_ribbon(aes(ymin = mean - 1.96 * sd, ymax = mean + 1.96 * sd,
-                  group = use_obs_model,
-                  alpha = .2 * as.numeric(!use_obs_model)),
-              fill = "black") +
-  geom_ribbon(aes(ymin = mean - sd, ymax = mean + sd,
-                  group = use_obs_model, 
-                  alpha = .2 * as.numeric(!use_obs_model)),
-              fill = "black") +
-  geom_line(aes(y = mean, group = use_obs_model, color = use_obs_model,
-                alpha = .5 * as.numeric(!use_obs_model))) +
-  facet_wrap(~ model) +
-  geom_line(aes(y = richness, alpha = .5 * (year < min(bound$year - 1)))) + 
-  geom_point(aes(y = richness), size = 1, shape = 21, fill = "red") +
-  scale_alpha(range = c(0, 1), guide = FALSE) + 
-  coord_cartesian(xlim = c(1980, 2014), ylim = c(44, 83), expand = FALSE) +
-  ylab("Richness") +
-  scale_fill_brewer(guide = FALSE) +
[…]

### 2017-07-28 manuscript/draft.Rmd
add figs to manuscript

--- 2017-07-27 manuscript/draft.Rmd
@@
 ## Abstract
-
-Developing forecasts for biodiversity is important for conservation, management,
-and evaluating how well current models characterize natural systems. While there
-are an increasing number of studies forecasting biodiversity, there is little
-information available on how well these forecasts work because these studies
-fail to follow good practice for making and evaluating forecasts. Most diversity
-forecasts do not conduct meaningful evaluations of the forecasts, fail to
-account for uncertainty, and do not use time-series data that captures the
-actual dynamics being studied. Using hindcasting we made and evaluated XXXX
-[number of years * number of models * number of sites] forecasts for breeding
-bird species richness at XXX sites throughout North America following best
-practices for forecasting. Forecasts were made for up to a decade into the
-future. Our results show that while all of the models we evaluated produce
-reasonable forecasts, none of them outperforms the simple baseline of long-term
-average richness. This means that at present our best forecast for future
-richness is simply to assume that it is unchanging. This calls into question the
-efficacy of current modeling approaches. Most of the best practices implemented
-in this study directly influence the outcome of the forecasts and result in
-different conclusions from those made in previous studies. This emphasizes the
-importance of following good practice in making ecological forecasts and of
-starting to actively make more forecasts in such a way that they can be quickly
-evaluated and improved on.
 
@@
 ### Site-level models
+
+![](../figures/model_predictions.png)
 
@@
 
+![](../figures/scatter.png)
+
 While the models generally did well in absolute terms (Figure 2), they tended
@@
 "Average" model.
+
+![](../figures/model_violins.png)
+
+![](../figures/performance_time.png)
 
@@
 attribute to its ability to model this variation with its latent variables.
+
+![](../figures/observers.png)
 

### 2017-07-28 figures/plots.R

@@ -262,12 +262,6 @@ list(ts_models, env_models) %>%
   
 ggsave(filename = "figures/model_predictions.png", width = 7.5, height = 4)
 
-
-make_ts_plots(c("average", "naive", "rf_sdm"), ylim = c(41, 91), 
-              use_obs_model = TRUE, sample_site_id = 72035,
-              main = "Controlling for observer differences")
-ggsave(filename = "figures/observer_predictions.png", width = 4, height = 4.5)
-
 # Numbers for the manuscript ----------------------------------------------
 
 
@@ -297,21 +291,26 @@ observer_error_data = bound %>%
   mutate(y = abs(no) - abs(yes))
 
 plot_grid(
-  make_violins(observer_error_data, 
-                main = "Absolute error reduction from observer model",
-                ylab = "Reduction in absolute\nrichness error (species)",
-                ylim = quantile(observer_error_data$y, c(.0005, .9995)), 
-                yintercept = 0,
-                adjust = 2),
-  make_violins(observer_deviance_data, 
-                main = "Posterior weight of model including observer effect",
-                ylab = "Posterior weight of\nobserver model",
-                ylim = c(0, 1), 
-                yintercept = 0.5, 
-                adjust = 5),
-  nrow = 2
+  make_ts_plots(c("average", "naive", "rf_sdm"), ylim = c(41, 91), 
+                use_obs_model = TRUE, sample_site_id = 72035,
+                main = "Controlling for observer differences"),
+  plot_grid(
+    make_violins(observer_error_data, 
+                 main = "Absolute error reduction from observer model",
+                 ylab = "Reduction in absolute\nrichness error (species)",
+                 ylim = quantile(observer_error_data$y, c(.0005, .9995)), 
+                 yintercept = 0,
+                 adjust = 2),
+    make_violins(observer_deviance_data, 
+                 main = "Posterior weight of model including observer effect",
+                 ylab = "Posterior weight of\nobserver model",
+                 ylim = c(0, 1), 
+                 yintercept = 0.5, 
+                 adjust = 5),
+    nrow = 2
+  )
 )
-ggsave("figures/observer_violins.png", width = 4, height = 5)
+ggsave("figures/observers.png", width = 8, height = 5)
 
 
 # Numerical odds and ends for manuscript ----------------------------------

### 2017-07-28 manuscript/draft.Rmd
Remove reference to "sites" at first mention of SDMs
Since lots of these are built using grid cells it's not quite right and I think it actually reads fine just by removing the specifics.
Fix `(Box 1)` related rendering
typo

--- 2017-07-27 manuscript/draft.Rmd
@@
 more generally, have focused primarily on building species distributions models.
-In general, these models describe individual species' occurrence patterns across
-sites as functions of the site's environment [@thomas2004; @urban2015]. By
-plugging forecasts of future environmental conditions into these functions,
-ecologists can make predictions about where each species will occur in the
-future. These species-level predictions can then be added up ("stacked") to
-generate predicted richness values [e.g. @calabrese2014]. Alternatively, models
-that directly relate spatial patterns of species richness to the environment
-have been developed and generally perform equivalently to species distribution
-modeling based methods [@algar2009; @distler2015]. This approach is sometimes
-referred to as "macroecological" modeling, because it models the larger-scale
-pattern (richness) more directly [@distler2015].
+In general, these models describe individual species' occurrence patterns as
+functions of the environment [@thomas2004; @urban2015]. By plugging forecasts of
+future environmental conditions into these functions, ecologists can make
+predictions about where each species will occur in the future. These
+species-level predictions can then be added up ("stacked") to generate predicted
+richness values [e.g. @calabrese2014]. Alternatively, models that directly
+relate spatial patterns of species richness to the environment have been
+developed and generally perform equivalently to species distribution modeling
+based methods [@algar2009; @distler2015]. This approach is sometimes referred to
+as "macroecological" modeling, because it models the larger-scale pattern
+(richness) more directly [@distler2015].
 
@@
 developing good forecasting models and understanding how accurately these models
-can predict the future. These components of good forecasting and evaluation (Box
-1) broadly involve: 1) expanding the use of data to include biological and
+can predict the future. These components of good forecasting and evaluation
+(Box 1) broadly involve: 1) expanding the use of data to include biological and
 environmental time-series [@tredennick2016]; 2) accounting for uncertainty in
@@
 
+![](../figures/scatter.png)
+
 While the models generally did well in absolute terms (Figure 2), none tended to
@@
 
+![](../figures/observers.png)
+
 ## Discussion
@@
 among sites may not indicate much about a model's ability to predict changes
-over time. This issue has also recently been recently highlighted as an issue
+over time. This issue has also recently been highlighted as an issue
 for using SDMs to predict the future locations of individual species

### 2017-07-28 manuscript/draft.Rmd
add example to sum up this paragraph
also remove note to check Petchey et al. They do not make this
statement about different models, their discussion of the this
is mostly about modeling at different grains

@@
 this approach for considering the potential importance of different models and
-processes for making forecasts at different scales. [[Should make sure Petchey et
-al. don't make this point]] Our results show that for forecasts 1-3 years in the
-future, baselines and time-series based approaches outperform all other methods
-using integrative metrics. At these time scales, environmental changes are
-relatively small and changes in the biota may lag behind changes in the
-environment. However, as the timescale of the forecast increases to a decade the
-JSDMs and environmental richness models get closer to the accuracy of the
-time-series methods and even surpass some of those methods. As the forecast
-distance increases the amount of environmental change is expected to increase
-and the system will have more time to respond, potentially leading to shifts
-towards improved relative performance of the models incorporating environmental
-and ecological information. If this shift with forecast distance in the relative
-importance of different processes and models continues, this could lead to JSDMs
-and environmental richness models outperforming time-series based approaches at
-sufficiently long time scales. These results suggest that assessment of how
-forecast performance changes with time lag is not only important for determining
-how far in the future to forecast, but also for determining models and processes
-are most relevant for making forecasts at different time scales.
+processes for making forecasts at different scales. Our results show that for
+forecasts 1-3 years in the future, baselines and time-series based approaches
+outperform all other methods using integrative metrics. At these time scales,
+environmental changes are relatively small and changes in the biota may lag
+behind changes in the environment. However, as the timescale of the forecast
+increases to a decade the JSDMs and environmental richness models get closer
+to the accuracy of the time-series methods and even surpass some of those
+methods. As the forecast distance increases the amount of environmental change
+is expected to increase and the system will have more time to respond,
+potentially leading to shifts towards improved relative performance of the
+models incorporating environmental and ecological information. If this shift
+with forecast distance in the relative importance of different processes and
+models continues, this could lead to JSDMs and environmental richness models
+outperforming time-series based approaches at sufficiently long time scales.
+These results suggest that assessment of how forecast performance changes with
+time lag is not only important for determining how far in the future to
+forecast, but also for determining which models and processes are most relevant
+for making forecasts at different time scales. For example ensemble forecasts
+could be created which incorporate these time lags into the weights for
+individual models.
 

### 2017-07-29 manuscript/draft.Rmd
add hindcasting notes

@@
 
-To evaluate the expected accuracy and uncertainty of forecasts assess the
-performance of these forecasts within existing time-series data. [[citaton?]]
+Evalutating a models predictive performance across time is critical. Hindcasting
+uses a temporal out of sample validation to mimic how well a model would have
+performed had it been run in the past. For example using occurance data from the
+early 20th century to model distributions which are validated with late
+20th century occurances. Dense timeseries, such as yearly observations, are
+desirable to also evalulate the forecast horizon (see #6), but this is not
+a strict requirement.
 

### 2017-07-31 manuscript/draft.Rmd
some manuscript changes

@@
 
-Developing forecasts for biodiversity is important for conservation, management,
-and evaluating how well current models characterize natural systems. While there
-are an increasing number of studies forecasting biodiversity, there is little
-information available on how well these forecasts work because these studies
-fail to follow good practice for making and evaluating forecasts. Most diversity
-forecasts do not conduct meaningful evaluations of the forecasts, fail to
-account for uncertainty, and do not use time-series data that captures the
-actual dynamics being studied. Using hindcasting we made and evaluated `r
-numbers$n_forecasts` forecasts for breeding bird species richness at `r
-numbers$N_predict_sites` sites throughout the continental United States,
-following best practices for forecasting. Forecasts were made for up to a decade
-after the final training observations. Our results show that while all of the
-models we evaluated produce reasonable forecasts, none of them consistently
-outperforms the simple baseline of long-term average richness. This means that
-at present, our best forecast for future richness is simply to assume that it is
-unchanging. This calls into question the efficacy of current modeling
-approaches. Most of the best practices implemented in this study directly
-influence the outcome of the forecasts and result in different conclusions from
-those made in previous studies. This emphasizes the importance of following good
-practice in making ecological forecasts and of starting to actively make more
-forecasts in such a way that they can be quickly evaluated and improved on.
+Biodiversity forecasts are important for conservation, management, and
+evaluating how well current models characterize natural systems. While there are
+an increasing number of studies forecasting biodiversity, there is little
+information available on how well these forecasts work. Most diversity forecasts
+do not conduct meaningful evaluations of the forecasts, fail to account for
+uncertainty, and do not use time-series data that captures the actual dynamics
+being studied. Using hindcasting we made and evaluated `r numbers$n_forecasts`
+forecasts for breeding bird species richness at `r numbers$N_predict_sites`
+sites throughout the continental United States, following ten suggested best
+practices for forecasting. Forecasts were made for up to a decade after the
+final training observations. While all of the models we evaluated produce
+reasonable forecasts, none of them consistently outperforms the simple baseline
+of constant long-term average richness. At present, our best forecast for future
+richness is thus to assume that it is unchanging, and current modeling
+approaches do not reliably improve on this simple baseline. Most of the best
+practices implemented in this study directly influence the outcome of the
+forecasts and result in different conclusions from those made in previous
+studies. This emphasizes the importance of following good practice in making
+ecological forecasts and of starting to actively make more forecasts in such a
+way that they can be quickly evaluated and improved on.
 
@@
 Forecasting the future state of ecological systems is increasingly important for
-planning, management, and evaluating how well ecological models capture the key
+planning, managing, and evaluating how well ecological models capture the key
 ecological processes governing a system [@clark2001; @houlahan2017;
-@dietze2017]. One area of particular interest for forecasting is biodiversity
-since it is important for ecosystem function, central to conservation planning,
-and expected to be influenced by anthropogenic global change [@cardinale2012;
-@diaz2015; @tilman2017]. Anticipating potential changes in biodiversity is
-crucial for large scale management and conservation, and for addressing debates
-regarding whether local scale diversity is declining and may continue to decline
-in response to anthropogenic influences [@diaz2015; @dornelas2014].
+@dietze2017]. Forecasting is biodiversity is especially important, due to its
+central role in conservation planning and its sensitivity to anthropogenic
+effects [@cardinale2012; @diaz2015; @tilman2017]. Anticipating potential changes
+in biodiversity is crucial for large scale management and conservation, and for
+addressing debates regarding whether local scale diversity is declining and may
+continue to decline in response to anthropogenic influences [@diaz2015;
+@dornelas2014].
 
 Previous efforts to predict future patterns of species richness, and diversity
-more generally, have focused primarily on building species distributions models.
-In general, these models describe individual species' occurrence patterns as
-functions of the environment [@thomas2004; @urban2015]. By plugging forecasts of
-future environmental conditions into these functions, ecologists can make
-predictions about where each species will occur in the future. These
+more generally, have focused primarily on building species distributions models
+(SDMs). In general, these models describe individual species' occurrence
+patterns as functions of the environment [@thomas2004; @urban2015]. By plugging
+forecasts of future environmental conditions into these functions, ecologists
+can make predictions about where each species will occur in the future. These
 species-level predictions can then be added up ("stacked") to generate predicted
@@
 relate spatial patterns of species richness to the environment have been
-developed and generally perform equivalently to species distribution modeling
-based methods [@algar2009; @distler2015]. This approach is sometimes referred to
-as "macroecological" modeling, because it models the larger-scale pattern
-(richness) more directly [@distler2015].
+developed and generally perform equivalently to stacked SDMs [@algar2009;
+@distler2015]. This approach is sometimes referred to as "macroecological"
+modeling, because it models the larger-scale pattern (richness) directly
+[@distler2015].
 
@@
 dynamics. This is due in part to the long time scales over which many ecological
-forecasts are applied and the resulting difficulty in assessing whether the
-predicted changes occurred [@dietze2016]. What we do know comes from a small
+forecasts are applied [and the resulting difficulty in assessing whether the
+predicted changes occurred @dietze2016]. What we do know comes from a small
 number of hindcasting studies, where models are built using data on species
@@
 approaches to spatial models of species richness, time-series methods, and
-simple baselines. All forecasts model uncertainty and observation, are evaluated
-across different time lags using hindcasting, and are publicly archived to allow
-future assessment. We discuss the implications of these practices for our
-understanding of, and confidence in, the resulting forecasts, and how we can
-continue to build on these approaches improve ecological forecasting in the
-future.
+simple baselines (Figure 1). All forecasts model uncertainty and observation
+error, are evaluated across different time lags using hindcasting, and are
+publicly archived to allow future assessment. We discuss the implications of
+these practices for our understanding of, and confidence in, the resulting
+forecasts, and how we can continue to build on these approaches improve
+ecological forecasting in the future.
+
+![](../figures/model_predictions.png)
 
@@
 model with observer as a random effect built in the Stan probabilistic
-programming language [@stan] and its R interface [@rstan]. Because observer and
-site are strongly related (observers tend to repeatedly sample the same site),
-site was also included as a random effect to ensure that inferred deviations
-were actually observer-related (as opposed to being related to the sites that a
-given observer happened to see). The resulting model partitions the variance in
-observed richness values into site-level variance, observer-level variance, and
-residual variance (e.g. variation within a site from year to year). The
-site-level estimates can also be used directly as the "average" baseline model
-(see below). The estimated observer effects can be subtracted from the richness
-values for a particular observer to provide an estimate of how many species
-would have been found by a "typical" observer. To incorporate uncertainty in
-these "corrected" richness values into the forecasting models we collected 500
-Monte Carlo samples from the model's posterior distribution, and fit each of the
-downstream models with each of the Monte Carlo samples.
+programming language [@stan]. Because observer and site are strongly related
+(observers tend to repeatedly sample the same site), site was also included as a
+random effect to ensure that inferred deviations were actually observer-related
+(as opposed to being related to the sites that a given observer happened to
+see). The resulting model partitions the variance in observed richness values
+into site-level variance, observer-level variance, and residual variance (e.g.
+variation within a site from year to year; Figure 3). The site-level estimates
+can also be used directly as the "average" baseline model (see below). The
+estimated observer effects can be subtracted from the richness values for a
+particular observer to provide an estimate of how many species would have been
+found by a "typical" observer. To incorporate uncertainty in these "corrected"
+richness values into the forecasting models we collected 500 Monte Carlo samples
+from the model's posterior distribution, and fit each of the downstream models
+with each of the Monte Carlo samples.
+
+![](../figures/observer_predictions.png)
 
 ### Site-level models
-
-![](../figures/model_predictions.png)
 
@@
 When comparing forecasts for richness across sites all methods performed well
-(Figure 2; all $R^2 > 0.5$). However SDMs (both stacked and joint) and the
+(Figure 4; all $R^2 > 0.5$). However SDMs (both stacked and joint) and the
 macroecological model all failed to successfully forecast the highest-richness
@@
 
-While the models generally did well in absolute terms (Figure 2), none tended to
+While the models generally did well in absolute terms (Figure 4), none tended to
 outperform the "average" baseline from the generalized linear mixed model
-(Figure 3A). [[Stats go here]].  The auto-ARIMA was generally the
+(Figure 4A). [[Stats go here]].  The auto-ARIMA was generally the
 best-performing non-baseline model, but in many cases (67% of the time), the
@@
 Most models produced confidence intervals that were too narrow, indicating
-overconfident predictions (Figure 3B). The random forest-based SDM stack was the
+overconfident predictions (Figure 4B). The random forest-based SDM stack was the
 most overconfident model, with only `r numbers$rf_coverage_pct`% of observations
@@
 predictive distribution also caused it to have notably higher deviance (Figure
-3C) than the next-worst model, even though its point estimates were not
+4C) than the next-worst model, even though its point estimates were not
 unusually bad in terms of RMSE (3A). As discussed elsewhere [@harris2015], this
@@
 the GBM-based "macroecological" model and especially the mistnet JSDM had
-reasonably well-calibrated uncertainty estimates (Figure 3B); as a result, their
+reasonably well-calibrated uncertainty estimates (Figure 4B); as a result, their
 relative performance was higher in terms of deviance than in terms of RMSE. The
 "naive" model was the only model whose confidence intervals were too wide
-(Figure 3B), which can be attributed to the rapid rate at which these intervals
+(Figure 4B), which can be attributed to the rapid rate at which these intervals
 expand as the time lag increases (Figure 1).
@@
 In general, each model's error increased more-or-less linearly with time,
-although the slopes and initial error rates varied across models (Figure 4). The
+although the slopes and initial error rates varied across models (Figure 5). The
 mistnet JSDM and GBM-based macroecological model's deviance tended to rise less
@@
 When the differences among observers were not explicitly accounted for, nearly
-all metrics of fit worsened (Figure 5). The increased error resulted from a
+all metrics of fit worsened (Figure 6). The increased error resulted from a
 small number of forecasts where observer turnover caused a large shift in the

### 2017-07-31 figures/plots.R

@@ -233,7 +233,7 @@ make_ts_plots = function(models, ylim, use_obs_model, sample_site_id,
     faceting +
     geom_line(aes(y = richness, alpha = .5 * (year < min(bound$year - 1)))) + 
     geom_point(aes(y = richness, fill = factor(observer_id)),
-               size = .8, shape = 21, stroke = .5) +
+               size = 1, shape = 21, stroke = .5) +
     scale_alpha(range = c(0, 1), guide = FALSE) + 
     coord_cartesian(ylim = ylim, expand = TRUE) +
     ylab("Richness") +
@@ -261,6 +261,12 @@ list(ts_models, env_models) %>%
   plot_grid(plotlist = ., nrow = 2)
   
 ggsave(filename = "figures/model_predictions.png", width = 7.5, height = 4)
+
+
+make_ts_plots(c("average", "naive", "rf_sdm"), ylim = c(41, 91), 
+              use_obs_model = TRUE, sample_site_id = 72035,
+              main = "Controlling for observer differences")
+ggsave(filename = "figures/observer_predictions.png", width = 4, height = 4)
 
 # Numbers for the manuscript ----------------------------------------------
 
@@ -290,25 +296,21 @@ observer_error_data = bound %>%
   spread(key = use_obs_model, value = diff) %>% 
   mutate(y = abs(no) - abs(yes))
 
+
 plot_grid(
-  make_ts_plots(c("average", "naive", "rf_sdm"), ylim = c(41, 91), 
-                use_obs_model = TRUE, sample_site_id = 72035,
-                main = "Controlling for observer differences"),
-  plot_grid(
-    make_violins(observer_error_data, 
-                 main = "Absolute error reduction from observer model",
-                 ylab = "Reduction in absolute\nrichness error (species)",
-                 ylim = quantile(observer_error_data$y, c(.0005, .9995)), 
-                 yintercept = 0,
-                 adjust = 2),
-    make_violins(observer_deviance_data, 
-                 main = "Posterior weight of model including observer effect",
-                 ylab = "Posterior weight of\nobserver model",
-                 ylim = c(0, 1), 
-                 yintercept = 0.5, 
-                 adjust = 5),
-    nrow = 2
-  )
+  make_violins(observer_error_data, 
+               main = "Absolute error reduction from observer model",
+               ylab = "Reduction in absolute\nrichness error (species)",
+               ylim = quantile(observer_error_data$y, c(.0005, .9995)), 
+               yintercept = 0,
[…]

### 2017-08-01 manuscript/draft.Rmd
more edits

@@
 csl: csl.csl
-geometry: margin=1in
+geometry: left=3.5cm, right=3.5cm, top=2.25cm, bottom=2.25cm, headheight=12pt, letterpaper
 fontsize: 12pt
@@
 
-![](../figures/model_predictions.png)
+![Example predictions from our six models for a single site. The data points
+from 1982 through 2003, connected by solid lines, were used for training all six
+models; the remaining points were used for evaluating the models' forecasts. In
+each panel, point estimates for each year are shown with lines; the darker
+ribbon indicates 1 standard deviation of uncertainty, and the lighter ribbon
+indicates 95% confidence intervals. **A.** The time series models models were
+trained using one site at a time, based solely on observed richness values
+shown. The first two models ("average" and "naive") served as our simple
+forecasting baselines. **B.** The environmental models were trained to predict
+richness based on elevation, climate, and NDVI; unlike the time series models,
+the environmental models' predictions change from year to year as environmental
+conditions change.](../figures/model_predictions.png)
 
@@
 
-![](../figures/observer_predictions.png)
+![**A.** When all observers are treated the same (black points), predictions can
+be poor. **B.** By accounting for systematic differences between observers
+(represented by the points' colors), most models can be made more robust to
+observer turnover. Note that the "naive" model is already less sensitive to
+observer turnover, and does not benefit as much from modeling
+it.](../figures/observer_predictions.png)
 
@@
 
-![](../figures/scatter.png)
+![Performance of the six models we evaluated when predicting one year into the
+future (2004) or ten years into the future (2013).  In general, the time series
+models (**A.**) outperformed the environmental models (**B.**), and the accuracy
+of the predictions declined as the forecast horizon was
+extended.](../figures/scatter.png)
 
@@
 
-![](../figures/performance_time.png)
+![Performance of all six models over time, according to three metrics. **A.**
+Root mean square error (rmse) describes the distance between each model's point
+estimates and the observed richness values. Note that the three environmental
+models tend to show the largest errors. **B.** Deviance (-2 times the Gaussian
+log-likelihood) describes the lack of fit of entire predictive distributions.
+Note that the stack of single-species random forest SDMs shows much higher error
+than the other models (since its predictions are too confident), and that the
+"naive" model's deviance grows relatively quickly (as its predictions become
+broader and less informative). **C.** Coverage of a model's 95% confidence
+intervals indicates how often the observed values fall inside the predicted
+range. For well-calibrated models, the coverage of the 95% confidence intervals
+should be 95%. As in the previous panel, the "naive" model's predictive
+distribution is too wide (capturing almost all of the data) and the stacked
+SDM's predictive distribution is too narrow (missing almost a third of the
+observed richness values by 2014).](../figures/performance_time.png)
 
@@
 
-![](../figures/observers.png)
+![Accounting for differences between observers typically has a small impact on
+accuracy (especially for the "naive" model), but occasonally changes the
+predicted richness values by 10+ species. These changes usually improve the
+model's predictions, but not always.](../figures/observers.png)
 

### 2017-08-01 figures/plots.R

@@ -9,6 +9,17 @@ prepend_timeframe = function(x) {
   paste0("results", "/", timeframe, "/", x)
 }
 
+base_size = 12 # 12-point font for figures
+
+# peerj PDFs are 14.59 cm wide
+my_ggsave = function(filename, plot, height, width = 14.59, units = "cm", ...){
+  ggsave(filename = filename, 
+         plot = plot, 
+         height = height, 
+         width = width, 
+         units = units, 
+         ...)
+}
 
 forecast_results = readRDS(prepend_timeframe("forecast.rds")) %>% 
   filter(!is.na(richness))
@@ -49,13 +60,13 @@ env_models = c("richness_gbm", "rf_sdm", "mistnet")
 models = c(ts_models, env_models)
 
 R2s = filter(bound, use_obs_model, year %in% c(2004, 2013)) %>% 
-  mutate(x = min(mean), y = max(richness)) %>% 
+  mutate(x = min(mean - 4), y = max(richness)) %>% 
   group_by(model, x, y, year) %>% 
   mutate(R2 = 1 - var(mean - richness) / var(richness)) %>% 
   mutate(R2_formatted = paste("R^2: ", 
                               format(R2,digits = 2, nsmall = 2)))
 
-make_scatterplots = function(models){
+make_scatterplots = function(models, main){
   x_range = range(bound$mean)
   ggplot(data = filter(bound, use_obs_model, year %in% c(2004, 2013),
                        model %in% models), 
@@ -69,19 +80,18 @@ make_scatterplots = function(models){
     geom_text(inherit.aes = FALSE,
               data = filter(R2s, model %in% models),
               aes(x = x, y = y, label = R2_formatted),
-              hjust = 0, vjust = 1,
-              size = 3) +
-    theme_light(base_size = 14) +
-    xlim(x_range)
+              hjust = 0, vjust = 1, size = 3) +
+    theme_light(base_size = base_size) +
+    xlim(x_range) + 
+    ggtitle(main)
 }
 
-plot_grid(
-  make_scatterplots(ts_models),
-  make_scatterplots(env_models),
+scatters = plot_grid(
+  make_scatterplots(ts_models, main = "A. Time-series models"),
+  make_scatterplots(env_models, main = "B. Environmental models"),
   align = "h",
-  nrow = 2,
-  labels = c("A. Time-series models", "B. Environmental models"))
-ggsave(filename = "figures/scatter.png", width = 7.5, height = 10)
+  nrow = 2)
[…]

### 2017-08-01 manuscript/draft.Rmd
more

@@
 being studied. Using hindcasting we made and evaluated `r numbers$n_forecasts`
-forecasts for breeding bird species richness at `r numbers$N_predict_sites`
-sites throughout the continental United States, following ten suggested best
-practices for forecasting. Forecasts were made for up to a decade after the
-final training observations. While all of the models we evaluated produce
-reasonable forecasts, none of them consistently outperforms the simple baseline
-of constant long-term average richness. At present, our best forecast for future
-richness is thus to assume that it is unchanging, and current modeling
-approaches do not reliably improve on this simple baseline. Most of the best
-practices implemented in this study directly influence the outcome of the
-forecasts and result in different conclusions from those made in previous
-studies. This emphasizes the importance of following good practice in making
-ecological forecasts and of starting to actively make more forecasts in such a
-way that they can be quickly evaluated and improved on.
+forecasts for breeding bird species richness from each of 6 modeling approaches
+at `r numbers$N_predict_sites` sites throughout the continental United States,
+following ten suggested best practices for forecasting. Forecasts were made for
+up to a decade after the final training observations. While all of the models we
+evaluated produce reasonable forecasts, none of them consistently outperforms
+the simple baseline of constant long-term average richness. At present, our best
+forecast for future richness is thus to assume that it is unchanging, and
+current modeling approaches do not reliably improve on this simple baseline.
+Most of the best practices implemented in this study directly influence the
+outcome of the forecasts and result in different conclusions from those made in
+previous studies. This emphasizes the importance of following good practice in
+making ecological forecasts and of starting to actively make more forecasts in
+such a way that they can be quickly evaluated and improved on.
 
@@
 developing good forecasting models and understanding how accurately these models
-can predict the future. These components of good forecasting and evaluation
-(Box 1) broadly involve: 1) expanding the use of data to include biological and
+can predict the future. These components of good forecasting and evaluation (Box
+1) broadly involve: 1) expanding the use of data to include biological and
 environmental time-series [@tredennick2016]; 2) accounting for uncertainty in
@@
 
-In contrast to the time-series models, most attempts to predict species richness
-focus on using correlative models to predict richness changes over time from
-environmental data. We tested three common variants of this approach - direct
-modeling of species richness, stacking individual species distribution models,
-and joint species distribution models (JSDMs). Following the standard approach,
-site-level random effects were not included in these models as predictors,
-meaning that this approach implicitly assumes that two sites with identical
-Bioclim, elevation, and NDVI values should have identical richness
-distributions. As above, we included observer effects & the associated
-uncertainty by running these models 500 times (once per MCMC sample).
+In contrast to the site-level time-series models, most attempts to predict
+species richness focus on using correlative models based on environmental
+variables. We tested three common variants of this approach: direct modeling of
+species richness; stacking individual species distribution models; and joint
+species distribution models (JSDMs). Following the standard approach, site-level
+random effects were not included in these models as predictors, meaning that
+this approach implicitly assumes that two sites with identical Bioclim,
+elevation, and NDVI values should have identical richness distributions. As
+above, we included observer effects & the associated uncertainty by running
+these models 500 times (once per MCMC sample).
 
@@
 number of trees used for prediction was selected using the "out of bag"
-estimator; this number averaged [[N]] for the non-observer data and [[N]] for
-the observer-corrected data.
+estimator; this number averaged `r numbers$n_gbm[1]` for the non-observer data
+and `r numbers$n_gbm[2]` for the observer-corrected data.
 
@@
 possibility that a site will be much more (or less) suitable for birds in
-general, or particular groups of birds, than one would expect based on the
+general (or particular groups of birds) than one would expect based on the
 available environmental measurements alone. As a result, JSDMs do a better job
@@
 underdispersed for individual sites), so these count models would have had
-difficulty fitting the data. Since richness had a relatively large mean ([[N]])
-and all models already produce continuous richness estimates (due to summing
-probabilities in the (J)SDM case), we defined model performance for all models
-in terms of continuous Gaussian errors instead of using discrete count
-distributions. When a model was run multiple times for the purpose of correcting
-for observer effects, we used the mean of those runs' point estimates as our
-final point estimate and we calculated the uncertainty using the law of total
-variance (i.e.  the average of the model runs' variance, plus the variance in
-the point estimates).
+difficulty fitting the data. Since richness had a relatively large mean 
+(`r numbers$richness_summary$mean`) and all models already produce continuous
+richness estimates (due to summing probabilities in the (J)SDM case), we defined
+model performance for all models in terms of continuous Gaussian errors instead
+of using discrete count distributions. When a model was run multiple times for
+the purpose of correcting for observer effects, we used the mean of those runs'
+point estimates as our final point estimate and we calculated the uncertainty
+using the law of total variance (i.e.  the average of the model runs' variance,
+plus the variance in the point estimates).
 
-We evaluated the performance of each model for forecasting using the data for
-each year between `r timeframe$last_train_year` and `r timeframe$end_yr`. We used
-three metrics for evaluating performance: 1) root-mean-square error (RMSE) to
-determine how far, on average, the models' predictions were from the observed
-value; 2) the 95% prediction interval coverage to determine how well the models
-predicted the range of possible outcomes; and 3) deviance (i.e. negative 2 times
-the Gaussian log-likelihood) as an integrative measure of fit incorporating good
-point estimates, precision, and coverage.  In addition to looking at forecast
+We evaluated each model's for forecasts using the data for each year between `r
+timeframe$last_train_year` and `r timeframe$end_yr`. We used three metrics for
+evaluating performance: 1) root-mean-square error (RMSE) to determine how far,
+on average, the models' predictions were from the observed value; 2) the 95%
+prediction interval coverage to determine how well the models predicted the
+range of possible outcomes; and 3) deviance (i.e. negative 2 times the Gaussian
+log-likelihood) as an integrative measure of fit incorporating good point
+estimates, precision, and coverage.  In addition to looking at forecast
 performance in general we evaluated how performance changed as the time horizon
@@
 different years. In the training set, the residuals had a standard deviation of
-about [[+/- 5]] species. After correcting for observer differences, there was
-little temporal autocorrelation in these residuals ([[about 0.04, +/-
+about `r numbers$resid_sd` species. After correcting for observer differences,
+there was little temporal autocorrelation in these residuals ([[about 0.04, +/-
 something]]), suggesting that richness is approximately stationary on these time
@@
 sites resulting in a notable clustering of predicted values near ~60 species and
-the poorest performance ($R^2$ `r numbers$env_R2s`, versus XX-XX for the
-within-site methods).
+the poorest performance ($R^2$=`r numbers$env_R2s`, versus $R^2$=`r numbers$ts_R2s` 
+for the within-site methods).
 
@@
 
-While the models generally did well in absolute terms (Figure 4), none tended to
-outperform the "average" baseline from the generalized linear mixed model
-(Figure 4A). [[Stats go here]].  The auto-ARIMA was generally the
+While the models generally did well in absolute terms (Figure 4), none
+consistently outperformed the "average" baseline from the generalized linear
+mixed model (Figure 4A). [[Stats go here]].  The auto-ARIMA was generally the
 best-performing non-baseline model, but in many cases (67% of the time), the
-best-fitting ARIMA model chosen by the model selection procedure had no
-autoregressive, drift, or moving average terms, making it similar to the
-"Average" model.
+auto.arima procedure automatically selected a model with no autoregressive,
+drift, or moving average terms, making it similar to the "Average" model.
 
-![](../figures/model_violins.png)
+![None of the models provided a consistent improvement over the "average"
+baseline, indicated by a horizontal line. **A.** Models' absolute error was
+generally similar or larger than "average" model, with large outliers in both
+directions. **B.** The other models' deviance was also generally higher than
+the "average" baseline.](../figures/model_violins.png)
 
@@
 species-level predictions are independent. Because they avoided this assumption,
-the GBM-based "macroecological" model and especially the mistnet JSDM had
-reasonably well-calibrated uncertainty estimates (Figure 4B); as a result, their
-relative performance was higher in terms of deviance than in terms of RMSE. The
-"naive" model was the only model whose confidence intervals were too wide
-(Figure 4B), which can be attributed to the rapid rate at which these intervals
-expand as the time lag increases (Figure 1).
+the GBM-based "macroecological" model and the mistnet JSDM had reasonably
+well-calibrated uncertainty estimates (Figure 4B); as a result, their relative
+performance was higher in terms of deviance than in terms of RMSE. The "naive"
+model was the only model whose confidence intervals were too wide (Figure 4B),
+which can be attributed to the rapid rate at which these intervals expand
+(Figure 1).
 
@@
 accuracy (especially for the "naive" model), but occasonally changes the
-predicted richness values by 10+ species. These changes usually improve the
-model's predictions, but not always.](../figures/observers.png)
+predicted richness values by more than 10 species. These changes usually improve
+the model's predictions, but not always.](../figures/observers.png)
 
@@
 
-Forecasting is an emerging imperative in ecology and as such the field needs to
+Forecasting is an emerging imperative in ecology; as such, the field needs to
 develop and follow best practices for conducting and evaluating ecological
@@
 the future, relative to other sites, it generally does. However, the forecasts
-perform relatively poorly at predicting variation in richness over time at a
-location. If a site is predicted to have a higher richness next year than it did
-last year there is little confidence that this will be the case. As a result,
-the baseline models, like the long-term average, along with site-level
-time-series models, provided some of the best forecasts for future biodiversity.
+perform relatively poorly at predicting variation in richness over time. If a
+site is predicted to have a higher richness next year than it did last year,
+there is little confidence that this will be the case. As a result, baseline
+models, like the long-term average, along with the other site-level time-series
+models, consistently provided the best forecasts for future biodiversity.
 
-The most commonly used method to forecast future biodiversity, stacked SDMS
+The most commonly used method to forecast future biodiversity, stacked SDMs
 [@urban2015], provided the worst overall forecasts of all methods evaluated. In
@@
 repeated observations at the same site. Since site-level richness is relatively
-stable, model evaluations that show high accuracy in predicting differences
-among sites may not indicate much about a model's ability to predict changes
-over time. This issue has also recently been highlighted as an issue
-for using SDMs to predict the future locations of individual species
-[@rapacciuolo2012]. This result is particularly sobering because this approach
-forms the main foundation for estimates of the predicted loss of biodiversity to
-climate change [@thomas2004; @urban2015]. The poor performance of SDM-based
-forecasts highlights the crucial importance of comparing multiple modeling
-approaches when conducting ecological forecasts, and in particular the value of
-comparing results to simple baselines to avoid over-interpreting the information
-present in these forecasts (Box 1).  Disciplines that have more mature forecasting
-cultures often do this by reporting "forecast skill", which is the improvement
-in the forecast relative to a simple baseline [@jolliffe2003]. We recommend
-following Ye et al. [@ye2015] and adopting this approach in future ecological
-forecasting research. It also highlights the need for using hindcasting and
-time-series data for assessing the effectiveness of forecasting methods, since
-this allows comparisons to baselines and assessment of predicted dynamics within
-sites or regions.
+stable, model evaluations that show high accuracy across spatial gradients may
+not indicate much about a model's ability to predict changes over time. This
+issue has also recently been highlighted as an issue for using SDMs to predict
+the future locations of individual species [@rapacciuolo2012]. This result is
+particularly sobering because this approach forms the main foundation for
+estimates of the predicted loss of biodiversity to climate change [@thomas2004;
+@urban2015]. The poor performance of SDM-based forecasts highlights the crucial
+importance of comparing multiple modeling approaches when conducting ecological
+forecasts, and in particular the value of comparing results to simple baselines
+to avoid over-interpreting the information present in these forecasts (Box 1).
+Disciplines that have more mature forecasting cultures often do this by
+reporting "forecast skill", i.e., the improvement in the forecast relative to a
+simple baseline [@jolliffe2003]. We recommend following Ye et al. [@ye2015] and
+adopting this approach in future ecological forecasting research. It also
+highlights the need for using hindcasting and time-series data for assessing the
+effectiveness of forecasting methods, since this allows comparisons to baselines
+and assessment of predicted dynamics within sites or regions.
 
-One of the main alternatives to summed SDMs is modeling richness directly using
+One of the main alternatives to stacked SDMs is modeling richness directly using
 environmental variables and using the resulting models to make forecasts
@@
 between stacked SDMs and direct richness modeling for point estimates, they also
-show that modeling richness directly results in much better estimation of the
-uncertainty of the forecast and therefore is a better overall approach to
+show that stacked SDMs dramatically understimate the uncertainty. For this
+reason, direct modeling of richness therefore is a better overall approach to
 forecasting richness. A similar result is seen when comparing joint species
@@
 SDMs, but the JSDMs provide much better estimates of uncertainty [@harris2015;
-others?]. In fact, JSDMs and modeling richness directly provide some of the best
+others?]. In fact, JSDMs and direct richness modeling provide some of the best
 estimates of uncertainty across all modeling approaches. This highlights the
 importance of evaluating models based on their uncertainty as well as their mean
-estimates (Box 1). Not doing so is particularly problematic in this case because the
-summed species distribution models are overly confident meaning that using them
-would lead to believing that richness would be restricted to a much narrower
-range than would actually occur.
+estimates (Box 1). Not doing so is particularly problematic in this case because
+the summed species distribution models are overly confident meaning that using
+them would lead to believing that richness would be restricted to a much
+narrower range than would actually occur.
 
@@
 environments or direct environmental constraints on richness may continue to
-produce forecasts that are worse than simply assuming the systems are
-static. This would be expected to occur if richness in these systems is not
-changing over the relevant time scales and therefore simple models of no
-directional change are most appropriate. Recent suggestions that local scale
-richness in some systems is not changing directionally at multi-decadal scales
-supports this possibility [@brown2001; @ernest2001; @dornelas2014;
-@vellend2013]. This lack of change may be expected even in the presence of
-substantial changes in environmental conditions and species composition at a
-site due to the replacement of species from the regional pool [@brown2001;
-@ernest2001]. On average the Breeding Bird Survey sites used in this study show
-little change in richness ([[Add summary stat here]]) [see also
-@lasorte2005]. This absence of rapid change is beneficial for the absolute
-accuracy of forecasts across sites, because if a past year's richness is known
-it is easy to estimate a future richness, and explains why stacked SDMs perform
-relatively well at this task despite failing to capture meaningful
-dynamics. However it makes it difficult to improve forecasts relative to simple
-baselines, since those baselines are close to representing what is actually
-occurring in the system. This suggests that simple time-series models and
-baselines should be actively considered for forecasts of richness and other
-stable aspects of biodiversity and also suggests that future efforts to
-understand and forecast biodiversity should also focus on composition since it
-is expected to be more dynamic [@ernest2001; @dornelas2014].
+produce forecasts that are worse than simply assuming the systems are static.
+This would be expected to occur if richness in these systems is not changing
+over the relevant time scales and therefore simple models of no directional
+change are most appropriate. Recent suggestions that local scale richness in
+some systems is not changing directionally at multi-decadal scales supports this
+possibility [@brown2001; @ernest2001; @dornelas2014; @vellend2013]. This lack of
+change may be expected even in the presence of substantial changes in
+environmental conditions and species composition at a site due to the
+replacement of species from the regional pool [@brown2001; @ernest2001]. On
+average the Breeding Bird Survey sites used in this study show little change in
+richness (site-level SD of `r numbers$resid_sd` species, after controlling for
+differences among observers) [see also @lasorte2005]. This absence of rapid
+change is beneficial for the absolute accuracy of forecasts across sites,
+because if a past year's richness is known it is easy to estimate a future
+richness, and explains why stacked SDMs perform relatively well at this task
+despite failing to capture meaningful dynamics. However it makes it difficult to
+improve forecasts relative to simple baselines, since those baselines are close
+to representing what is actually occurring in the system. This suggests that
+simple time-series models and baselines should be actively considered for
+forecasts of richness and other stable aspects of biodiversity and also suggests
+that future efforts to understand and forecast biodiversity should also focus on
+composition since it is expected to be more dynamic [@ernest2001;
+@dornelas2014].
 

### 2017-08-01 figures/plots.R

@@ -123,16 +123,17 @@ make_violins = function(data, ylab = "", ylim, yintercept, adjust = 1, main){
 # of a small number of extreme outliers
 model_violins = plot_grid(
   for_violins %>% 
-    mutate(model = model.other, y = -abs_diff) %>% 
+    mutate(model = model.other, y = abs_diff) %>% 
     make_violins(yintercept = 0, adjust = 5,
                  ylim = quantile(for_violins$abs_diff, c(.0005, .9995)),
-                 main = "A. Model improvement over\n\"Average\" baseline",
-                 ylab = "Absolute error reduction"),
+                 main = "A. Absolute error increase versus \"Average\" baseline",
+                 ylab = "Absolute error increase (species)"),
   for_violins %>% 
-    mutate(model = model.other, y = plogis(-dev_diff / 2)) %>%
-    make_violins(yintercept = 0.5, adjust = 1, ylim = c(0, 1), 
-                 main = "B. Model posterior weight versus\n\"Average\" baseline",
-                 ylab = "Posterior weight"),
+    mutate(model = model.other, y = dev_diff) %>%
+    make_violins(yintercept = 0, adjust = 1, 
+                 ylim = quantile(for_violins$abs_diff, c(.0005, .9995)), 
+                 main = "B. Deviance increase versus \"Average\" baseline",
+                 ylab = "Deviance increase"),
   nrow = 2
 )
 my_ggsave(file = "figures/model_violins.png", 
@@ -326,7 +327,7 @@ observer_deviance_data = bound %>%
                                              yes = "TRUE")) %>% 
   select(site_id, year, model, use_obs_model, deviance) %>% 
   spread(key = use_obs_model, value = deviance) %>% 
-  mutate(y = plogis(0.5 * (no - yes)))
+  mutate(y = no - yes)
 
 observer_error_data = bound %>% 
   mutate(use_obs_model = forcats::fct_recode(factor(use_obs_model), 
@@ -345,10 +346,10 @@ obs_violins = plot_grid(
                yintercept = 0,
                adjust = 2),
   make_violins(observer_deviance_data, 
-               main = "Posterior weight of model including observer effect",
+               main = "Deviance reduction from observer model",
                ylab = "Posterior weight of\nobserver model",
-               ylim = c(0, 1), 
[…]

### 2017-08-02 manuscript/draft.Rmd
More changes.

@@
 
-The site-observer model found that 70% of the variance in richness in the
+The site-observer mixed model found that 70% of the variance in richness in the
 training set could be explained by differences among sites, and 21% could be
@@
 about `r numbers$resid_sd` species. After correcting for observer differences,
-there was little temporal autocorrelation in these residuals ([[about 0.04, +/-
-something]]), suggesting that richness is approximately stationary on these time
-scales.
+there was little temporal autocorrelation in these residuals (i.e. the residuals
+in one year explain `r round(100 *numbers$residual_autocor^2, 1)`% of the
+variance in the residuals of the following year), suggesting that richness was
+approximately stationary between 1982 and 2003.
 

### 2017-08-02 figures/plots.R

@@ -2,6 +2,8 @@ devtools::load_all()
 library(tidyverse)
 library(cowplot)
 library(ggjoy)
+library(mvtnorm)
+
 settings = yaml::yaml.load_file("settings.yaml")
 
 timeframe = "train_22"
@@ -51,6 +53,19 @@ bound = bind_rows(forecast_results, gbm_results, average_results,
          p = pnorm(richness, mean, sd),
          deviance = -2 * dnorm(richness, mean, sd, log = TRUE)) %>% 
   ungroup()
+
+# Divide squared error by predictive variance to get the optimal sd factor
+# for the RF model
+rf2 = bound %>% 
+  filter(use_obs_model, model == "rf_sdm")
+rf_sd_factor = rf2 %>% 
+  summarize(sqrt(mean(diff^2 / sd^2))) %>% 
+  pull()
+rf2$sd = rf2$sd * rf_sd_factor
+
+rf2 = rf2 %>% 
+  mutate(p = pnorm(richness, mean, sd),
+         deviance = -2 * dnorm(richness, mean, sd, log = TRUE))
 
 
 # scatterplot -------------------------------------------------------------
@@ -105,7 +120,7 @@ for_violins = bound %>%
          squared_diff = diff.other^2 - diff.avg^2,
          abs_diff = abs(diff.other) - abs(diff.avg))
 
-make_violins = function(data, ylab = "", ylim, yintercept, adjust = 1, main){
+make_violins = function(data, ylab = "", ylim, yintercept, adjust = 2, main){
   data %>% 
     ggplot(aes(x = model, y = y)) + 
     geom_hline(yintercept = yintercept, color = alpha(1, .25), size = 1/2) +
@@ -124,13 +139,13 @@ make_violins = function(data, ylab = "", ylim, yintercept, adjust = 1, main){
 model_violins = plot_grid(
   for_violins %>% 
     mutate(model = model.other, y = abs_diff) %>% 
-    make_violins(yintercept = 0, adjust = 5,
+    make_violins(yintercept = 0, 
                  ylim = quantile(for_violins$abs_diff, c(.0005, .9995)),
                  main = "A. Absolute error increase versus \"Average\" baseline",
                  ylab = "Absolute error increase (species)"),
   for_violins %>% 
     mutate(model = model.other, y = dev_diff) %>%
-    make_violins(yintercept = 0, adjust = 1, 
+    make_violins(yintercept = 0, 
[…]

### 2017-08-02 manuscript/draft.Rmd
Stuff

@@
 consistently outperformed the "average" baseline from the generalized linear
-mixed model (Figure 4A). [[Stats go here]].  The auto-ARIMA was generally the
+mixed model (Figure 4). The auto-ARIMA was generally the
 best-performing non-baseline model, but in many cases (67% of the time), the
@@
 drift, or moving average terms, making it similar to the "Average" model.
+All five alternatives to the "average" model achieved lower error on some
+of the sites in some years, but each one had a higher mean absolute error
+and higher mean deviance.
 
@@
 Most models produced confidence intervals that were too narrow, indicating
-overconfident predictions (Figure 4B). The random forest-based SDM stack was the
+overconfident predictions (Figure 5C). The random forest-based SDM stack was the
 most overconfident model, with only `r numbers$rf_coverage_pct`% of observations
@@
 predictive distribution also caused it to have notably higher deviance (Figure
-4C) than the next-worst model, even though its point estimates were not
-unusually bad in terms of RMSE (3A). As discussed elsewhere [@harris2015], this
+5B) than the next-worst model, even though its point estimates were not
+unusually bad in terms of RMSE (5A). As discussed elsewhere [@harris2015], this
 overconfidence is a product of the assumption in stacked SDMs that errors in the
@@
 the GBM-based "macroecological" model and the mistnet JSDM had reasonably
-well-calibrated uncertainty estimates (Figure 4B); as a result, their relative
+well-calibrated uncertainty estimates (Figure 5B); as a result, their relative
 performance was higher in terms of deviance than in terms of RMSE. The "naive"
-model was the only model whose confidence intervals were too wide (Figure 4B),
+model was the only model whose confidence intervals were too wide (Figure 5C),
 which can be attributed to the rapid rate at which these intervals expand

### 2017-08-02 figures/plots.R

@@ -300,7 +300,7 @@ make_ts_plots = function(models, ylim, use_obs_model, sample_site_id,
 # before 2004.
 model_predictions = list(ts_models, env_models) %>% 
   map(~make_ts_plots(.x, 
-                     ylim = c(42, 73), 
+                     ylim = c(42, 85), 
                      use_obs_model = FALSE, 
                      sample_site_id = 91034,
                      main = ifelse(

### 2017-08-06 manuscript/draft.Rmd
barcharts etc

@@
 for our uncertainty in the true values of those effects. All of these models use
-a Gaussian error distribution.
+a Gaussian error distribution (rather than a count distribution) for the reasons
+discussed below .
 
@@
 of forecasting increased by plotting performance metrics against the forecast
-horizon.
+horizon. Finally, we decomposed each model's squared error into two components:
+one described the squared error associated with site-level means, while the
+other described the squared error associated with annual fluctuations in
+richness within a site. This decomposition tells us the extent to which each
+model's error depends on consistent differences among sites versus changes in
+site-level richness from year to year.
 
@@
 
-In general, each model's error increased more-or-less linearly with time,
-although the slopes and initial error rates varied across models (Figure 5). The
-mistnet JSDM and GBM-based macroecological model's deviance tended to rise less
-rapidly over time than the baseline models' deviance did; for the ten-year
-forecasts, these environment-based models outperformed the Naive model and
-nearly matched the performance of the Average model.
+Figure 6 shows each model's squared error, divided into consistent site-level
+biases, versus errors that fluctuated year-to-year. Across all models, most of
+the error came from errors in the site-level mean. The "average" model, which
+was based on the site-level mean, had the lowest error in this regard. In
+contrast, the three environmental models showed the largest biases at the site
+level; this makes sense, given that they could not explicitly distinguish among
+sites with similar climate, NDVI, and elevation.
+
+![Each model's squared error can be divided into two parts. If a model
+consistently over-estimates or under-estimates the richness at a given site,
+this contributes to its error in predicting the site-level mean. The second
+component is based on errors in predicting fluctuations in a site's richness
+over time. Both components of the mean squared error were lower for the time
+series models than for the environmental models.](../figures/barcharts.png)
 
 When the differences among observers were not explicitly accounted for, nearly
-all metrics of fit worsened (Figure 6). The increased error resulted from a
+all metrics of fit worsened (Figure 7). The increased error resulted from a
 small number of forecasts where observer turnover caused a large shift in the
@@
 ![Accounting for differences between observers typically has a small impact on
-accuracy (especially for the "naive" model), but occasonally changes the
-predicted richness values by more than 10 species. These changes usually improve
-the model's predictions, but not always.](../figures/observers.png)
+accuracy (especially for the "naive" model), but occasonally changes a
+prediction's accuracy by more than 10 species. These changes are beneficial, on
+average.](../figures/observers.png)
 
@@
 
-The most commonly used method to forecast future biodiversity, stacked SDMs
+The most commonly used method to forecast future biodiversity, SDMs
 [@urban2015], provided the worst overall forecasts of all methods evaluated. In
@@
 narrower range than would actually occur.
-
-For the error metric that evaluates both models' point estimates and their
-uncertainty simultaneously, the JSDM and the nonlinear richness model converge
-towards the time-series baselines at the longest time-scales assessed. This
-convergence at longer time scales highlights the importance of considering how
-forecasting method performance changes with the distance into the future being
-predicted (Box 1). The value of this approach, and associated forecast horizons,
-has been raised for knowing how far into the future a model can be effectively
-used for forecasting [@petchey2015], but our results suggest a broader value to
-this approach for considering the potential importance of different models and
-processes for making forecasts at different scales. Our results show that for
-forecasts 1-3 years in the future, baselines and time-series based approaches
-outperform all other methods using integrative metrics. At these time scales,
-environmental changes are relatively small and changes in the biota may lag
-behind changes in the environment. However, as the timescale of the forecast
-increases to a decade the JSDMs and environmental richness models get closer
-to the accuracy of the time-series methods and even surpass some of those
-methods. As the forecast distance increases the amount of environmental change
-is expected to increase and the system will have more time to respond,
-potentially leading to shifts towards improved relative performance of the
-models incorporating environmental and ecological information. If this shift
-with forecast distance in the relative importance of different processes and
-models continues, this could lead to JSDMs and environmental richness models
-outperforming time-series based approaches at sufficiently long time scales.
-These results suggest that assessment of how forecast performance changes with
-time lag is not only important for determining how far in the future to
-forecast, but also for determining which models and processes are most relevant
-for making forecasts at different time scales. For example ensemble forecasts
-could be created which incorporate these time lags into the weights for
-individual models.
 

### 2017-08-06 figures/plots.R

@@ -52,7 +52,18 @@ bound = bind_rows(forecast_results, gbm_results, average_results,
   mutate(diff = richness - mean, z = diff / sd,
          p = pnorm(richness, mean, sd),
          deviance = -2 * dnorm(richness, mean, sd, log = TRUE)) %>% 
-  ungroup()
+  ungroup() %>% 
+  mutate(
+    model_name = forcats::fct_recode(
+      factor(model), 
+      Average = "average", 
+      Naive = "naive",
+      Auto.arima = "auto.arima", 
+      `Stacked random forest SDMs` = "rf_sdm",
+      `Mistnet JSDM` = "mistnet",
+      `GBM richness regression` = "richness_gbm"
+    )
+  )
 
 # Divide squared error by predictive variance to get the optimal sd factor
 # for the RF model
@@ -73,6 +84,7 @@ rf2 = rf2 %>%
 ts_models = c("average", "naive", "auto.arima")
 env_models = c("richness_gbm", "rf_sdm", "mistnet")
 models = c(ts_models, env_models)
+model_names = unique(bound$model_name)
 
 R2s = filter(bound, use_obs_model, year %in% c(2004, 2013)) %>% 
   mutate(x = min(mean - 4), y = max(richness)) %>% 
@@ -208,6 +220,65 @@ time_error = plot_grid(
 my_ggsave(file = "figures/performance_time.png", plot = time_error, height = 15)
 
 
+
+# Decomposed squared error ------------------------------------------------
+
+barchart_data = bound %>% 
+  rename(y = richness, x = mean) %>% 
+  filter(use_obs_model) %>% 
+  group_by(site_id, model) %>% 
+  mutate(
+    x_bar = mean(x),
+    y_bar = mean(y),
+    delta_x = x - x_bar, 
+    delta_y = y - y_bar
+  ) %>% 
+  group_by(model) %>% 
+  summarize(`Site-level mean` = var(y_bar - x_bar), 
+            `Annual fluctuations` = var(delta_y - delta_x)) %>% 
+  gather(key = `Error component`, value = value, -1) %>% 
+  mutate(
+    env = model %in% env_models,
+    model = factor(model, levels = !!models),
+    `Error component` = factor(`Error component`, 
+                               levels = rev(sort(unique(`Error component`))))
+  )
+
+
+make_error_barchart = function(use_env){
+  filter(barchart_data, env == use_env) %>% 
[…]

### 2017-08-12 manuscript/draft.Rmd
table, other edits

@@
 ---
+
+```{r, echo=FALSE, message=FALSE}
+library(tidyverse, quietly = TRUE)
+library(kableExtra, quietly = TRUE)
+```
 
@@
 
+\begin{singlespace}
+
+```{r table, echo=FALSE}
+
+table_caption = "A summary of the six models evaluated here. The time series
+models were trained site-by-site, without access to any environmental
+information. The environmental models were trained at a continental scale,
+without information regarding which transects occurred at which site or during
+which year. Most of the models were trained to predict richness directly.
+However, the random forest SDMs were each trained to predict the occurrence of
+individual species at each site, and these species-level predictions were added
+up to produce richness estimates. The mistnet JSDM was trained to predict the
+full species composition at each site, and the number of species in its
+predictions was used as an estimate of richness."
+
+table_data = read_csv("../figures/table.csv",col_types = "ccccc")
+
+table_data %>% 
+  knitr::kable(
+    format = "latex", 
+    escape = FALSE, 
+    booktabs = TRUE,
+    caption = table_caption
+  ) %>% 
+  add_header_above(c(" " = 2, "Predictors" = 3)) %>% 
+  group_rows("Time series models", 1, 3) %>% 
+  group_rows("Environmental models", 4, 6) %>% 
+  kable_styling(position = "center")
+```
+
+\end{singlespace}
+
 In this paper, we attempt to forecast the species richness of breeding birds at
@@
 approaches to spatial models of species richness, time-series methods, and
-simple baselines (Figure 1). All forecasts model uncertainty and observation
+simple baselines (Table 1). All forecasts model uncertainty and observation
 error, are evaluated across different time lags using hindcasting, and are
@@
 forecasts of biodiversity in the form of species richness. The results of this
-effort are both promising and humbling. On an absolute basis, when comparing
-forecasts across sites, many different approaches to forecasting produce
-reasonable forecasts. If a site is predicted to have a high number of species in
-the future, relative to other sites, it generally does. However, the forecasts
-perform relatively poorly at predicting variation in richness over time. If a
-site is predicted to have a higher richness next year than it did last year,
-there is little confidence that this will be the case. As a result, baseline
-models, like the long-term average, along with the other site-level time-series
-models, consistently provided the best forecasts for future biodiversity.
+effort are both promising and humbling. When comparing forecasts across sites,
+many different approaches to forecasting produce reasonable forecasts. If a site
+is predicted to have a high number of species in the future, relative to other
+sites, it generally does. However (after controlling for observer differences),
+none of the methods we evaluated could reliably tell us whether site-level
+richness would increase or decrease over time (Figure 6), which is generally the
+stated purpose of these forecasts. As a result, our baseline models, which did
+not attempt to anticipate changes in richness over time, generally provided
+the best forecasts for future biodiversity.
 
-The most commonly used method to forecast future biodiversity, SDMs
-[@urban2015], provided the worst overall forecasts of all methods evaluated. In
-general, SDM-based forecasts are rarely evaluated, particularly in terms of
-repeated observations at the same site. Since site-level richness is relatively
-stable, model evaluations that show high accuracy across spatial gradients may
-not indicate much about a model's ability to predict changes over time. This
-issue has also recently been highlighted as an issue for using SDMs to predict
-the future locations of individual species [@rapacciuolo2012]. This result is
-particularly sobering because this approach forms the main foundation for
-estimates of the predicted loss of biodiversity to climate change [@thomas2004;
-@urban2015]. The poor performance of SDM-based forecasts highlights the crucial
-importance of comparing multiple modeling approaches when conducting ecological
-forecasts, and in particular the value of comparing results to simple baselines
-to avoid over-interpreting the information present in these forecasts (Box 1).
-Disciplines that have more mature forecasting cultures often do this by
-reporting "forecast skill", i.e., the improvement in the forecast relative to a
-simple baseline [@jolliffe2003]. We recommend following Ye et al. [@ye2015] and
-adopting this approach in future ecological forecasting research. It also
-highlights the need for using hindcasting and time-series data for assessing the
-effectiveness of forecasting methods, since this allows comparisons to baselines
-and assessment of predicted dynamics within sites or regions.
+These results raise a number of important issues. First, we should be skeptical
+of predictions about future biodiversity from models that have not been
+validated in terms of repeated observations at the same site. Since site-level
+richness is relatively stable, model evaluations that show high accuracy across
+spatial gradients may not indicate much about a model's ability to predict
+changes over time. This issue has also recently been highlighted as an issue for
+using SDMs to predict the future locations of individual species
+[@rapacciuolo2012; @oedekoven2017]. This result is particularly sobering because
+SDMs form the main foundation for estimates of the predicted loss of
+biodiversity to climate change [@thomas2004; @urban2015].
+
+Second, our results highlight the importance of comparing multiple modeling
+approaches when conducting ecological forecasts, and in particular, the value of
+comparing results to simple baselines to avoid over-interpreting the information
+present in these forecasts (Box 1). Disciplines that have more mature
+forecasting cultures often do this by reporting "forecast skill", i.e., the
+improvement in the forecast relative to a simple baseline [@jolliffe2003]. We
+recommend following Ye et al. [@ye2015] and adopting this approach in future
+ecological forecasting research. It also highlights the need for using
+hindcasting and time-series data for assessing the effectiveness of forecasting
+methods, since this allows comparisons to baselines and assessment of predicted
+dynamics within sites or regions.
 

### 2017-08-13 manuscript/draft.Rmd
abstract & title

@@
 ---
-title: Using best practices for forecasting biodiversity changes in breeding birds
+title: Best practices for forecasting changes in biodiversity: an example using breeding birds
 author:
@@
 an increasing number of studies forecasting biodiversity, there is little
-information available on how well these forecasts work. Most diversity forecasts
-do not conduct meaningful evaluations of the forecasts, fail to account for
-uncertainty, and do not use time-series data that captures the actual dynamics
-being studied. Using hindcasting we made and evaluated `r numbers$n_forecasts`
-forecasts for breeding bird species richness from each of 6 modeling approaches
-at `r numbers$N_predict_sites` sites throughout the continental United States,
-following ten suggested best practices for forecasting. Forecasts were made for
-up to a decade after the final training observations. While all of the models we
-evaluated produce reasonable forecasts, none of them consistently outperforms
-the simple baseline of constant long-term average richness. At present, our best
-forecast for future richness is thus to assume that it is unchanging, and
-current modeling approaches do not reliably improve on this simple baseline.
-Most of the best practices implemented in this study directly influence the
-outcome of the forecasts and result in different conclusions from those made in
-previous studies. This emphasizes the importance of following good practice in
-making ecological forecasts and of starting to actively make more forecasts in
-such a way that they can be quickly evaluated and improved on.
+information available on how well these forecasts work. Most biodiversity
+forecasts are not evaluated in terms of their ability to predict future
+diversity, fail to account for uncertainty, and do not use time-series data that
+captures the actual dynamics being studied. We compiled a list of ten best
+practices for forecasting biodiversity and used hindcasting to evaluate six
+different modeling approaches for predicting breeding bird species richness.
+Each method was evaluated at `r numbers$N_predict_sites` sites distributed
+throughout the continental United States. Forecasts were evaluated for up to a
+decade after the final training observations. While all of the models that we
+evaluated explained most of the variance in richness, none of them consistently
+outperforms a simple baseline model where each site's species richness
+fluctuates randomly around a constant value. At present, our best forecast for
+future richness is thus to assume that it is unchanging, and current modeling
+approaches do not reliably improve on this simple baseline. Most of the best
+practices implemented in this study (e.g., evaluating confidence intervals as
+well as point estimates) directly influence the outcome of the forecasts; some
+even led to large changes in the relative performance of different modeling
+approaches. We emphasize the importance of following specific best practices
+when making ecological forecasts, and of robust evaluation of model predictions
+to facilitate rapid improvement in our biodiversity forecasts.
 

### 2017-08-13 manuscript/draft.Rmd
fix knitting syntax
more abstract

@@
 ---
-title: Best practices for forecasting changes in biodiversity: an example using breeding birds
+title: "Best practices for forecasting changes in biodiversity: an example using breeding birds"
 author:
@@
 Biodiversity forecasts are important for conservation, management, and
-evaluating how well current models characterize natural systems. While there are
-an increasing number of studies forecasting biodiversity, there is little
-information available on how well these forecasts work. Most biodiversity
-forecasts are not evaluated in terms of their ability to predict future
-diversity, fail to account for uncertainty, and do not use time-series data that
-captures the actual dynamics being studied. We compiled a list of ten best
-practices for forecasting biodiversity and used hindcasting to evaluate six
-different modeling approaches for predicting breeding bird species richness.
-Each method was evaluated at `r numbers$N_predict_sites` sites distributed
-throughout the continental United States. Forecasts were evaluated for up to a
-decade after the final training observations. While all of the models that we
-evaluated explained most of the variance in richness, none of them consistently
-outperforms a simple baseline model where each site's species richness
-fluctuates randomly around a constant value. At present, our best forecast for
-future richness is thus to assume that it is unchanging, and current modeling
-approaches do not reliably improve on this simple baseline. Most of the best
+evaluating how well current models characterize natural systems. While
+biodiversity forecasts are common in the literature, there is little information
+available on how well these forecasts work. Most biodiversity forecasts are not
+evaluated in terms of their ability to predict future diversity, fail to account
+for uncertainty, and do not use time-series data that captures the actual
+dynamics being studied. We compiled a list of ten best practices for forecasting
+biodiversity, and used hindcasting to evaluate six different modeling approaches
+for predicting breeding bird species richness. Each method was evaluated at `r
+numbers$N_predict_sites` sites distributed throughout the continental United
+States. Forecasts were evaluated for up to a decade after the final training
+observations. While each model could explain most of the variance in richness,
+none of them consistently outperformed a baseline model that simply predicted
+constant richness at each site. In particular, we found no evidence that current
+methods (such as species distribution models) can turn spatial data into useful
+temporal predictions about biodiversity. This result may call into question the
+validity of many high-profile forecasts from the last 15 years. Most of the best
 practices implemented in this study (e.g., evaluating confidence intervals as
@@
 even led to large changes in the relative performance of different modeling
-approaches. We emphasize the importance of following specific best practices
-when making ecological forecasts, and of robust evaluation of model predictions
-to facilitate rapid improvement in our biodiversity forecasts.
+approaches. To facilitate the rapid improvement of biodiversity forecasts, we
+emphasize the value of specific best practices in making and evaluating
+different forecasting methods.
 

### 2017-08-15 manuscript/draft.Rmd
Anna comments

@@
     - \usepackage{setspace}
+    - \usepackage{booktabs}
     - \doublespacing
@@
 evaluating how well current models characterize natural systems. While
-biodiversity forecasts are common in the literature, there is little information
-available on how well these forecasts work. Most biodiversity forecasts are not
-evaluated in terms of their ability to predict future diversity, fail to account
-for uncertainty, and do not use time-series data that captures the actual
-dynamics being studied. We compiled a list of ten best practices for forecasting
+biodiversity forecasts are common, there is little information available on how
+well these forecasts work. Most biodiversity forecasts are not evaluated in
+terms of their ability to predict future diversity, fail to account for
+uncertainty, and do not use time-series data that captures the actual dynamics
+being studied. We compiled a list of ten best practices for forecasting
 biodiversity, and used hindcasting to evaluate six different modeling approaches
@@
 constant richness at each site. In particular, we found no evidence that current
-methods (such as species distribution models) can turn spatial data into useful
-temporal predictions about biodiversity. This result may call into question the
-validity of many high-profile forecasts from the last 15 years. Most of the best
-practices implemented in this study (e.g., evaluating confidence intervals as
-well as point estimates) directly influence the outcome of the forecasts; some
-even led to large changes in the relative performance of different modeling
-approaches. To facilitate the rapid improvement of biodiversity forecasts, we
-emphasize the value of specific best practices in making and evaluating
-different forecasting methods.
+methods (such as species distribution models) can successfully turn spatial data
+into useful temporal predictions about biodiversity. This result may call into
+question the validity of many high-profile forecasts from the last 15 years.
+Most of the best practices implemented in this study (e.g., evaluating
+confidence intervals as well as point estimates) directly influence the outcome
+of the forecasts; some even led to large changes in the relative performance of
+different modeling approaches. To facilitate the rapid improvement of
+biodiversity forecasts, we emphasize the value of specific best practices in
+making and evaluating different forecasting methods.
 
@@
 Forecasting the future state of ecological systems is increasingly important for
-planning, managing, and evaluating how well ecological models capture the key
-ecological processes governing a system [@clark2001; @houlahan2017;
-@dietze2017]. Forecasting is biodiversity is especially important, due to its
-central role in conservation planning and its sensitivity to anthropogenic
-effects [@cardinale2012; @diaz2015; @tilman2017]. Anticipating potential changes
-in biodiversity is crucial for large scale management and conservation, and for
-addressing debates regarding whether local scale diversity is declining and may
-continue to decline in response to anthropogenic influences [@diaz2015;
-@dornelas2014].
+planning and management, and also for quantitatively evaluating how well our
+models capture the key ecological processes governing these systems [@clark2001;
+@houlahan2017; @dietze2017]. Forecasts regarding biodiversity are especially
+important, due to biodiversity's central role in conservation planning and its
+sensitivity to anthropogenic effects [@cardinale2012; @diaz2015; @tilman2017].
+High-profile studies forecasting large biodiversity declines over the coming
+decades have played a large role in shaping ecologists' priorities [as well as
+those of policymakers; @ipcc2014], but it is inherently difficult to evaluate
+such long-term predictions before the projected biodiversity declines have
+occurred.
 
@@
 Despite the emerging interest in forecasting species richness and other aspects
-of biodiversity, little is known about how effectively we can anticipate their
-dynamics. This is due in part to the long time scales over which many ecological
-forecasts are applied [and the resulting difficulty in assessing whether the
-predicted changes occurred @dietze2016]. What we do know comes from a small
-number of hindcasting studies, where models are built using data on species
-occurrence and richness from the past and evaluated on their ability to predict
-contemporary patterns [e.g., @algar2009; @distler2015]. These studies are a
-valuable first step, but lack several components that are important for
-developing good forecasting models and understanding how accurately these models
-can predict the future. These components of good forecasting and evaluation (Box
-1) broadly involve: 1) expanding the use of data to include biological and
-environmental time-series [@tredennick2016]; 2) accounting for uncertainty in
-observations, processes, model choice, and forecast evaluation [@clark2001,
-@dietze2017]; and 3) conducting meaningful evaluations of the forecasts by
-hindcasting, archiving short-term forecasts, and comparing forecasts to
-baselines to determine whether the forecasts are more accurate than simply
-assuming the system is basically static [@perretti2013].
+of biodiversity [[refs]], little is known about how effectively we can
+anticipate these dynamics. This is due in part to the long time scales over
+which many ecological forecasts are applied [and the resulting difficulty in
+assessing whether the predicted changes occurred; @dietze2016]. What we do know
+comes from a small number of hindcasting studies, where models are built using
+data on species occurrence and richness from the past and evaluated on their
+ability to predict contemporary patterns [e.g., @algar2009; @distler2015]. These
+studies are a valuable first step, but lack several components that are
+important for developing forecasting models with high predictive accuracy, and
+for understanding how well different methods can predict the future. The
+components an effective forecasting and evaluation strategy (Box 1) broadly
+involve: 1) expanding the use of data to include biological and environmental
+time-series [@tredennick2016]; 2) accounting for uncertainty in observations,
+processes, model choice, and forecast evaluation [@clark2001; @dietze2017]; and
+3) conducting meaningful evaluations of the forecasts by hindcasting, archiving
+short-term forecasts, and comparing forecasts to baselines to determine whether
+the forecasts are more accurate than simply assuming the system is basically
+static [@perretti2013].
 
@@
 In this paper, we attempt to forecast the species richness of breeding birds at
-over 1,000 of locations throughout North America while following the best
+over 1,000 of sites located throughout North America, while following the best
 practices in Box 1 for ecological forecasting. To do this, we combine 30 years
 of time-series data on bird distributions from annual surveys with monthly
-time-series of climate data and satellite-based remote-sensing. This 30 year
-time-scale has only recently become possible for large-scale time-series based
-forecasting and allows us to model and assess changes a decade or more into the
-future in the presence of shifts in environmental conditions on par with
-predicted climate change. We compare traditional distribution modeling based
-approaches to spatial models of species richness, time-series methods, and
-simple baselines (Table 1). All forecasts model uncertainty and observation
-error, are evaluated across different time lags using hindcasting, and are
-publicly archived to allow future assessment. We discuss the implications of
-these practices for our understanding of, and confidence in, the resulting
-forecasts, and how we can continue to build on these approaches improve
-ecological forecasting in the future.
+time-series of climate data and satellite-based remote-sensing. Datasets that
+span a time scale of 30 years or more have only recently become available for
+large-scale time-series based forecasting. A dataset of this size allows us to
+model and assess changes a decade or more into the future in the presence of
+shifts in environmental conditions on par with predicted climate change. We
+compare traditional distribution modeling based approaches to spatial models of
+species richness, time-series methods, and two simple baselines that predict
+constant richness for each site, on average (Table 1). All of our forecasting
+models account for uncertainty and observation error, are evaluated across
+different time lags using hindcasting, and are publicly archived to allow future
+assessment. We discuss the implications of these practices for our understanding
+of, and confidence in, the resulting forecasts, and how we can continue to build
+on these approaches to improve ecological forecasting in the future.
 
@@
 
+We evaluated 6 types of forecasting models (Table 1) at two time-scales. In the
+first one, we divided our [30] years of data into [22] years of training data
+and retained [9] years of data for evaluation via hindcasting. Second, we used
+the full data set for training and made predictions through the year 2050. For
+both time scales, we evaluated versions of each model with and without
+correcting for observer effects, as described below.
+
 ### Data
 
-Bird species richness was obtained from the North American Breeding Bird Survey
-(BBS) [@pardieck2017] using the Data Retriever software [@morris2013] and
-rdataretriever R package [@mcglinn2017]. The BBS data was filtered to exclude
-all nocturnal, cepuscular, and aquatic species [since these species are not well
-sampled by BBS methods; @hurlbert2005], as well as unidentified species, and
-hybrids. All data from routes that did not meet BBS acceptable criteria were
-also excluded.
+**Richness data.** Bird species richness was obtained from the North American
+Breeding Bird Survey (BBS) [@pardieck2017] using the Data Retriever software
+[@morris2013] and rdataretriever R package [@mcglinn2017]. The BBS data was
+filtered to exclude all nocturnal, cepuscular, and aquatic species [since these
+species are not well sampled by BBS methods; @hurlbert2005], as well as
+unidentified species, and hybrids. All data from routes that did not meet BBS
+acceptable criteria were also excluded.
 
-Environmental data included a combination of elevation, bioclimatic variables
-and a remotely sensed vegetation index (the normalized difference vegetation
-index; NDVI), all of which are known to influence richness and distribution in
-the BBS data. We used the 4 km resolution PRISM data [@prism] to calculate eight
-bioclimatic variables identified as relevant to bird distributions [@harris2015]
-for each year in the dataset: mean diurnal range, isothermality, max temperature
-of the warmest month, mean temperature of the wettest quarter, mean temperature
-of the driest quarter, precipitation seasonality, precipitation of the wettest
-quarter, precipitation of the warmest quarter. Satellite-derived NDVI, a primary
-correlate of richness in BBS data [@hurlbert2002], with an 8 km resolution was
-obtained from the NDIV3g dataset [@pinzon2014] and was available from the time
+**Past environmental data.** Environmental data included a combination of
+elevation, bioclimatic variables and a remotely sensed vegetation index (the
+normalized difference vegetation index; NDVI), all of which are known to
+influence richness and distribution in the BBS data [[ref]]. For each year in
+the dataset, we used the 4 km resolution PRISM data [@prism] to calculate eight
+bioclimatic variables identified as relevant to bird distributions
+[@harris2015]: mean diurnal range, isothermality, max temperature of the warmest
+month, mean temperature of the wettest quarter, mean temperature of the driest
+quarter, precipitation seasonality, precipitation of the wettest quarter, and
+precipitation of the warmest quarter. Satellite-derived NDVI, a primary
+correlate of richness in BBS data [@hurlbert2002], was obtained from the NDIV3g
+dataset with an 8 km resolution [@pinzon2014] and was available from the time
 period 1981-2013. Average summer (May, June, July) and winter (December,
@@
 SRTM 90m elevation dataset [@jarvis2008] obtained using the R package raster
-[@hijmans2016]. All environmental variables were aggregated to the average
-values within a 40 km radius (the length of a BBS route) of the starting point
-of each BBS route.
+[@hijmans2016]. Because BBS routes are 40-km transects rather than point counts,
+we used the average value of each environmental variable within a 40 km radius
+of each BBS route's starting point.
 
-We made long term forecasts from 2014-2050 using CMIP5 multi-model ensemble
-dataset as the source for climatic variables. Precipitation and temperature from
-37 downscaled model runs [@brekke2013, see Table S1] using the RCP6.0 scenario
-were averaged together to create a single ensemble used to calculate the
-bioclimatic variables for North America. For NDVI we used the per-site average
-values from 2000-2013 as a simple forecast. For observer effects (see below) each
-site was set to have zero observer bias.
+**Future environmental projections.** We made long term forecasts from 2014-2050
+using the CMIP5 multi-model ensemble dataset as the source for climatic
+variables [@brekke2013]. Precipitation and temperature from 37 downscaled model
+runs [@brekke2013, see Table S1] using the RCP6.0 scenario were averaged
+together to create a single ensemble used to calculate the bioclimatic variables
+for North America. For NDVI we used the per-site average values from 2000-2013
+as a simple forecast. For observer effects (see below) each site was set to have
+zero observer bias.
 
-We used observed richness values from 1982 (the first complete year of NDVI
-data) to 2003 for training the models, and data from 2004 to 2013 was used for
-testing their performance. We only used BBS routes from the continental United
-States (i.e. routes where PRISM data was available), and restricted the analysis
-to routes that were sampled during 70% of the years in the training period
-(i.e., routes with at least 16 annual observations). The resulting dataset
-included `r numbers$N_runs` transect runs of `r numbers$N_sites` unique sites,
-and included `r numbers$N_species` species. Site-level richness varied from `r
-numbers$richness_summary$min` to `r numbers$richness_summary$max` with an
-average richness of `r numbers$richness_summary$mean` species.
+**[[xxx]].** We used observed richness values from 1982 (the first complete year
+of NDVI data) to 2003 for training the models, and data from 2004 to 2013 was
+used for testing their performance. We only used BBS routes from the continental
+United States (i.e. routes where PRISM data was available), and restricted the
+analysis to routes that were sampled during 70% of the years in the training
+period (i.e., routes with at least 16 annual observations). The resulting
+dataset included `r numbers$N_runs` transect runs of `r numbers$N_sites` unique
+sites, and included `r numbers$N_species` species. Site-level richness varied
+from `r numbers$richness_summary$min` to `r numbers$richness_summary$max` with
+an average richness of `r numbers$richness_summary$mean` species.
 
@@
 
+For each forecasting approach, we trained two versions of the corresponding
+model: one with corrections for differences among observers, and one without.
 Observer effects are inherent in large data sets collected by different
@@
 from the model's posterior distribution, and fit each of the downstream models
-with each of the Monte Carlo samples.
+with each of the Monte Carlo samples. Each Monte Carlo sample represented a
+different possible set of observer-level and site-level random effect values
+across the full 30-year dataset.
 
@@
 
-### Site-level models
+### Models: site-level models
 
-We fit three sets of models that were fit to each site independently, with no
-environmental information. These models were fit to each BBS route both on the
-raw richness values and on the residuals from our observer model. When
-correcting for observer effects, we averaged across 500 models that were fit
-separately to the 500 Monte Carlo estimates of the observer effects, to account
-for our uncertainty in the true values of those effects. All of these models use
-a Gaussian error distribution (rather than a count distribution) for the reasons
-discussed below .
+We fit three sets of models that were fit to each site separately, with no
+environmental information (Table 1). These models were fit to each BBS route
+twice: once using the residuals from the observer model, and once using the raw
+richness values. When correcting for observer effects, we averaged across 500
+models that were fit separately to the 500 Monte Carlo estimates of the observer
+effects, to account for our uncertainty in the true values of those effects. All
+of these models use a Gaussian error distribution (rather than a count
+distribution) for the reasons discussed below (see "Model evaluation").
 
-**Baseline models.** We used two simple models as baselines for comparison to
-the fits of more complex models. These baselines treated site-level richness
-observations either as uncorrelated noise around a site-level constant (the
-"average" model) or as an autoregressive model with a single year of history
-(the "naive" model) [@hyndman2014]. Predictions from the "average" model are
-centered on the average richness observed during training, and the confidence
-intervals are narrow and constant-width. The "naive" model, in contrast, 
-predicts that future observations will be similar to the final observed value
-(e.g. the value observed in `r timeframe$last_train_year`), and the confidence
-intervals expand rapidly as the predictions extend farther into the future.
+**Baseline models.** We used two simple baseline models as a basis for
+comparison with the more complex models (Figure 2A). These baselines treated
+site-level richness observations either as uncorrelated noise around a
+site-level constant (the "average" model) or as an autoregressive model with a
+single year of history (the "naive" model) [@hyndman2014]. Predictions from the
+"average" model are centered on the average richness observed during training,
+and the confidence intervals are narrow and constant-width. The "naive" model,
+in contrast, predicts that future observations will be similar to the final
+observed value (e.g. the value observed in `r timeframe$last_train_year`), and
+the confidence intervals expand rapidly as the predictions extend farther into
+the future. Both models' richness predictions are centered on a constant value,
+so neither model can anticipate any trends in richness or any responses to
+future environmental changes.
 
@@
 
-### Continental-scale models
+### Models: continental-scale models
 

### 2017-08-15 manuscript/draft.Rmd
more anna comments

@@
 time-series, we did not include a season component in these models. Otherwise we
-used the default settings for this function.
+used the default settings for this function [@forecast1].
 
-### Models: continental-scale models
+[[Note: we sometimes refer to the baselines as "time series models", but (1)
+we're also using that name for just auto.arima and (2) the "average" model
+doesn't use any time series.  Other possible names: site-level models (as
+above); richness-only models; environment-free models...]]
+
+### Models: environmental models
 
@@
 elevation, and NDVI values should have identical richness distributions. As
-above, we included observer effects & the associated uncertainty by running
+above, we included observer effects and the associated uncertainty by running
 these models 500 times (once per MCMC sample).
 
-**"Macroecological" model.** We used a boosted regression tree model using the
-`gbm` package [@ridgeway2017] to directly model species richness as a function
-of environmental variables. Boosted regression trees are a form of tree based
-modeling that work by fitting thousands of small tree-structured models
-sequentially, with each tree optimized to reduce the error of its predecessors.
-They are flexible models that are considered well suited for prediction
-[@elith2008]. This model was optimized using a Gaussian likelihood, with a
-maximum interaction depth of 5, shrinkage of 0.015, and up to 10,000 trees. The
-number of trees used for prediction was selected using the "out of bag"
-estimator; this number averaged `r numbers$n_gbm[1]` for the non-observer data
-and `r numbers$n_gbm[2]` for the observer-corrected data.
+**"Macroecological" model: richness GBM.** We used a boosted regression tree
+model using the `gbm` package [@ridgeway2017] to directly model species richness
+as a function of environmental variables. Boosted regression trees are a form of
+tree-based modeling that work by fitting thousands of small tree-structured
+models sequentially, with each tree optimized to reduce the error of its
+predecessors. They are flexible models that are considered well suited for
+prediction [@elith2008]. This model was optimized using a Gaussian likelihood,
+with a maximum interaction depth of 5, shrinkage of 0.015, and up to 10,000
+trees. The number of trees used for prediction was selected using the "out of
+bag" estimator; this number averaged `r numbers$n_gbm[1]` for the non-observer
+data and `r numbers$n_gbm[2]` for the observer-corrected data. [[add comma separators for thousands place]]
 
-**Species Distribution Model.** Species distribution models (SDMs) predict
-individual species' occurrence probabilities using environmental variables.
-Species-level models are used to predict richness by summing the predicted
-probability of occupancy across all species at a site. This avoids known
-problems with the use of thresholds for determining whether or not a species
-will be present at a site [@calabrese2014; pellissier2013]. Following
-@calabrese2014, we calculated the uncertainty in our richness estimate by
-treating richness as a sum over independent Bernoulli random variables:
+**Species Distribution Model: stacked random forests.** Species distribution
+models (SDMs) predict individual species' occurrence probabilities using
+environmental variables. Species-level models are used to predict richness by
+summing the predicted probability of occupancy across all species at a site.
+This avoids known problems with the use of thresholds for determining whether or
+not a species will be present at a site [@calabrese2014; @pellissier2013].
+Following @calabrese2014, we calculated the uncertainty in our richness estimate
+by treating richness as a sum over independent Bernoulli random variables:
 $\sigma^2_{richness} = \sum_i{p_i (1 - p_i)}$, where $i$ indexes species. By
@@
 for each of the species-level probabilities instead of just one, propagating the
-uncertainty forward. We obtained these estimates using using random forests, a
-common approach in the species distribution modeling literature based on fitting
-hundreds of independent regression trees to randomly-perturbed versions of the
-data [@cutler2007; @caruana2008]. When correcting for observer effects, each of
-the 500 trees in our species-level random forests used a different Monte Carlo
-estimate of the observer effects as a predictor variable.
+uncertainty forward. We obtained these estimates using random forests, a common
+approach in the species distribution modeling literature. Random forests are
+constructed by fitting hundreds of independent regression trees to
+randomly-perturbed versions of the data [@cutler2007; @caruana2008]. When
+correcting for observer effects, each of the 500 trees in our species-level
+random forests used a different Monte Carlo estimate of the observer effects as
+a predictor variable.
 
-**Joint Species Distribution Models.** Joint species distribution models (JSDMs)
-are a new approach that makes predictions about the full composition of a
-community instead of modeling each species independently as above. JSDMs remove
-the assumed independence among species and explicitly account for the
+**Joint Species Distribution Model: mistnet.** Joint species distribution models
+(JSDMs) are a new approach that makes predictions about the full composition of
+a community instead of modeling each species independently as above. JSDMs
+remove the assumed independence among species and explicitly account for the
 possibility that a site will be much more (or less) suitable for birds in
@@
 
-Variance in species richness within sites was lower than predicted by several
-common count models, such as the Poisson or binomial (i.e. richness was
-underdispersed for individual sites), so these count models would have had
-difficulty fitting the data. Since richness had a relatively large mean 
-(`r numbers$richness_summary$mean`) and all models already produce continuous
-richness estimates (due to summing probabilities in the (J)SDM case), we defined
-model performance for all models in terms of continuous Gaussian errors instead
-of using discrete count distributions. When a model was run multiple times for
-the purpose of correcting for observer effects, we used the mean of those runs'
-point estimates as our final point estimate and we calculated the uncertainty
-using the law of total variance (i.e.  the average of the model runs' variance,
-plus the variance in the point estimates).
+We defined model performance for all models in terms of continuous Gaussian
+errors, instead of using discrete count distributions. Variance in species
+richness within sites was lower than predicted by several common count models,
+such as the Poisson or binomial (i.e. richness was underdispersed for individual
+sites), so these count models would have had difficulty fitting the data [cf.
+@calabrese2014]. The use of a continuous distribution is adequate here, since
+richness had a relatively large mean (`r numbers$richness_summary$mean`) and all
+models already produce continuous richness estimates (due to summing SDM or JSDM
+probabilities). When a model was run multiple times for the purpose of
+correcting for observer effects, we used the mean of those runs' point estimates
+as our final point estimate and we calculated the uncertainty using the law of
+total variance (i.e. the average of the model runs' variance, plus the variance
+in the point estimates).
 
-We evaluated each model's for forecasts using the data for each year between `r
+We evaluated each model's forecasts using the data for each year between `r
 timeframe$last_train_year` and `r timeframe$end_yr`. We used three metrics for
@@
 log-likelihood) as an integrative measure of fit incorporating good point
-estimates, precision, and coverage.  In addition to looking at forecast
-performance in general we evaluated how performance changed as the time horizon
-of forecasting increased by plotting performance metrics against the forecast
-horizon. Finally, we decomposed each model's squared error into two components:
-one described the squared error associated with site-level means, while the
-other described the squared error associated with annual fluctuations in
-richness within a site. This decomposition tells us the extent to which each
-model's error depends on consistent differences among sites versus changes in
-site-level richness from year to year.
+estimates, precision, and coverage.  In addition to evaluating forecast
+performance in general, we evaluated how performance changed as the time horizon
+of forecasting increased by plotting performance metrics against year. Finally,
+we decomposed each model's squared error into two components: one described the
+squared error associated with site-level means, while the other described the
+squared error associated with annual fluctuations in richness within a site.
+This decomposition tells us the extent to which each model's error depends on
+consistent differences among sites versus changes in site-level richness from
+year to year.
 
@@
 is available on GitHub (https://github.com/weecology/bbs-forecasting) and
-archived on Zenodo ().
+archived on Zenodo (https://doi.org/10.5281/zenodo.839581).
 
@@
 When comparing forecasts for richness across sites all methods performed well
-(Figure 4; all $R^2 > 0.5$). However SDMs (both stacked and joint) and the
+(Figure 3; all $R^2 > 0.5$). However SDMs (both stacked and joint) and the
 macroecological model all failed to successfully forecast the highest-richness
-sites resulting in a notable clustering of predicted values near ~60 species and
-the poorest performance ($R^2$=`r numbers$env_R2s`, versus $R^2$=`r numbers$ts_R2s` 
-for the within-site methods).
+sites, resulting in a notable clustering of predicted values near ~60 species
+and the poorest model performance ($R^2$=`r numbers$env_R2s`, versus $R^2$=`r
+numbers$ts_R2s` for the within-site methods).
 
@@
 future (2004) or ten years into the future (2013).  In general, the time series
-models (**A.**) outperformed the environmental models (**B.**), and the accuracy
-of the predictions declined as the forecast horizon was
+models (**A.**) outperformed the environmental models. (**B.**) The accuracy of
+the predictions generally declined as the forecast horizon was
 extended.](../figures/scatter.png)
 
-While the models generally did well in absolute terms (Figure 4), none
-consistently outperformed the "average" baseline from the generalized linear
-mixed model (Figure 4). The auto-ARIMA was generally the
-best-performing non-baseline model, but in many cases (67% of the time), the
-auto.arima procedure automatically selected a model with no autoregressive,
-drift, or moving average terms, making it similar to the "Average" model.
-All five alternatives to the "average" model achieved lower error on some
-of the sites in some years, but each one had a higher mean absolute error
-and higher mean deviance.
+While the models generally did well in absolute terms (Figure 3), none
+consistently outperformed the "average" baseline (Figure 4). The auto-ARIMA was
+generally the best-performing non-baseline model, but in many cases (67% of the
+time), the auto.arima procedure automatically selected a model with only an
+intercept term (i.e. no autoregressive terms, no drift, and no moving average
+terms), making it similar to the "average" model. All five alternatives to the
+"average" model achieved lower error on some of the sites in some years, but
+each one had a higher mean absolute error and higher mean deviance (Figure 4).
 
@@
 small number of forecasts where observer turnover caused a large shift in the
-reported richness values. The Naive model was less sensitive to these shifts,
+reported richness values. The naive baseline was less sensitive to these shifts,
 because it largely ignored the richness values reported by observers that had
-retired by the end of the training period (Figure 1). The Average model, which
+retired by the end of the training period (Figure 1). The average model, which
 gave equal weight to observations from the whole training period, showed a
-larger decline in performance -- especially in terms of coverage. The mistnet
-JSDM was notable here because the coverage of its prediction intervals retained
-very good coverage in the presence of this observer-based noise, which we
-attribute to its ability to model this variation with its latent variables.
+larger decline in performance -- especially in terms of coverage. The
+performance of the mistnet JSDM was notable here, because its prediction
+intervals retained very good coverage in the presence of uncorrected
+observer-based noise, which we attribute to the JSDM's ability to model this
+variation with its latent variables.
 
@@
 American Breeding Bird Survey, NASA, the World Climate Research Programme's
-Working Group on Coupled Modelling and its working groups, the
-U.S. Department of Energy's Program for Climate Model Diagnosis and
-Intercomparison, and the Global Organization for Earth System Science Portals.
+Working Group on Coupled Modelling and its working groups, the U.S. Department
+of Energy's Program for Climate Model Diagnosis and Intercomparison, and the
+Global Organization for Earth System Science Portals. A. C. Perry provided 
+valuable comments that improved the clarity of this manuscript.
 

### 2017-08-15 manuscript/draft.Rmd
more Anna

@@
 validated in terms of repeated observations at the same site. Since site-level
-richness is relatively stable, model evaluations that show high accuracy across
-spatial gradients may not indicate much about a model's ability to predict
-changes over time. This issue has also recently been highlighted as an issue for
-using SDMs to predict the future locations of individual species
-[@rapacciuolo2012; @oedekoven2017]. This result is particularly sobering because
-SDMs form the main foundation for estimates of the predicted loss of
+richness is relatively stable [[refs]], model evaluations that show high
+accuracy across spatial gradients may not indicate much about a model's ability
+to predict changes over time. This issue has also recently been highlighted as a
+potential downside of using SDMs to predict the future locations of individual
+species [@rapacciuolo2012; @oedekoven2017]. This result is particularly sobering
+because SDMs form the main foundation for estimates of the predicted loss of
 biodiversity to climate change [@thomas2004; @urban2015].
@@
 improvement in the forecast relative to a simple baseline [@jolliffe2003]. We
-recommend following Ye et al. [@ye2015] and adopting this approach in future
-ecological forecasting research. It also highlights the need for using
+recommend following the advice of @ye2015 and adopting this approach in future
+ecological forecasting research. Our results also highlight the need for using
 hindcasting and time-series data for assessing the effectiveness of forecasting
-methods, since this allows comparisons to baselines and assessment of predicted
-dynamics within sites or regions.
+methods. This practice allows comparisons to baseline models and also
+facilitates the assessment of predicted dynamics within sites or regions.
 
 One of the main alternatives to stacked SDMs is modeling richness directly using
-environmental variables and using the resulting models to make forecasts
-[@algar2009; @distler2015]. Two comparisons of the the SDM and direct richness
-modeling approaches reported that the methods yielded equivalent results for
-forecasting diversity. While our results generally support rough equivalence
-between stacked SDMs and direct richness modeling for point estimates, they also
-show that stacked SDMs dramatically understimate the uncertainty. For this
-reason, direct modeling of richness therefore is a better overall approach to
-forecasting richness. A similar result is seen when comparing joint species
-distribution models (JSDMs) to stacked single species distribution models. For
-point estimates the joint distribution models are roughly equivalent to stacked
-SDMs, but the JSDMs provide much better estimates of uncertainty [@harris2015;
-others?]. In fact, JSDMs and direct richness modeling provide some of the best
-estimates of uncertainty across all modeling approaches. This highlights the
-importance of evaluating models based on their uncertainty as well as their mean
-estimates (Box 1). Not doing so is particularly problematic in this case because
-the summed species distribution models are overly confident meaning that using
-them would lead to believing that richness would be restricted to a much
-narrower range than would actually occur.
+environmental variables and using the resulting models to make forecasts. Two
+studied compared the SDM and direct richness modeling approaches, and reported
+that the methods yielded equivalent results for forecasting diversity
+[@algar2009; @distler2015]. While our results generally support the idea of
+rough equivalence between stacked SDMs and direct richness modeling for point
+estimates, they also show that stacked SDMs dramatically understimate the
+uncertainty. For this reason, direct modeling of richness is a better overall
+approach to forecasting richness. A similar result is seen when comparing joint
+species distribution models (JSDMs) to stacked single species distribution
+models. For point estimates, the joint distribution models are roughly
+equivalent to stacked SDMs, but the JSDMs provide much better estimates of
+uncertainty [@harris2015; others?]. In fact, JSDMs and direct richness modeling
+provide some of the best estimates of uncertainty across all modeling
+approaches. This highlights the importance of evaluating models based on their
+uncertainty as well as their mean estimates (Box 1). Not doing so is
+particularly problematic in this case, because the summed species distribution
+models are overly confident, which means that these models could mislead
+ecologists into believing that richness would be restricted to a much narrower
+range than would actually occur.
 
 It is also possible that models that include species' relationships to their
-environments or direct environmental constraints on richness may continue to
+environments or direct environmental constraints on richness will continue to
 produce forecasts that are worse than simply assuming the systems are static.
 This would be expected to occur if richness in these systems is not changing
-over the relevant time scales and therefore simple models of no directional
-change are most appropriate. Recent suggestions that local scale richness in
-some systems is not changing directionally at multi-decadal scales supports this
-possibility [@brown2001; @ernest2001; @dornelas2014; @vellend2013]. This lack of
-change may be expected even in the presence of substantial changes in
-environmental conditions and species composition at a site due to the
-replacement of species from the regional pool [@brown2001; @ernest2001]. On
-average the Breeding Bird Survey sites used in this study show little change in
-richness (site-level SD of `r numbers$resid_sd` species, after controlling for
-differences among observers) [see also @lasorte2005]. This absence of rapid
-change is beneficial for the absolute accuracy of forecasts across sites,
-because if a past year's richness is known it is easy to estimate a future
-richness, and explains why stacked SDMs perform relatively well at this task
-despite failing to capture meaningful dynamics. However it makes it difficult to
-improve forecasts relative to simple baselines, since those baselines are close
-to representing what is actually occurring in the system. This suggests that
-simple time-series models and baselines should be actively considered for
-forecasts of richness and other stable aspects of biodiversity and also suggests
-that future efforts to understand and forecast biodiversity should also focus on
-composition since it is expected to be more dynamic [@ernest2001;
-@dornelas2014].
+over the relevant time scales (e.g. 10-30 years in this study), which would make
+simpler models with no directional change more appropriate. Recent suggestions
+that local scale richness in some systems is not changing directionally at
+multi-decadal scales supports this possibility [@brown2001; @ernest2001;
+@dornelas2014; @vellend2013]. This lack of change in richness may be expected
+even in the presence of substantial changes in environmental conditions and
+species composition at a site due to the replacement of species from the
+regional pool [@brown2001; @ernest2001]. On average, the Breeding Bird Survey
+sites used in this study show little change in richness [site-level SD of `r
+numbers$resid_sd` species, after controlling for differences among observers;
+see also @lasorte2005]. The absence of rapid change in this dataset is
+beneficial for the absolute accuracy of forecasts across different sites: when a
+past year's richness is already known, it is easy to estimate a future richness.
+The site-level stability of the BBS data also explains why stacked SDMs perform
+relatively well at predicting future richness, despite failing to capture
+changes in richness over time. However, this stability also makes it difficult
+to improve forecasts relative to simple baselines, since those baselines are
+already close to representing what is actually occurring in the system. These
+results suggest that simple time-series models and baselines should be actively
+considered for forecasts of richness and other stable aspects of biodiversity.
+Our results also suggest that future efforts to understand and forecast
+biodiversity should also focus on species composition, since lower-level
+processes are expected to be more dynamic [@ernest2001; @dornelas2014] and
+contain more useful information [@harris2015].
 
-In addition to consideration of the different process models used for
-forecasting it is important to consider the observation models. When working
-with any ecological dataset there are imperfections in sampling that have the
-potential to influence results. With large scale survey and citizen science
-datasets like the Breeding Bird Survey these issues are potentially magnified by
-the large number of different observers and major differences in habitat and
-species [@sauer1994]. We included an observation model for one of the two major
-observation issues known for the Breeding Bird Survey, differences among
-observers. Accounting for differences in observers reduced the average error in
-point estimates and also improved the coverage of the confidence intervals. In
-addition, it resulted in changes in which models performed best, most notably
-reducing the relative performance of the naive model for point estimates. This
-suggests that the naive model performed well in part because it was capable of
-accommodating rapid shifts in estimated richness introduced by changes in the
-observer. These kinds of rapid changes were difficult for the other time-series
-models to accommodate and so the average performance of the ARIMA and average
-models improved once this source of observation error was addressed. This
-demonstrates that properly modeling observation error can be important for
-properly estimating and reducing uncertainty in forecasts and can also lead to
-changes in the best methods for forecasting. We did not address differences in
-detection probability across species and sites [@boulinier1998] since there is
-no clear way to address this issue without making strong assumptions about the
-data (i.e., assuming there is no biological variation in stops along a route),
-but this would be a valuable addition to future forecasting models.
+**Observer effects.** In addition to consideration of the different process
+models used for forecasting it is important to consider the observation models.
+When working with any ecological dataset, there are imperfections in the
+sampling process that have the potential to influence results. With large scale
+surveys and citizen science datasets, such as the Breeding Bird Survey, these
+issues are potentially magnified by the large number of different observers and
+by major differences in the habitats and species being surveyed [@sauer1994].
+Accounting for differences in observers reduced the average error in our point
+estimates and also improved the coverage of the confidence intervals. In
+addition, controlling for observer effects resulted in changes in which models
+performed best, most notably reducing the relative performance of the naive
+model's point estimates. This suggests that, prior to accounting for observer
+effects, the naive model performed well because it was capable of accommodating
+rapid shifts in estimated richness introduced by changes in the observer. These
+kinds of rapid changes were difficult for the other time-series models to
+accommodate, and so the average performance of the ARIMA and average models
+improved once this source of observation error was addressed. This demonstrates
+that modeling observation error can be important for properly estimating and
+reducing uncertainty in forecasts and can also lead to changes in the best
+methods for forecasting. We did not address differences in detection probability
+across species and sites [@boulinier1998] since there is no clear way to address
+this issue without making strong assumptions about the data (i.e., assuming
+there is no biological variation in stops along a route), but this would be a
+valuable addition to future forecasting models.
 
-Future efforts will also need to begin to address the additional uncertainty
-that comes from error in forecasting the environmental conditions themselves. In
-this and other hindcasting studies the environmental conditions for the "future"
-are known because the data has already been observed. In real forecasts the
-environmental conditions themselves have to be forecast and those forecasts have
-uncertainty and bias. This will cause ecological forecasts that use
-environmental data to be more uncertain than observed in current hindcasting
-efforts. It is important to correctly incorporate this uncertainty in the
-predictor variables into forecasting models [@clark2001; @dietze2017]. Difficulty
-in forecasting future environmental conditions at small scales will present
-continued challenges for models incorporating these conditions and this may
-result in a continued advantage to simple time-series based approaches.
+**Environmental uncertainty.** Future biodiversity forecasting efforts will also
+need to begin to address the additional uncertainty introduced by the error
+inherent in the process of forecasting the environmental conditions that
+ecologists use as predictor variables. In this, and other hindcasting studies,
+the environmental conditions for the "future" are known because the data has
+already been observed. However, in real forecasts the environmental conditions
+themselves have to be predicted, and environmental forecasts will also have
+uncertainty and bias. Ultimately, ecological forecasts that use environmental
+data will therefore be more uncertain than our current hindcasting efforts.
+Thus, it is important to correctly incorporate this uncertainty in the predictor
+variables into forecasting models [@clark2001; @dietze2017], as we did with the
+uncertainty in observer-level effects. Difficulty in forecasting future
+environmental conditions at small scales will present continued challenges for
+models incorporating environmental variables, and this may result in a continued
+advantage for simple time-series based approaches.
 
-Another challenge for ecological forecasting is the timescales of typical
-forecasts. We evaluated forecasts using yearly timesteps up to a decade into the
-future. In contrast, many ecological forecasts are temporally aggregated to 5-30
-year timesteps and projected up to a century into the future. These are commonly
-beyond the career or even lifespan of the researchers, making reasonable model
-assessment impossible. Currently assessments at these long time-scales can
-only be made with a small number of opportune datasets. For examples hindcasts
-have been used to assess models of species richness [@algar2009, @distler2015]
-and distribution [@rapacciuolo2012, @moran-ordonez2017, @araujo2005,
-@eskildsen2013] using data aggregated across similar time scales. More
-evaluation studies like these are needed but there is a paucity of adequate long
-term data available. Short term forecasts will allow for relatively rapid
-validation of models by researchers without needing to wait for enough data to
-accumulate [@tredennick2016; @dietze2016]. This does not mean that they should
-replace large scale long-term forecasts. As discussed above, drivers of species
-richness are likely to differ at different temporal scales [@rosenzweig1995,
-@white2004, @white2007]. Thus short-term forecasts should be expected to add
-value to end users and inform, not replace, long-term forecasts. Hopefully by
-evaluating trends in the shorter term forecasts as done here it will be
-possible to also use the information from short-term forecasting to improve
-longer-term estimates.
+**Multi-timescale forecasts.** Another challenge for ecological forecasting is
+the timescales of typical forecasts. We evaluated forecasts using yearly
+timesteps up to a decade into the future. In contrast, many ecological forecasts
+are temporally aggregated to 5-30 year timesteps and projected up to a century
+into the future. These are commonly beyond the career or even lifespan of the
+researchers, which makes it impossible to directly validate the models'
+predictions. Currently, assessments at these long time-scales can only be made
+with a small number of opportune datasets. For example, hindcasts have been used
+to assess models of species richness [@algar2009; @distler2015] and distribution
+[@rapacciuolo2012; @moran-ordonez2017; @araujo2005; @eskildsen2013] using data
+aggregated across similar time scales to those studied here. More evaluation
+studies like these are needed but there is a paucity of adequate long term data
+available. Short term forecasts will allow for relatively rapid validation of
+models by researchers without needing to wait for enough data to accumulate
+[@tredennick2016; @dietze2016]. This does not mean that they should replace
+large scale long-term forecasts. As discussed above, drivers of species richness
+are likely to differ at different temporal scales [@rosenzweig1995; @white2004;
+@white2007]. Thus, short-term forecasts should be expected to add value to
+scientists and policymakers, and inform (not replace) long-term forecasts.
+Hopefully, by evaluating trends in the shorter term forecasts as done here it
+will be possible to also use the information from short-term forecasting to
+improve longer-term estimates.
 
-The science of forecasting biodiversity, and ecology more broadly, remains in
-its infancy and it is important to consider the general inability of
-forecasting methods to improve on simple baselines in that context. When weather
-forecasting first started the forecasts were likewise worse than simple
-baselines [@mcgill2012]. One of the things that helped weather forecasts improve
-was large numbers of forecasts were made in public, which allowed different
-approaches to forecasting to be regularly assessed and improved [@mcgill2012;
-@silver2012]. This suggests that it is important for ecologists to start
-regularly making and evaluating true ecological forecasts, even if they perform
-poorly, and to make these forecasts publicly available for assessment. These
-forecasts should include both short-term predictions, which can be assessed
-quickly, and mid to long-term forecasts to help assess long time-scale processes
-and determine how far into the future we can successfully forecast
+**Conclusions.** The science of forecasting biodiversity, and ecology more
+broadly, remains in its infancy and it is important to consider the general
+inability of forecasting methods to improve on simple baselines in that context.
+When weather forecasting first started the forecasts were likewise worse than
+simple baselines [@mcgill2012]. One of the things that helped weather forecasts
+improve was large numbers of forecasts were made in public, which allowed
+different approaches to forecasting to be regularly assessed and improved
+[@mcgill2012; @silver2012]. This suggests that it is important for ecologists to
+start regularly making and evaluating true ecological forecasts, even if they
+perform poorly, and to make these forecasts publicly available for assessment.
+These forecasts should include both short-term predictions, which can be
+assessed quickly, and mid to long-term forecasts to help assess long time-scale
+processes and determine how far into the future we can successfully forecast
 [@tredennick2016; @dietze2017]. Forecasts based on the models in this paper from

### 2017-08-15 manuscript/draft.Rmd
minor

@@
 biodiversity, and used hindcasting to evaluate six different modeling approaches
-for predicting breeding bird species richness. Each method was evaluated at `r
-numbers$N_predict_sites` sites distributed throughout the continental United
+for predicting breeding bird species richness. Each method was evaluated at `r numbers$N_predict_sites` sites distributed throughout the continental United
 States. Forecasts were evaluated for up to a decade after the final training
@@
 
-table_caption = "A summary of the six models evaluated here. The time series
+table_caption = "A summary of the six models evaluated here. The single-site
 models were trained site-by-site, without access to any environmental
@@
   add_header_above(c(" " = 2, "Predictors" = 3)) %>% 
-  group_rows("Time series models", 1, 3) %>% 
+  group_rows("Single-site models", 1, 3) %>% 
   group_rows("Environmental models", 4, 6) %>% 
@@
 ribbon indicates 1 standard deviation of uncertainty, and the lighter ribbon
-indicates 95% confidence intervals. **A.** The time series models models were
+indicates 95% confidence intervals. **A.** The single-site models models were
 trained using one site at a time, based solely on observed richness values
@@
 forecasting baselines. **B.** The environmental models were trained to predict
-richness based on elevation, climate, and NDVI; unlike the time series models,
+richness based on elevation, climate, and NDVI; unlike the single-site models,
 the environmental models' predictions change from year to year as environmental
@@
 acceptable criteria were also excluded.
+
+We used observed richness values from 1982 (the first year of complete
+environmental data) to 2003 for training the models, and data from 2004 to 2013
+was used for testing their performance. We only used BBS routes from the
+continental United States (i.e. routes where climate data was available @prism),
+and we restricted the analysis to routes that were sampled during 70% of the
+years in the training period (i.e., routes with at least 16 annual
+observations). The resulting dataset included `r numbers$N_runs` transect runs
+of `r numbers$N_sites` unique sites, and included `r numbers$N_species` species.
+Site-level richness varied from `r numbers$richness_summary$min` to `r
+numbers$richness_summary$max` with an average richness of `r
+numbers$richness_summary$mean` species.
 
@@
 zero observer bias.
-
-**[[xxx]].** We used observed richness values from 1982 (the first complete year
-of NDVI data) to 2003 for training the models, and data from 2004 to 2013 was
-used for testing their performance. We only used BBS routes from the continental
-United States (i.e. routes where PRISM data was available), and restricted the
-analysis to routes that were sampled during 70% of the years in the training
-period (i.e., routes with at least 16 annual observations). The resulting
-dataset included `r numbers$N_runs` transect runs of `r numbers$N_sites` unique
-sites, and included `r numbers$N_species` species. Site-level richness varied
-from `r numbers$richness_summary$min` to `r numbers$richness_summary$max` with
-an average richness of `r numbers$richness_summary$mean` species.
 
@@
 
-[[Note: we sometimes refer to the baselines as "time series models", but (1)
-we're also using that name for just auto.arima and (2) the "average" model
-doesn't use any time series.  Other possible names: site-level models (as
-above); richness-only models; environment-free models...]]
-
 ### Models: environmental models
 
-In contrast to the site-level time-series models, most attempts to predict
+In contrast to the single-site models, most attempts to predict
 species richness focus on using correlative models based on environmental
@@
 
-We evaluated each model's forecasts using the data for each year between `r
-timeframe$last_train_year` and `r timeframe$end_yr`. We used three metrics for
+We evaluated each model's forecasts using the data for each year between `r timeframe$last_train_year` and `r timeframe$end_yr`. We used three metrics for
 evaluating performance: 1) root-mean-square error (RMSE) to determine how far,
@@
 sites, resulting in a notable clustering of predicted values near ~60 species
-and the poorest model performance ($R^2$=`r numbers$env_R2s`, versus $R^2$=`r
-numbers$ts_R2s` for the within-site methods).
+and the poorest model performance ($R^2$=`r numbers$env_R2s`, versus $R^2$=`r numbers$ts_R2s` for the within-site methods).
 
 ![Performance of the six models we evaluated when predicting one year into the
-future (2004) or ten years into the future (2013).  In general, the time series
+future (2004) or ten years into the future (2013).  In general, the single-site
 models (**A.**) outperformed the environmental models. (**B.**) The accuracy of
@@
 component is based on errors in predicting fluctuations in a site's richness
-over time. Both components of the mean squared error were lower for the time
-series models than for the environmental models.](../figures/barcharts.png)
+over time. Both components of the mean squared error were lower for the
+single-site models than for the environmental models.](../figures/barcharts.png)
 
@@
 regional pool [@brown2001; @ernest2001]. On average, the Breeding Bird Survey
-sites used in this study show little change in richness [site-level SD of `r
-numbers$resid_sd` species, after controlling for differences among observers;
+sites used in this study show little change in richness [site-level SD of `r numbers$resid_sd` species, after controlling for differences among observers;
 see also @lasorte2005]. The absence of rapid change in this dataset is
@@
 already close to representing what is actually occurring in the system. These
-results suggest that simple time-series models and baselines should be actively
-considered for forecasts of richness and other stable aspects of biodiversity.
-Our results also suggest that future efforts to understand and forecast
-biodiversity should also focus on species composition, since lower-level
-processes are expected to be more dynamic [@ernest2001; @dornelas2014] and
-contain more useful information [@harris2015].
+results suggest that single-site models should be actively considered for
+forecasts of richness and other stable aspects of biodiversity. Our results also
+suggest that future efforts to understand and forecast biodiversity should also
+focus on species composition, since lower-level processes are expected to be
+more dynamic [@ernest2001; @dornelas2014] and contain more useful information
+[@harris2015].
 
@@
 rapid shifts in estimated richness introduced by changes in the observer. These
-kinds of rapid changes were difficult for the other time-series models to
+kinds of rapid changes were difficult for the other single-site models to
 accommodate, and so the average performance of the ARIMA and average models
@@
 models incorporating environmental variables, and this may result in a continued
-advantage for simple time-series based approaches.
+advantage for simple single-site approaches.
 
@@
 early 20th century to model distributions which are validated with late
-20th century occurances. Dense timeseries, such as yearly observations, are
+20th century occurances. Dense time series, such as yearly observations, are
 desirable to also evalulate the forecast horizon (see #6), but this is not

### 2017-08-15 manuscript/draft.Rmd
stuff

@@
 biodiversity, and used hindcasting to evaluate six different modeling approaches
-for predicting breeding bird species richness. Each method was evaluated at `r numbers$N_predict_sites` sites distributed throughout the continental United
+for predicting breeding bird species richness. Each method was evaluated at `r
+numbers$N_predict_sites` sites distributed throughout the continental United
 States. Forecasts were evaluated for up to a decade after the final training
@@
 decades have played a large role in shaping ecologists' priorities [as well as
-those of policymakers; @ipcc2014], but it is inherently difficult to evaluate
-such long-term predictions before the projected biodiversity declines have
-occurred.
+those of policymakers; e.g. @ipcc2014], but it is inherently difficult to
+evaluate such long-term predictions before the projected biodiversity declines
+have occurred.
 
@@
   group_rows("Single-site models", 1, 3) %>% 
-  group_rows("Environmental models", 4, 6) %>% 
-  kable_styling(position = "center")
+  group_rows("Environmental models", 4, 6) %>% kable_styling(position = "center")
 ```
@@
 
-In contrast to the single-site models, most attempts to predict
-species richness focus on using correlative models based on environmental
-variables. We tested three common variants of this approach: direct modeling of
-species richness; stacking individual species distribution models; and joint
-species distribution models (JSDMs). Following the standard approach, site-level
-random effects were not included in these models as predictors, meaning that
-this approach implicitly assumes that two sites with identical Bioclim,
-elevation, and NDVI values should have identical richness distributions. As
-above, we included observer effects and the associated uncertainty by running
-these models 500 times (once per MCMC sample).
+In contrast to the single-site models, most attempts to predict species richness
+focus on using correlative models based on environmental variables. We tested
+three common variants of this approach: direct modeling of species richness;
+stacking individual species distribution models; and joint species distribution
+models (JSDMs). Following the standard approach, site-level random effects were
+not included in these models as predictors, meaning that this approach
+implicitly assumes that two sites with identical Bioclim, elevation, and NDVI
+values should have identical richness distributions. As above, we included
+observer effects and the associated uncertainty by running these models 500
+times (once per MCMC sample).
 
@@
 bag" estimator; this number averaged `r numbers$n_gbm[1]` for the non-observer
-data and `r numbers$n_gbm[2]` for the observer-corrected data. [[add comma separators for thousands place]]
+data and `r numbers$n_gbm[2]` for the observer-corrected data. [[add comma
+separators for thousands place]]
 
@@
 
-We evaluated each model's forecasts using the data for each year between `r timeframe$last_train_year` and `r timeframe$end_yr`. We used three metrics for
+We evaluated each model's forecasts using the data for each year between `r
+timeframe$last_train_year` and `r timeframe$end_yr`. We used three metrics for
 evaluating performance: 1) root-mean-square error (RMSE) to determine how far,
@@
 sites, resulting in a notable clustering of predicted values near ~60 species
-and the poorest model performance ($R^2$=`r numbers$env_R2s`, versus $R^2$=`r numbers$ts_R2s` for the within-site methods).
+and the poorest model performance ($R^2$=`r numbers$env_R2s`, versus $R^2$=`r
+numbers$ts_R2s` for the within-site methods).
 
@@
 generally similar or larger than "average" model, with large outliers in both
-directions. **B.** The other models' deviance was also generally higher than
-the "average" baseline.](../figures/model_violins.png)
+directions. **B.** The other models' deviance was also generally higher than the
+"average" baseline.](../figures/model_violins.png)
 
@@
 stated purpose of these forecasts. As a result, our baseline models, which did
-not attempt to anticipate changes in richness over time, generally provided
-the best forecasts for future biodiversity.
+not attempt to anticipate changes in richness over time, generally provided the
+best forecasts for future biodiversity.
 
@@
 regional pool [@brown2001; @ernest2001]. On average, the Breeding Bird Survey
-sites used in this study show little change in richness [site-level SD of `r numbers$resid_sd` species, after controlling for differences among observers;
+sites used in this study show little change in richness [site-level SD of `r
+numbers$resid_sd` species, after controlling for differences among observers;
 see also @lasorte2005]. The absence of rapid change in this dataset is
@@
 When weather forecasting first started the forecasts were likewise worse than
-simple baselines [@mcgill2012]. One of the things that helped weather forecasts
-improve was large numbers of forecasts were made in public, which allowed
-different approaches to forecasting to be regularly assessed and improved
+simple baselines [@mcgill2012]. One practice that helped weather forecasts to
+improve was publicly making a large number of predictions, which allowed
+different forecasting approaches to be regularly assessed and refined
 [@mcgill2012; @silver2012]. This suggests that it is important for ecologists to
-start regularly making and evaluating true ecological forecasts, even if they
-perform poorly, and to make these forecasts publicly available for assessment.
-These forecasts should include both short-term predictions, which can be
-assessed quickly, and mid to long-term forecasts to help assess long time-scale
-processes and determine how far into the future we can successfully forecast
-[@tredennick2016; @dietze2017]. Forecasts based on the models in this paper from
-now until 2050 are openly archived on Dryad so that we and others can assess how
-well they perform and we plan to evaluate these forecasts and report the results
-as each new year of BBS data becomes available. Weather forecasting has
-continually improved throughout its history [@bauer2015] in part due to making
-and evaluating public forecasts [@mcgill2012] and we hope this will help ecology
-do the same.
+start regularly making and evaluating ecological forecasts on real data, even if
+they perform poorly, and to make these forecasts publicly available for
+assessment. These forecasts should include both short-term predictions, which
+can be assessed quickly, and mid- to long-term forecasts, which can help
+ecologists to assess long time-scale processes and determine how far into the
+future we can successfully forecast [@tredennick2016; @dietze2017]. Forecasts
+based on the models in this paper (from now until 2050) are openly archived on
+Zenodo, so that we and others can assess how well they perform. We also plan to
+evaluate these forecasts and report the results as each new year of BBS data
+becomes available at the same URL [@zenodo]. Weather forecasting has continually
+improved throughout its history [@bauer2015], in part due to making and
+evaluating public forecasts [@mcgill2012], and we hope this practice will help
+ecology do the same.
 
@@
 systems, and we currently lack the large amounts of data that produce effective
-forecasts through machine learning. Despite this we believe that progress can be
+forecasts through machine learning. Despite this, we believe that progress can be
 made if we build an active forecasting culture in ecology that builds and
@@
 can learn more rapidly about the best approaches for making predictions for a
-given ecological quantity. This includes comparing process based and data-driven
-models and comparing the accuracy of forecasts to simple baselines to determine
-if the modeled forecasts are more accurate than the naive assumption that the
-world is static [@ye2015].
+given ecological quantity. This includes comparing process-based [e.g. [[refs]]]
+and data-driven models, as well as comparing the accuracy of forecasts to simple
+baselines to determine if the modeled forecasts are more accurate than the naive
+assumption that the world is static [@ye2015].
 
@@
 areas of ecological forecasting focus primarily on time-series data, others
-primarily focus on using spatial models and space-for time substitutions. Using
+primarily focus on using spatial models and space-for-time substitutions. Using
 ecological and environmental time-series data allows the consideration of actual
@@
 possible [@clark2001; @dietze2017]. Evaluations of forecasts should assess the
-accuracy of uncertainties as well as point estimates [@dietze2017].
+accuracy of models' estimated uncertainties as well as their point estimates
+[@dietze2017].
 
@@
 
-In general the accuracy of forecasts decreases with the length of time into the
-future being forecast [@petchey2015]. This decay in accuracy and the potential for
-different rates of decay to result in different relative model performance at
-different lead times should be considered when evaluating forecasts and
-comparing models.
+In general, the accuracy of forecasts decreases with the length of time into the
+future being forecast [@petchey2015]. This decay in accuracy (and the potential
+for different rates of decay to result in different relative model performance
+at different lead times) should be considered when evaluating forecasts
+and comparing models.
 
@@
 
-Ecological observations are influenced by both the underlying processes and how
-the system is sampled. When possible forecasts should model the factors
-influencing the observation of the data. [[citation?]]
+Ecological observations are influenced by both the underlying biological
+processes (e.g. resource limitation) and how the system is sampled. When
+possible, forecasts should model the factors influencing the observation of the
+data. [[citation?]]
 
@@
 
-Evalutating a models predictive performance across time is critical. Hindcasting
-uses a temporal out of sample validation to mimic how well a model would have
-performed had it been run in the past. For example using occurance data from the
-early 20th century to model distributions which are validated with late
-20th century occurances. Dense time series, such as yearly observations, are
-desirable to also evalulate the forecast horizon (see #6), but this is not
-a strict requirement.
+Evalutating a model's predictive performance across time is critical.
+Hindcasting uses a temporal out-of-sample validation approach to mimic how well
+a model would have performed had it been run in the past. For example, using
+occurance data from the early 20th century to model distributions which are
+validated with late 20th century occurances. Dense time series, such as yearly
+observations, are desirable to also evalulate the forecast horizon (see #6), but
+this is not a strict requirement.
 
@@
 
-To allow the future evaluation of the accuracy and uncertainty of forecasts the
-forecast values and/or models should be archived so that they can be assessed
+Forecast values and/or models should be archived so that they can be assessed
 after new data is generated [@mcgill2012; @silver2012]. Enough information
-should be provided to allow an unambiguous assessment of the forecast
-performance.
+should be provided in the archive to allow ecologists to make an unambiguous
+assessment of each forecast's performance.
 
-### 10. Make short-term and long-term predictions
+### 10. Make both short-term and long-term predictions
 
-In cases where long-term predictions are the primary goal, short-term should
-also be made to accommodate the time-scales of planning and management decisions
-and to allow the accuracy of the forecasts to be quickly evaluated
-[@tredennick2016].
+In cases where long-term predictions are the primary goal, short-term
+predictions should also be made to accommodate the time-scales of planning and
+management decisions and to allow the accuracy of the forecasts to be quickly
+evaluated [@tredennick2016].
 

### 2017-08-15 manuscript/knit.R

@@ -1,0 +1,15 @@
+library(knitr)
+library(rmarkdown)
+library(tidyverse)
+
+# Eliminate newlines inside of R code marked by `r code_goes_here`.
+readLines("manuscript/draft.Rmd") %>% 
+  paste(collapse = "\n") %>% 
+  gsub("`r\n", "`r ", .) %>% 
+  cat(file = "manuscript/draft-modified.Rmd")
+
+render("manuscript/draft-modified.Rmd", 
+       output_format = "pdf_document", 
+       output_dir = "manuscript",
+       output_file = "draft.pdf")
+file.remove("manuscript/draft-modified.Rmd")

### 2017-08-16 manuscript/draft.Rmd

@@
 
+[[Note: this paragraph may no longer be accurate!]] For the error metric that
+evaluates both models' point estimates and their uncertainty simultaneously, the
+JSDM and the nonlinear richness model converge towards the time-series baselines
+at the longest time-scales assessed. This convergence at longer time scales
+highlights the importance of considering how forecasting method performance
+changes with the distance into the future being predicted (Box 1). The value of
+this approach, and associated forecast horizons, has been raised for knowing how
+far into the future a model can be effectively used for forecasting
+[@petchey2015], but our results suggest a broader value to this approach for
+considering the potential importance of different models and processes for
+making forecasts at different scales. Our results show that for forecasts 1-3
+years in the future, baselines and time-series based approaches outperform all
+other methods using integrative metrics. At these time scales, environmental
+changes are relatively small and changes in the biota may lag behind changes in
+the environment. However, as the timescale of the forecast increases to a decade
+the JSDMs and environmental richness models get closer to the accuracy of the
+time-series methods and even surpass some of those methods. As the forecast
+distance increases the amount of environmental change is expected to increase
+and the system will have more time to respond, potentially leading to shifts
+towards improved relative performance of the models incorporating environmental
+and ecological information. If this shift with forecast distance in the relative
+importance of different processes and models continues, this could lead to JSDMs
+and environmental richness models outperforming time-series based approaches at
+sufficiently long time scales. These results suggest that assessment of how
+forecast performance changes with time lag is not only important for determining
+how far in the future to forecast, but also for determining which models and
+processes are most relevant for making forecasts at different time scales. For
+example ensemble forecasts could be created which incorporate these time lags
+into the weights for individual models.
+
 It is also possible that models that include species' relationships to their

### 2017-08-16 figures/plots.R
improve time-series figures

@@ -308,18 +308,10 @@ my_ggsave("figures/observer_uncertainty.png", obs_joy, height = 11)
 
 # Time series -------------------------------------------------------------
 
-
-# sample_site_id = 88005
-# sample_site_id = 91012
-# sample_site_id = 2024
-# sample_site_id = 91062
-sample_site_id = 38032
-sample_site_id = 72035
-
 make_ts_plots = function(models, ylim, use_obs_model, sample_site_id, 
                          main = ""){
   grid = crossing(model = c(ts_models, env_models),
-                  use_obs_model = unique(c(use_obs_model, FALSE)),
+                  use_obs_model = use_obs_model,
                   year = unique(obs_model$data$year))
   
   time_series_data = obs_model$data %>% 
@@ -333,7 +325,7 @@ make_ts_plots = function(models, ylim, use_obs_model, sample_site_id,
     select(site_id, year, model, use_obs_model, mean, sd, richness, observer_id,
            deviance)
   
-  faceting = if (use_obs_model) {
+  faceting = if (length(use_obs_model) == 2) {
     facet_grid(model ~ use_obs_model)
   } else {
     (facet_grid(~ model))
@@ -353,12 +345,14 @@ make_ts_plots = function(models, ylim, use_obs_model, sample_site_id,
     geom_line(aes(y = mean), color = "gray30") +
     faceting +
     geom_line(aes(y = richness, alpha = .5 * (year < min(bound$year - 1)))) + 
-    geom_point(aes(y = richness, fill = factor(observer_id)),
-               size = 1.25, shape = 21, stroke = .5) +
+    geom_point(aes(y = richness, fill = factor(observer_id),
+                   shape = factor(observer_id)),
+               size = 1.25, stroke = .5) +
     scale_alpha(range = c(0, 1), guide = FALSE) + 
     coord_cartesian(ylim = ylim, expand = TRUE) +
+    scale_shape_manual(values = c(16, 22, 23), guide = FALSE) + 
     ylab("Richness") +
-    viridis::scale_fill_viridis(discrete = TRUE, option = "A", guide = FALSE) + 
+    scale_fill_manual(values = c("white", "#fdcc8a", "#d7301f"), guide = FALSE) + 
     theme_light(base_size = base_size) + 
     theme(plot.margin = unit(c(10, 7, 1, 7), units = "pt"),
           strip.background = element_rect(fill="white"), 
[…]

### 2017-08-17 R/observer_model.R
delete broken save command
(not needed anyway)

@@ -73,7 +73,6 @@ fit_observer_model = function(settings, output_dir,
   message("R-hat values should be close to 1; values larger than about 1.05\nindicate convergence problems")
   
   saveRDS(obs_model, file = output_file)
-  saveRDS(samples, file = paste0("rstan_", output_file))
   
   NULL
 }

### 2017-08-18 manuscript/draft.Rmd
fix year counts (#165)

@@
 over 1,000 of sites located throughout North America, while following the best
-practices in Box 1 for ecological forecasting. To do this, we combine 30 years
+practices in Box 1 for ecological forecasting. To do this, we combine 32 years
 of time-series data on bird distributions from annual surveys with monthly
@@
 We evaluated 6 types of forecasting models (Table 1) at two time-scales. In the
-first one, we divided our [30] years of data into [22] years of training data
-and retained [9] years of data for evaluation via hindcasting. Second, we used
+first one, we divided our 32 years of data into 22 years of training data
+and retained 10 years of data for evaluation via hindcasting. Second, we used
 the full data set for training and made predictions through the year 2050. For
@@
 different possible set of observer-level and site-level random effect values
-across the full 30-year dataset.
+across the full 32-year dataset.
 
@@
 timesteps up to a decade into the future. In contrast, many ecological forecasts
-are temporally aggregated to 5-30 year timesteps and projected up to a century
+are temporally aggregated to 5-32 year timesteps and projected up to a century
 into the future. These are commonly beyond the career or even lifespan of the

### 2017-08-18 manuscript/draft.Rmd
Editorial changes to abstract

--- 2017-08-16 manuscript/draft.Rmd
@@
 Biodiversity forecasts are important for conservation, management, and
-evaluating how well current models characterize natural systems. While
-biodiversity forecasts are common, there is little information available on how
-well these forecasts work. Most biodiversity forecasts are not evaluated in
-terms of their ability to predict future diversity, fail to account for
-uncertainty, and do not use time-series data that captures the actual dynamics
-being studied. We compiled a list of ten best practices for forecasting
-biodiversity, and used hindcasting to evaluate six different modeling approaches
-for predicting breeding bird species richness. Each method was evaluated at `r
-numbers$N_predict_sites` sites distributed throughout the continental United
-States. Forecasts were evaluated for up to a decade after the final training
-observations. While each model could explain most of the variance in richness,
-none of them consistently outperformed a baseline model that simply predicted
-constant richness at each site. In particular, we found no evidence that current
-methods (such as species distribution models) can successfully turn spatial data
-into useful temporal predictions about biodiversity. This result may call into
-question the validity of many high-profile forecasts from the last 15 years.
-Most of the best practices implemented in this study (e.g., evaluating
-confidence intervals as well as point estimates) directly influence the outcome
-of the forecasts; some even led to large changes in the relative performance of
-different modeling approaches. To facilitate the rapid improvement of
-biodiversity forecasts, we emphasize the value of specific best practices in
-making and evaluating different forecasting methods.
+evaluating how well current models characterize natural systems. While the
+number of forecasts for biodiversity are increasing, there is little information
+available on how well these forecasts work. Most biodiversity forecasts are not
+evaluated to determine how well they predict future diversity, fail to account
+for uncertainty, and do not use time-series data that captures the actual
+dynamics being studied. We addressed these limitations by using best practices
+to explore our ability to forecast the species richness of breeding birds in
+North America. We used hindcasting to evaluate six different modeling approaches
+for predicting richness. Hindcasts for each method were evaluated annually for a
+decade at `r numbers$N_predict_sites` sites distributed throughout the
+continental United States. While each model could explain most of the variance
+in richness, none of them consistently outperformed a baseline model that
+predicted constant richness at each site. In particular, we found no evidence
+that current methods (such as species distribution models) can successfully turn
+spatial data into useful temporal predictions about biodiversity at decadale
+time-scales. The best practices implemented in this study directly influence the
+forecasts, the relative performance of different modeling approaches, and the
+conclusions about the current state of biodiversity forecasting . To facilitate
+the rapid improvement of biodiversity forecasts, we emphasize the value of
+specific best practices in making forecasts and evaluating forecasting methods.
 

### 2017-08-18 manuscript/draft.Rmd
Editorial work on Intro
Includes moving Table 1 into methods where the detail present in it makes more
sense. Replaces the reference to that table in the intro with a reference to
Figure 1, which works better there.

--- 2017-08-18 manuscript/draft.Rmd
@@
 Forecasting the future state of ecological systems is increasingly important for
-planning and management, and also for quantitatively evaluating how well our
-models capture the key ecological processes governing these systems [@clark2001;
-@houlahan2017; @dietze2017]. Forecasts regarding biodiversity are especially
-important, due to biodiversity's central role in conservation planning and its
-sensitivity to anthropogenic effects [@cardinale2012; @diaz2015; @tilman2017].
-High-profile studies forecasting large biodiversity declines over the coming
-decades have played a large role in shaping ecologists' priorities [as well as
-those of policymakers; e.g. @ipcc2014], but it is inherently difficult to
-evaluate such long-term predictions before the projected biodiversity declines
-have occurred.
+planning and management, and also for quantitatively evaluating how well
+ecological models capture the key processes governing natural systems
+[@clark2001; @houlahan2017; @dietze2017]. Forecasts regarding biodiversity are
+especially important, due to biodiversity's central role in conservation
+planning and its sensitivity to anthropogenic effects [@cardinale2012;
+@diaz2015; @tilman2017]. High-profile studies forecasting large biodiversity
+declines over the coming decades have played a large role in shaping ecologists'
+priorities [as well as those of policymakers; e.g. @ipcc2014], but it is
+inherently difficult to evaluate such long-term predictions before the projected
+biodiversity declines have occurred.
 
@@
 more generally, have focused primarily on building species distributions models
-(SDMs). In general, these models describe individual species' occurrence
-patterns as functions of the environment [@thomas2004; @urban2015]. By plugging
-forecasts of future environmental conditions into these functions, ecologists
-can make predictions about where each species will occur in the future. These
-species-level predictions can then be added up ("stacked") to generate predicted
-richness values [e.g. @calabrese2014]. Alternatively, models that directly
-relate spatial patterns of species richness to the environment have been
+[SDMs; @thomas2004; @urban2015]. In general, these models describe individual
+species' occurrence patterns as functions of the environment. Predictions about
+where each species will occur in the future are made by using forecasts of
+future environmental conditions to generate predictions from these models. These
+species-level predictions are then combined ("stacked") to generate forecasts
+for species richness [e.g. @calabrese2014]. Alternatively, models that directly
+relate spatial patterns of species richness to environment conditions have been
 developed and generally perform equivalently to stacked SDMs [@algar2009;
@@
 important for developing forecasting models with high predictive accuracy, and
-for understanding how well different methods can predict the future. The
-components an effective forecasting and evaluation strategy (Box 1) broadly
-involve: 1) expanding the use of data to include biological and environmental
-time-series [@tredennick2016]; 2) accounting for uncertainty in observations,
-processes, model choice, and forecast evaluation [@clark2001; @dietze2017]; and
-3) conducting meaningful evaluations of the forecasts by hindcasting, archiving
+for understanding how well different methods can predict the future. These "best
+practices" for effective forecasting and evaluation (Box 1) broadly involve: 1)
+expanding the use of data to include biological and environmental time-series
+[@tredennick2016]; 2) accounting for uncertainty in observations, processes,
+model choice, and forecast evaluation [@clark2001; @dietze2017]; and 3)
+conducting meaningful evaluations of the forecasts by hindcasting, archiving
 short-term forecasts, and comparing forecasts to baselines to determine whether
-the forecasts are more accurate than simply assuming the system is basically
-static [@perretti2013].
-
-\begin{singlespace}
-
-```{r table, echo=FALSE}
-
-table_caption = "A summary of the six models evaluated here. The single-site
-models were trained site-by-site, without access to any environmental
-information. The environmental models were trained at a continental scale,
-without information regarding which transects occurred at which site or during
-which year. Most of the models were trained to predict richness directly.
-However, the random forest SDMs were each trained to predict the occurrence of
-individual species at each site, and these species-level predictions were added
-up to produce richness estimates. The mistnet JSDM was trained to predict the
-full species composition at each site, and the number of species in its
-predictions was used as an estimate of richness."
-
-table_data = read_csv("../figures/table.csv",col_types = "ccccc")
-
-table_data %>% 
-  knitr::kable(
-    format = "latex", 
-    escape = FALSE, 
-    booktabs = TRUE, 
-    caption = table_caption
-  ) %>% 
-  add_header_above(c(" " = 2, "Predictors" = 3)) %>%
-  group_rows("Single-site models", 1, 3) %>% 
-  group_rows("Environmental models", 4, 6) %>% kable_styling(position = "center")
-```
-
-\end{singlespace}
+the forecasts are more accurate than assuming the system is basically static
+[@perretti2013].
 
 In this paper, we attempt to forecast the species richness of breeding birds at
-over 1,000 of sites located throughout North America, while following the best
-practices in Box 1 for ecological forecasting. To do this, we combine 32 years
-of time-series data on bird distributions from annual surveys with monthly
+over 1,200 of sites located throughout North America, while following best
+practices for ecological forecasting (Box 1). To do this, we combine 32 years of
+time-series data on bird distributions from annual surveys with monthly
 time-series of climate data and satellite-based remote-sensing. Datasets that
@@
 species richness, time-series methods, and two simple baselines that predict
-constant richness for each site, on average (Table 1). All of our forecasting
+constant richness for each site, on average (Figure 1). All of our forecasting
 models account for uncertainty and observation error, are evaluated across
@@
 distribution) for the reasons discussed below (see "Model evaluation").
+
+\begin{singlespace}
+
+```{r table, echo=FALSE}
+
+table_caption = "A summary of the six models evaluated here. The single-site
+models were trained site-by-site, without access to any environmental
+information. The environmental models were trained at a continental scale,
+without information regarding which transects occurred at which site or during
+which year. Most of the models were trained to predict richness directly.
+However, the random forest SDMs were each trained to predict the occurrence of
+individual species at each site, and these species-level predictions were added
+up to produce richness estimates. The mistnet JSDM was trained to predict the
+full species composition at each site, and the number of species in its
+predictions was used as an estimate of richness."
+
+table_data = read_csv("../figures/table.csv",col_types = "ccccc")
+
+table_data %>%
+  knitr::kable(
+    format = "latex",
+    escape = FALSE,
+    booktabs = TRUE,
+    caption = table_caption
+  ) %>%
+  add_header_above(c(" " = 2, "Predictors" = 3)) %>%
+  group_rows("Single-site models", 1, 3) %>%
+  group_rows("Environmental models", 4, 6) %>% kable_styling(position = "center")
+```
+
+\end{singlespace}
 
@@
 timesteps up to a decade into the future. In contrast, many ecological forecasts
-are temporally aggregated to 5-32 year timesteps and projected up to a century
+are temporally aggregated to 5-30 year timesteps and projected up to a century
 into the future. These are commonly beyond the career or even lifespan of the

### 2017-08-18 manuscript/draft.Rmd
Editorial work on Intro
Includes moving Table 1 into methods where the detail present in it makes more
sense. Replaces the reference to that table in the intro with a reference to
Figure 1, which works better there.

@@
 over 1,200 of sites located throughout North America, while following best
-practices for ecological forecasting (Box 1). To do this, we combine 32 years of
+practices for ecological forecasting (Box 1). To do this, we combine 30 years of
 time-series data on bird distributions from annual surveys with monthly
@@
 We evaluated 6 types of forecasting models (Table 1) at two time-scales. In the
-first one, we divided our 32 years of data into 22 years of training data
-and retained 10 years of data for evaluation via hindcasting. Second, we used
+first one, we divided our [30] years of data into [22] years of training data
+and retained [9] years of data for evaluation via hindcasting. Second, we used
 the full data set for training and made predictions through the year 2050. For
@@
 different possible set of observer-level and site-level random effect values
-across the full 32-year dataset.
+across the full 30-year dataset.
 

### 2017-08-18 manuscript/draft.Rmd
Edits to Figure 1 legend

--- 2017-08-18 manuscript/draft.Rmd
@@
 Biodiversity forecasts are important for conservation, management, and
-evaluating how well current models characterize natural systems. While the
-number of forecasts for biodiversity are increasing, there is little information
-available on how well these forecasts work. Most biodiversity forecasts are not
-evaluated to determine how well they predict future diversity, fail to account
-for uncertainty, and do not use time-series data that captures the actual
-dynamics being studied. We addressed these limitations by using best practices
-to explore our ability to forecast the species richness of breeding birds in
-North America. We used hindcasting to evaluate six different modeling approaches
-for predicting richness. Hindcasts for each method were evaluated annually for a
-decade at `r numbers$N_predict_sites` sites distributed throughout the
-continental United States. While each model could explain most of the variance
-in richness, none of them consistently outperformed a baseline model that
-predicted constant richness at each site. In particular, we found no evidence
-that current methods (such as species distribution models) can successfully turn
-spatial data into useful temporal predictions about biodiversity at decadale
-time-scales. The best practices implemented in this study directly influence the
-forecasts, the relative performance of different modeling approaches, and the
-conclusions about the current state of biodiversity forecasting . To facilitate
-the rapid improvement of biodiversity forecasts, we emphasize the value of
-specific best practices in making forecasts and evaluating forecasting methods.
+evaluating how well current models characterize natural systems. While
+biodiversity forecasts are common, there is little information available on how
+well these forecasts work. Most biodiversity forecasts are not evaluated in
+terms of their ability to predict future diversity, fail to account for
+uncertainty, and do not use time-series data that captures the actual dynamics
+being studied. We compiled a list of ten best practices for forecasting
+biodiversity, and used hindcasting to evaluate six different modeling approaches
+for predicting breeding bird species richness. Each method was evaluated at `r
+numbers$N_predict_sites` sites distributed throughout the continental United
+States. Forecasts were evaluated for up to a decade after the final training
+observations. While each model could explain most of the variance in richness,
+none of them consistently outperformed a baseline model that simply predicted
+constant richness at each site. In particular, we found no evidence that current
+methods (such as species distribution models) can successfully turn spatial data
+into useful temporal predictions about biodiversity. This result may call into
+question the validity of many high-profile forecasts from the last 15 years.
+Most of the best practices implemented in this study (e.g., evaluating
+confidence intervals as well as point estimates) directly influence the outcome
+of the forecasts; some even led to large changes in the relative performance of
+different modeling approaches. To facilitate the rapid improvement of
+biodiversity forecasts, we emphasize the value of specific best practices in
+making and evaluating different forecasting methods.
 
@@
 
-![Example predictions from our six models for a single site. The data points
-from 1982 through 2003, connected by solid lines, were used for training all six
-models; the remaining points were used for evaluating the models' forecasts. In
-each panel, point estimates for each year are shown with lines; the darker
-ribbon indicates 1 standard deviation of uncertainty, and the lighter ribbon
-indicates 95% confidence intervals. **A.** The single-site models models were
-trained using one site at a time, based solely on observed richness values
-shown. The first two models ("average" and "naive") served as our simple
-forecasting baselines. **B.** The environmental models were trained to predict
-richness based on elevation, climate, and NDVI; unlike the single-site models,
-the environmental models' predictions change from year to year as environmental
-conditions change.](../figures/model_predictions.png)
+![Example predictions from six forecasting models for a single site. Data from
+1982 through 2003, connected by solid lines, were used for training the models;
+the remaining points were used for evaluating the models' forecasts. In each
+panel, point estimates for each year are shown with lines; the darker ribbon
+indicates the 68% prediction interval (1 standard deviation of uncertainty), and
+the lighter ribbon indicates the 95% prediction interval. **A.** Single-site
+models were trained independently on each site's observed richness values. The
+first two models ("average" and "naive") served as baselines. **B.** The
+environmental models were trained to predict richness based on elevation,
+climate, and NDVI; the environmental models' predictions change from year to
+year as environmental conditions change.](../figures/model_predictions.png)
 

### 2017-08-18 manuscript/draft.Rmd
30, not 32

@@
 
-![Example predictions from six forecasting models for a single site. Data from
-1982 through 2003, connected by solid lines, were used for training the models;
-the remaining points were used for evaluating the models' forecasts. In each
-panel, point estimates for each year are shown with lines; the darker ribbon
-indicates the 68% prediction interval (1 standard deviation of uncertainty), and
-the lighter ribbon indicates the 95% prediction interval. **A.** Single-site
-models were trained independently on each site's observed richness values. The
-first two models ("average" and "naive") served as baselines. **B.** The
-environmental models were trained to predict richness based on elevation,
-climate, and NDVI; the environmental models' predictions change from year to
-year as environmental conditions change.](../figures/model_predictions.png)
+![Example predictions from our six models for a single site. The data points
+from 1982 through 2003, connected by solid lines, were used for training all six
+models; the remaining points were used for evaluating the models' forecasts. In
+each panel, point estimates for each year are shown with lines; the darker
+ribbon indicates 1 standard deviation of uncertainty, and the lighter ribbon
+indicates 95% confidence intervals. **A.** The single-site models models were
+trained using one site at a time, based solely on observed richness values
+shown. The first two models ("average" and "naive") served as our simple
+forecasting baselines. **B.** The environmental models were trained to predict
+richness based on elevation, climate, and NDVI; unlike the single-site models,
+the environmental models' predictions change from year to year as environmental
+conditions change.](../figures/model_predictions.png)
 

### 2017-08-18 manuscript/draft.Rmd
Editorial work on Methods section

--- 2017-08-18 manuscript/draft.Rmd
@@
 over 1,200 of sites located throughout North America, while following best
-practices for ecological forecasting (Box 1). To do this, we combine 30 years of
+practices for ecological forecasting (Box 1). To do this, we combine 32 years of
 time-series data on bird distributions from annual surveys with monthly
@@
 **Richness data.** Bird species richness was obtained from the North American
-Breeding Bird Survey (BBS) [@pardieck2017] using the Data Retriever software
-[@morris2013] and rdataretriever R package [@mcglinn2017]. The BBS data was
-filtered to exclude all nocturnal, cepuscular, and aquatic species [since these
-species are not well sampled by BBS methods; @hurlbert2005], as well as
-unidentified species, and hybrids. All data from routes that did not meet BBS
-acceptable criteria were also excluded.
+Breeding Bird Survey (BBS) [@pardieck2017] using the Data Retriever Python
+package [@morris2013] and rdataretriever R package [@mcglinn2017]. The BBS data
+was filtered to exclude all nocturnal, cepuscular, and aquatic species [since
+these species are not well sampled by BBS methods; @hurlbert2005], as well as
+unidentified species, and hybrids. All data from surveys that did not meet BBS
+quality criteria were also excluded.
 
 We used observed richness values from 1982 (the first year of complete
-environmental data) to 2003 for training the models, and data from 2004 to 2013
-was used for testing their performance. We only used BBS routes from the
-continental United States (i.e. routes where climate data was available @prism),
-and we restricted the analysis to routes that were sampled during 70% of the
-years in the training period (i.e., routes with at least 16 annual
-observations). The resulting dataset included `r numbers$N_runs` transect runs
-of `r numbers$N_sites` unique sites, and included `r numbers$N_species` species.
-Site-level richness varied from `r numbers$richness_summary$min` to `r
-numbers$richness_summary$max` with an average richness of `r
-numbers$richness_summary$mean` species.
+environmental data) to 2003 to train the models, and from 2004 to 2013 to test
+their performance. We only used BBS routes from the continental United States
+(i.e. routes where climate data was available @prism), and we restricted the
+analysis to routes that were sampled during 70% of the years in the training
+period (i.e., routes with at least 16 annual observations). The resulting
+dataset included `r numbers$N_runs` annual surveys of `r numbers$N_sites` unique
+sites, and included `r numbers$N_species` species.  Site-level richness varied
+from `r numbers$richness_summary$min` to `r numbers$richness_summary$max` with
+an average richness of `r numbers$richness_summary$mean` species.
 
@@
 correlate of richness in BBS data [@hurlbert2002], was obtained from the NDIV3g
-dataset with an 8 km resolution [@pinzon2014] and was available from the time
-period 1981-2013. Average summer (May, June, July) and winter (December,
-January, Feburary) NDVI values were used as predictors. Elevation was from the
-SRTM 90m elevation dataset [@jarvis2008] obtained using the R package raster
+dataset with an 8 km resolution [@pinzon2014] and was available from 1981-2013.
+Average summer (May, June, July) and winter (December, January, Feburary) NDVI
+values were used as predictors. Elevation was from the SRTM 90m elevation
+dataset [@jarvis2008] obtained using the R package raster
 [@hijmans2016]. Because BBS routes are 40-km transects rather than point counts,
@@
 **Future environmental projections.** We made long term forecasts from 2014-2050
-using the CMIP5 multi-model ensemble dataset as the source for climatic
+using the CMIP5 multi-model ensemble dataset as the source for climate
 variables [@brekke2013]. Precipitation and temperature from 37 downscaled model
@@
 
-For each forecasting approach, we trained two versions of the corresponding
-model: one with corrections for differences among observers, and one without.
 Observer effects are inherent in large data sets collected by different
-observers, and are known to occur in BBS [@sauer1994]. We estimated the observer
-effects (and associated uncertainty about those effects) with a linear mixed
-model with observer as a random effect built in the Stan probabilistic
-programming language [@stan]. Because observer and site are strongly related
-(observers tend to repeatedly sample the same site), site was also included as a
-random effect to ensure that inferred deviations were actually observer-related
-(as opposed to being related to the sites that a given observer happened to
-see). The resulting model partitions the variance in observed richness values
-into site-level variance, observer-level variance, and residual variance (e.g.
-variation within a site from year to year; Figure 3). The site-level estimates
-can also be used directly as the "average" baseline model (see below). The
-estimated observer effects can be subtracted from the richness values for a
-particular observer to provide an estimate of how many species would have been
-found by a "typical" observer. To incorporate uncertainty in these "corrected"
-richness values into the forecasting models we collected 500 Monte Carlo samples
-from the model's posterior distribution, and fit each of the downstream models
-with each of the Monte Carlo samples. Each Monte Carlo sample represented a
-different possible set of observer-level and site-level random effect values
-across the full 30-year dataset.
+observers, and are known to occur in BBS [@sauer1994]. For each forecasting
+approach, we trained two versions of the corresponding model: one with
+corrections for differences among observers, and one without (Figure 2). We
+estimated the observer effects (and associated uncertainty about those effects)
+with a linear mixed model, with observer as a random effect, built in the Stan
+probabilistic programming language [@stan]. Because observer and site are
+strongly related (observers tend to repeatedly sample the same site), site was
+also included as a random effect to ensure that inferred deviations were
+actually observer-related (as opposed to being related to the sites that a given
+observer happened to see). The resulting model partitions the variance in
+observed richness values into site-level variance, observer-level variance, and
+residual variance (e.g.  variation within a site from year to year). The
+site-level estimates can also be used directly as the "average" baseline model
+(see below). The estimated observer effects can be subtracted from the richness
+values for a particular observer to provide an estimate of how many species
+would have been found by a "typical" observer. To incorporate uncertainty in
+these "corrected" richness values into the forecasting models we collected 500
+Monte Carlo samples from the model's posterior distribution, and fit each of the
+downstream models with each of the Monte Carlo samples. Each Monte Carlo sample
+represented a different possible set of observer-level and site-level random
+effect values across the full 32-year dataset.
 
-![**A.** When all observers are treated the same (black points), predictions can
-be poor. **B.** By accounting for systematic differences between observers
-(represented by the points' colors), most models can be made more robust to
-observer turnover. Note that the "naive" model is already less sensitive to
+![**A.** Model predictions for [[STATE]] route [[ROUTE NUMBER]] when all
+observers are treated the same (black points). **B.** Model predictions for the
+same route when accounting for systematic differences between observers
+(represented by the points' colors). In this example most models are made more
+robust to observer turnover. Note that the "naive" model is less sensitive to
 observer turnover, and does not benefit as much from modeling
@@
 
-We fit three sets of models that were fit to each site separately, with no
+Three of models used in this study were fit to each site separately, with no
 environmental information (Table 1). These models were fit to each BBS route
@@
 of these models use a Gaussian error distribution (rather than a count
-distribution) for the reasons discussed below (see "Model evaluation").
+distribution) for reasons discussed below (see "Model evaluation").
 
@@
 
-table_caption = "A summary of the six models evaluated here. The single-site
-models were trained site-by-site, without access to any environmental
-information. The environmental models were trained at a continental scale,
-without information regarding which transects occurred at which site or during
-which year. Most of the models were trained to predict richness directly.
-However, the random forest SDMs were each trained to predict the occurrence of
-individual species at each site, and these species-level predictions were added
-up to produce richness estimates. The mistnet JSDM was trained to predict the
-full species composition at each site, and the number of species in its
-predictions was used as an estimate of richness."
+table_caption = "Six forecasting models. Single-site models were trained
+site-by-site, without environmental data. Environmental models were trained
+using all sites together, without information regarding which transects occurred
+at which site or during which year. Most of the models were trained to predict
+richness directly. This mirrors the standard application of these
+techniques. Separate random forest SDMs were fit for each species and used to
+predict the probability of that species occurring at each site. The
+species-level probabilities at a site were summed to predict richness. The
+mistnet JSDM was trained to predict the full species composition at each site,
+and the number of species in its predictions was used as an estimate of
+richness."
 
@@
 site-level constant (the "average" model) or as an autoregressive model with a
-single year of history (the "naive" model) [@hyndman2014]. Predictions from the
+single year of history [the "naive" model, @hyndman2014]. Predictions from the
 "average" model are centered on the average richness observed during training,
@@
 in contrast, predicts that future observations will be similar to the final
-observed value (e.g. the value observed in `r timeframe$last_train_year`), and
-the confidence intervals expand rapidly as the predictions extend farther into
-the future. Both models' richness predictions are centered on a constant value,
-so neither model can anticipate any trends in richness or any responses to
-future environmental changes.
+observed value (e.g., in our hindcasts the value observed in `r
+timeframe$last_train_year`), and the confidence intervals expand rapidly as the
+predictions extend farther into the future. Both models' richness predictions
+are centered on a constant value, so neither model can anticipate any trends in
+richness or any responses to future environmental changes.
 
 **Time series models.** We used Auto-ARIMA models [based on the `auto.arima`
-function in @forecast1] to represent a array of different time-series modeling
+function in @forecast1] to represent an array of different time-series modeling
 approaches. These models can include an autoregressive component (as in the
@@
 bag" estimator; this number averaged `r numbers$n_gbm[1]` for the non-observer
-data and `r numbers$n_gbm[2]` for the observer-corrected data. [[add comma
-separators for thousands place]]
+data and `r numbers$n_gbm[2]` for the observer-corrected data.
 
@@
 (JSDMs) are a new approach that makes predictions about the full composition of
-a community instead of modeling each species independently as above. JSDMs
-remove the assumed independence among species and explicitly account for the
-possibility that a site will be much more (or less) suitable for birds in
-general (or particular groups of birds) than one would expect based on the
-available environmental measurements alone. As a result, JSDMs do a better job
-of representing our uncertainty about richness, while stacked SDMs underestimate
-it [@harris2015; [[other refs, e.g., Clark]]]. We used the `mistnet` package
+a community instead of modeling each species independently as above
+[@warton2015]. JSDMs remove the assumed independence among species and
+explicitly account for the possibility that a site will be much more (or less)
+suitable for birds in general (or particular groups of birds) than one would
+expect based on the available environmental measurements alone. As a result,
+JSDMs do a better job of representing uncertainty about richness than stacked
+SDMs [@harris2015; [[other refs, e.g., Clark]]]. We used the `mistnet` package
 [@harris2015] because it is the only JSDM that describes species' environmental
@@
 richness had a relatively large mean (`r numbers$richness_summary$mean`) and all
-models already produce continuous richness estimates (due to summing SDM or JSDM
-probabilities). When a model was run multiple times for the purpose of
-correcting for observer effects, we used the mean of those runs' point estimates
-as our final point estimate and we calculated the uncertainty using the law of
-total variance (i.e. the average of the model runs' variance, plus the variance
-in the point estimates).
+models produce continuous richness estimates. When a model was run multiple
+times for the purpose of correcting for observer effects, we used the mean of
+those runs' point estimates as our final point estimate and we calculated the
+uncertainty using the law of total variance (i.e. the average of the model runs'
+variance, plus the variance in the point estimates).
 
 We evaluated each model's forecasts using the data for each year between `r
-timeframe$last_train_year` and `r timeframe$end_yr`. We used three metrics for
-evaluating performance: 1) root-mean-square error (RMSE) to determine how far,
-on average, the models' predictions were from the observed value; 2) the 95%
-prediction interval coverage to determine how well the models predicted the
+timeframe$last_train_year + 1` and `r timeframe$end_yr`. We used three metrics
+for evaluating performance: 1) root-mean-square error (RMSE) to determine how
+far, on average, the models' predictions were from the observed value; 2) the
+95% prediction interval coverage to determine how well the models predicted the
 range of possible outcomes; and 3) deviance (i.e. negative 2 times the Gaussian
@@
 of forecasting increased by plotting performance metrics against year. Finally,
-we decomposed each model's squared error into two components: one described the
-squared error associated with site-level means, while the other described the
-squared error associated with annual fluctuations in richness within a site.
-This decomposition tells us the extent to which each model's error depends on
-consistent differences among sites versus changes in site-level richness from
-year to year.
+we decomposed each model's squared error into two components: the squared error
+associated with site-level means and the squared error associated with annual
+fluctuations in richness within a site. This decomposition describes the extent
+to which each model's error depends on consistent differences among sites versus
+changes in site-level richness from year to year.
 

### 2017-08-18 manuscript/draft.Rmd

@@
 
-We evaluated 6 types of forecasting models (Table 1) at two time-scales. In the
-first one, we divided our 32 years of data into 22 years of training data
-and retained 10 years of data for evaluation via hindcasting. Second, we used
-the full data set for training and made predictions through the year 2050. For
-both time scales, we evaluated versions of each model with and without
-correcting for observer effects, as described below.
+We evaluated 6 types of forecasting models (Table 1) using hindcasting by
+dividing the [30] years of data into [22] years of training data and [9] years
+of data for evaluation. We also used the full data set to train all 6 models and
+made annual forecasts through the year 2050. For both time scales, we made
+forecasts using versions of each model with and without correcting for observer
+effects, as described below.
 
@@
 represented a different possible set of observer-level and site-level random
-effect values across the full 32-year dataset.
+effect values across the full 30-year dataset.
 

### 2017-08-19 figures/plots.R
plot fixes

@@ -4,6 +4,11 @@ library(cowplot)
 library(ggjoy)
 library(mvtnorm)
 
+ts_models = c("average", "naive", "auto.arima")
+env_models = c("richness_gbm", "rf_sdm", "mistnet")
+models = c(ts_models, env_models)
+
+
 settings = yaml::yaml.load_file("settings.yaml")
 
 timeframe = "train_22"
@@ -20,6 +25,7 @@ my_ggsave = function(filename, plot, height, width = 14.59, units = "cm", ...){
          height = height, 
          width = width, 
          units = units, 
+         dpi = 500,
          ...)
 }
 
@@ -52,18 +58,17 @@ bound = bind_rows(forecast_results, gbm_results, average_results,
   mutate(diff = richness - mean, z = diff / sd,
          p = pnorm(richness, mean, sd),
          deviance = -2 * dnorm(richness, mean, sd, log = TRUE)) %>% 
-  ungroup() %>% 
-  mutate(
-    model_name = forcats::fct_recode(
-      factor(model), 
-      Average = "average", 
-      Naive = "naive",
-      Auto.arima = "auto.arima", 
-      `Stacked random forest SDMs` = "rf_sdm",
-      `Mistnet JSDM` = "mistnet",
-      `GBM richness regression` = "richness_gbm"
-    )
-  )
+  ungroup()
+
+formal_names = read_csv("figures/table.csv", col_types = "ccccc") %>% 
+  pull(Model)
+formal_names = formal_names
+
+bound = mutate(bound) %>% 
+  mutate(formal_name = factor(model, levels = models, 
+                              labels = formal_names))
+
+distinct(bound, model, formal_name)
 
 # Divide squared error by predictive variance to get the optimal sd factor
 # for the RF model
@@ -81,11 +86,6 @@ rf2 = rf2 %>%
 
 # scatterplot -------------------------------------------------------------
 
-ts_models = c("average", "naive", "auto.arima")
-env_models = c("richness_gbm", "rf_sdm", "mistnet")
-models = c(ts_models, env_models)
-model_names = unique(bound$model_name)
-
 R2s = filter(bound, use_obs_model, year %in% c(2004, 2013)) %>% 
   mutate(x = min(mean - 4), y = max(richness)) %>% 
   group_by(model, x, y, year) %>% 
@@ -100,7 +100,7 @@ make_scatterplots = function(models, main){
          aes(x = mean, y = richness)) +
     geom_hex() + 
[…]

### 2017-08-23 manuscript/draft.Rmd

--- 2017-08-18 manuscript/draft.Rmd
@@
 
-We evaluated 6 types of forecasting models (Table 1) using hindcasting by
-dividing the [30] years of data into [22] years of training data and [9] years
-of data for evaluation. We also used the full data set to train all 6 models and
-made annual forecasts through the year 2050. For both time scales, we made
-forecasts using versions of each model with and without correcting for observer
-effects, as described below.
+We evaluated 6 types of forecasting models (Table 1) at two time-scales. In the
+first one, we divided our 32 years of data into 22 years of training data
+and retained 10 years of data for evaluation via hindcasting. Second, we used
+the full data set for training and made predictions through the year 2050. For
+both time scales, we evaluated versions of each model with and without
+correcting for observer effects, as described below.
 
@@
 represented a different possible set of observer-level and site-level random
-effect values across the full 30-year dataset.
+effect values across the full 32-year dataset.
 

### 2017-08-23 manuscript/draft.Rmd
Minor wording suggestions
Tweaks of Ethan's wording in 2 places

@@
 evaluating how well current models characterize natural systems. While the
-number of forecasts for biodiversity are increasing, there is little information
+number of forecasts for biodiversity is increasing, there is little information
 available on how well these forecasts work. Most biodiversity forecasts are not
@@
 [SDMs; @thomas2004; @urban2015]. In general, these models describe individual
-species' occurrence patterns as functions of the environment. Predictions about
-where each species will occur in the future are made by using forecasts of
-future environmental conditions to generate predictions from these models. These
-species-level predictions are then combined ("stacked") to generate forecasts
-for species richness [e.g. @calabrese2014]. Alternatively, models that directly
-relate spatial patterns of species richness to environment conditions have been
-developed and generally perform equivalently to stacked SDMs [@algar2009;
-@distler2015]. This approach is sometimes referred to as "macroecological"
-modeling, because it models the larger-scale pattern (richness) directly
-[@distler2015].
+species' occurrence patterns as functions of the environment. Given forecasted
+environmental conditions, these models can predict where each species will occur 
+in the future. These species-level predictions are then combined ("stacked") to 
+generate forecasts for species richness [e.g. @calabrese2014]. Alternatively, 
+models that directly relate spatial patterns of species richness to environment 
+conditions have been developed and generally perform equivalently to stacked SDMs 
+[@algar2009; @distler2015]. This approach is sometimes referred to as 
+"macroecological" modeling, because it models the larger-scale pattern (richness) 
+directly [@distler2015].
 

### 2017-08-23 manuscript/draft.Rmd
Minor wording suggestions (#177)
* Minor wording suggestions

Tweaks of Ethan's wording in 2 places

* Reword forecasts for future

@@
 [SDMs; @thomas2004; @urban2015]. In general, these models describe individual
-species' occurrence patterns as functions of the environment. Given forecasted
+species' occurrence patterns as functions of the environment. Given forecasts for
 environmental conditions, these models can predict where each species will occur 

### 2017-08-23 manuscript/draft.Rmd
refs w/o conflicts

--- 2017-08-23 manuscript/draft.Rmd
@@
 more generally, have focused primarily on building species distributions models
-[SDMs; @thomas2004; @urban2015]. In general, these models describe individual
-species' occurrence patterns as functions of the environment. Predictions about
-where each species will occur in the future are made by using forecasts of
-future environmental conditions to generate predictions from these models. These
-species-level predictions are then combined ("stacked") to generate forecasts
-for species richness [e.g. @calabrese2014]. Alternatively, models that directly
-relate spatial patterns of species richness to environment conditions have been
-developed and generally perform equivalently to stacked SDMs [@algar2009;
-@distler2015]. This approach is sometimes referred to as "macroecological"
-modeling, because it models the larger-scale pattern (richness) directly
-[@distler2015].
+[SDMs; @thomas2004; @thuiller2011; @urban2015]. In general, these models
+describe individual species' occurrence patterns as functions of the
+environment. Predictions about where each species will occur in the future are
+made by using forecasts of future environmental conditions to generate
+predictions from these models. These species-level predictions are then combined
+("stacked") to generate forecasts for species richness [e.g. @calabrese2014].
+Alternatively, models that directly relate spatial patterns of species richness
+to environment conditions have been developed and generally perform equivalently
+to stacked SDMs [@algar2009; @distler2015]. This approach is sometimes referred
+to as "macroecological" modeling, because it models the larger-scale pattern
+(richness) directly [@distler2015].
 
 Despite the emerging interest in forecasting species richness and other aspects
-of biodiversity [[refs]], little is known about how effectively we can
-anticipate these dynamics. This is due in part to the long time scales over
-which many ecological forecasts are applied [and the resulting difficulty in
-assessing whether the predicted changes occurred; @dietze2016]. What we do know
-comes from a small number of hindcasting studies, where models are built using
-data on species occurrence and richness from the past and evaluated on their
-ability to predict contemporary patterns [e.g., @algar2009; @distler2015]. These
-studies are a valuable first step, but lack several components that are
-important for developing forecasting models with high predictive accuracy, and
-for understanding how well different methods can predict the future. These "best
-practices" for effective forecasting and evaluation (Box 1) broadly involve: 1)
-expanding the use of data to include biological and environmental time-series
-[@tredennick2016]; 2) accounting for uncertainty in observations, processes,
-model choice, and forecast evaluation [@clark2001; @dietze2017]; and 3)
-conducting meaningful evaluations of the forecasts by hindcasting, archiving
-short-term forecasts, and comparing forecasts to baselines to determine whether
-the forecasts are more accurate than assuming the system is basically static
-[@perretti2013].
+of biodiversity [@jetz2007; @thuiller2011], little is known about how
+effectively we can anticipate these dynamics. This is due in part to the long
+time scales over which many ecological forecasts are applied [and the resulting
+difficulty in assessing whether the predicted changes occurred; @dietze2016].
+What we do know comes from a small number of hindcasting studies, where models
+are built using data on species occurrence and richness from the past and
+evaluated on their ability to predict contemporary patterns [e.g., @algar2009;
+@distler2015]. These studies are a valuable first step, but lack several
+components that are important for developing forecasting models with high
+predictive accuracy, and for understanding how well different methods can
+predict the future. These "best practices" for effective forecasting and
+evaluation (Box 1) broadly involve: 1) expanding the use of data to include
+biological and environmental time-series [@tredennick2016]; 2) accounting for
+uncertainty in observations, processes, model choice, and forecast evaluation
+[@clark2001; @dietze2017]; and 3) conducting meaningful evaluations of the
+forecasts by hindcasting, archiving short-term forecasts, and comparing
+forecasts to baselines to determine whether the forecasts are more accurate than
+assuming the system is basically static [@perretti2013].
 
@@
 normalized difference vegetation index; NDVI), all of which are known to
-influence richness and distribution in the BBS data [[ref]]. For each year in
-the dataset, we used the 4 km resolution PRISM data [@prism] to calculate eight
-bioclimatic variables identified as relevant to bird distributions
+influence richness and distribution in the BBS data [@kent2014]. For each year
+in the dataset, we used the 4 km resolution PRISM data [@prism] to calculate
+eight bioclimatic variables identified as relevant to bird distributions
 [@harris2015]: mean diurnal range, isothermality, max temperature of the warmest
@@
 values were used as predictors. Elevation was from the SRTM 90m elevation
-dataset [@jarvis2008] obtained using the R package raster
-[@hijmans2016]. Because BBS routes are 40-km transects rather than point counts,
-we used the average value of each environmental variable within a 40 km radius
-of each BBS route's starting point.
+dataset [@jarvis2008] obtained using the R package raster [@hijmans2016].
+Because BBS routes are 40-km transects rather than point counts, we used the
+average value of each environmental variable within a 40 km radius of each BBS
+route's starting point.
 
@@
 JSDMs do a better job of representing uncertainty about richness than stacked
-SDMs [@harris2015; [[other refs, e.g., Clark]]]. We used the `mistnet` package
+SDMs [@harris2015; @warton2015]. We used the `mistnet` package
 [@harris2015] because it is the only JSDM that describes species' environmental
@@
 because SDMs form the main foundation for estimates of the predicted loss of
-biodiversity to climate change [@thomas2004; @urban2015].
+biodiversity to climate change [@thomas2004; @thuiller2011; @urban2015].
 
@@
 equivalent to stacked SDMs, but the JSDMs provide much better estimates of
-uncertainty [@harris2015; others?]. In fact, JSDMs and direct richness modeling
+uncertainty [@harris2015; @warton2015]. In fact, JSDMs and direct richness modeling
 provide some of the best estimates of uncertainty across all modeling
@@
 can learn more rapidly about the best approaches for making predictions for a
-given ecological quantity. This includes comparing process-based [e.g. [[refs]]]
-and data-driven models, as well as comparing the accuracy of forecasts to simple
-baselines to determine if the modeled forecasts are more accurate than the naive
-assumption that the world is static [@ye2015].
+given ecological quantity. This includes comparing process-based [e.g.
+@kearney2009] and data-driven models, as well as comparing the accuracy of
+forecasts to simple baselines to determine if the modeled forecasts are more
+accurate than the naive assumption that the world is static [@ye2015].
 
@@
 variables may have at least as many benefits as using more sophisticated
-modeling techniques. [[Need a citation or two here]]
+modeling techniques [@kent2014].
 
@@
 possible, forecasts should model the factors influencing the observation of the
-data. [[citation?]]
+data. [@yu2010; @hutchinson2011; @schurr2012]
 

### 2017-08-23 manuscript/draft.Rmd

@@
 to as "macroecological" modeling, because it models the larger-scale pattern
+(richness) directly [@distler2015]. [SDMs; @thomas2004; @thuiller2011;
+@urban2015]. In general, these models describe individual species' occurrence
+patterns as functions of the environment. Given forecasts for environmental
+conditions, these models can predict where each species will occur in the
+future. These species-level predictions are then combined ("stacked") to
+generate forecasts for species richness [e.g. @calabrese2014]. Alternatively,
+models that directly relate spatial patterns of species richness to environment
+conditions have been developed and generally perform equivalently to stacked
+SDMs [@algar2009; @distler2015]. This approach is sometimes referred to as
+"macroecological" modeling, because it models the larger-scale pattern
 (richness) directly [@distler2015].

### 2017-08-23 manuscript/draft.Rmd
??
intro edit
We aren't accounting for uncertainty in model choice and eval

@@
 describe individual species' occurrence patterns as functions of the
-environment. Predictions about where each species will occur in the future are
-made by using forecasts of future environmental conditions to generate
-predictions from these models. These species-level predictions are then combined
-("stacked") to generate forecasts for species richness [e.g. @calabrese2014].
-Alternatively, models that directly relate spatial patterns of species richness
-to environment conditions have been developed and generally perform equivalently
-to stacked SDMs [@algar2009; @distler2015]. This approach is sometimes referred
-to as "macroecological" modeling, because it models the larger-scale pattern
-(richness) directly [@distler2015]. [SDMs; @thomas2004; @thuiller2011;
-@urban2015]. In general, these models describe individual species' occurrence
-patterns as functions of the environment. Given forecasts for environmental
-conditions, these models can predict where each species will occur in the
-future. These species-level predictions are then combined ("stacked") to
-generate forecasts for species richness [e.g. @calabrese2014]. Alternatively,
-models that directly relate spatial patterns of species richness to environment
-conditions have been developed and generally perform equivalently to stacked
-SDMs [@algar2009; @distler2015]. This approach is sometimes referred to as
-"macroecological" modeling, because it models the larger-scale pattern
-(richness) directly [@distler2015].
+environment. Given forecasts for environmental conditions, these models can 
+predict where each species will occur in the future. These species-level 
+predictions are then combined ("stacked") to generate forecasts for species 
+richness [e.g. @calabrese2014]. Alternatively, models that directly relate spatial 
+patterns of species richness to environment conditions have been developed and 
+generally perform equivalently to stacked SDMs [@algar2009; @distler2015]. 
+This approach is sometimes referred to as "macroecological" modeling, because it 
+models the larger-scale pattern (richness) directly [@distler2015].
 
@@
 biological and environmental time-series [@tredennick2016]; 2) accounting for
-uncertainty in observations, processes, model choice, and forecast evaluation
-[@clark2001; @dietze2017]; and 3) conducting meaningful evaluations of the
-forecasts by hindcasting, archiving short-term forecasts, and comparing
-forecasts to baselines to determine whether the forecasts are more accurate than
-assuming the system is basically static [@perretti2013].
+uncertainty in observations and processes, [@harris2015; @yu2010]; and 3)
+conducting meaningful evaluations of the forecasts by hindcasting, archiving
+short-term forecasts, and comparing forecasts to baselines to determine whether
+the forecasts are more accurate than assuming the system is basically static
+[@perretti2013].
 

### 2017-08-23 manuscript/draft.Rmd
change stan reference
I think this ref is more appropriate for the core stan software.

@@
 with a linear mixed model, with observer as a random effect, built in the Stan
-probabilistic programming language [@stan]. Because observer and site are
+probabilistic programming language [@carpenter2017]. Because observer and site are
 strongly related (observers tend to repeatedly sample the same site), site was

### 2017-08-23 manuscript/draft.Rmd
fix gbm ref
remove duplicate, and "other" authors which was causing issues.

@@
 **"Macroecological" model: richness GBM.** We used a boosted regression tree
-model using the `gbm` package [@ridgeway2017] to directly model species richness
+model using the `gbm` package [@gbm] to directly model species richness
 as a function of environmental variables. Boosted regression trees are a form of

### 2017-08-23 manuscript/draft.Rmd
clarify fig 3 text
some added clarification to remind readers. this also isn't the strict definition of forecast horison.

@@
 
-![Performance of the six models we evaluated when predicting one year into the
-future (2004) or ten years into the future (2013).  In general, the single-site
-models (**A.**) outperformed the environmental models. (**B.**) The accuracy of
-the predictions generally declined as the forecast horizon was
-extended.](../figures/scatter.png)
+![Performance of the six models, trained with data from 1982-2003, when 
+predicting one year into the future (2004) or ten years into the future (2013).
+In general, the single-site models (**A.**) outperformed the environmental
+models (**B.**). The accuracy of the predictions generally declined as the
+timescale of the forecast was extended from 2004 to 2013.](../figures/scatter.png)
 

### 2017-08-23 manuscript/draft.Rmd
fig 4 clarify

@@
 
-![None of the models provided a consistent improvement over the "average"
-baseline, indicated by a horizontal line. **A.** Models' absolute error was
-generally similar or larger than "average" model, with large outliers in both
-directions. **B.** The other models' deviance was also generally higher than the
-"average" baseline.](../figures/model_violins.png)
+![Site/year difference between the average model and all other models in absolute
+error (**A.**) and deviance (**B.**). All testing years (2004-2013) are used here. 
+None of the five other models provided a consistent improvement over the "average"
+baseline, indicated by a horizontal line. Models' absolute error was generally
+similar or larger than "average" model, with large outliers in both directions.
+The other models' deviance was also generally higher than the "average"
+baseline.](../figures/model_violins.png)
 

### 2017-08-23 manuscript/draft.Rmd
fig 5 clarify

@@
 Note that the stack of single-species random forest SDMs shows much higher error
-than the other models (since its predictions are too confident), and that the
-"naive" model's deviance grows relatively quickly (as its predictions become
-broader and less informative). **C.** Coverage of a model's 95% confidence
-intervals indicates how often the observed values fall inside the predicted
-range. For well-calibrated models, the coverage of the 95% confidence intervals
-should be 95%. As in the previous panel, the "naive" model's predictive
-distribution is too wide (capturing almost all of the data) and the stacked
-SDM's predictive distribution is too narrow (missing almost a third of the
-observed richness values by 2014).](../figures/performance_time.png)
+than the other models (since its predictions are too confident, see Fig. 1), and 
+that the "naive" model's deviance grows relatively quickly (as its predictions
+become broader and less informative, see Fig. 1). **C.** Coverage of a model's
+95% confidence intervals indicates how often the observed values fall inside the
+predicted range. For well-calibrated models, the coverage of the 95% confidence
+intervals should be 95% (indicated by the black line). As in the previous panel,
+the "naive" model's predictive distribution is too wide (capturing almost all of
+the data) and the stacked SDM's predictive distribution is too narrow (missing
+almost a third of the observed richness values by 2014).](../figures/performance_time.png)
 

### 2017-08-24 manuscript/draft.Rmd

@@
 **"Macroecological" model: richness GBM.** We used a boosted regression tree
-model using the `gbm` package [@gbm] to directly model species richness
-as a function of environmental variables. Boosted regression trees are a form of
-tree-based modeling that work by fitting thousands of small tree-structured
-models sequentially, with each tree optimized to reduce the error of its
-predecessors. They are flexible models that are considered well suited for
-prediction [@elith2008]. This model was optimized using a Gaussian likelihood,
-with a maximum interaction depth of 5, shrinkage of 0.015, and up to 10,000
-trees. The number of trees used for prediction was selected using the "out of
-bag" estimator; this number averaged `r numbers$n_gbm[1]` for the non-observer
-data and `r numbers$n_gbm[2]` for the observer-corrected data.
+model using the `gbm` package [Ridgeway *et al.* -@gbm] to directly model
+species richness as a function of environmental variables. Boosted regression
+trees are a form of tree-based modeling that work by fitting thousands of small
+tree-structured models sequentially, with each tree optimized to reduce the
+error of its predecessors. They are flexible models that are considered well
+suited for prediction [@elith2008]. This model was optimized using a Gaussian
+likelihood, with a maximum interaction depth of 5, shrinkage of 0.015, and up to
+10,000 trees. The number of trees used for prediction was selected using the
+"out of bag" estimator; this number averaged `r numbers$n_gbm[1]` for the
+non-observer data and `r numbers$n_gbm[2]` for the observer-corrected data.
 
@@
 [@harris2015], viridis [@viridis], rstan [@rstan], yaml [@yaml], purrr [@purrr],
-gbm [@gbm], randomForest [@randomForest]. Code to fully reproduce this analysis
-is available on GitHub (https://github.com/weecology/bbs-forecasting) and
-archived on Zenodo (https://doi.org/10.5281/zenodo.839581).
+gbm [Ridgeway *et al.* -@gbm], randomForest [@randomForest]. Code to fully
+reproduce this analysis is available on GitHub
+(https://github.com/weecology/bbs-forecasting) and archived on Zenodo
+(https://doi.org/10.5281/zenodo.839581).
 

### 2017-08-24 manuscript/draft.Rmd
no ref needed here.
This is explained further in the 6th discussion paragraph

@@
 validated in terms of repeated observations at the same site. Since site-level
-richness is relatively stable [[refs]], model evaluations that show high
+richness is relatively stable, model evaluations that show high
 accuracy across spatial gradients may not indicate much about a model's ability

### 2017-08-24 manuscript/draft.Rmd
remove duplicate ref, tidy others
minor edit
this didn't make sense

@@
 values were used as predictors. Elevation was from the SRTM 90m elevation
-dataset [@jarvis2008] obtained using the R package raster [@hijmans2016].
+dataset [@jarvis2008] obtained using the R package raster [@raster].
 Because BBS routes are 40-km transects rather than point counts, we used the
@@
 [@rapacciuolo2012; @moran-ordonez2017; @araujo2005; @eskildsen2013] using data
-aggregated across similar time scales to those studied here. More evaluation
+aggregated across similar time scales. More evaluation
 studies like these are needed but there is a paucity of adequate long term data

### 2017-08-24 manuscript/draft.Rmd
fix methods intro
we did not evaluate the long term forecasts

@@
 
-We evaluated 6 types of forecasting models (Table 1) at two time-scales. In the
-first one, we divided our 32 years of data into 22 years of training data
-and retained 10 years of data for evaluation via hindcasting. Second, we used
-the full data set for training and made predictions through the year 2050. For
-both time scales, we evaluated versions of each model with and without
+We evaluated 6 types of forecasting models (Table 1) by dividing the 32 years
+of data into 22 years of training data and retained 10 years of data for 
+evaluation via hindcasting. We also made long term forecasts by using the full
+data set for training and making forecasts through the year 2050. For
+both time scales, we made forecasts using each model with and without
 correcting for observer effects, as described below.

### 2017-08-24 manuscript/draft.Rmd
minor edit for #4

@@
 include predictor variables that are expected to relate to the ecological
-quantity being forecast and dynamic time-series of predictors instead of
-long-term averages. Investing time in identifying and acquiring better predictor
+quantity being forecast. Time-series of predictors, instead of long-term
+averages, are also preferable to match the ecologial data (see #2).
+Investing time in identifying and acquiring better predictor
 variables may have at least as many benefits as using more sophisticated

### 2017-08-24 manuscript/draft.Rmd
typos

@@
 
-Three of models used in this study were fit to each site separately, with no
+Three of the models used in this study were fit to each site separately, with no
 environmental information (Table 1). These models were fit to each BBS route
@@
 environmental variables and using the resulting models to make forecasts. Two
-studied compared the SDM and direct richness modeling approaches, and reported
+studies compared the SDM and direct richness modeling approaches, and reported
 that the methods yielded equivalent results for forecasting diversity

### 2017-08-25 manuscript/draft.Rmd
Minor ms edit for grammar and clarity

@@
 We evaluated 6 types of forecasting models (Table 1) by dividing the 32 years
-of data into 22 years of training data and retained 10 years of data for 
-evaluation via hindcasting. We also made long term forecasts by using the full
-data set for training and making forecasts through the year 2050. For
+of data into 22 years of training data and 10 years of data for 
+evaluating forecasts using hindcasting. We also made long term forecasts by using
+the full data set for training and making forecasts through the year 2050. For
 both time scales, we made forecasts using each model with and without

### 2017-08-26 figures/plots.R
update figures w/ mistnet etc.

@@ -188,12 +188,12 @@ d = bound %>%
             coverage = mean(p > .025 & p < .975)) %>% 
   filter(use_obs_model) %>% 
   gather(key = "variable", value = "value", rmse, mean_deviance, coverage) %>% 
-  mutate(variable = forcats::fct_relevel(variable, "rmse", "mean_deviance", 
-                                         "coverage"),
+  mutate(variable = forcats::fct_relevel(variable, "rmse", "coverage", 
+                                         "mean_deviance"),
          variable = factor(variable, 
                            labels = c("Root mean squared error (RMSE)",
-                                      "Mean deviance",
-                                      "Coverage")))
+                                      "Coverage",
+                                      "Mean deviance")))
 
 # "Color Universal Design" by Okabe and Ito
 # http://jfly.iam.u-tokyo.ac.jp/color/,
@@ -226,7 +226,7 @@ plotlist = pmap(
 time_error_legend = get_legend(plotlist[[1]])
 plotlist = map(plotlist, ~.x + theme(legend.position = "none"))
 
-plotlist[[3]] = plotlist[[3]] + 
+plotlist[[2]] = plotlist[[2]] + 
   geom_hline(yintercept = 0.95) +
   scale_y_continuous(limits = c(.65, 1.0025), expand = c(0, 0))
 
@@ -482,28 +482,28 @@ observer_deviance_data = bound %>%
   mutate(use_obs_model = forcats::fct_recode(factor(use_obs_model), 
                                              no = "FALSE", 
                                              yes = "TRUE")) %>% 
-  select(site_id, year, model, use_obs_model, deviance) %>% 
+  select(site_id, year, formal_name, use_obs_model, deviance) %>% 
   spread(key = use_obs_model, value = deviance) %>% 
-  mutate(y = no - yes)
+  mutate(y = no - yes, formal_name.other = gsub(" ", "\n", formal_name))
 
 observer_error_data = bound %>% 
   mutate(use_obs_model = forcats::fct_recode(factor(use_obs_model), 
                                              no = "FALSE", 
                                              yes = "TRUE")) %>% 
-  select(site_id, year, model, use_obs_model, diff) %>% 
+  select(site_id, year, formal_name, use_obs_model, diff) %>% 
   spread(key = use_obs_model, value = diff) %>% 
-  mutate(y = abs(no) - abs(yes))
+  mutate(y = abs(no) - abs(yes), formal_name.other = gsub(" ", "\n", formal_name))
 
 
 obs_violins = plot_grid(
[…]

### 2017-08-26 manuscript/draft.Rmd

@@
 
-![Performance of the six models, trained with data from 1982-2003, when 
-predicting one year into the future (2004) or ten years into the future (2013).
-In general, the single-site models (**A.**) outperformed the environmental
-models (**B.**). The accuracy of the predictions generally declined as the
-timescale of the forecast was extended from 2004 to 2013.](../figures/scatter.png)
+![Performance of six forecasting models for predicting species richness one year
+(2004) and ten years into the future (2013). Plots show observed vs. predicted
+values for species richness. Models were trained with data from 1982-2003.  In
+general, the single-site models (**A**) outperformed the environmental models
+(**B**). The accuracy of the predictions generally declined as the timescale of
+the forecast was extended from 2004 to 2013.](../figures/scatter.png)
 
-While the models generally did well in absolute terms (Figure 3), none
+While all models generally performed well in absolute terms (Figure 3), none
 consistently outperformed the "average" baseline (Figure 4). The auto-ARIMA was
 generally the best-performing non-baseline model, but in many cases (67% of the
-time), the auto.arima procedure automatically selected a model with only an
-intercept term (i.e. no autoregressive terms, no drift, and no moving average
-terms), making it similar to the "average" model. All five alternatives to the
-"average" model achieved lower error on some of the sites in some years, but
-each one had a higher mean absolute error and higher mean deviance (Figure 4).
+time), the auto.arima procedure selected a model with only an intercept term
+(i.e. no autoregressive terms, no drift, and no moving average terms), making it
+similar to the "average" model. All five alternatives to the "average" model
+achieved lower error on some of the sites in some years, but each one had a
+higher mean absolute error and higher mean deviance (Figure 4).
 
-![Site/year difference between the average model and all other models in absolute
-error (**A.**) and deviance (**B.**). All testing years (2004-2013) are used here. 
-None of the five other models provided a consistent improvement over the "average"
-baseline, indicated by a horizontal line. Models' absolute error was generally
-similar or larger than "average" model, with large outliers in both directions.
-The other models' deviance was also generally higher than the "average"
-baseline.](../figures/model_violins.png)
+![Difference between the forecast error of models and the error of the average
+baseline using both absolute error (**A.**) and deviance (**B.**). Differences
+are taken for each site and testing year so that errors for the same forecast
+are directly compared. The error of the average baseline is by definition zero
+and is indicated by the horizontal gray line.  None of the five models provided
+a consistent improvement over the average baseline. The absolute error of the
+models was generally similar or larger than that of the "average" model, with
+large outliers in both directions.  The deviance of the models was also
+generally higher than the "average" baseline.](../figures/model_violins.png)
 
-![Performance of all six models over time, according to three metrics. **A.**
-Root mean square error (rmse) describes the distance between each model's point
-estimates and the observed richness values. Note that the three environmental
-models tend to show the largest errors. **B.** Deviance (-2 times the Gaussian
-log-likelihood) describes the lack of fit of entire predictive distributions.
-Note that the stack of single-species random forest SDMs shows much higher error
-than the other models (since its predictions are too confident, see Fig. 1), and 
-that the "naive" model's deviance grows relatively quickly (as its predictions
-become broader and less informative, see Fig. 1). **C.** Coverage of a model's
-95% confidence intervals indicates how often the observed values fall inside the
-predicted range. For well-calibrated models, the coverage of the 95% confidence
-intervals should be 95% (indicated by the black line). As in the previous panel,
-the "naive" model's predictive distribution is too wide (capturing almost all of
-the data) and the stacked SDM's predictive distribution is too narrow (missing
-almost a third of the observed richness values by 2014).](../figures/performance_time.png)
+![Change in performance of the six forecasting models with the time scale of the
+forecast (1-10 years into the future). **A.** Root mean square error (rmse; the
+error in the point estimates) shows the three environmental models tending to
+show the largest errors at all time scales and the models getting worse as they
+forecast further into the future at approximately the same rate. **B.** Deviance
+(lack of fit of the entire predictive distribution) shows the stacked species
+distribution models with much higher error than other models and shows that the
+"naive" model's deviance grows relatively quickly. **C.** Coverage of a model's
+95% confidence intervals (how often the observed values fall inside the
+predicted range; the black line indicates ideal performance) shows that the
+"naive" model's predictive distribution is too wide (capturing almost all of the
+data) and the stacked SDM's predictive distribution is too narrow (missing
+almost a third of the observed richness values by
+2014).](../figures/performance_time.png)
 
@@
 falling inside its 95% confidence intervals. This stacked SDM's narrow
-predictive distribution also caused it to have notably higher deviance (Figure
-5B) than the next-worst model, even though its point estimates were not
-unusually bad in terms of RMSE (5A). As discussed elsewhere [@harris2015], this
+predictive distribution caused it to have notably higher deviance (Figure 5B)
+than the next-worst model, even though its point estimates were not unusually
+bad in terms of RMSE (5A). As discussed elsewhere [@harris2015], this
 overconfidence is a product of the assumption in stacked SDMs that errors in the
-species-level predictions are independent. Because they avoided this assumption,
-the GBM-based "macroecological" model and the mistnet JSDM had reasonably
-well-calibrated uncertainty estimates (Figure 5B); as a result, their relative
-performance was higher in terms of deviance than in terms of RMSE. The "naive"
-model was the only model whose confidence intervals were too wide (Figure 5C),
-which can be attributed to the rapid rate at which these intervals expand
-(Figure 1).
+species-level predictions are independent. The GBM-based "macroecological" model
+and the mistnet JSDM had the best calibrated uncertainty estimates (Figure
+5B)and therefore their relative performance was higher in terms of deviance than
+in terms of RMSE. The "naive" model was the only model whose confidence
+intervals were too wide (Figure 5C), which can be attributed to the rapid rate
+at which these intervals expand (Figure 1).
 
-Figure 6 shows each model's squared error, divided into consistent site-level
-biases, versus errors that fluctuated year-to-year. Across all models, most of
-the error came from errors in the site-level mean. The "average" model, which
-was based on the site-level mean, had the lowest error in this regard. In
-contrast, the three environmental models showed the largest biases at the site
-level; this makes sense, given that they could not explicitly distinguish among
-sites with similar climate, NDVI, and elevation.
+Partitioning each model's squared error into consistent errors across sites and
+errors that fluctuated year-to-year shows that the majority of the error comes
+from errors in the site-level mean (Figure 6). The "average" model, which was
+based on the site-level mean, had the lowest error in this regard. In contrast,
+the three environmental models showed the largest biases at the site level; this
+makes sense, given that they could not explicitly distinguish among sites with
+similar climate, NDVI, and elevation.
 
-![Each model's squared error can be divided into two parts. If a model
-consistently over-estimates or under-estimates the richness at a given site,
-this contributes to its error in predicting the site-level mean. The second
-component is based on errors in predicting fluctuations in a site's richness
-over time. Both components of the mean squared error were lower for the
-single-site models than for the environmental models.](../figures/barcharts.png)
+![Partitioning of the squared error for each model into site and year
+components. The site-level mean component shows consistent over or under
+estimates of richness at a site across years. The annual fluctuation compoonent
+shows errors in predicting fluctuations in a site's richness over time. Both
+components of the mean squared error were lower for the single-site models than
+for the environmental models.](../figures/barcharts.png)
 
-When the differences among observers were not explicitly accounted for, nearly
-all metrics of fit worsened (Figure 7). The increased error resulted from a
-small number of forecasts where observer turnover caused a large shift in the
-reported richness values. The naive baseline was less sensitive to these shifts,
-because it largely ignored the richness values reported by observers that had
-retired by the end of the training period (Figure 1). The average model, which
-gave equal weight to observations from the whole training period, showed a
-larger decline in performance -- especially in terms of coverage. The
-performance of the mistnet JSDM was notable here, because its prediction
-intervals retained very good coverage in the presence of uncorrected
-observer-based noise, which we attribute to the JSDM's ability to model this
+Accounting for differences among observers generally improved measures of model
+fit (Figure 7). Improvements primarily resulted from a small number of forecasts
+where observer turnover caused a large shift in the reported richness
+values. The naive baseline was less sensitive to these shifts, because it
+largely ignored the richness values reported by observers that had retired by
+the end of the training period (Figure 1). The average model, which gave equal
+weight to observations from the whole training period, showed a larger decline
+in performance when not accounting for observer effects -- especially in terms
+of coverage. The performance of the mistnet JSDM was notable here, because its
+prediction intervals retained good coverage even when not correcting for
+observer differences, which we attribute to the JSDM's ability to model this
 variation with its latent variables.
@@
 accuracy (especially for the "naive" model), but occasonally changes a
-prediction's accuracy by more than 10 species. These changes are beneficial, on
-average.](../figures/observers.png)
+prediction's accuracy by more than 10 species. As a result the inclusion of
+observer effects results in average improvements in overall model
+fit.](../figures/observers.png)
 

### 2017-08-28 manuscript/draft.Rmd
Edits to Discussion
* Rework first 4 paragraphs back into 3 for better flow
* Combine two time-scale paragraphs and reduce emphasis on observed convergence
* Remove paragraph level headers and improve overall structure and flow instead
* Add reference for not being able to address imperfect detection
* General editorial work throughout

@@
 Breeding Bird Survey (BBS) [@pardieck2017] using the Data Retriever Python
-package [@morris2013] and rdataretriever R package [@mcglinn2017]. The BBS data
-was filtered to exclude all nocturnal, cepuscular, and aquatic species [since
-these species are not well sampled by BBS methods; @hurlbert2005], as well as
-unidentified species, and hybrids. All data from surveys that did not meet BBS
-quality criteria were also excluded.
+package [@morris2013] and rdataretriever R package [@rdataretriever]. The BBS
+data was filtered to exclude all nocturnal, cepuscular, and aquatic species
+[since these species are not well sampled by BBS methods; @hurlbert2005], as
+well as unidentified species, and hybrids. All data from surveys that did not
+meet BBS quality criteria were also excluded.
 
@@
 develop and follow best practices for conducting and evaluating ecological
-forecasts. We have laid out ideas for a number of these practices (Box 1) and
-attempted to implement them in a single study that builds and evaluates
-forecasts of biodiversity in the form of species richness. The results of this
-effort are both promising and humbling. When comparing forecasts across sites,
-many different approaches to forecasting produce reasonable forecasts. If a site
-is predicted to have a high number of species in the future, relative to other
-sites, it generally does. However (after controlling for observer differences),
-none of the methods we evaluated could reliably tell us whether site-level
-richness would increase or decrease over time (Figure 6), which is generally the
-stated purpose of these forecasts. As a result, our baseline models, which did
-not attempt to anticipate changes in richness over time, generally provided the
-best forecasts for future biodiversity.
+forecasts [@clark2001]. We have used a number of these practices (Box 1) in a
+single study that builds and evaluates forecasts of biodiversity in the form of
+species richness. The results of this effort are both promising and
+humbling. When comparing forecasts across sites, many different approaches to
+forecasting produce reasonable forecasts (Figure 3). If a site is predicted to
+have a high number of species in the future, relative to other sites, it
+generally does. However, none of the methods evaluated reliably determined how
+site-level richness changes over time (Figure 6), which is generally the stated
+purpose of these forecasts. As a result, baseline models, which did not attempt
+to anticipate changes in richness over time, generally provided the best
+forecasts for future biodiversity. It should be noted that the scope of
+inference for our results is limited to local scale forecasts for breeding birds
+in North America. As such we cannot draw meaningful conclusions about global
+scale forecasts of biodiversity loss or extinction rates, or about about
+forecasting biodiversity changes in areas of the world like the tropics that are
+expected to be heavily impacted by global change.
 
-These results raise a number of important issues. First, we should be skeptical
-of predictions about future biodiversity from models that have not been
-validated in terms of repeated observations at the same site. Since site-level
-richness is relatively stable, model evaluations that show high
-accuracy across spatial gradients may not indicate much about a model's ability
-to predict changes over time. This issue has also recently been highlighted as a
-potential downside of using SDMs to predict the future locations of individual
-species [@rapacciuolo2012; @oedekoven2017]. This result is particularly sobering
-because SDMs form the main foundation for estimates of the predicted loss of
-biodiversity to climate change [@thomas2004; @thuiller2011; @urban2015].
+The most commonly used methods for forecasting future biodiversity, SDMs and
+macroecological models, produced worse forecasts than time-series models and
+simple baselines. This highlights the need for using hindcasting and time-series
+data for assessing the effectiveness of forecasting methods and suggests that
+predictions about future biodiversity change that have not been validated in
+this way should be viewed with skepticism. Since site-level richness is
+relatively stable, model evaluations that show high accuracy across spatial
+gradients may not indicate that a model is able to predict changes over
+time. This issue has also recently been highlighted as a potential downside of
+using SDMs to predict the future locations of individual species
+[@rapacciuolo2012; @oedekoven2017]. This result is particularly sobering because
+SDMs are one of the main foundations for estimates of the predicted loss of
+biodiversity to climate change [@thomas2004; @thuiller2011; @urban2015]. These
+results also highlight the importance of comparing multiple modeling approaches
+when conducting ecological forecasts, and in particular, the value of comparing
+results to simple baselines to avoid over-interpreting the information present
+in these forecasts [Box 1]. Disciplines that have more mature forecasting
+cultures often do this by reporting "forecast skill", i.e., the improvement in
+the forecast relative to a simple baseline [@jolliffe2003]. We recommend
+following the example of @ye2015 and adopting this approach in future ecological
+forecasting research.
 
-Second, our results highlight the importance of comparing multiple modeling
-approaches when conducting ecological forecasts, and in particular, the value of
-comparing results to simple baselines to avoid over-interpreting the information
-present in these forecasts (Box 1). Disciplines that have more mature
-forecasting cultures often do this by reporting "forecast skill", i.e., the
-improvement in the forecast relative to a simple baseline [@jolliffe2003]. We
-recommend following the advice of @ye2015 and adopting this approach in future
-ecological forecasting research. Our results also highlight the need for using
-hindcasting and time-series data for assessing the effectiveness of forecasting
-methods. This practice allows comparisons to baseline models and also
-facilitates the assessment of predicted dynamics within sites or regions.
-
-One of the main alternatives to stacked SDMs is modeling richness directly using
-environmental variables and using the resulting models to make forecasts. Two
-studies compared the SDM and direct richness modeling approaches, and reported
-that the methods yielded equivalent results for forecasting diversity
-[@algar2009; @distler2015]. While our results generally support the idea of
-rough equivalence between stacked SDMs and direct richness modeling for point
-estimates, they also show that stacked SDMs dramatically understimate the
-uncertainty. For this reason, direct modeling of richness is a better overall
-approach to forecasting richness. A similar result is seen when comparing joint
+When comparing different methods for forecasting our results demonstrate the
+importance of considering uncertainty [Box 1; @clark2001; @dietz2016]. Previous
+comparisons between stacked SDMs and macroecological models reported that the
+methods yielded equivalent results for forecasting diversity [@algar2009;
+@distler2015]. While our results support this equivalence for point estimates,
+they also show that stacked SDMs dramatically underestimate uncertainty. For
+this reason, direct modeling of richness is a better overall approach to
+forecasting richness in this case. A similar result is seen when comparing joint
 species distribution models (JSDMs) to stacked single species distribution
 models. For point estimates, the joint distribution models are roughly
-equivalent to stacked SDMs, but the JSDMs provide much better estimates of
-uncertainty [@harris2015; @warton2015]. In fact, JSDMs and direct richness modeling
-provide some of the best estimates of uncertainty across all modeling
-approaches. This highlights the importance of evaluating models based on their
-uncertainty as well as their mean estimates (Box 1). Not doing so is
-particularly problematic in this case, because the summed species distribution
-models are overly confident, which means that these models could mislead
-ecologists into believing that richness would be restricted to a much narrower
+equivalent to stacked SDMs, but the JSDMs provide better estimates of
+uncertainty [@harris2015; @warton2015]. In fact, JSDMs and direct richness
+modeling provide some of the best estimates of uncertainty across all modeling
+approaches. Failing to consider uncertainty in model comparisons can result
+changes in the choice of methods and can lead to over confident models, which
+could result in predictions that biodiversity would be restricted to a narrower
 range than would actually occur.
 
-[[Note: this paragraph may no longer be accurate!]] For the error metric that
-evaluates both models' point estimates and their uncertainty simultaneously, the
-JSDM and the nonlinear richness model converge towards the time-series baselines
-at the longest time-scales assessed. This convergence at longer time scales
-highlights the importance of considering how forecasting method performance
-changes with the distance into the future being predicted (Box 1). The value of
-this approach, and associated forecast horizons, has been raised for knowing how
-far into the future a model can be effectively used for forecasting
-[@petchey2015], but our results suggest a broader value to this approach for
-considering the potential importance of different models and processes for
-making forecasts at different scales. Our results show that for forecasts 1-3
-years in the future, baselines and time-series based approaches outperform all
-other methods using integrative metrics. At these time scales, environmental
-changes are relatively small and changes in the biota may lag behind changes in
-the environment. However, as the timescale of the forecast increases to a decade
-the JSDMs and environmental richness models get closer to the accuracy of the
-time-series methods and even surpass some of those methods. As the forecast
-distance increases the amount of environmental change is expected to increase
-and the system will have more time to respond, potentially leading to shifts
-towards improved relative performance of the models incorporating environmental
-and ecological information. If this shift with forecast distance in the relative
-importance of different processes and models continues, this could lead to JSDMs
-and environmental richness models outperforming time-series based approaches at
-sufficiently long time scales. These results suggest that assessment of how
-forecast performance changes with time lag is not only important for determining
-how far in the future to forecast, but also for determining which models and
-processes are most relevant for making forecasts at different time scales. For
-example ensemble forecasts could be created which incorporate these time lags
-into the weights for individual models.
+The accuracy for forecasts is expected to decline as forecasts are made further
+into the future, so it is important to evaluate forecasts at different lead
+times [Box 1; @petchey2015]. For deviance (the error metric that evaluates both
+models' point estimates and their uncertainty simultaneously), there is some
+indication that the JSDM and the nonlinear richness model converge towards the
+time-series and baseline models as the lead time of the forecast increases. This
+potential convergence is biologically plausible because at short lead times,
+environmental changes are relatively small and changes in the biota may lag
+behind changes in the environment. However, as the lead time of the forecast
+increases the amount of environmental change should increase and the systems
+will have more time to respond, potentially resulting in shifts towards improved
+relative performance of models incorporating environmental and ecological
+information. This is potentially important for interpreting our results because
+we only evaluated annual forecasts up to a decade into the future. In contrast,
+many ecological forecasts are temporally aggregated to 5-30 year timesteps and
+have lead times of up to a century. These long-term forecasts are difficult to
+evaluate given the small number of century scale datasets, but are important for
+understanding changes in biodiversity at some of the lead times relevant for
+conservation and management. Two studies have assessed models of species
+richness at longer lead times [@algar2009; @distler2015], but the results were
+not compared to baseline or time-series models (in part due to data limitations)
+making them difficult to compare to our results. Given the potential changes
+with lead time of the models and processes are most relevant to forecasting, it
+is important that more long lead time evaluation studies are conducted and that
+the inferences drawn from short-term and long-term forecasting studies be
+compared. For true forecasts, short term forecasts will allow for relatively
+rapid validation of models by researchers without needing to wait decades for
+data to accumulate [@tredennick2016; @dietze2016]. However, since drivers of
+species richness differ at different temporal scales [@rosenzweig1995;
+@white2004; @white2007], short-term forecasts should be viewed as informing, not
+replacing, long-term forecasts [@tredennick2016].
 
-It is also possible that models that include species' relationships to their
-environments or direct environmental constraints on richness will continue to
-produce forecasts that are worse than simply assuming the systems are static.
-This would be expected to occur if richness in these systems is not changing
-over the relevant time scales (e.g. 10-30 years in this study), which would make
-simpler models with no directional change more appropriate. Recent suggestions
-that local scale richness in some systems is not changing directionally at
+While it is possible that models that include species' relationships to their
+environments or direct environmental constraints on richness will provide better
+fits at longer lead times, it is also possible that they will continue to
+produce forecasts that are worse than baselines assuming that the systems are
+static.  This would be expected to occur if richness in these systems is not
+changing over the relevant multi-decadal time scales, which would make simpler
+models with no directional change more appropriate. Recent suggestions that
+local scale richness in some systems is not changing directionally at
 multi-decadal scales supports this possibility [@brown2001; @ernest2001;
-@dornelas2014; @vellend2013]. This lack of change in richness may be expected
-even in the presence of substantial changes in environmental conditions and
-species composition at a site due to the replacement of species from the
-regional pool [@brown2001; @ernest2001]. On average, the Breeding Bird Survey
-sites used in this study show little change in richness [site-level SD of `r
-numbers$resid_sd` species, after controlling for differences among observers;
-see also @lasorte2005]. The absence of rapid change in this dataset is
-beneficial for the absolute accuracy of forecasts across different sites: when a
-past year's richness is already known, it is easy to estimate a future richness.
-The site-level stability of the BBS data also explains why stacked SDMs perform
-relatively well at predicting future richness, despite failing to capture
-changes in richness over time. However, this stability also makes it difficult
-to improve forecasts relative to simple baselines, since those baselines are
-already close to representing what is actually occurring in the system. These
-results suggest that single-site models should be actively considered for
-forecasts of richness and other stable aspects of biodiversity. Our results also
-suggest that future efforts to understand and forecast biodiversity should also
-focus on species composition, since lower-level processes are expected to be
-more dynamic [@ernest2001; @dornelas2014] and contain more useful information
-[@harris2015].
+@vellend2013; @dornelas2014]. A lack of change in richness may be expected even
+in the presence of substantial changes in environmental conditions and species
+composition at a site due to replacement of species from the regional pool
+[@brown2001; @ernest2001]. On average, the Breeding Bird Survey sites used in
+this study show little change in richness [site-level SD of `r numbers$resid_sd`
+species, after controlling for differences among observers; see also
+@lasorte2005]. The absence of rapid change in this dataset is beneficial for the
+absolute accuracy of forecasts across different sites: when a past year's
+richness is already known, it is easy to estimate future richness.  The
+site-level stability of the BBS data also explains why SDMs and macroecological
+models perform relatively well at predicting future richness, despite failing to
+capture changes in richness over time. However, this stability also makes it
+difficult to improve forecasts relative to simple baselines, since those
+baselines are already close to representing what is actually occurring in the
+system. These results suggest that single-site models should be actively
+considered for forecasts of richness and other stable aspects of
+biodiversity. Our results also suggest that future efforts to understand and
+forecast biodiversity should incorporate species composition, since lower-level
+processes are expected to be more dynamic [@ernest2001; @dornelas2014] and
+contain more useful information [@harris2015].
 
-**Observer effects.** In addition to consideration of the different process
-models used for forecasting it is important to consider the observation models.
-When working with any ecological dataset, there are imperfections in the
-sampling process that have the potential to influence results. With large scale
-surveys and citizen science datasets, such as the Breeding Bird Survey, these
-issues are potentially magnified by the large number of different observers and
-by major differences in the habitats and species being surveyed [@sauer1994].
-Accounting for differences in observers reduced the average error in our point
-estimates and also improved the coverage of the confidence intervals. In
-addition, controlling for observer effects resulted in changes in which models
-performed best, most notably reducing the relative performance of the naive
-model's point estimates. This suggests that, prior to accounting for observer
-effects, the naive model performed well because it was capable of accommodating
-rapid shifts in estimated richness introduced by changes in the observer. These
-kinds of rapid changes were difficult for the other single-site models to
-accommodate, and so the average performance of the ARIMA and average models
-improved once this source of observation error was addressed. This demonstrates
-that modeling observation error can be important for properly estimating and
-reducing uncertainty in forecasts and can also lead to changes in the best
-methods for forecasting. We did not address differences in detection probability
-across species and sites [@boulinier1998] since there is no clear way to address
-this issue without making strong assumptions about the data (i.e., assuming
-there is no biological variation in stops along a route), but this would be a
-valuable addition to future forecasting models.
-
-**Environmental uncertainty.** Future biodiversity forecasting efforts will also
-need to begin to address the additional uncertainty introduced by the error
-inherent in the process of forecasting the environmental conditions that
-ecologists use as predictor variables. In this, and other hindcasting studies,
-the environmental conditions for the "future" are known because the data has
-already been observed. However, in real forecasts the environmental conditions
+Future biodiversity forecasting efforts also need to address the uncertainty
+introduced by the error in forecasting the environmental conditions that are
+used as predictor variables. In this, and other hindcasting studies, the
+environmental conditions for the "future" are known because the data has already
+been observed. However, in real forecasts the environmental conditions
 themselves have to be predicted, and environmental forecasts will also have
 uncertainty and bias. Ultimately, ecological forecasts that use environmental
-data will therefore be more uncertain than our current hindcasting efforts.
-Thus, it is important to correctly incorporate this uncertainty in the predictor
-variables into forecasting models [@clark2001; @dietze2017], as we did with the
-uncertainty in observer-level effects. Difficulty in forecasting future
-environmental conditions at small scales will present continued challenges for
-models incorporating environmental variables, and this may result in a continued
-advantage for simple single-site approaches.
+data will therefore be more uncertain than our current hindcasting efforts. It
+is important to correctly incorporate this uncertainty in the predictor
+variables into forecasting models [@clark2001; @dietze2017]. Limitations in
+forecasting future environmental conditions, particularly at small scales, will
+present continued challenges for models incorporating environmental variables,
+and this may result in a continued advantage for simple single-site approaches.
 
-**Multi-timescale forecasts.** Another challenge for ecological forecasting is
-the timescales of typical forecasts. We evaluated forecasts using yearly
-timesteps up to a decade into the future. In contrast, many ecological forecasts
-are temporally aggregated to 5-30 year timesteps and projected up to a century
-into the future. These are commonly beyond the career or even lifespan of the
-researchers, which makes it impossible to directly validate the models'
-predictions. Currently, assessments at these long time-scales can only be made
-with a small number of opportune datasets. For example, hindcasts have been used
-to assess models of species richness [@algar2009; @distler2015] and distribution
-[@rapacciuolo2012; @moran-ordonez2017; @araujo2005; @eskildsen2013] using data
-aggregated across similar time scales. More evaluation
-studies like these are needed but there is a paucity of adequate long term data
-available. Short term forecasts will allow for relatively rapid validation of
-models by researchers without needing to wait for enough data to accumulate
-[@tredennick2016; @dietze2016]. This does not mean that they should replace
-large scale long-term forecasts. As discussed above, drivers of species richness
-are likely to differ at different temporal scales [@rosenzweig1995; @white2004;
-@white2007]. Thus, short-term forecasts should be expected to add value to
-scientists and policymakers, and inform (not replace) long-term forecasts.
-Hopefully, by evaluating trends in the shorter term forecasts as done here it
-will be possible to also use the information from short-term forecasting to
-improve longer-term estimates.
+In addition to comparing and improving the process models used for forecasting
+it is important to consider the observation models.  When working with any
+ecological dataset, there are imperfections in the sampling process that have
+the potential to influence results. With large scale surveys and citizen science
+datasets, such as the Breeding Bird Survey, these issues are potentially
+magnified by the large number of different observers and by major differences in
+the habitats and species being surveyed [@sauer1994]. Accounting for differences
+in observers reduced the average error in our point estimates and also improved
+the coverage of the confidence intervals. In addition, controlling for observer
+effects resulted in changes in which models performed best, most notably
+reducing the relative performance of the naive model's point estimates. This
+suggests that, prior to accounting for observer effects, the naive model
+performed well because it was capable of accommodating rapid shifts in estimated
+richness introduced by changes in the observer. These kinds of rapid changes
+were difficult for the other single-site models to accommodate, and so the
+average performance of the ARIMA and average models improved once this source of
+observation error was addressed. This demonstrates that modeling observation
+error can be important for properly estimating and reducing uncertainty in
+forecasts and can also lead to changes in the best methods for forecasting [Box
+1]. Another key aspect of an ideal observation model is imperfect detection. In
+this study, we did not address differences in detection probability across
+species and sites [@boulinier1998] since there is no clear way to address this
+issue using NA Breeding Bird Survey data without making strong assumptions about
+the data [i.e., assuming there is no biological variation in stops along a
+route; @white2010], but this would be a valuable addition to future forecasting
+models.
 
-**Conclusions.** The science of forecasting biodiversity, and ecology more
-broadly, remains in its infancy and it is important to consider the general
-inability of forecasting methods to improve on simple baselines in that context.
-When weather forecasting first started the forecasts were likewise worse than
-simple baselines [@mcgill2012]. One practice that helped weather forecasts to
-improve was publicly making a large number of predictions, which allowed
-different forecasting approaches to be regularly assessed and refined
-[@mcgill2012; @silver2012]. This suggests that it is important for ecologists to
-start regularly making and evaluating ecological forecasts on real data, even if
-they perform poorly, and to make these forecasts publicly available for
-assessment. These forecasts should include both short-term predictions, which
-can be assessed quickly, and mid- to long-term forecasts, which can help
-ecologists to assess long time-scale processes and determine how far into the
-future we can successfully forecast [@tredennick2016; @dietze2017]. Forecasts
-based on the models in this paper (from now until 2050) are openly archived on
-Zenodo, so that we and others can assess how well they perform. We also plan to
-evaluate these forecasts and report the results as each new year of BBS data
-becomes available at the same URL [@zenodo]. Weather forecasting has continually
-improved throughout its history [@bauer2015], in part due to making and
-evaluating public forecasts [@mcgill2012], and we hope this practice will help
-ecology do the same.
+The science of forecasting biodiversity remains in its infancy and it is
+important to consider weaknesses in current forecasting methods in that
+context. In the beginning weather forecasts were also worse than simple
+baselines, but these forecasts have continually improved throughout the history
+of the field [@mcgill2012; @silver2012; @bauer2015]. One practice that lead to
+improvements in weather forecasts was that large numbers of forecasts were made
+publicly, allowing different approaches to be regularly assessed and refined
+[@mcgill2012; @silver2012]. To facilitate this kind of improvement, it is
+important for ecologists to start regularly making and evaluating real
+ecological forecasts, even if they perform poorly, and to make these forecasts
+openly available for assessment [@mcgill2012; @dietze2016]. These forecasts
+should include both short-term predictions, which can be assessed quickly, and
+mid- to long-term forecasts, which can help ecologists to assess long time-scale
+processes and determine how far into the future we can successfully forecast
+[@dietze2016; @tredennick2016]. Forecasts out to 2050 based on the models in
+this paper have been openly archived [@zenodo], so that we and others can assess
+how well they perform. We plan to evaluate these forecasts and report the
+results as each new year of BBS data becomes available, and make iterative
+improvements to the forecasting models in response to these assessments.
 
 Making successful ecological forecasts will be challenging. Ecological systems
-are complex, fundamental theory is less refined than simpler physical/chemical
-systems, and we currently lack the large amounts of data that produce effective
-forecasts through machine learning. Despite this, we believe that progress can be
-made if we build an active forecasting culture in ecology that builds and
-assesses forecasts in ways that will allow us to improve the effectiveness of
-ecological forecasts most rapidly (Box 1). This includes expanding the scope of
-the ecological and environmental data we work with, paying attention to
-uncertainty in both model building and forecast evaluation, and rigorously
-assessing forecasts using a combination of hindcasting, archived forecasts, and
-comparisons to simple baselines.
+are complex, our fundamental theory is less refined than for simpler physical
+and chemical systems, and we currently lack the scale of data that often
+produces effective forecasts through machine learning. Despite this, we believe
+that progress can be made if we develop an active forecasting culture in ecology
+that builds and assesses forecasts in ways that will allow us to improve the
+effectiveness of ecological forecasts more rapidly [Box 1; @mcgill2012;
+@dietze2016]. This includes expanding the scope of the ecological and
+environmental data we work with, paying attention to uncertainty in both model
+building and forecast evaluation, and rigorously assessing forecasts using a
+combination of hindcasting, archived forecasts, and comparisons to simple
+baselines.
 

### 2017-08-30 manuscript/draft.Rmd
Minor edits to the box
1. Change title to "Best practices" since that's how they are described in the
   paper
2. Add some references
3. Minor editorial work

--- 2017-08-26 manuscript/draft.Rmd
@@
 
-## Box 1: Ten simple rules for making and evaluating ecological forecasts
+## Box 1: Best practices for making and evaluating ecological forecasts
 
@@
 can learn more rapidly about the best approaches for making predictions for a
-given ecological quantity. This includes comparing process-based [e.g.
-@kearney2009] and data-driven models, as well as comparing the accuracy of
-forecasts to simple baselines to determine if the modeled forecasts are more
-accurate than the naive assumption that the world is static [@ye2015].
+given ecological quantity [@clark2001; @ward2014]. This includes comparing
+process-based [e.g., @kearney2009] and data-driven models [e.g., @ward2014], as
+well as comparing the accuracy of forecasts to simple baselines to determine if
+the modeled forecasts are more accurate than the naive assumption that the world
+is static [@jolliffe2003; @ye2015].
 
@@
 Forecasts describe how systems are expected to change through time. While some
-areas of ecological forecasting focus primarily on time-series data, others
-primarily focus on using spatial models and space-for-time substitutions. Using
-ecological and environmental time-series data allows the consideration of actual
-dynamics from both a process and error structure perspective [@tredennick2016].
+areas of ecological forecasting focus primarily on time-series data [@ward2014],
+others primarily focus on using spatial models and space-for-time substitutions
+[ref?]. Using ecological and environmental time-series data allows the
+consideration of actual dynamics from both a process and error structure
+perspective [@tredennick2016].
 
@@
 the average or expected outcome. Failing to account for uncertainty can result
-in overconfidence in highly uncertain outcomes leading to poor decision making
-and erosion of confidence in ecological forecasts. Models should explicitly
-include sources of uncertainty and propagate them through the forecast where
-possible [@clark2001; @dietze2017]. Evaluations of forecasts should assess the
-accuracy of models' estimated uncertainties as well as their point estimates
-[@dietze2017].
+in overconfidence in uncertain outcomes leading to poor decision making and
+erosion of confidence in ecological forecasts [@clark2001]. Models should
+explicitly include sources of uncertainty and propagate them through the
+forecast where possible [@clark2001; @dietze2017]. Evaluations of forecasts
+should assess the accuracy of models' estimated uncertainties as well as their
+point estimates [@dietze2017].
 
@@
 In general, the accuracy of forecasts decreases with the length of time into the
-future being forecast [@petchey2015]. This decay in accuracy (and the potential
-for different rates of decay to result in different relative model performance
-at different lead times) should be considered when evaluating forecasts
-and comparing models.
+future being forecast [@petchey2015]. This decay in accuracy should be
+considered when evaluating forecasts. In addition to simple decreases in
+forecast accuracy the potential for different rates of decay to result in
+different relative model performance at different lead times should be
+considered.
 
@@
 possible, forecasts should model the factors influencing the observation of the
-data. [@yu2010; @hutchinson2011; @schurr2012]
+data [@yu2010; @hutchinson2011; @schurr2012].
 
@@
 
-Evalutating a model's predictive performance across time is critical.
-Hindcasting uses a temporal out-of-sample validation approach to mimic how well
-a model would have performed had it been run in the past. For example, using
-occurance data from the early 20th century to model distributions which are
-validated with late 20th century occurances. Dense time series, such as yearly
-observations, are desirable to also evalulate the forecast horizon (see #6), but
-this is not a strict requirement.
+Evalutating a model's predictive performance across time is critical for
+understanding if it is useful for forecasting the future. Hindcasting uses a
+temporal out-of-sample validation approach to mimic how well a model would have
+performed had it been run in the past. For example, using occurance data from
+the early 20th century to model distributions which are validated with late 20th
+century occurances. Dense time series, such as yearly observations, are
+desirable to also evalulate the forecast horizon (see #6), but this is not a
+strict requirement.
 
@@
 Forecast values and/or models should be archived so that they can be assessed
-after new data is generated [@mcgill2012; @silver2012]. Enough information
-should be provided in the archive to allow ecologists to make an unambiguous
-assessment of each forecast's performance.
+after new data is generated [@mcgill2012; @silver2012; @dietze2016]. Enough
+information should be provided in the archive to allow unambiguous assessment of
+each forecast's performance [@tetlock2016].
 
@@
 
-In cases where long-term predictions are the primary goal, short-term
+Even in cases where long-term predictions are the primary goal, short-term
 predictions should also be made to accommodate the time-scales of planning and
 management decisions and to allow the accuracy of the forecasts to be quickly
-evaluated [@tredennick2016].
+evaluated [@dietze2016; @tredennick2016].
 

### 2017-09-01 manuscript/draft.Rmd
Add citation to Blonder 2017 regarding lags in prediction

--- 2017-08-28 manuscript/draft.Rmd
@@
 relative performance of models incorporating environmental and ecological
-information. This is potentially important for interpreting our results because
-we only evaluated annual forecasts up to a decade into the future. In contrast,
-many ecological forecasts are temporally aggregated to 5-30 year timesteps and
-have lead times of up to a century. These long-term forecasts are difficult to
-evaluate given the small number of century scale datasets, but are important for
-understanding changes in biodiversity at some of the lead times relevant for
-conservation and management. Two studies have assessed models of species
-richness at longer lead times [@algar2009; @distler2015], but the results were
-not compared to baseline or time-series models (in part due to data limitations)
-making them difficult to compare to our results. Given the potential changes
-with lead time of the models and processes are most relevant to forecasting, it
-is important that more long lead time evaluation studies are conducted and that
-the inferences drawn from short-term and long-term forecasting studies be
-compared. For true forecasts, short term forecasts will allow for relatively
-rapid validation of models by researchers without needing to wait decades for
-data to accumulate [@tredennick2016; @dietze2016]. However, since drivers of
-species richness differ at different temporal scales [@rosenzweig1995;
-@white2004; @white2007], short-term forecasts should be viewed as informing, not
-replacing, long-term forecasts [@tredennick2016].
+information [@blonder2017]. This is potentially important for interpreting our
+results because we only evaluated annual forecasts up to a decade into the
+future. In contrast, many ecological forecasts are temporally aggregated to 5-30
+year timesteps and have lead times of up to a century. These long-term forecasts
+are difficult to evaluate given the small number of century scale datasets, but
+are important for understanding changes in biodiversity at some of the lead
+times relevant for conservation and management. Two studies have assessed models
+of species richness at longer lead times [@algar2009; @distler2015], but the
+results were not compared to baseline or time-series models (in part due to data
+limitations) making them difficult to compare to our results. Given the
+potential changes with lead time of the models and processes are most relevant
+to forecasting, it is important that more long lead time evaluation studies are
+conducted and that the inferences drawn from short-term and long-term
+forecasting studies be compared. For true forecasts, short term forecasts will
+allow for relatively rapid validation of models by researchers without needing
+to wait decades for data to accumulate [@tredennick2016; @dietze2016]. However,
+since drivers of species richness differ at different temporal scales
+[@rosenzweig1995; @white2004; @white2007], short-term forecasts should be viewed
+as informing, not replacing, long-term forecasts [@tredennick2016].
 

### 2017-09-01 manuscript/draft.Rmd
Add citation to Blois et al. 2013 for space-for-time box item

@@
 areas of ecological forecasting focus primarily on time-series data, others
-primarily focus on using spatial models and space-for-time substitutions. Using
-ecological and environmental time-series data allows the consideration of actual
-dynamics from both a process and error structure perspective [@tredennick2016].
+primarily focus on using spatial models and space-for-time substitutions
+[@blois2013]. Using ecological and environmental time-series data allows the
+consideration of actual dynamics from both a process and error structure
+perspective [@tredennick2016].
 

### 2017-09-01 manuscript/draft.Rmd

--- 2017-09-01 manuscript/draft.Rmd
@@
 Forecasts describe how systems are expected to change through time. While some
-areas of ecological forecasting focus primarily on time-series data, others
-primarily focus on using spatial models and space-for-time substitutions. Using
-ecological and environmental time-series data allows the consideration of actual
-dynamics from both a process and error structure perspective [@tredennick2016].
+areas of ecological forecasting focus primarily on time-series data [@ward2014], 
+others primarily focus on using spatial models and space-for-time substitutions
+[@blois2013]. Using ecological and environmental time-series data allows the
+consideration of actual dynamics from both a process and error structure
+perspective [@tredennick2016].
 

### 2017-09-03 manuscript/draft.Rmd
early discussion

@@
 that current methods (such as species distribution models) can successfully turn
-spatial data into useful temporal predictions about biodiversity at decadale
+spatial data into useful temporal predictions about biodiversity at decadal
 time-scales. The best practices implemented in this study directly influence the
@@
 single study that builds and evaluates forecasts of biodiversity in the form of
-species richness. The results of this effort are both promising and
-humbling. When comparing forecasts across sites, many different approaches to
-forecasting produce reasonable forecasts (Figure 3). If a site is predicted to
-have a high number of species in the future, relative to other sites, it
-generally does. However, none of the methods evaluated reliably determined how
-site-level richness changes over time (Figure 6), which is generally the stated
-purpose of these forecasts. As a result, baseline models, which did not attempt
-to anticipate changes in richness over time, generally provided the best
-forecasts for future biodiversity. It should be noted that the scope of
-inference for our results is limited to local scale forecasts for breeding birds
-in North America. As such we cannot draw meaningful conclusions about global
-scale forecasts of biodiversity loss or extinction rates, or about about
-forecasting biodiversity changes in areas of the world like the tropics that are
-expected to be heavily impacted by global change.
+species richness. The results of this effort are both promising and humbling.
+When comparing forecasts across sites, many different approaches to forecasting
+produce reasonable forecasts (Figure 3). If a site is predicted to have a high
+number of species in the future, relative to other sites, it generally does.
+However, none of the methods evaluated reliably determined how site-level
+richness changes over time (Figure 6), which is generally the stated purpose of
+these forecasts. As a result, baseline models, which did not attempt to
+anticipate changes in richness over time, generally provided the best forecasts
+for future biodiversity. While this study was based on a single data set, its
+results are consistent with a growing literature on the limits of ecological 
+forecasting, as discussed below.
 
 The most commonly used methods for forecasting future biodiversity, SDMs and
-macroecological models, produced worse forecasts than time-series models and
-simple baselines. This highlights the need for using hindcasting and time-series
-data for assessing the effectiveness of forecasting methods and suggests that
-predictions about future biodiversity change that have not been validated in
-this way should be viewed with skepticism. Since site-level richness is
-relatively stable, model evaluations that show high accuracy across spatial
-gradients may not indicate that a model is able to predict changes over
-time. This issue has also recently been highlighted as a potential downside of
-using SDMs to predict the future locations of individual species
-[@rapacciuolo2012; @oedekoven2017]. This result is particularly sobering because
-SDMs are one of the main foundations for estimates of the predicted loss of
-biodiversity to climate change [@thomas2004; @thuiller2011; @urban2015]. These
-results also highlight the importance of comparing multiple modeling approaches
-when conducting ecological forecasts, and in particular, the value of comparing
-results to simple baselines to avoid over-interpreting the information present
-in these forecasts [Box 1]. Disciplines that have more mature forecasting
-cultures often do this by reporting "forecast skill", i.e., the improvement in
-the forecast relative to a simple baseline [@jolliffe2003]. We recommend
-following the example of @ye2015 and adopting this approach in future ecological
-forecasting research.
+macroecological models, both produced worse forecasts than time-series models
+and simple baselines. This weakness suggests that predictions about future
+biodiversity change should be viewed with skepticism unless the underlying
+models have been validated temporally, via hindcasting. Since site-level
+richness is relatively stable, spatial validation is not enough: a model can
+have high accuracy across spatial gradients without being able to predict
+changes over time. This gap between spatial and temporal accuracy is known to be
+important for species-level predictions [@rapacciuolo2012; @oedekoven2017]; our
+results indicate that it is substantial for higher-level patterns like richness
+as well. SDMs' poor temporal predictions are particularly sobering, as these
+models have been one of the main foundations for estimates of the predicted loss
+of biodiversity to climate change over the past decade or so [@thomas2004;
+@thuiller2011; @urban2015]. Our results also highlight the importance of
+comparing multiple modeling approaches when conducting ecological forecasts, and
+in particular, the value of comparing results to simple baselines to avoid
+over-interpreting the information present in these forecasts [Box 1].
+Disciplines that have more mature forecasting cultures often do this by
+reporting "forecast skill", i.e., the improvement in the forecast relative to a
+simple baseline [@jolliffe2003]. We recommend following the example of @ye2015
+and adopting this approach in future ecological forecasting research.
 
@@
 @distler2015]. While our results support this equivalence for point estimates,
-they also show that stacked SDMs dramatically underestimate uncertainty. For
-this reason, direct modeling of richness is a better overall approach to
-forecasting richness in this case. A similar result is seen when comparing joint
-species distribution models (JSDMs) to stacked single species distribution
-models. For point estimates, the joint distribution models are roughly
-equivalent to stacked SDMs, but the JSDMs provide better estimates of
-uncertainty [@harris2015; @warton2015]. In fact, JSDMs and direct richness
-modeling provide some of the best estimates of uncertainty across all modeling
-approaches. Failing to consider uncertainty in model comparisons can result
-changes in the choice of methods and can lead to over confident models, which
-could result in predictions that biodiversity would be restricted to a narrower
-range than would actually occur.
+they also show that stacked SDMs dramatically underestimate the range of
+possible outcomes; after ten years, more than a third of the observed richness
+values fell outside the stacked SDMs' 95% prediction intervals. Consistent with
+@harris2015 and @warton2015, we found that JSDMs' wider prediction intervals
+enabled them to avoid this problem. Macroecological models appear to share this
+advantage, while being considerably easier to implement.
 

### 2017-09-03 manuscript/draft.Rmd
mention baselines earlier

@@
 biodiversity change should be viewed with skepticism unless the underlying
-models have been validated temporally, via hindcasting. Since site-level
-richness is relatively stable, spatial validation is not enough: a model can
-have high accuracy across spatial gradients without being able to predict
-changes over time. This gap between spatial and temporal accuracy is known to be
-important for species-level predictions [@rapacciuolo2012; @oedekoven2017]; our
-results indicate that it is substantial for higher-level patterns like richness
-as well. SDMs' poor temporal predictions are particularly sobering, as these
-models have been one of the main foundations for estimates of the predicted loss
-of biodiversity to climate change over the past decade or so [@thomas2004;
-@thuiller2011; @urban2015]. Our results also highlight the importance of
-comparing multiple modeling approaches when conducting ecological forecasts, and
-in particular, the value of comparing results to simple baselines to avoid
-over-interpreting the information present in these forecasts [Box 1].
-Disciplines that have more mature forecasting cultures often do this by
-reporting "forecast skill", i.e., the improvement in the forecast relative to a
-simple baseline [@jolliffe2003]. We recommend following the example of @ye2015
-and adopting this approach in future ecological forecasting research.
+models have been validated temporally, via hindcasting and comparison with
+simple baselines. Since site-level richness is relatively stable, spatial
+validation is not enough: a model can have high accuracy across spatial
+gradients without being able to predict changes over time. This gap between
+spatial and temporal accuracy is known to be important for species-level
+predictions [@rapacciuolo2012; @oedekoven2017]; our results indicate that it is
+substantial for higher-level patterns like richness as well. SDMs' poor temporal
+predictions are particularly sobering, as these models have been one of the main
+foundations for estimates of the predicted loss of biodiversity to climate
+change over the past decade or so [@thomas2004; @thuiller2011; @urban2015]. Our
+results also highlight the importance of comparing multiple modeling approaches
+when conducting ecological forecasts, and in particular, the value of comparing
+results to simple baselines to avoid over-interpreting the information present
+in these forecasts [Box 1]. Disciplines that have more mature forecasting
+cultures often do this by reporting "forecast skill", i.e., the improvement in
+the forecast relative to a simple baseline [@jolliffe2003]. We recommend
+following the example of @ye2015 and adopting this approach in future ecological
+forecasting research.
 

### 2017-09-03 manuscript/draft.Rmd
update figure captions

--- 2017-09-01 manuscript/draft.Rmd
@@
 
-![**A.** Model predictions for [[STATE]] route [[ROUTE NUMBER]] when all
-observers are treated the same (black points). **B.** Model predictions for the
-same route when accounting for systematic differences between observers
-(represented by the points' colors). In this example most models are made more
-robust to observer turnover. Note that the "naive" model is less sensitive to
-observer turnover, and does not benefit as much from modeling
-it.](../figures/observer_predictions.png)
+![**A.** Model predictions for route 72035 when all observers are treated the
+same (black points). **B.** Model predictions for the same route when accounting
+for systematic differences between observers (represented by the points'
+colors). In this example most models are made more robust to observer turnover.
+Note that the "naive" model is less sensitive to observer turnover, and does not
+benefit as much from modeling it.](../figures/observer_predictions.png)
 
@@
 fit (Figure 7). Improvements primarily resulted from a small number of forecasts
-where observer turnover caused a large shift in the reported richness
-values. The naive baseline was less sensitive to these shifts, because it
-largely ignored the richness values reported by observers that had retired by
-the end of the training period (Figure 1). The average model, which gave equal
-weight to observations from the whole training period, showed a larger decline
-in performance when not accounting for observer effects -- especially in terms
-of coverage. The performance of the mistnet JSDM was notable here, because its
-prediction intervals retained good coverage even when not correcting for
-observer differences, which we attribute to the JSDM's ability to model this
-variation with its latent variables.
+where observer turnover caused a large shift in the reported richness values
+(Appendix [[Figure Number]]). The naive baseline was less sensitive to these
+shifts, because it largely ignored the richness values reported by observers
+that had retired by the end of the training period (Figure 1). The average
+model, which gave equal weight to observations from the whole training period,
+showed a larger decline in performance when not accounting for observer effects
+-- especially in terms of coverage. The performance of the mistnet JSDM was
+notable here, because its prediction intervals retained good coverage even when
+not correcting for observer differences, which we attribute to the JSDM's
+ability to model this variation with its latent variables.
 
-![Accounting for differences between observers typically has a small impact on
-accuracy (especially for the "naive" model), but occasonally changes a
-prediction's accuracy by more than 10 species. As a result the inclusion of
-observer effects results in average improvements in overall model
-fit.](../figures/observers.png)
+![Controlling for differences among observers generally improved each model's
+predictions, on average. The magnitude of this effect was negligible for the
+Naive baseline, however.](../figures/obs_arrows.png)
 

### 2017-09-04 manuscript/draft.Rmd
fix route number
make caveat more specific

@@
 anticipate changes in richness over time, generally provided the best forecasts
-for future biodiversity. While this study was based on a single data set, its
-results are consistent with a growing literature on the limits of ecological 
-forecasting, as discussed below.
+for future biodiversity. While this study is restricted to breeding birds in
+North America, its results are consistent with a growing literature on the
+limits of ecological forecasting, as discussed below.
 

### 2017-09-04 manuscript/draft.Rmd
fix dietze spelling

@@
 
-![**A.** Model predictions for route 72035 when all observers are treated the
-same (black points). **B.** Model predictions for the same route when accounting
-for systematic differences between observers (represented by the points'
-colors). In this example most models are made more robust to observer turnover.
-Note that the "naive" model is less sensitive to observer turnover, and does not
-benefit as much from modeling it.](../figures/observer_predictions.png)
+![**A.** Model predictions for Pennsylvania route 35 when all observers are
+treated the same (black points). **B.** Model predictions for the same route
+when accounting for systematic differences between observers (represented by the
+points' colors). In this example most models are made more robust to observer
+turnover. Note that the "naive" model is less sensitive to observer turnover,
+and does not benefit as much from modeling
+it.](../figures/observer_predictions.png)
 
@@
 When comparing different methods for forecasting our results demonstrate the
-importance of considering uncertainty [Box 1; @clark2001; @dietz2016]. Previous
+importance of considering uncertainty [Box 1; @clark2001; @dietze2016]. Previous
 comparisons between stacked SDMs and macroecological models reported that the

### 2017-09-05 manuscript/draft.Rmd
Discussion edits

@@
 
-The accuracy for forecasts is expected to decline as forecasts are made further
-into the future, so it is important to evaluate forecasts at different lead
-times [Box 1; @petchey2015]. For deviance (the error metric that evaluates both
-models' point estimates and their uncertainty simultaneously), there is some
-indication that the JSDM and the nonlinear richness model converge towards the
-time-series and baseline models as the lead time of the forecast increases. This
-potential convergence is biologically plausible because at short lead times,
-environmental changes are relatively small and changes in the biota may lag
-behind changes in the environment. However, as the lead time of the forecast
-increases the amount of environmental change should increase and the systems
-will have more time to respond, potentially resulting in shifts towards improved
-relative performance of models incorporating environmental and ecological
-information [@blonder2017]. This is potentially important for interpreting our
-results because we only evaluated annual forecasts up to a decade into the
-future. In contrast, many ecological forecasts are temporally aggregated to 5-30
-year timesteps and have lead times of up to a century. These long-term forecasts
-are difficult to evaluate given the small number of century scale datasets, but
-are important for understanding changes in biodiversity at some of the lead
-times relevant for conservation and management. Two studies have assessed models
-of species richness at longer lead times [@algar2009; @distler2015], but the
-results were not compared to baseline or time-series models (in part due to data
-limitations) making them difficult to compare to our results. Given the
-potential changes with lead time of the models and processes are most relevant
-to forecasting, it is important that more long lead time evaluation studies are
-conducted and that the inferences drawn from short-term and long-term
-forecasting studies be compared. For true forecasts, short term forecasts will
-allow for relatively rapid validation of models by researchers without needing
-to wait decades for data to accumulate [@tredennick2016; @dietze2016]. However,
+We have only evaluated annual forecasts up to a decade into the future, but
+forecasts are often made with a lead time of 50 years or more. These long-term
+forecasts are difficult to evaluate given the small number of century-scale
+datasets, but are important for understanding changes in biodiversity at some of
+the lead times relevant for conservation and management. Two studies have
+assessed models of species richness at longer lead times [@algar2009;
+@distler2015], but the results were not compared to baseline or time-series
+models (in part due to data limitations) making them difficult to compare to our
+results directly. Given the potential changes with lead time of the models and
+processes are most relevant to forecasting, it is important that more forecasts
+with longer lead times are evaluated and that the inferences drawn from
+short-term and long-term forecasting studies be compared. By evaluating
+shorter-term forecasts, researchers can avoid waiting so many decades to
+evaluate their methods [@petchey2015; @tredennick2016; @dietze2016]. However,
 since drivers of species richness differ at different temporal scales
-[@rosenzweig1995; @white2004; @white2007], short-term forecasts should be viewed
-as informing, not replacing, long-term forecasts [@tredennick2016].
+[@rosenzweig1995; @white2004; @white2007], short-term evaluations should be
+viewed as informing, not replacing, long-term evaluations [@tredennick2016]. In
+particular, we might expect environmental and ecological information to become
+more important at longer time scales, and thus for the performance of simple
+baseline forecasts to degrade faster than forecasts from SDMs and other similar
+models [@blonder2017].  On the other hand, there is evidence that species
+respond differently to spatial versus temporal variation in the environment
+[@oedekoven2017], implying that most SDMs are not using environmental
+information well enough to make accurate forecasts.
 
@@
 fits at longer lead times, it is also possible that they will continue to
-produce forecasts that are worse than baselines assuming that the systems are
+produce forecasts that are worse than baselines that assume the systems are
 static.  This would be expected to occur if richness in these systems is not
@@
 system. These results suggest that single-site models should be actively
-considered for forecasts of richness and other stable aspects of
-biodiversity. Our results also suggest that future efforts to understand and
-forecast biodiversity should incorporate species composition, since lower-level
-processes are expected to be more dynamic [@ernest2001; @dornelas2014] and
-contain more useful information [@harris2015].
+considered for forecasts of richness and other stable aspects of biodiversity.
+Our results also suggest that future efforts to understand and forecast
+biodiversity should incorporate species composition, since lower-level processes
+are expected to be more dynamic [@ernest2001; @dornelas2014] and contain more
+useful information [@harris2015].
 
@@
 uncertainty and bias. Ultimately, ecological forecasts that use environmental
-data will therefore be more uncertain than our current hindcasting efforts. It
-is important to correctly incorporate this uncertainty in the predictor
-variables into forecasting models [@clark2001; @dietze2017]. Limitations in
-forecasting future environmental conditions, particularly at small scales, will
-present continued challenges for models incorporating environmental variables,
-and this may result in a continued advantage for simple single-site approaches.
+data will therefore be more uncertain than our current hindcasting efforts, and
+it is important to correctly incorporate this uncertainty into our models
+[@clark2001; @dietze2017]. Limitations in forecasting future environmental
+conditions---particularly at small scales---will present continued challenges
+for models incorporating environmental variables, and this may result in a
+continued advantage for simple single-site approaches.
 
@@
 effects resulted in changes in which models performed best, most notably
-reducing the relative performance of the naive model's point estimates. This
-suggests that, prior to accounting for observer effects, the naive model
-performed well because it was capable of accommodating rapid shifts in estimated
-richness introduced by changes in the observer. These kinds of rapid changes
-were difficult for the other single-site models to accommodate, and so the
-average performance of the ARIMA and average models improved once this source of
-observation error was addressed. This demonstrates that modeling observation
-error can be important for properly estimating and reducing uncertainty in
-forecasts and can also lead to changes in the best methods for forecasting [Box
-1]. Another key aspect of an ideal observation model is imperfect detection. In
-this study, we did not address differences in detection probability across
-species and sites [@boulinier1998] since there is no clear way to address this
-issue using NA Breeding Bird Survey data without making strong assumptions about
-the data [i.e., assuming there is no biological variation in stops along a
-route; @white2010], but this would be a valuable addition to future forecasting
-models.
+improving most models' point estimates relative to the naive baseline [Box 1].
+This suggests that, prior to accounting for observer effects, the naive model
+performed well largely because it was capable of accommodating rapid shifts in
+estimated richness introduced by changes in the observer. These kinds of rapid
+changes were difficult for the other single-site models to accommodate. Another
+key aspect of an ideal observation model is imperfect detection. In this study,
+we did not address differences in detection probability across species and sites
+[@boulinier1998] since there is no clear way to address this issue using North
+American Breeding Bird Survey data without making strong assumptions about the
+data [i.e., assuming there is no biological variation in stops along a route;
+@white2010], but this would be a valuable addition to future forecasting models.
 
 The science of forecasting biodiversity remains in its infancy and it is
-important to consider weaknesses in current forecasting methods in that
-context. In the beginning weather forecasts were also worse than simple
-baselines, but these forecasts have continually improved throughout the history
-of the field [@mcgill2012; @silver2012; @bauer2015]. One practice that lead to
-improvements in weather forecasts was that large numbers of forecasts were made
-publicly, allowing different approaches to be regularly assessed and refined
-[@mcgill2012; @silver2012]. To facilitate this kind of improvement, it is
-important for ecologists to start regularly making and evaluating real
-ecological forecasts, even if they perform poorly, and to make these forecasts
-openly available for assessment [@mcgill2012; @dietze2016]. These forecasts
-should include both short-term predictions, which can be assessed quickly, and
-mid- to long-term forecasts, which can help ecologists to assess long time-scale
-processes and determine how far into the future we can successfully forecast
-[@dietze2016; @tredennick2016]. Forecasts out to 2050 based on the models in
-this paper have been openly archived [@zenodo], so that we and others can assess
-how well they perform. We plan to evaluate these forecasts and report the
-results as each new year of BBS data becomes available, and make iterative
-improvements to the forecasting models in response to these assessments.
+important to consider weaknesses in current forecasting methods in that context.
+In the beginning, weather forecasts were also worse than simple baselines, but
+these forecasts have continually improved throughout the history of the field
+[@mcgill2012; @silver2012; @bauer2015]. One practice that lead to improvements
+in weather forecasts was that large numbers of forecasts were made publicly,
+allowing different approaches to be regularly assessed and refined [@mcgill2012;
+@silver2012]. To facilitate this kind of improvement, it is important for
+ecologists to start regularly making and evaluating real ecological forecasts,
+even if they perform poorly, and to make these forecasts openly available for
+assessment [@mcgill2012; @dietze2016]. These forecasts should include both
+short-term predictions, which can be assessed quickly, and mid- to long-term
+forecasts, which can help ecologists to assess long time-scale processes and
+determine how far into the future we can successfully forecast [@dietze2016;
+@tredennick2016]. We have openly archived forecasts from all six models through
+the year 2050 [@zenodo], so that we and others can assess how well they perform.
+We plan to evaluate these forecasts and report the results as each new year of
+BBS data becomes available, and make iterative improvements to the forecasting
+models in response to these assessments.
 

### 2017-09-05 manuscript/draft.Rmd
variance partitioning

--- 2017-09-04 manuscript/draft.Rmd
@@
 
-Partitioning each model's squared error into consistent errors across sites and
-errors that fluctuated year-to-year shows that the majority of the error comes
-from errors in the site-level mean (Figure 6). The "average" model, which was
-based on the site-level mean, had the lowest error in this regard. In contrast,
-the three environmental models showed the largest biases at the site level; this
-makes sense, given that they could not explicitly distinguish among sites with
-similar climate, NDVI, and elevation.
+Partitioning each model's squared error shows that the majority of the residual
+error was attributed to errors in estimating site-level means, rather than
+errors in tracking year-to-year fluctuations (Figure 6). The "average" model,
+which was based entirely on site-level means, had the lowest error in this
+regard. In contrast, the three environmental models showed larger biases at the
+site level, though they still explained most of the variance in this component.
+This makes sense, given that they could not explicitly distinguish among sites
+with similar climate, NDVI, and elevation. Interestingly, the environmental
+models had higher squared error than the baselines did for tracking year-to-year
+fluctuations in richness as well.
 

### 2017-09-05 manuscript/draft.Rmd
fix timescale discussion based on our meeting

--- 2017-09-05 manuscript/draft.Rmd
@@
 models (in part due to data limitations) making them difficult to compare to our
-results directly. Given the potential changes with lead time of the models and
-processes are most relevant to forecasting, it is important that more forecasts
-with longer lead times are evaluated and that the inferences drawn from
-short-term and long-term forecasting studies be compared. By evaluating
-shorter-term forecasts, researchers can avoid waiting so many decades to
-evaluate their methods [@petchey2015; @tredennick2016; @dietze2016]. However,
-since drivers of species richness differ at different temporal scales
-[@rosenzweig1995; @white2004; @white2007], short-term evaluations should be
-viewed as informing, not replacing, long-term evaluations [@tredennick2016]. In
-particular, we might expect environmental and ecological information to become
-more important at longer time scales, and thus for the performance of simple
-baseline forecasts to degrade faster than forecasts from SDMs and other similar
-models [@blonder2017].  On the other hand, there is evidence that species
-respond differently to spatial versus temporal variation in the environment
-[@oedekoven2017], implying that most SDMs are not using environmental
-information well enough to make accurate forecasts.
+results directly. Studies on shorter time scales, such as ours, provide one way
+to evaluate our forecasting methods without having to wait several decades to
+observe the effects of environmental change on biodiversity [@petchey2015;
+@tredennick2016; @dietze2016], but cannot fully replace longer-term evaluations
+[@tredennick2016]. In general, drivers of species richness can differ at
+different temporal scales [@rosenzweig1995; @white2004; @white2007], so
+different methods may perform better for different lead times. In particular, we
+might expect environmental and ecological information to become more important
+at longer time scales, and thus for the performance of simple baseline forecasts
+to degrade faster than forecasts from SDMs and other similar models
+[@blonder2017].  We did observe a small trend in this direction: deviance for
+the auto-ARIMA models and for the average baseline grew faster than for two of
+the environmental models (the JSDM and the macroecological model), although this
+growth was not statistically significant for the average baseline.
 

### 2017-09-05 manuscript/draft.Rmd
cite Ward in discussion. fixes #190
fix blonder citation per Shawn's advice

@@
 [@tredennick2016]. In general, drivers of species richness can differ at
-different temporal scales [@rosenzweig1995; @white2004; @white2007], so
-different methods may perform better for different lead times. In particular, we
-might expect environmental and ecological information to become more important
-at longer time scales, and thus for the performance of simple baseline forecasts
-to degrade faster than forecasts from SDMs and other similar models
-[@blonder2017].  We did observe a small trend in this direction: deviance for
+different temporal scales [@rosenzweig1995; @white2004; @white2007;
+@blonder2017], so different methods may perform better for different lead times.
+In particular, we might expect environmental and ecological information to
+become more important at longer time scales, and thus for the performance of
+simple baseline forecasts to degrade faster than forecasts from SDMs and other
+similar models.  We did observe a small trend in this direction: deviance for
 the auto-ARIMA models and for the average baseline grew faster than for two of
@@
 absolute accuracy of forecasts across different sites: when a past year's
-richness is already known, it is easy to estimate future richness.  The
-site-level stability of the BBS data also explains why SDMs and macroecological
-models perform relatively well at predicting future richness, despite failing to
+richness is already known, it is easy to estimate future richness. Ward et al.
+(2014) found similar patterns in time series of fisheries stocks, where
+relatively stable time series were best predicted by simple models and more
+complex models were only beneficial with dynamic time series. The site-level
+stability of the BBS data also explains why SDMs and macroecological models
+perform relatively well at predicting future richness, despite failing to
 capture changes in richness over time. However, this stability also makes it

### 2017-09-05 manuscript/draft.Rmd
move Box 1 citation

@@
 In addition to comparing and improving the process models used for forecasting
-it is important to consider the observation models.  When working with any
-ecological dataset, there are imperfections in the sampling process that have
-the potential to influence results. With large scale surveys and citizen science
-datasets, such as the Breeding Bird Survey, these issues are potentially
+it is important to consider the observation models [Box 1].  When working with
+any ecological dataset, there are imperfections in the sampling process that
+have the potential to influence results. With large scale surveys and citizen
+science datasets, such as the Breeding Bird Survey, these issues are potentially
 magnified by the large number of different observers and by major differences in
@@
 effects resulted in changes in which models performed best, most notably
-improving most models' point estimates relative to the naive baseline [Box 1].
-This suggests that, prior to accounting for observer effects, the naive model
+improving most models' point estimates relative to the naive baseline. This
+suggests that, prior to accounting for observer effects, the naive model
 performed well largely because it was capable of accommodating rapid shifts in

### 2017-09-05 manuscript/draft.Rmd

@@
 In addition to comparing and improving the process models used for forecasting
-it is important to consider the observation models [Box 1].  When working with
-any ecological dataset, there are imperfections in the sampling process that
-have the potential to influence results. With large scale surveys and citizen
-science datasets, such as the Breeding Bird Survey, these issues are potentially
+it is important to consider the observation models.  When working with any
+ecological dataset, there are imperfections in the sampling process that have
+the potential to influence results. With large scale surveys and citizen science
+datasets, such as the Breeding Bird Survey, these issues are potentially
 magnified by the large number of different observers and by major differences in
@@
 improving most models' point estimates relative to the naive baseline. This
-suggests that, prior to accounting for observer effects, the naive model
-performed well largely because it was capable of accommodating rapid shifts in
-estimated richness introduced by changes in the observer. These kinds of rapid
-changes were difficult for the other single-site models to accommodate. Another
-key aspect of an ideal observation model is imperfect detection. In this study,
-we did not address differences in detection probability across species and sites
-[@boulinier1998] since there is no clear way to address this issue using North
-American Breeding Bird Survey data without making strong assumptions about the
-data [i.e., assuming there is no biological variation in stops along a route;
-@white2010], but this would be a valuable addition to future forecasting models.
+demonstrates that modeling observation error can be important for properly
+estimating and reducing uncertainty in forecasts and can also lead to changes in
+the best methods for forecasting [Box 1]. This suggests that, prior to
+accounting for observer effects, the naive model performed well largely because
+it was capable of accommodating rapid shifts in estimated richness introduced by
+changes in the observer. These kinds of rapid changes were difficult for the
+other single-site models to accommodate. Another key aspect of an ideal
+observation model is imperfect detection. In this study, we did not address
+differences in detection probability across species and sites [@boulinier1998]
+since there is no clear way to address this issue using North American Breeding
+Bird Survey data without making strong assumptions about the data [i.e.,
+assuming there is no biological variation in stops along a route; @white2010],
+but this would be a valuable addition to future forecasting models.
 

### 2017-09-11 manuscript/draft.Rmd
Minor pre-submission changes
* Citations
* Don't promise non-existent appendix

@@
 (https://github.com/weecology/bbs-forecasting) and archived on Zenodo
-(https://doi.org/10.5281/zenodo.839581).
+[@zenodoCode].
 
@@
 fit (Figure 7). Improvements primarily resulted from a small number of forecasts
-where observer turnover caused a large shift in the reported richness values
-(Appendix [[Figure Number]]). The naive baseline was less sensitive to these
-shifts, because it largely ignored the richness values reported by observers
-that had retired by the end of the training period (Figure 1). The average
-model, which gave equal weight to observations from the whole training period,
-showed a larger decline in performance when not accounting for observer effects
--- especially in terms of coverage. The performance of the mistnet JSDM was
-notable here, because its prediction intervals retained good coverage even when
-not correcting for observer differences, which we attribute to the JSDM's
-ability to model this variation with its latent variables.
+where observer turnover caused a large shift in the reported richness values.
+The naive baseline was less sensitive to these shifts, because it largely
+ignored the richness values reported by observers that had retired by the end of
+the training period (Figure 1). The average model, which gave equal weight to
+observations from the whole training period, showed a larger decline in
+performance when not accounting for observer effects -- especially in terms of
+coverage. The performance of the mistnet JSDM was notable here, because its
+prediction intervals retained good coverage even when not correcting for
+observer differences, which we attribute to the JSDM's ability to model this
+variation with its latent variables.
 
@@
 @tredennick2016]. We have openly archived forecasts from all six models through
-the year 2050 [@zenodo], so that we and others can assess how well they perform.
-We plan to evaluate these forecasts and report the results as each new year of
-BBS data becomes available, and make iterative improvements to the forecasting
-models in response to these assessments.
+the year 2050 [@zenodoPredict], so that we and others can assess how well they
+perform. We plan to evaluate these forecasts and report the results as each new
+year of BBS data becomes available, and make iterative improvements to the
+forecasting models in response to these assessments.
 

### 2017-09-19 manuscript/draft.Rmd
Fix title page

@@
 ---
-title: "Best practices for forecasting changes in biodiversity: an example using breeding birds"
+title: "Forecasting biodiversity in breeding birds using best practices"
 author:
-- David J. Harris
-- Shawn D. Taylor
-- Ethan P. White
+- David J. Harris$^1$ (corresponding author)
+- Shawn D. Taylor$^2$
+- Ethan P. White$^1$
 bibliography: refs.bib
@@
 output:
-  pdf_document: default
+  pdf_document: 
+    keep_tex: true
   html_document: default
 ---
+
+$^1$ Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, FL, United States
+
+$^2$ School of Natural Resources and Environment, University of Florida Gainesville, FL, United States
 
@@
 `r numbers = yaml::yaml.load_file("numbers.yaml")`
+
+\newpage
 

### 2017-11-06 manuscript/reviewer_reply.Rmd
add reviewer reply template

## Editors Comments

The reviewers and I agree that this manuscript is interesting, timely and well-written. I will add a few of my own observations here. First, and least important - normally submitted manuscripts do not hyphenate and do not right justify.

Next, in your figure 1, which forms the basis for comparison of model accuracy over time, I really expected you to choose data WITH a temporal pattern. After all, it looks to me like you have a time series in which species richness simply varies around a more or less stable average value, and the variation can be due to nothing (random), observer bias, and any number of species population trends (and combinations thereof). Thus, you have a situation of years of no trend, and use those data to predict no trend into the future? I would have thought you would choose data with a definite time trend that was already potentially correlated with, say, climate change, and then use hindcasting and forecasting to test whether the models accurately predict a trend into the "future"! To me, it is not surprising that models do not "...successfully turn spatial data into useful temporal predictions about biodiversity at decadal time-scales." But, I wonder if you chose data WITH a trend, you would have gotten other results.

Finally, your figure two seems to reverse panels A and B, neither of which is marked in the text.

I will leave the remainder of the comments to the two reviewers.

## Reviewer 1: Eric Ward
Basic reporting

I found the paper very clear and easy to follow, with a good intro/discussion that tied the current work into previous studies. All data shared - in addition to code. More minor comments in General Comments section.
Experimental design

This question and forecasting competition was well designed. I had minor comments to improve clarity with models. The analysis is all documented/replicable on the Github repo. More minor comments in General Comments section.
Validity of the findings

All data is robust, and analyses are sound. Conclusions and recommendations are well stated. The authors have really done a great job with transparency - this is a model of what all papers should strive for. See additional comments in General Comments section
Comments for the author

Review of Harris et al. Forecasting biodiversity in breeding birds using best practices

In general, I thought this paper was interesting and well written. These results add to the increasing literature supporting the idea that ecological forecasting in time/space is difficult. I think the recommendations/best practice bullet points described by the authors are essential for anyone interested in forecasting this kind of data. All of my points below can be considered minor issues, which I think will improve the paper’s clarity.

Introduction:
Line 41: I think this is generally more common in terrestrial ecology – in fisheries these SDM models with more of a theoretical background don’t seem to be used as much

Line 62: I really like the recommendations / clarity in Box 1, and think these will be accessible to most ecologists

Line 71-73: I’m re-reading this after commenting on data, and am now a little confused. The bird responses are annual snapshots during the breeding season? And I thought the environmental data you used was annual or quarterly, not monthly?
https://www.mbr-pwrc.usgs.gov/bbs/grass/genintro.htm

Methods:
Line 92: For those not familiar with the BBS protocol, maybe describe in a sentence. I could be mis-remembering, but isn’t each site made up of 50 stops, and presumably you aggregated richness across these stops?

Line 105: I assume all these species have been similarly ID’d over the course of the dataset – e.g. birds split into sub-species haven’t been included.

Line 107: For the environmental variables, do you have any that match up with the temporal resolution of the response data (monthly?). It seems like all the variables are annual / quarterly / etc.

Line 124: I’m not as familiar with terrestrial projections, but are the CMIP5 projections relatively unbiased / precise when compared to recent data?

Line 138: Maybe also cite any of the interfaces you used, like RStan:
http://mc-stan.org/users/citations/

Line 140: I’m curious why you didn’t consider spatial structure in the random effects here, using GMRF models or similar? One hand these models might be overkill if the habitat/environmental variables (which are spatially correlated) explain variability in richness, but the spatial correlation may pick up other sources of variability not accounted for by those variables.

Line 141-143: I think the paper thus far is pretty clear. I was hoping to see a few equations here – it helps me especially when components start to become hierarchical!

Line 163-164: again, I think equations could really help here to describe the baseline models. I think it’s also important to use these to clarify key details – for example, by training models on a site-by-site basis, you’re estimating separate parameters by site. For example with the uncorrelated noise model, you’re estimating unique variances by site right? Maybe say something like this, in addition to including equations.

Line 180 – Maybe also mention this is in the forecast package – as you do with gbm below?

Line 183: Because these defaults may change over time, it’d be good to specify things like the ranges of AR / MA order that you searched over, which criteria was used to compare models, etc. And whether or not you included a drift parameter?

Table 1 caption: without getting to more methods (below), I don’t understand the sentence ‘Environmental models were trained using all sites together, without information regarding which transects occurred at which site or during which year’.

Table 1: Just my personal preference, but I’d include the check marks and replace the ‘Xs’ with blank cells

Line 194-221: It seems for each approach you only considered the full models with all environmental covariates? I’m curious whether any of the covariates appeared to not improve explanatory power, and whether you considered models that dropped these uninformative predictors?

Line 216: Cite randomForest / caret/ party R package – or whatever else you used?

Line 242: Again, I think including the equation here would be useful

Line 245: In our 2014 forecasting paper, we used MASE over RMSE because of rationale in Hyndmand and Koehler (2006). Using RMSE here is probably ok because you probably don’t have zeros/huge values/etc

Line 249: Can you elaborate on what you did with the deviance? By itself mean deviance is just a measure of average fit (I think most of the models are giving you this). Var(deviance) should incorporate precision – but I’ve only seen this extracted from Bayesian models

Line 266: I think you guys have done a great job putting together this public repository + made the analysis totally transparent. My one hang up with the repo was that it’d be awesome to include a short vignette with example data from this project that is more easily accessible to readers without having to install Jupyter/R packages/etc. For example – I was travelling when I started this review and couldn’t install one of the R packages on my government computer without making an appointment with our IT staff (so it took ~ a day). This isn’t your fault of course, but much harder for me to dig through the repo.

Line 269: See my above comments about spatially correlated random effects / GMRF models. One way to confirm these other approaches are overkill would be to make some maps of the estimated random effects, and fit some simple models to estimate variogram parameters – e.g. even gls() in R can do this

Line 276: Is one of the potential reasons that richness is stationary that presence/absence on the BBS varies less than other measures – maybe abundance, or other measures of diversity are more variable?

Line 294: Is the narrow predictive distribution for the stacked SDM also why this exhibits the strongest trend in decreasing coverge as a function of forecast length (Fig 5)?

Line 329 – maybe reword, you use forecast 3x in this sentence

Line 414: Agreed – doing this same kind of forecasting competition with compositional data, or other species diversity measures (Simpson’s) would be interesting

Line 461: Another thread with conservation implications is that back in the 1990s, it was pretty common for single species PVA models to be forecasted 100-200 years. This seems less common, in part because some predictions may have not performed well, and others are on a time horizon that can’t be evaluated. In Steve Beissinger’s PVA book he cites Goldwasser (2000) “Variability and measurement error in extinction risk analysis...” for this point (p 133).

## 
Reviewer 2
Basic reporting

see general comments
Experimental design

see general comments
Validity of the findings

see general comments
Comments for the author

In this paper, the authors use the incredibly rich data from the Breeding Bird Survey to examine how well various models perform at predicting bird richness at sites throughout North America. They train models on data from 1982-2003 and attempt to predict data from 2004-2013. They examine the performance of six different model types, two of which are null or baseline models.

I find it a bit surprising that the authors think that any of these richness models could predict a high level of accuracy for theses ten years. I think of richness estimations as a very coarse technique for looking at broad shifts in richness over relatively long time periods rather than these 5-species (~5-7%) shifts in richness on annual bases indicated in Figure 2. The authors don’t provide a lot of their raw data on sites other than as examples, but my impression is that they are pursuing an unrealistic level of accuracy and detail. That being said, given what the paper does, it’s okay. The models seem to be performed well. The authors give all of the keywords to indicate that they know what they are doing.

Here are some comments:
- Birds are not the best system for doing any sort of distribution analyses. They are incredibly vagile, hard to detect, and the distributions of breeding birds are affected by many components besides environmental variables (particularly interspecific interactions). No assessment of detectability is included here other than accounting for observer identity. However, I do realize that BBS data are detailed and compelling data to use given what a large amount of detailed time-series it provides.
- Abstract: The authors forecast the data and then test it using known data. Hindcast only refers to back-in-time predictions.
- 25-29: rather than relay the type of information that the paper will provide, the authors should summarize their results and findings
- 57-58: this is not the definition of hindcasting. This is the definition of testing. Also, the authors exclude the many deeper-time hindcasting studies that have been performed on these types of models.
- Methods: the authors perform their analyses on data that are collected from an incredibly dynamic time period. I wouldn’t expect species to have settled into a richness pattern given the incredibly disrupted nature of the last 32 years. Some strategies to try to handle these issues might be to train and predict at different temporal scales and to both forecast and hindcast the predictions. (e.g. train on 2004-2013 and try to predict 1982-1992 data + 1993-2003 data). If the authors were to make splits based on expected periods of dynamism, they could better tell whether the drivers that are included in the models are accurate predictors.
- 124-130: I didn’t see where future projections were performed/used in this paper. Why is this here?
- Fig. 2: Panels not labeled. No key for colors. Somewhat confusing.
- Fig. 6 caption: “compoonents”
- 466: only other mention of future forecasts… not sure why this is included in this paper.

### 2017-11-06 manuscript/reviewer_reply.Rmd
clarify hindcasting to reviewer

@@
 - Abstract: The authors forecast the data and then test it using known data. Hindcast only refers to back-in-time predictions.
+
+The definition of hindcasting varies by field.
+We use it here as it's used meteorology, where any prediction of observations already available is generally a hindcast, and forecasts are only for predictions which are actually in the future.
+See Jolliffe & Stephenson 2003 (in manuscript) section 1.4.2 for a longer discussion.
+
 - 25-29: rather than relay the type of information that the paper will provide, the authors should summarize their results and findings
@@
 This is the definition of testing.
-Also, the authors exclude the many deeper-time hindcasting studies that have been performed on these types of models.
+Also, the authors exclude the many deeper-time hindcasting studies that have been performed on these types of models
+
+See comment above on hindcasting.
+
 - Methods: the authors perform their analyses on data that are collected from an incredibly dynamic time period.

### 2017-11-06 manuscript/reviewer_reply.Rmd
typo and clarify statement

@@
 The definition of hindcasting varies by field.
-We use it here as it's used meteorology, where any prediction of observations already available is generally a hindcast, and forecasts are only for predictions which are actually in the future.
+We use it here as it's used in meteorology, where any prediction of observations already available is generally a hindcast, and forecasts are only for predictions which are actually in the future.
 See Jolliffe & Stephenson 2003 (in manuscript) section 1.4.2 for a longer discussion.
+We've clarified this in the text.
 

### 2017-11-06 manuscript/draft.Rmd
define hindcasting

--- 2017-09-07 manuscript/draft.Rmd
@@
 We evaluated 6 types of forecasting models (Table 1) by dividing the 32 years
-of data into 22 years of training data and 10 years of data for 
-evaluating forecasts using hindcasting. We also made long term forecasts by using
-the full data set for training and making forecasts through the year 2050. For
-both time scales, we made forecasts using each model with and without
-correcting for observer effects, as described below.
+of data into 22 years of training data and 10 years of data for evaluating
+forecasts using hindcasting. Here we use definitions from meteorology, where a
+hindcast is generally any prediction for an event that has already happened,
+while forecasts are predictions for actual future events [@jolliffe2003]. We
+also made long term forecasts by using the full data set for training and
+making forecasts through the year 2050. For both time frames, we made forecasts
+using each model with and without correcting for observer effects, as described
+below.
 

### 2017-11-08 manuscript/reviewer_reply.Rmd
comment on timescale mismatch

--- 2017-11-06 manuscript/reviewer_reply.Rmd
@@
 
-Line 71-73: I’m re-reading this after commenting on data, and am now a little confused.
+"Line 71-73: I’m re-reading this after commenting on data, and am now a little confused.
 The bird responses are annual snapshots during the breeding season?
 And I thought the environmental data you used was annual or quarterly, not monthly?
-https://www.mbr-pwrc.usgs.gov/bbs/grass/genintro.htm
+https://www.mbr-pwrc.usgs.gov/bbs/grass/genintro.htm"
+
+--BBS data are generally viewed as annual data, even though the observations happen on a single day.
+The presence of an individual species is primarily a function of suitable habitat and, for many species, timing of migration.
+Environmental variables with a shorter temporal scale could likely improve models, such as incorporating cool weather days into the observation model, and would be worth pursuing in a different study.
 
@@
 
-Line 107: For the environmental variables, do you have any that match up with the temporal resolution of the response data (monthly?). It seems like all the variables are annual / quarterly / etc.
+"Line 107: For the environmental variables, do you have any that match up with the temporal resolution of the response data (monthly?). It seems like all the variables are annual / quarterly / etc."
+
+--See comment above on using shorter time scale environmental variables.
 

### 2017-11-08 manuscript/reviewer_reply.Rmd
note on rstan
clarify bbs methods

@@
 Methods:
-Line 92: For those not familiar with the BBS protocol, maybe describe in a sentence.
-I could be mis-remembering, but isn’t each site made up of 50 stops, and presumably you aggregated richness across these stops?
+"Line 92: For those not familiar with the BBS protocol, maybe describe in a sentence.
+I could be mis-remembering, but isn’t each site made up of 50 stops, and presumably you aggregated richness across these stops?"
 
-Line 105: I assume all these species have been similarly ID’d over the course of the dataset – e.g. birds split into sub-species haven’t been included.
+--This is correct, we've added clarification to the Methods.
+
+"Line 105: I assume all these species have been similarly ID’d over the course of the dataset – e.g. birds split into sub-species haven’t been included."
+
+--This is correct, see the section "Richness Data" in the Methods.
 
@@
 
-Line 138: Maybe also cite any of the interfaces you used, like RStan:
-http://mc-stan.org/users/citations/
+"Line 138: Maybe also cite any of the interfaces you used, like RStan:
+http://mc-stan.org/users/citations/"
+
+--RStan is cited in the last paragraph of the Methods.
 

### 2017-11-08 manuscript/draft.Rmd
clarify bbs methods

@@
 We evaluated 6 types of forecasting models (Table 1) by dividing the 32 years
-of data into 22 years of training data and 10 years of data for evaluating
-forecasts using hindcasting. Here we use definitions from meteorology, where a
-hindcast is generally any prediction for an event that has already happened,
-while forecasts are predictions for actual future events [@jolliffe2003]. We
-also made long term forecasts by using the full data set for training and
-making forecasts through the year 2050. For both time frames, we made forecasts
-using each model with and without correcting for observer effects, as described
-below.
+of data into 22 years of training data and 10 years of data for 
+evaluating forecasts using hindcasting. We also made long term forecasts by using
+the full data set for training and making forecasts through the year 2050. For
+both time scales, we made forecasts using each model with and without
+correcting for observer effects, as described below.
 
@@
 Breeding Bird Survey (BBS) [@pardieck2017] using the Data Retriever Python
-package [@morris2013] and rdataretriever R package [@rdataretriever]. The BBS
-data was filtered to exclude all nocturnal, cepuscular, and aquatic species
-[since these species are not well sampled by BBS methods; @hurlbert2005], as
-well as unidentified species, and hybrids. All data from surveys that did not
-meet BBS quality criteria were also excluded.
+package [@morris2013] and rdataretriever R package [@rdataretriever]. BBS
+observations are three minute point counts made at 50 fixed locations along a
+40km route. Here we denote each route as a site and summarize richness as the
+total species observed at all 50 locations in each surveyed year. Prior to
+summarizing the data was filtered to exclude all nocturnal, cepuscular, and
+aquatic species [since these species are not well sampled by BBS methods;
+@hurlbert2005], as well as unidentified species, and hybrids. All data from
+surveys that did not meet BBS quality criteria were also excluded.
 

### 2017-11-08 manuscript/reviewer_reply.Rmd
cite randomforest directly where its used

@@
 
-Line 180 – Maybe also mention this is in the forecast package – as you do with gbm below?
+"Line 180 – Maybe also mention this is in the forecast package – as you do with gbm below?"
+
+--We have added this.
 
@@
 
-Line 216: Cite randomForest / caret/ party R package – or whatever else you used?
+"Line 216: Cite randomForest / caret/ party R package – or whatever else you used?"
+
+--We've cited this here.
 

### 2017-11-08 manuscript/draft.Rmd
cite randomforest directly where its used

@@
 **Time series models.** We used Auto-ARIMA models [based on the `auto.arima`
-function in @forecast1] to represent an array of different time-series modeling
-approaches. These models can include an autoregressive component (as in the
-"naive" model, but with the possibility of longer-term dependencies in the
-underlying process), a moving average component (where the noise can have serial
-autocorrelation) and an integration/differencing component (so that the analysis
-could be performed on sequential differences of the raw data, accommodating more
-complex patterns including trends). The `auto.arima` function chooses whether to
-include each of these components (and how many terms to include for each one)
-using AICc [@forecast1]. Since there is no seasonal component to the BBS
-time-series, we did not include a season component in these models. Otherwise we
-used the default settings for this function [@forecast1].
+function in the package `forecast`; @forecast1] to represent an array of
+different time-series modeling approaches. These models can include an
+autoregressive component (as in the "naive" model, but with the possibility of
+longer-term dependencies in the underlying process), a moving average component
+(where the noise can have serial autocorrelation) and an
+integration/differencing component (so that the analysis could be performed on
+sequential differences of the raw data, accommodating more complex patterns
+including trends). The `auto.arima` function chooses whether to include each of
+these components (and how many terms to include for each one) using AICc
+[@forecast1]. Since there is no seasonal component to the BBS time-series, we
+did not include a season component in these models. Otherwise we used the
+default settings for this function [@forecast1].
 
@@
 for each of the species-level probabilities instead of just one, propagating the
-uncertainty forward. We obtained these estimates using random forests, a common
-approach in the species distribution modeling literature. Random forests are
-constructed by fitting hundreds of independent regression trees to
-randomly-perturbed versions of the data [@cutler2007; @caruana2008]. When
-correcting for observer effects, each of the 500 trees in our species-level
-random forests used a different Monte Carlo estimate of the observer effects as
-a predictor variable.
+uncertainty forward. We obtained these estimates using random forests
+[@randomForest], a common approach in the species distribution modeling
+literature. Random forests are constructed by fitting hundreds of independent
+regression trees to randomly-perturbed versions of the data [@cutler2007;
+@caruana2008]. When correcting for observer effects, each of the 500 trees in
+our species-level random forests used a different Monte Carlo estimate of the
+observer effects as a predictor variable.
 

### 2017-11-17 manuscript/reviewer_reply.Rmd
update randomforest comment

@@
 
-"Line 216: Cite randomForest / caret/ party R package – or whatever else you used?"
+> "Line 216: Cite randomForest / caret/ party R package – or whatever else you used?"
 
---We've cited this here.
+**The randomForest package was cited in the final methods paragraph. We've cited it where it is first mentioned as well. **
 
@@
 - Abstract: The authors forecast the data and then test it using known data. Hindcast only refers to back-in-time predictions.
+
+The definition of hindcasting varies by field.
+We use it here as it's used in meteorology, where any prediction of observations already available is generally a hindcast, and forecasts are only for predictions which are actually in the future.
+See Jolliffe & Stephenson 2003 (in manuscript) section 1.4.2 for a longer discussion.
+We've clarified this in the text.
+
 - 25-29: rather than relay the type of information that the paper will provide, the authors should summarize their results and findings

### 2017-11-17 manuscript/draft.Rmd

@@
 Breeding Bird Survey (BBS) [@pardieck2017] using the Data Retriever Python
-package [@morris2013] and rdataretriever R package [@rdataretriever]. The BBS
-data was filtered to exclude all nocturnal, cepuscular, and aquatic species
-[since these species are not well sampled by BBS methods; @hurlbert2005], as
-well as unidentified species, and hybrids. All data from surveys that did not
-meet BBS quality criteria were also excluded.
+package [@morris2013] and rdataretriever R package [@rdataretriever]. BBS
+observations are three-minute point counts made at 50 fixed locations along a
+40km route. Here we denote each route as a site and summarize richness as the
+total species observed at all 50 locations in each surveyed year. Prior to
+summarizing the data was filtered to exclude all nocturnal, cepuscular, and
+aquatic species [since these species are not well sampled by BBS methods;
+@hurlbert2005], as well as unidentified species, and hybrids. All data from
+surveys that did not meet BBS quality criteria were also excluded.
 

### 2017-11-17 manuscript/reviewer_reply.Rmd
markdown throughout the reviewer reply
italics for reviewer comments, bold for our replies, headers in different sections.

@@
 
-The reviewers and I agree that this manuscript is interesting, timely and well-written.
+*The reviewers and I agree that this manuscript is interesting, timely and well-written.
 I will add a few of my own observations here.
-First, and least important - normally submitted manuscripts do not hyphenate and do not right justify.
+First, and least important - normally submitted manuscripts do not hyphenate and do not right justify.*
 
-Next, in your figure 1, which forms the basis for comparison of model accuracy over time, I really expected you to choose data WITH a temporal pattern.
+*Next, in your figure 1, which forms the basis for comparison of model accuracy over time, I really expected you to choose data WITH a temporal pattern.
 After all, it looks to me like you have a time series in which species richness simply varies around a more or less stable average value, and the variation can be due to nothing (random), observer bias, and any number of species population trends (and combinations thereof).
@@
 I would have thought you would choose data with a definite time trend that was already potentially correlated with, say, climate change, and then use hindcasting and forecasting to test whether the models accurately predict a trend into the "future"!
-To me, it is not surprising that models do not "...successfully turn spatial data into useful temporal predictions about biodiversity at decadal time-scales." But, I wonder if you chose data WITH a trend, you would have gotten other results.
+To me, it is not surprising that models do not "...successfully turn spatial data into useful temporal predictions about biodiversity at decadal time-scales." But, I wonder if you chose data WITH a trend, you would have gotten other results.*
 
-Finally, your figure two seems to reverse panels A and B, neither of which is marked in the text.
+*Finally, your figure two seems to reverse panels A and B, neither of which is marked in the text.*
 
-I will leave the remainder of the comments to the two reviewers.
+*I will leave the remainder of the comments to the two reviewers.*
 
 ## Reviewer 1: Eric Ward
-Basic reporting
+#### Basic reporting
 
-I found the paper very clear and easy to follow, with a good intro/discussion that tied the current work into previous studies.
+*I found the paper very clear and easy to follow, with a good intro/discussion that tied the current work into previous studies.
 All data shared - in addition to code.
-More minor comments in General Comments section.
-Experimental design
+More minor comments in General Comments section.*
+#### Experimental design
 
-This question and forecasting competition was well designed.
+*This question and forecasting competition was well designed.
 I had minor comments to improve clarity with models.
 The analysis is all documented/replicable on the Github repo.
-More minor comments in General Comments section.
-Validity of the findings
+More minor comments in General Comments section.*
+#### Validity of the findings
 
-All data is robust, and analyses are sound.
+*All data is robust, and analyses are sound.
 Conclusions and recommendations are well stated.
 The authors have really done a great job with transparency - this is a model of what all papers should strive for.
-See additional comments in General Comments section
-Comments for the author
+See additional comments in General Comments section*
 
-Review of Harris et al. Forecasting biodiversity in breeding birds using best practices
+#### Comments for the author
 
-In general, I thought this paper was interesting and well written.
+#### Review of Harris et al. Forecasting biodiversity in breeding birds using best practices
+
+*In general, I thought this paper was interesting and well written.
 These results add to the increasing literature supporting the idea that ecological forecasting in time/space is difficult.
 I think the recommendations/best practice bullet points described by the authors are essential for anyone interested in forecasting this kind of data.
-All of my points below can be considered minor issues, which I think will improve the paper’s clarity.
+All of my points below can be considered minor issues, which I think will improve the paper’s clarity.*
 
-Introduction:
-Line 41: I think this is generally more common in terrestrial ecology – in fisheries these SDM models with more of a theoretical background don’t seem to be used as much
+#### Introduction:
+*Line 41: I think this is generally more common in terrestrial ecology – in fisheries these SDM models with more of a theoretical background don’t seem to be used as much*
 
-Line 62: I really like the recommendations / clarity in Box 1, and think these will be accessible to most ecologists
+*Line 62: I really like the recommendations / clarity in Box 1, and think these will be accessible to most ecologists*
 
-"Line 71-73: I’m re-reading this after commenting on data, and am now a little confused.
+*Line 71-73: I’m re-reading this after commenting on data, and am now a little confused.
 The bird responses are annual snapshots during the breeding season?
-And I thought the environmental data you used was annual or quarterly, not monthly?
-https://www.mbr-pwrc.usgs.gov/bbs/grass/genintro.htm"
+And I thought the environmental data you used was annual or quarterly, not monthly?* https://www.mbr-pwrc.usgs.gov/bbs/grass/genintro.htm
 
---BBS data are generally viewed as annual data, even though the observations happen on a single day.
+**BBS data are generally viewed as annual data, even though the observations happen on a single day.
 The presence of an individual species is primarily a function of suitable habitat and, for many species, timing of migration.
-Environmental variables with a shorter temporal scale could likely improve models, such as incorporating cool weather days into the observation model, and would be worth pursuing in a different study.
+Environmental variables with a shorter temporal scale could likely improve models, such as incorporating cool weather days into the observation model, and would be worth pursuing in a different study.**
 
-Methods:
-"Line 92: For those not familiar with the BBS protocol, maybe describe in a sentence.
-I could be mis-remembering, but isn’t each site made up of 50 stops, and presumably you aggregated richness across these stops?"
+#### Methods:
 
---This is correct, we've added clarification to the Methods.
+*Line 92: For those not familiar with the BBS protocol, maybe describe in a sentence.
+I could be mis-remembering, but isn’t each site made up of 50 stops, and presumably you aggregated richness across these stops?*
 
-"Line 105: I assume all these species have been similarly ID’d over the course of the dataset – e.g. birds split into sub-species haven’t been included."
+**This is correct, we've added clarification to the Methods. **
 
---This is correct, see the section "Richness Data" in the Methods.
+*Line 105: I assume all these species have been similarly ID’d over the course of the dataset – e.g. birds split into sub-species haven’t been included.*
 
-"Line 107: For the environmental variables, do you have any that match up with the temporal resolution of the response data (monthly?). It seems like all the variables are annual / quarterly / etc."
+**This is correct, see the section "Richness Data" in the Methods. **
 
---See comment above on using shorter time scale environmental variables.
+*Line 107: For the environmental variables, do you have any that match up with the temporal resolution of the response data (monthly?). It seems like all the variables are annual / quarterly / etc.*
 
-Line 124: I’m not as familiar with terrestrial projections, but are the CMIP5 projections relatively unbiased / precise when compared to recent data?
+**See comment above on using shorter time scale environmental variables. **
 
-"Line 138: Maybe also cite any of the interfaces you used, like RStan:
-http://mc-stan.org/users/citations/"
+*Line 124: I’m not as familiar with terrestrial projections, but are the CMIP5 projections relatively unbiased / precise when compared to recent data?*
 
---RStan is cited in the last paragraph of the Methods.
+*Line 138: Maybe also cite any of the interfaces you used, like RStan: http://mc-stan.org/users/citations/ *
 
-Line 140: I’m curious why you didn’t consider spatial structure in the random effects here, using GMRF models or similar?
-One hand these models might be overkill if the habitat/environmental variables (which are spatially correlated) explain variability in richness, but the spatial correlation may pick up other sources of variability not accounted for by those variables.
+**RStan is cited in the last paragraph of the Methods. **
 
-Line 141-143: I think the paper thus far is pretty clear. I was hoping to see a few equations here – it helps me especially when components start to become hierarchical!
+*Line 140: I’m curious why you didn’t consider spatial structure in the random effects here, using GMRF models or similar?
+One hand these models might be overkill if the habitat/environmental variables (which are spatially correlated) explain variability in richness, but the spatial correlation may pick up other sources of variability not accounted for by those variables.*
 
-Line 163-164: again, I think equations could really help here to describe the baseline models.
+*Line 141-143: I think the paper thus far is pretty clear. I was hoping to see a few equations here – it helps me especially when components start to become hierarchical!*
+
+*Line 163-164: again, I think equations could really help here to describe the baseline models.
 I think it’s also important to use these to clarify key details – for example, by training models on a site-by-site basis, you’re estimating separate parameters by site.
 For example with the uncorrelated noise model, you’re estimating unique variances by site right?
-Maybe say something like this, in addition to including equations.
+Maybe say something like this, in addition to including equations.*
 
-"Line 180 – Maybe also mention this is in the forecast package – as you do with gbm below?"
+*Line 180 – Maybe also mention this is in the forecast package – as you do with gbm below?*
 
---We have added this.
+**We have added this. **
 
-Line 183: Because these defaults may change over time, it’d be good to specify things like the ranges of AR / MA order that you searched over, which criteria was used to compare models, etc.
-And whether or not you included a drift parameter?
+*Line 183: Because these defaults may change over time, it’d be good to specify things like the ranges of AR / MA order that you searched over, which criteria was used to compare models, etc.
+And whether or not you included a drift parameter?*
 
-Table 1 caption: without getting to more methods (below), I don’t understand the sentence ‘Environmental models were trained using all sites together, without information regarding which transects occurred at which site or during which year’.
+*Table 1 caption: without getting to more methods (below), I don’t understand the sentence ‘Environmental models were trained using all sites together, without information regarding which transects occurred at which site or during which year’.*
 
-Table 1: Just my personal preference, but I’d include the check marks and replace the ‘Xs’ with blank cells
+*Table 1: Just my personal preference, but I’d include the check marks and replace the ‘Xs’ with blank cells*
 
-Line 194-221: It seems for each approach you only considered the full models with all environmental covariates?
-I’m curious whether any of the covariates appeared to not improve explanatory power, and whether you considered models that dropped these uninformative predictors?
+*Line 194-221: It seems for each approach you only considered the full models with all environmental covariates?
+I’m curious whether any of the covariates appeared to not improve explanatory power, and whether you considered models that dropped these uninformative predictors?*
 
-> "Line 216: Cite randomForest / caret/ party R package – or whatever else you used?"
+*Line 216: Cite randomForest / caret/ party R package – or whatever else you used?*
 
@@
 
-Line 242: Again, I think including the equation here would be useful
+*Line 242: Again, I think including the equation here would be useful*
 
-Line 245: In our 2014 forecasting paper, we used MASE over RMSE because of rationale in Hyndmand and Koehler (2006).
-Using RMSE here is probably ok because you probably don’t have zeros/huge values/etc
+*Line 245: In our 2014 forecasting paper, we used MASE over RMSE because of rationale in Hyndmand and Koehler (2006).
+Using RMSE here is probably ok because you probably don’t have zeros/huge values/etc*
 
-Line 249: Can you elaborate on what you did with the deviance?
+*Line 249: Can you elaborate on what you did with the deviance?
 By itself mean deviance is just a measure of average fit (I think most of the models are giving you this).
-Var(deviance) should incorporate precision – but I’ve only seen this extracted from Bayesian models
+Var(deviance) should incorporate precision – but I’ve only seen this extracted from Bayesian models*
 
-Line 266: I think you guys have done a great job putting together this public repository + made the analysis totally transparent.
+*Line 266: I think you guys have done a great job putting together this public repository + made the analysis totally transparent.
 My one hang up with the repo was that it’d be awesome to include a short vignette with example data from this project that is more easily accessible to readers without having to install Jupyter/R packages/etc.
 For example – I was travelling when I started this review and couldn’t install one of the R packages on my government computer without making an appointment with our IT staff (so it took ~ a day).
-This isn’t your fault of course, but much harder for me to dig through the repo.
+This isn’t your fault of course, but much harder for me to dig through the repo.*
 
-Line 269: See my above comments about spatially correlated random effects / GMRF models.
+*Line 269: See my above comments about spatially correlated random effects / GMRF models.
 One way to confirm these other approaches are overkill would be to make some maps of the estimated random effects, and fit some simple models to estimate variogram parameters – e.g.
-even gls() in R can do this
+even gls() in R can do this*
 
-Line 276: Is one of the potential reasons that richness is stationary that presence/absence on the BBS varies less than other measures – maybe abundance, or other measures of diversity are more variable?
+*Line 276: Is one of the potential reasons that richness is stationary that presence/absence on the BBS varies less than other measures – maybe abundance, or other measures of diversity are more variable?*
 
-Line 294: Is the narrow predictive distribution for the stacked SDM also why this exhibits the strongest trend in decreasing coverge as a function of forecast length (Fig 5)?
+*Line 294: Is the narrow predictive distribution for the stacked SDM also why this exhibits the strongest trend in decreasing coverge as a function of forecast length (Fig 5)?*
 
-Line 329 – maybe reword, you use forecast 3x in this sentence
+*Line 329 – maybe reword, you use forecast 3x in this sentence*
 
-Line 414: Agreed – doing this same kind of forecasting competition with compositional data, or other species diversity measures (Simpson’s) would be interesting
+*Line 414: Agreed – doing this same kind of forecasting competition with compositional data, or other species diversity measures (Simpson’s) would be interesting*
 
-Line 461: Another thread with conservation implications is that back in the 1990s, it was pretty common for single species PVA models to be forecasted 100-200 years.
+*Line 461: Another thread with conservation implications is that back in the 1990s, it was pretty common for single species PVA models to be forecasted 100-200 years.
 This seems less common, in part because some predictions may have not performed well, and others are on a time horizon that can’t be evaluated.
-In Steve Beissinger’s PVA book he cites Goldwasser (2000) “Variability and measurement error in extinction risk analysis...” for this point (p 133).
+In Steve Beissinger’s PVA book he cites Goldwasser (2000) “Variability and measurement error in extinction risk analysis...” for this point (p 133).*
 
-##
-Reviewer 2
-Basic reporting
+## Reviewer 2
 
-see general comments
-Experimental design
+#### Basic reporting
 
-see general comments
-Validity of the findings
+*see general comments*
 
-see general comments
-Comments for the author
+#### Experimental design
 
-In this paper, the authors use the incredibly rich data from the Breeding Bird Survey to examine how well various models perform at predicting bird richness at sites throughout North America.
+*see general comments*
+
+#### Validity of the findings
+
+*see general comments*
+
+#### Comments for the author
+
+*In this paper, the authors use the incredibly rich data from the Breeding Bird Survey to examine how well various models perform at predicting bird richness at sites throughout North America.
 They train models on data from 1982-2003 and attempt to predict data from 2004-2013.
-They examine the performance of six different model types, two of which are null or baseline models.
+They examine the performance of six different model types, two of which are null or baseline models.*
 
-I find it a bit surprising that the authors think that any of these richness models could predict a high level of accuracy for theses ten years.
+*I find it a bit surprising that the authors think that any of these richness models could predict a high level of accuracy for theses ten years.
 I think of richness estimations as a very coarse technique for looking at broad shifts in richness over relatively long time periods rather than these 5-species (~5-7%) shifts in richness on annual bases indicated in Figure 2.
@@
 The models seem to be performed well.
-The authors give all of the keywords to indicate that they know what they are doing.
+The authors give all of the keywords to indicate that they know what they are doing.*
 
-Here are some comments:
-- Birds are not the best system for doing any sort of distribution analyses.
+*Here are some comments:    *
+
+*Birds are not the best system for doing any sort of distribution analyses.
 They are incredibly vagile, hard to detect, and the distributions of breeding birds are affected by many components besides environmental variables (particularly interspecific interactions).
 No assessment of detectability is included here other than accounting for observer identity.
-However, I do realize that BBS data are detailed and compelling data to use given what a large amount of detailed time-series it provides.
-- Abstract: The authors forecast the data and then test it using known data. Hindcast only refers to back-in-time predictions.
+However, I do realize that BBS data are detailed and compelling data to use given what a large amount of detailed time-series it provides.*
 
-The definition of hindcasting varies by field.
+*Abstract: The authors forecast the data and then test it using known data. Hindcast only refers to back-in-time predictions.*
+
+**The definition of hindcasting varies by field.
 We use it here as it's used in meteorology, where any prediction of observations already available is generally a hindcast, and forecasts are only for predictions which are actually in the future.
@@
 We've clarified this in the text.
+**
 
-- 25-29: rather than relay the type of information that the paper will provide, the authors should summarize their results and findings
-- 57-58: this is not the definition of hindcasting.
+*25-29: rather than relay the type of information that the paper will provide, the authors should summarize their results and findings  *
+
+*57-58: this is not the definition of hindcasting.
 This is the definition of testing.
-Also, the authors exclude the many deeper-time hindcasting studies that have been performed on these types of models
+Also, the authors exclude the many deeper-time hindcasting studies that have been performed on these types of models*
 
-See comment above on hindcasting.
+**See comment above on hindcasting.**
 
-- Methods: the authors perform their analyses on data that are collected from an incredibly dynamic time period.
+#### Methods:
+
+*the authors perform their analyses on data that are collected from an incredibly dynamic time period.
 I wouldn’t expect species to have settled into a richness pattern given the incredibly disrupted nature of the last 32 years.
@@
 train on 2004-2013 and try to predict 1982-1992 data + 1993-2003 data).
-If the authors were to make splits based on expected periods of dynamism, they could better tell whether the drivers that are included in the models are accurate predictors.
-- 124-130: I didn’t see where future projections were performed/used in this paper. Why is this here?
-- Fig. 2: Panels not labeled. No key for colors. Somewhat confusing.
-- Fig. 6 caption: “compoonents”
-- 466: only other mention of future forecasts… not sure why this is included in this paper.
+If the authors were to make splits based on expected periods of dynamism, they could better tell whether the drivers that are included in the models are accurate predictors.*
+
+*124-130: I didn’t see where future projections were performed/used in this paper. Why is this here?*
+
+*Fig. 2: Panels not labeled. No key for colors. Somewhat confusing.*
+
+*Fig. 6 caption: “compoonents”*
+
+*466: only other mention of future forecasts… not sure why this is included in this paper.*

### 2017-11-20 manuscript/reviewer_reply.Rmd
remove spaces
these prevented GitHub's Markdown from rendering some markup from rendering

@@
 
-**This is correct, we've added clarification to the Methods. **
+**This is correct, we've added clarification to the Methods.**
 
@@
 
-**This is correct, see the section "Richness Data" in the Methods. **
+**This is correct, see the section "Richness Data" in the Methods.**
 
@@
 
-**See comment above on using shorter time scale environmental variables. **
+**See comment above on using shorter time scale environmental variables.**
 
@@
 
-*Line 138: Maybe also cite any of the interfaces you used, like RStan: http://mc-stan.org/users/citations/ *
+*Line 138: Maybe also cite any of the interfaces you used, like RStan: http://mc-stan.org/users/citations/*
 
-**RStan is cited in the last paragraph of the Methods. **
+**RStan is cited in the last paragraph of the Methods.**
 
@@
 
-**We have added this. **
+**We have added this.**
 
@@
 
-**The randomForest package was cited in the final methods paragraph. We've cited it where it is first mentioned as well. **
+**The randomForest package was cited in the final methods paragraph. We've cited it where it is first mentioned as well.**
 
@@
 
-*Here are some comments:    *
+*Here are some comments:*
 
@@
 See Jolliffe & Stephenson 2003 (in manuscript) section 1.4.2 for a longer discussion.
-We've clarified this in the text.
-**
+We've clarified this in the text.**
 
-*25-29: rather than relay the type of information that the paper will provide, the authors should summarize their results and findings  *
+*25-29: rather than relay the type of information that the paper will provide, the authors should summarize their results and findings*
 

### 2017-11-20 manuscript/draft.Rmd
minor table formatting

@@
 
-table_data = read_csv("../figures/table.csv",col_types = "ccccc")
+table_data = read_csv("../figures/table.csv",col_types = "ccccc", na = "na")
 
@@
     booktabs = TRUE,
-    caption = table_caption
+    caption = table_caption,
+    align = c("l", "l", "c", "c", "c")
   ) %>%

### 2017-11-20 manuscript/reviewer_reply.Rmd
law of total variance

@@
 *Finally, your figure two seems to reverse panels A and B, neither of which is marked in the text.*
+
+**We have switched the panels and labeled them. Thank you for noticing this error**.
 
@@
 
+**Thank you.**
+
 #### Introduction:
@@
 *Line 62: I really like the recommendations / clarity in Box 1, and think these will be accessible to most ecologists*
+
+**Thank you.**
 
@@
 
+**We have made this change**
+
 *Line 194-221: It seems for each approach you only considered the full models with all environmental covariates?
@@
 *Line 242: Again, I think including the equation here would be useful*
+
+**We have added an in-line equation and clarified the wording**
 
@@
 
+**Thank you for pointing out that we were unclear here.  The methods and purpose for the future predictions are now explained in more detail on lines [[XX]].**
+
 *Fig. 2: Panels not labeled. No key for colors. Somewhat confusing.*
@@
 
+**Thank you for pointing out this error. We have fixed it**
+
 *466: only other mention of future forecasts… not sure why this is included in this paper.*

### 2017-11-20 figures/plots.R
fix fig 2 panel order/names

@@ -366,8 +366,8 @@ make_ts_plots = function(models, ylim, use_obs_model, sample_site_id,
            deviance) %>% 
     left_join(distinct(bound, model, formal_name), "model") %>% 
     mutate(use_obs_model = ifelse(use_obs_model, 
-                                  "With observer model",
-                                  "Without observer model"))
+                                  "B. With observer model",
+                                  "A. Without observer model"))
 
   faceting = if (length(use_obs_model) == 2) {
     facet_grid(formal_name ~ use_obs_model)

### 2017-12-04 manuscript/reviewer_reply.Rmd
Spatial autocorrelation

--- 2017-11-20 manuscript/reviewer_reply.Rmd
@@
 
+**We agree that accounting for spatial autocorrelation in the observer model would have improved our estimates of the site-level random effects.
+However, we do not believe that this change would affect the paper's main results.
+Improvements would primarily benefit the "average" baseline, which is based directly on the observer model.
+As a result, it would be even harder for the three environmental models to outperform the baseline---especially since (as you noted), spatial and environmental explanations for variation in species richness trade off with one another.**
+
 *Line 141-143: I think the paper thus far is pretty clear. I was hoping to see a few equations here – it helps me especially when components start to become hierarchical!*
@@
 *Table 1: Just my personal preference, but I’d include the check marks and replace the ‘Xs’ with blank cells*
+
+** We have made this change**
 

### 2017-12-06 manuscript/draft.Rmd
law of total variance
minor

@@
 ---
+\raggedright
 
@@
 those runs' point estimates as our final point estimate and we calculated the
-uncertainty using the law of total variance (i.e. the average of the model runs'
-variance, plus the variance in the point estimates).
+uncertainty using the law of total variance (i.e. $\mathrm{Var}(\bar{y}) + \mathop{{}\mathbb{E}}\big[\mathrm{Var}(y)\big]$, or the variance in point estimates plus the average residual variance).
 
@@
 components. The site-level mean component shows consistent over or under
-estimates of richness at a site across years. The annual fluctuation compoonent
+estimates of richness at a site across years. The annual fluctuation component
 shows errors in predicting fluctuations in a site's richness over time. Both

### 2017-12-06 manuscript/reviewer_reply.Rmd
reviewer comments: minor changes
misc updates/responses

--- 2017-11-20 manuscript/reviewer_reply.Rmd
@@
 
+**This is a good point. We have added the word "terrestrial" to this line.**
+
 *Line 62: I really like the recommendations / clarity in Box 1, and think these will be accessible to most ecologists*
@@
 
+**We have added the equations and Stan code as a supplemental file.**
+
 *Line 163-164: again, I think equations could really help here to describe the baseline models.
@@
 
+**We chose not to drop uninformative predictors for two reasons.
+First, boosted trees are relatively robust to their inclusion ("A working guide to boosted regression trees", Elith et al.
+2008), so it seemed unlikely to affect our results.
+Second, it would have added complexity to the model-fitting procedure, which already involved 500 rounds of cross-validation (one per Monte Carlo sample of the observer effect).**
+
 *Line 216: Cite randomForest / caret/ party R package – or whatever else you used?*
@@
 
-**We have added an in-line equation and clarified the wording**
+**We have added an equation and clarified the wording here.**
 
@@
 Using RMSE here is probably ok because you probably don’t have zeros/huge values/etc*
+
+**We agree.**
 
@@
 
+**Thank you for pointing this out.  As the Editor noted, we had inadvertently reversed the two panels. We have fixed this, and also labeled them as you suggested.**
+
 *Fig. 6 caption: “compoonents”*

### 2017-12-06 manuscript/draft.Rmd
misc updates/responses

--- 2017-11-20 manuscript/draft.Rmd
@@
 
-Previous efforts to predict future patterns of species richness, and diversity
-more generally, have focused primarily on building species distributions models
-[SDMs; @thomas2004; @thuiller2011; @urban2015]. In general, these models
-describe individual species' occurrence patterns as functions of the
-environment. Given forecasts for environmental conditions, these models can 
+Previous efforts to predict future patterns of terrestrial species richness, and
+diversity more generally, have focused primarily on building species
+distributions models [SDMs; @thomas2004; @thuiller2011; @urban2015]. In general,
+these models describe individual species' occurrence patterns as functions of
+the environment. Given forecasts for environmental conditions, these models can
 predict where each species will occur in the future. These species-level 
 predictions are then combined ("stacked") to generate forecasts for species 
-richness [e.g. @calabrese2014]. Alternatively, models that directly relate spatial 
-patterns of species richness to environment conditions have been developed and 
-generally perform equivalently to stacked SDMs [@algar2009; @distler2015]. 
-This approach is sometimes referred to as "macroecological" modeling, because it 
-models the larger-scale pattern (richness) directly [@distler2015].
+richness [e.g. @calabrese2014]. Alternatively, models that directly relate
+spatial patterns of species richness to environment conditions have been
+developed and generally perform equivalently to stacked SDMs [@algar2009;
+@distler2015]. This approach is sometimes referred to as "macroecological"
+modeling, because it models the larger-scale pattern (richness) directly
+[@distler2015].
 
@@
 estimated the observer effects (and associated uncertainty about those effects)
-with a linear mixed model, with observer as a random effect, built in the Stan
-probabilistic programming language [@carpenter2017]. Because observer and site are
-strongly related (observers tend to repeatedly sample the same site), site was
-also included as a random effect to ensure that inferred deviations were
+using a linear mixed model, with observer as a random effect, built in the Stan
+probabilistic programming language [@carpenter2017]. Because observer and site
+are strongly related (observers tend to repeatedly sample the same site),
+site-level random effects were included to ensure that inferred deviations were
 actually observer-related (as opposed to being related to the sites that a given
-observer happened to see). The resulting model partitions the variance in
-observed richness values into site-level variance, observer-level variance, and
-residual variance (e.g.  variation within a site from year to year). The
-site-level estimates can also be used directly as the "average" baseline model
-(see below). The estimated observer effects can be subtracted from the richness
-values for a particular observer to provide an estimate of how many species
-would have been found by a "typical" observer. To incorporate uncertainty in
-these "corrected" richness values into the forecasting models we collected 500
-Monte Carlo samples from the model's posterior distribution, and fit each of the
-downstream models with each of the Monte Carlo samples. Each Monte Carlo sample
-represented a different possible set of observer-level and site-level random
-effect values across the full 32-year dataset.
+observer happened to see). The resulting model is described mathematically and
+with code in Supplement S1. The model partitions the variance in observed
+richness values into site-level variance, observer-level variance, and residual
+variance (e.g. variation within a site from year to year).
+
+Across our six modeling approaches (described below), we used estimates from the
+observer model in three different ways.  First, the expected values for
+site-level richness were used directly as our "average" baseline model (see
+below). For the two models that made species-level predictions, the estimated
+observer effects were included alongside the environmental variables as
+predictors.  Finally, we trained the remaining models to predict
+observer-corrected richness values (i.e. observed richness minus the observer
+effect, or the number of species that would have been recorded by a "typical"
+observer). Since the site-level and observer-level random effects are not known
+precisely, we represented the range of possible values using 500 Monte Carlo
+samples from the posterior distribution over these effects. Each downstream
+model was then trained 500 times using different possible values for the random
+effects.
 

### 2017-12-07 manuscript/observer-equations.Rmd

---
output: pdf_document
fontsize: 10pt
geometry: left=2.5cm, right=2.5cm, top=1.5cm, bottom=2cm, headheight=0pt, letterpaper
---

\begin{center}

\textbf{Supplement to "Forecasting biodiversity in breeding birds using best practices"}

by David J. Harris, Shawn D. Taylor, and Ethan P. White

\end{center}

\raggedright

\vspace{3mm}

**Observer model description**

Richness at site $i$, as recorded by observer $j$ is estimated using a linear mixed model.  Thus, the response variable was modeled as $y_{ij} \sim \mathcal{N}(\mu_{ij}, \sigma^\mathrm{residual}).$

Here, $\mu_{ij}$ is defined as an intercept term, plus a site-level effect, plus an observer-level effect: $\mu_{ij} = \alpha + \alpha^\mathrm{site}_i + \alpha^\mathrm{observer}_j.$

These two effects are each drawn from zero-mean Gaussians: $\alpha^\mathrm{site} \sim \mathcal{N}(0, \sigma^\mathrm{site})$ and $\alpha^\mathrm{observer} \sim \mathcal{N}(0, \sigma^\mathrm{observer}).$

Prior distributions on $\alpha$, $\sigma^\mathrm{residual}$, $\sigma^\mathrm{site}$, and $\sigma^\mathrm{observer}$ are included below.

\vspace{3mm}

**Stan code**

```{text, code=readLines("../mixed.stan")}
```

### 2017-12-07 manuscript/reviewer_reply.Rmd

--- 2017-11-20 manuscript/reviewer_reply.Rmd
@@
 First, and least important - normally submitted manuscripts do not hyphenate and do not right justify.*
+
+**Thank you. We have removed the right justification and hyphenation.**
 
@@
 *Line 124: I’m not as familiar with terrestrial projections, but are the CMIP5 projections relatively unbiased / precise when compared to recent data?*
+
+**The accuracy of CMIP5 models varies by the variable considered.
+Temperature is represented very well over North America.
+Precipitation is challenging, especially over mountainous terrain, but the CMIP5 ensemble represents precipitation across North America relatively well ("Evaluation of precipitation and temperature simulation performance of the CMIP3 and CMIP5 historical experiments", Koutroulis et al.
+2016).**
 

### 2017-12-07 manuscript/reviewer_reply.Rmd
rewrote "uninformative predictors"
You two were right and I misread: he did ask about all environmental models, not just boosting.

--- 2017-12-06 manuscript/reviewer_reply.Rmd
@@
 
-**We chose not to drop uninformative predictors for two reasons.
-First, boosted trees are relatively robust to their inclusion ("A working guide to boosted regression trees", Elith et al.
-2008), so it seemed unlikely to affect our results.
-Second, it would have added complexity to the model-fitting procedure, which already involved 500 rounds of cross-validation (one per Monte Carlo sample of the observer effect).**
+**We considered dropping uninformative predictors, but decided not to because of the complexity it would add to our analyses.
+In particular, it wasn't clear how to assess variable importance in a consistent way across three different model types and 500 Monte Carlo samples, and we did not want to bias the results by choosing a variable selection method that worked better for one method than another.**
 
@@
 
-**We have added an equation and clarified the wording here.**
+**We have added an in-line equation and clarified the wording**
 

### 2017-12-07 manuscript/reviewer_reply.Rmd
indent our responses

--- 2017-12-07 manuscript/reviewer_reply.Rmd
@@
 First, and least important - normally submitted manuscripts do not hyphenate and do not right justify.*
-
-**Thank you. We have removed the right justification and hyphenation.**
 
@@
 
-**We have switched the panels and labeled them. Thank you for noticing this error**.
+>We have switched the panels and labeled them. Thank you for noticing this error**.
 
@@
 
-**Thank you.**
+>Thank you.
 
@@
 
-**BBS data are generally viewed as annual data, even though the observations happen on a single day.
+>BBS data are generally viewed as annual data, even though the observations happen on a single day.
 The presence of an individual species is primarily a function of suitable habitat and, for many species, timing of migration.
-Environmental variables with a shorter temporal scale could likely improve models, such as incorporating cool weather days into the observation model, and would be worth pursuing in a different study.**
+Environmental variables with a shorter temporal scale could likely improve models, such as incorporating cool weather days into the observation model, and would be worth pursuing in a different study.
 
@@
 
-**This is correct, we've added clarification to the Methods.**
+>This is correct, we've added clarification to the Methods.
 
@@
 
-**This is correct, see the section "Richness Data" in the Methods.**
+>This is correct, see the section "Richness Data" in the Methods.
 
@@
 
-**See comment above on using shorter time scale environmental variables.**
+>See comment above on using shorter time scale environmental variables.
 
@@
 
-**The accuracy of CMIP5 models varies by the variable considered.
-Temperature is represented very well over North America.
-Precipitation is challenging, especially over mountainous terrain, but the CMIP5 ensemble represents precipitation across North America relatively well ("Evaluation of precipitation and temperature simulation performance of the CMIP3 and CMIP5 historical experiments", Koutroulis et al.
-2016).**
-
 *Line 138: Maybe also cite any of the interfaces you used, like RStan: http://mc-stan.org/users/citations/*
 
-**RStan is cited in the last paragraph of the Methods.**
+>RStan is cited in the last paragraph of the Methods.
 
@@
 
-**We have added this.**
+>We have added this.
 
@@
 
-**We have made this change**
+>We have made this change
 
@@
 
-**The randomForest package was cited in the final methods paragraph. We've cited it where it is first mentioned as well.**
+>The randomForest package was cited in the final methods paragraph. We've cited it where it is first mentioned as well.
 
 *Line 242: Again, I think including the equation here would be useful*
-
-**We have added an in-line equation and clarified the wording**
 
@@
 
-**The definition of hindcasting varies by field.
+>The definition of hindcasting varies by field.
 We use it here as it's used in meteorology, where any prediction of observations already available is generally a hindcast, and forecasts are only for predictions which are actually in the future.
 See Jolliffe & Stephenson 2003 (in manuscript) section 1.4.2 for a longer discussion.
-We've clarified this in the text.**
+We've clarified this in the text.
 
@@
 
-**See comment above on hindcasting.**
+>See comment above on hindcasting.
 
@@
 
-**Thank you for pointing out that we were unclear here.  The methods and purpose for the future predictions are now explained in more detail on lines [[XX]].**
+>Thank you for pointing out that we were unclear here.  The methods and purpose for the future predictions are now explained in more detail on lines [[XX]].
 
@@
 
-**Thank you for pointing out this error. We have fixed it**
+>Thank you for pointing out this error. We have fixed it
 

### 2017-12-07 manuscript/reviewer_reply.Rmd

@@
 
+**The accuracy of CMIP5 models varies by the variable considered.
+Temperature is represented very well over North America.
+Precipitation is challenging, especially over mountainous terrain, but the CMIP5 ensemble represents precipitation across North America relatively well ("Evaluation of precipitation and temperature simulation performance of the CMIP3 and CMIP5 historical experiments", Koutroulis et al.
+2016).**
+
 *Line 138: Maybe also cite any of the interfaces you used, like RStan: http://mc-stan.org/users/citations/*

### 2017-12-07 manuscript/reviewer_reply.Rmd

@@
 
+**We considered dropping uninformative predictors, but decided not to because of the complexity it would add to our analyses.
+In particular, it wasn't clear how to assess variable importance in a consistent way across three different model types and 500 Monte Carlo samples, and we did not want to bias the results by choosing a variable selection method that worked better for one method than another.**
+
 *Line 216: Cite randomForest / caret/ party R package – or whatever else you used?*

### 2017-12-07 manuscript/reviewer_reply.Rmd
Update spatial autocorrelation discussion
Basically used Ethan's version, but with improved punctuation and adding "for the broader goals of the paper. Moreover"

--- 2017-12-07 manuscript/reviewer_reply.Rmd
@@
 *Finally, your figure two seems to reverse panels A and B, neither of which is marked in the text.*
-
->We have switched the panels and labeled them. Thank you for noticing this error**.
 
@@
 
->Thank you.
-
 #### Introduction:
@@
 
->Thank you.
-
 *Line 71-73: I’m re-reading this after commenting on data, and am now a little confused.
@@
 
->BBS data are generally viewed as annual data, even though the observations happen on a single day.
+**BBS data are generally viewed as annual data, even though the observations happen on a single day.
 The presence of an individual species is primarily a function of suitable habitat and, for many species, timing of migration.
-Environmental variables with a shorter temporal scale could likely improve models, such as incorporating cool weather days into the observation model, and would be worth pursuing in a different study.
+Environmental variables with a shorter temporal scale could likely improve models, such as incorporating cool weather days into the observation model, and would be worth pursuing in a different study.**
 
@@
 
->This is correct, we've added clarification to the Methods.
+**This is correct, we've added clarification to the Methods.**
 
@@
 
->This is correct, see the section "Richness Data" in the Methods.
+**This is correct, see the section "Richness Data" in the Methods.**
 
@@
 
->See comment above on using shorter time scale environmental variables.
+**See comment above on using shorter time scale environmental variables.**
 
@@
 
->RStan is cited in the last paragraph of the Methods.
+**RStan is cited in the last paragraph of the Methods.**
 
@@
 One hand these models might be overkill if the habitat/environmental variables (which are spatially correlated) explain variability in richness, but the spatial correlation may pick up other sources of variability not accounted for by those variables.*
+
+>The goal of this model is to control for observer effects, not to provide a focal model of the system (though we do use it for the average model as a convenience).
+Because observers typically observe the same site year after year it was necessary to include site in this model to avoid classifying variation among sites as differences among observers and incorrectly removing it before fitting other models.
+Given the use of this model, it is unclear what including spatial structure would accomplish for the broader goals of the paper.
+Moreover, it seems possible that it could end up correcting out meaningful ecological patterns and therefore confuse the comparison of more complex models to the baselines.
 
@@
 
->We have added this.
+**We have added this.**
 
@@
 
->We have made this change
+** We have made this change**
 
@@
 
->The randomForest package was cited in the final methods paragraph. We've cited it where it is first mentioned as well.
+**The randomForest package was cited in the final methods paragraph. We've cited it where it is first mentioned as well.**
 
@@
 
->The definition of hindcasting varies by field.
+**The definition of hindcasting varies by field.
 We use it here as it's used in meteorology, where any prediction of observations already available is generally a hindcast, and forecasts are only for predictions which are actually in the future.
 See Jolliffe & Stephenson 2003 (in manuscript) section 1.4.2 for a longer discussion.
-We've clarified this in the text.
+We've clarified this in the text.**
 
@@
 
->See comment above on hindcasting.
+**See comment above on hindcasting.**
 
@@
 
->Thank you for pointing out that we were unclear here.  The methods and purpose for the future predictions are now explained in more detail on lines [[XX]].
-
 *Fig. 2: Panels not labeled. No key for colors. Somewhat confusing.*
@@
 
->Thank you for pointing out this error. We have fixed it
-
 *466: only other mention of future forecasts… not sure why this is included in this paper.*

### 2017-12-08 manuscript/reviewer_reply.Rmd

--- 2017-12-07 manuscript/reviewer_reply.Rmd
@@
 
-**Thank you. We have removed the right justification and hyphenation.**
+>Thank you. We have removed the right justification and hyphenation.
 
@@
 
->We have switched the panels and labeled them. Thank you for noticing this error**.
+>We have switched the panels and labeled them. Thank you for noticing this error.
 
@@
 
+>This is a good point. We have added the word "terrestrial" to this line.
+
 *Line 62: I really like the recommendations / clarity in Box 1, and think these will be accessible to most ecologists*
@@
 
-**The accuracy of CMIP5 models varies by the variable considered.
+>The accuracy of CMIP5 models varies by the variable considered.
 Temperature is represented very well over North America.
 Precipitation is challenging, especially over mountainous terrain, but the CMIP5 ensemble represents precipitation across North America relatively well ("Evaluation of precipitation and temperature simulation performance of the CMIP3 and CMIP5 historical experiments", Koutroulis et al.
-2016).**
+2016).
 
@@
 *Line 141-143: I think the paper thus far is pretty clear. I was hoping to see a few equations here – it helps me especially when components start to become hierarchical!*
+
+>We have added the equations and Stan code as a supplemental file.
 
@@
 
+>We considered dropping uninformative predictors, but decided not to because of the complexity it would add to our analyses.
+In particular, it wasn't clear how to assess variable importance in a consistent way across three different model types and 500 Monte Carlo samples, and we did not want to bias the results by choosing a variable selection method that worked better for one method than another.
+
 *Line 216: Cite randomForest / caret/ party R package – or whatever else you used?*
@@
 
-**We have added an in-line equation and clarified the wording**
+>We have added an in-line equation and clarified the wording
 
@@
 Using RMSE here is probably ok because you probably don’t have zeros/huge values/etc*
+
+>We agree.
 
@@
 
+>Thank you for pointing this out.  As the Editor noted, we had inadvertently reversed the two panels. We have fixed this, and also labeled them as you suggested.
+
 *Fig. 6 caption: “compoonents”*

### 2017-12-08 manuscript/observer-equations.Rmd
auto-arima details in the suppliment

@@
 ```
+
+**Details of the auto.arima model**
+
+We used non-seasonal time series model $ARIMA(p,d,q)$, with drift allowed, in the function 'auto.arima' of the R package `forecast`.
+The search space for $p$ and $q$ were $(2,5)$ and $(2,5)$, respectively.
+$d$ was selected via KPSS unit-root tests as described in Hyndman and Khandakar (2008) and with a maximum value of $2$.
+Stepwise model selection was done using AICc as the model criteria.

### 2017-12-08 manuscript/reviewer_reply.Rmd
reviewer answer and suppliment comment

--- 2017-12-08 manuscript/reviewer_reply.Rmd
@@
 
->Thank you. We have removed the right justification and hyphenation.
+**Thank you. We have removed the right justification and hyphenation.**
 
@@
 
->We have switched the panels and labeled them. Thank you for noticing this error.
+>We have switched the panels and labeled them. Thank you for noticing this error**.
 
@@
 
->This is a good point. We have added the word "terrestrial" to this line.
-
 *Line 62: I really like the recommendations / clarity in Box 1, and think these will be accessible to most ecologists*
@@
 
->The accuracy of CMIP5 models varies by the variable considered.
+**The accuracy of CMIP5 models varies by the variable considered.
 Temperature is represented very well over North America.
 Precipitation is challenging, especially over mountainous terrain, but the CMIP5 ensemble represents precipitation across North America relatively well ("Evaluation of precipitation and temperature simulation performance of the CMIP3 and CMIP5 historical experiments", Koutroulis et al.
-2016).
+2016).**
 
@@
 *Line 141-143: I think the paper thus far is pretty clear. I was hoping to see a few equations here – it helps me especially when components start to become hierarchical!*
-
->We have added the equations and Stan code as a supplemental file.
 
@@
 
+>We agree that software can evolve over time and make methods non-repricable. We have add details of the auto.arima model in the suppliment.
+
 *Table 1 caption: without getting to more methods (below), I don’t understand the sentence ‘Environmental models were trained using all sites together, without information regarding which transects occurred at which site or during which year’.*
@@
 
->We considered dropping uninformative predictors, but decided not to because of the complexity it would add to our analyses.
-In particular, it wasn't clear how to assess variable importance in a consistent way across three different model types and 500 Monte Carlo samples, and we did not want to bias the results by choosing a variable selection method that worked better for one method than another.
-
 *Line 216: Cite randomForest / caret/ party R package – or whatever else you used?*
@@
 
->We have added an in-line equation and clarified the wording
+**We have added an in-line equation and clarified the wording**
 
@@
 Using RMSE here is probably ok because you probably don’t have zeros/huge values/etc*
-
->We agree.
 
@@
 
->Thank you for pointing this out.  As the Editor noted, we had inadvertently reversed the two panels. We have fixed this, and also labeled them as you suggested.
-
 *Fig. 6 caption: “compoonents”*

### 2017-12-08 manuscript/draft.Rmd
reviewer answer and suppliment comment

--- 2017-12-06 manuscript/draft.Rmd
@@
 points' colors). In this example most models are made more robust to observer 
-turnover. Note that the "naive" model is less sensitive to observer turnover,
-and does not benefit as much from modeling
+turnover by including an observer model. Note that the "naive" model is less 
+sensitive to observer turnover, and does not benefit as much from modeling
 it.](../figures/observer_predictions.png)
@@
 did not include a season component in these models. Otherwise we used the
-default settings for this function [@forecast1].
+default settings for this function (See suppliment for details).
 

### 2017-12-08 manuscript/draft.Rmd
typos

--- 2017-12-07 manuscript/draft.Rmd
@@
 did not include a season component in these models. Otherwise we used the
-default settings for this function [@forecast1].
+default settings for this function (See supplement for details).
 

### 2017-12-08 manuscript/draft.Rmd
more results in the abstract?

@@
 spatial data into useful temporal predictions about biodiversity at decadal
-time-scales. The best practices implemented in this study directly influence the
-forecasts, the relative performance of different modeling approaches, and the
-conclusions about the current state of biodiversity forecasting . To facilitate
-the rapid improvement of biodiversity forecasts, we emphasize the value of
-specific best practices in making forecasts and evaluating forecasting methods.
+time-scales. Most methods' prediction intervals were also too narrow: twice as
+many observations fell outside the predicted 95% interval as should be expected.
+The intervals produced by "stacking" species distribution models---perhaps the
+most common forecasting approach for terrestrial species richness---were
+especially misleading, failing to contain the true value at 4-6 times the
+nominal rate.  The best practices implemented in this study directly influence
+the forecasts, the relative performance of different modeling approaches, and
+the conclusions about the current state of biodiversity forecasting. To
+facilitate the rapid improvement of biodiversity forecasts, we emphasize the
+value of specific best practices in making forecasts and evaluating forecasting
+methods.
 

### 2017-12-08 manuscript/draft.Rmd
is this unfair to SDMs?
in response to reviewer 2

--- 2017-12-08 manuscript/draft.Rmd
@@
 for future biodiversity. While this study is restricted to breeding birds in
-North America, its results are consistent with a growing literature on the
-limits of ecological forecasting, as discussed below.
+simple baselines. Since site-level richness is relatively stable in so many
+systems [@dornelas2014], spatial validation will not be enough: a model can have
+high accuracy across spatial gradients without being able to predict changes
+over time. This gap between spatial and temporal accuracy is known to be
+important for species-level predictions [@rapacciuolo2012; @oedekoven2017]; our
+results indicate that it is substantial for higher-level patterns like richness
+as well. SDMs' poor temporal predictions are particularly sobering, as these
+models have been one of the main foundations for estimates of the predicted loss
+of biodiversity to climate change over the past decade or so [@thomas2004;
+@thuiller2011; @urban2015]. One could argue that these conclusions are unfair to
+SDMs and related methods, which are generally used to predict large directional
+changes over multi-decade timescales rather than fluctuations within a single
+decade. It is certainly possible that these models' relative performance would
+improve over time or with larger shifts in richness, and future research should
+explore this possibility.
 
-The most commonly used methods for forecasting future biodiversity, SDMs and
-macroecological models, both produced worse forecasts than time-series models
-and simple baselines. This weakness suggests that predictions about future
-biodiversity change should be viewed with skepticism unless the underlying
-models have been validated temporally, via hindcasting and comparison with
-simple baselines. Since site-level richness is relatively stable, spatial
-validation is not enough: a model can have high accuracy across spatial
-gradients without being able to predict changes over time. This gap between
-spatial and temporal accuracy is known to be important for species-level
-predictions [@rapacciuolo2012; @oedekoven2017]; our results indicate that it is
-substantial for higher-level patterns like richness as well. SDMs' poor temporal
-predictions are particularly sobering, as these models have been one of the main
-foundations for estimates of the predicted loss of biodiversity to climate
-change over the past decade or so [@thomas2004; @thuiller2011; @urban2015]. Our
-results also highlight the importance of comparing multiple modeling approaches
-when conducting ecological forecasts, and in particular, the value of comparing
-results to simple baselines to avoid over-interpreting the information present
-in these forecasts [Box 1]. Disciplines that have more mature forecasting
-cultures often do this by reporting "forecast skill", i.e., the improvement in
-the forecast relative to a simple baseline [@jolliffe2003]. We recommend
-following the example of @ye2015 and adopting this approach in future ecological
-forecasting research.
+Our results also highlight the importance of comparing multiple modeling
+approaches when conducting ecological forecasts, and in particular, the value of
+comparing results to simple baselines to avoid over-interpreting the information
+present in these forecasts [Box 1]. Disciplines that have more mature
+forecasting cultures often do this by reporting "forecast skill", i.e., the
+improvement in the forecast relative to a simple baseline [@jolliffe2003]. We
+recommend following the example of @ye2015 and adopting this approach in future
+ecological forecasting research.
 

### 2017-12-08 manuscript/reviewer_reply.Rmd
typos
Update reviewer_reply.Rmd

--- 2017-12-08 manuscript/reviewer_reply.Rmd
@@
 To me, it is not surprising that models do not "...successfully turn spatial data into useful temporal predictions about biodiversity at decadal time-scales." But, I wonder if you chose data WITH a trend, you would have gotten other results.*
+
+>>We added a "future directions" paragraph to address this.
+Non-directional changes, such as fluctuations from year-to-year, are forecastable in principle, although we agree that they may be harder.
+Our goal was to evaluate how well this worked for as many time series as possible.
+We agree that it would be interesting to examine the subset of locations that showed strong directional change and see if those are easier to forecast.
+We plan on coming back to investigate these questions at a later date.
 
@@
 And whether or not you included a drift parameter?*
+
+>We agree that software can evolve over time and make methods non-repricable. We have added details of the auto.arima model in the supplement.
 
@@
 
+>We agree that a comparison of model performance with different train/test splits would be interesting, and plan to make such comparisons in future work.
+
 *124-130: I didn’t see where future projections were performed/used in this paper. Why is this here?*

### 2017-12-08 manuscript/reviewer_reply.Rmd
revise editor reply

--- 2017-12-08 manuscript/reviewer_reply.Rmd
@@
 To me, it is not surprising that models do not "...successfully turn spatial data into useful temporal predictions about biodiversity at decadal time-scales." But, I wonder if you chose data WITH a trend, you would have gotten other results.*
+
+>We now address this in lines [[xx]].
+We agree that non-directional changes, such as fluctuations from year-to-year, will generally be harder to predict, although it should still be possible to predict some of them.
+Some of the sites did show directional changes, and we agree that it would be interesting to examine this subset to see if those are easier to forecast.
+We plan on coming back to investigate these questions at a later date.
 

### 2017-12-08 manuscript/reviewer_reply.Rmd
Update reviewer_reply.Rmd
reviewer answer

@@
 To me, it is not surprising that models do not "...successfully turn spatial data into useful temporal predictions about biodiversity at decadal time-scales." But, I wonder if you chose data WITH a trend, you would have gotten other results.*
-
->We now address this in lines [[xx]].
-We agree that non-directional changes, such as fluctuations from year-to-year, will generally be harder to predict, although it should still be possible to predict some of them.
-Some of the sites did show directional changes, and we agree that it would be interesting to examine this subset to see if those are easier to forecast.
-We plan on coming back to investigate these questions at a later date.
 
@@
 
+>We have clarified the Table 1 caption to be more clear about model covariates.
+
 *Table 1: Just my personal preference, but I’d include the check marks and replace the ‘Xs’ with blank cells*
@@
 
->We agree that a comparison of model performance with different train/test splits would be interesting, and plan to make such comparisons in future work.
-
 *124-130: I didn’t see where future projections were performed/used in this paper. Why is this here?*

### 2017-12-08 manuscript/draft.Rmd

--- 2017-12-08 manuscript/draft.Rmd
@@
 table_caption = "Six forecasting models. Single-site models were trained
-site-by-site, without environmental data. Environmental models were trained
-using all sites together, without information regarding which transects occurred
-at which site or during which year. Most of the models were trained to predict
-richness directly. This mirrors the standard application of these
+site-by-site, without environmental data. Environmental models were trained at
+the continental scale, using only environmental variables (as opposed to site
+or time series information) as predictors. Most of the models were trained to 
+predict richness directly. This mirrors the standard application of these
 techniques. Separate random forest SDMs were fit for each species and used to

### 2017-12-08 manuscript/draft.Rmd
Update draft.Rmd

--- 2017-12-08 manuscript/draft.Rmd
@@
 results indicate that it is substantial for higher-level patterns like richness
-as well. SDMs' poor temporal predictions are particularly sobering, as these
-models have been one of the main foundations for estimates of the predicted loss
-of biodiversity to climate change over the past decade or so [@thomas2004;
+as well.
+
+SDMs' poor temporal predictions are particularly sobering, as these models have
+been one of the main foundations for estimates of the predicted loss of
+biodiversity to climate change over the past decade or so [@thomas2004;
 @thuiller2011; @urban2015]. One could argue that these conclusions are unfair to
@@
 improve over time or with larger shifts in richness, and future research should
-explore this possibility.
+explore this possibility. Nevertheless, our results have important implications
+for ecologists' long-term predictions regarding large changes in species
+richness, for two reasons. First, after controlling for observer effects, we
+found no evidence that these models have any ability to track changes in
+richness over time at all (Figure 6). Given that prediction error steadily
+increases over time (Figure 5), we expect that long-term SDM-based forecasts
+will generally be at least as poor as decade-scale ones.
 
@@
 @harris2015 and @warton2015, we found that JSDMs' wider prediction intervals
-enabled them to avoid this problem. Macroecological models appear to share this
-advantage, while being considerably easier to implement.
+mostly enabled them to avoid this problem. Macroecological models appear to share 
+this advantage, while being considerably easier to implement.
 

### 2017-12-08 manuscript/reviewer_reply.Rmd
Revise Discussion in response to questions about forecastability
This addresses the concerns of both the associate editor and Reviewer 2 about
the forecastability of BBS richness in the chosen time period. In adds a new
paragraph to the Discussion that expands on our previous discussion of these
points and calls for future research in this area.

Includes associated updates to the letter.

@@
 To me, it is not surprising that models do not "...successfully turn spatial data into useful temporal predictions about biodiversity at decadal time-scales." But, I wonder if you chose data WITH a trend, you would have gotten other results.*
+
+>We chose the North American Breeding Bird Survey data because it is one of the largest scale most long running spatiotemporal datasets available, thus making this kind of quantitative assessment of both spatial and temporally focused forecasting methods possible.
+We agree that the relative stability of these time-series is important, which is why we dedicated a large discussion paragraph to this point in the initial submission.
+To further emphasize this point and to highlight the idea that assessing these methods on selectively chosen data exhibiting significant trends is interesting we have added an additional paragraph to the discussion addressing this point directly and recommending it for future research.
 
@@
 
+>We agree that the relative shortness of the forecasting window and the lack of major directional shifts in richness (on average) are important to consider in interpretting our results.
+This is why we included long paragraphs on both topics in the initial submission.
+To further emphasize this point and to highlight the importance of longer time scales analyses and assessments on data showing larger changes in richness we have added an additional paragraph to the discussion addressing these points directly and recommending them for future research.
+
 *Here are some comments:*
@@
 However, I do realize that BBS data are detailed and compelling data to use given what a large amount of detailed time-series it provides.*
+
+>As the reviewer notes the North American Breeding Bird Survey data is one of the largest scale most long running spatiotemporal datasets available.
+We chose it because these features make this kind of quantitative assessment of both spatial and temporally focused forecasting methods possible.
+While there are downsides to any dataset we do not agree that birds uniquely problematic for this type of analysis.
+In addition to large amount of available data the vagility of birds potentially allows them to respond more quickly to changes in environmental conditions due to their low dispersal limitation.
+Imperfect detection and non-environmental drivers are common challenges in most animal species.
 
@@
 
+>We agree that understanding the conditions under which forecasts are likely to perform well is an important step for ecological forecasting.
+We have added a sentence to the discussion making this point.
+That said, most ecological forecasts are intended to be applied to the future, which at least at the moment we would expect to remain "incredibly dynamic".
+As such the dynamics of the last decade are likely a reasonable representation of the kinds of dynamics we are trying to forecast.
+
 *124-130: I didn’t see where future projections were performed/used in this paper. Why is this here?*

### 2017-12-08 manuscript/draft.Rmd
Revise Discussion in response to questions about forecastability
This addresses the concerns of both the associate editor and Reviewer 2 about
the forecastability of BBS richness in the chosen time period. In adds a new
paragraph to the Discussion that expands on our previous discussion of these
points and calls for future research in this area.

Includes associated updates to the letter.

--- 2017-12-08 manuscript/draft.Rmd
@@
 perform relatively well at predicting future richness, despite failing to
-capture changes in richness over time. However, this stability also makes it
+capture changes in richness over time.
+
+The relatively stable nature of the BBS richness time-series also makes it
 difficult to improve forecasts relative to simple baselines, since those
 baselines are already close to representing what is actually occurring in the
-system. These results suggest that single-site models should be actively
-considered for forecasts of richness and other stable aspects of biodiversity.
-Our results also suggest that future efforts to understand and forecast
-biodiversity should incorporate species composition, since lower-level processes
-are expected to be more dynamic [@ernest2001; @dornelas2014] and contain more
-useful information [@harris2015].
+system. It is possible that in systems exhibiting directional changes in
+richness and other biodiversity measures that models based on spatial patterns
+may yield better forecasts. Future research in this area should determine if
+regions or time periods exhibiting strong directional changes in biodiveristy
+are better predicted by these models and also extend our forecast horizon
+analyses to longer timescales where possible. Our results also suggest that
+future efforts to understand and forecast biodiversity should incorporate
+species composition, since lower-level processes are expected to be more dynamic
+[@ernest2001; @dornelas2014] and contain more information about how the systems
+are changing [@harris2015]. More generally, determining the forecastability of
+different aspects of ecological systems under different conditions is an
+important next step for the future of ecological forecasting.
 

### 2017-12-08 manuscript/draft.Rmd

--- 2017-12-08 manuscript/draft.Rmd
@@
 decade at `r numbers$N_predict_sites` sites distributed throughout the
-continental United States. While each model could explain most of the variance
-in richness, none of them consistently outperformed a baseline model that
-predicted constant richness at each site. In particular, we found no evidence
-that current methods (such as species distribution models) can successfully turn
-spatial data into useful temporal predictions about biodiversity at decadal
-time-scales. Most methods' prediction intervals were also too narrow: twice as
-many observations fell outside the predicted 95% interval as should be expected.
-The intervals produced by "stacking" species distribution models---perhaps the
-most common forecasting approach for terrestrial species richness---were
-especially misleading, failing to contain the true value at 4-6 times the
-nominal rate.  The best practices implemented in this study directly influence
-the forecasts, the relative performance of different modeling approaches, and
-the conclusions about the current state of biodiversity forecasting. To
-facilitate the rapid improvement of biodiversity forecasts, we emphasize the
-value of specific best practices in making forecasts and evaluating forecasting
-methods.
+continental United States. All models explained more than 50% of the variance in
+richness, but none of them consistently outperformed a baseline model that
+predicted constant richness at each site. The best practices implemented in this
+study directly influenced the forecasts and evaluations. Stacked species
+distribution models and "naive" forecasts produced poor estimates of uncertainty
+and accounting for this resulted in these models dropping in the relative
+performance compared to other models. Accounting for observer effects improved
+model performance overall, but also changed the rank ordering of models because
+it did not improve the accuracy of the "naive" model. Considering the forecast
+horizon revealed that the prediction accuracy decreased across all models as the
+time horizon of the forecast increased. To facilitate the rapid improvement of
+biodiversity forecasts, we emphasize the value of specific best practices in
+making forecasts and evaluating forecasting methods.
 

### 2017-12-08 manuscript/reviewer_reply.Rmd
More Ward replies

--- 2017-12-08 manuscript/reviewer_reply.Rmd
@@
 
+>See our comments about spatial autocorrelation above
+
 *Line 276: Is one of the potential reasons that richness is stationary that presence/absence on the BBS varies less than other measures – maybe abundance, or other measures of diversity are more variable?*
 
+>We agree that this is likely. We make this point on [[lines]]
+
 *Line 294: Is the narrow predictive distribution for the stacked SDM also why this exhibits the strongest trend in decreasing coverge as a function of forecast length (Fig 5)?*
+
+>Yes, exactly. More and more observations fall outside the narrow prediction intervals.
 
@@
 
+>We are currently working on a few projects along these lines.
+
 *Line 461: Another thread with conservation implications is that back in the 1990s, it was pretty common for single species PVA models to be forecasted 100-200 years.
@@
 In Steve Beissinger’s PVA book he cites Goldwasser (2000) “Variability and measurement error in extinction risk analysis...” for this point (p 133).*
+
+>We agree that this is an apt comparion. However, we are unfamiliar with this literature and UF does not have access to the paper you cited.
 

### 2017-12-08 manuscript/reviewer_reply.Rmd

@@
 
->We agree that this is an apt comparion. However, we are unfamiliar with this literature and UF does not have access to the paper you cited.
+>We agree that this is an apt comparison.
+However, we are unfamiliar with this literature and UF does not have access to the paper you cited.
+The existing citations should prove sufficient to justify these points.
 
@@
 The authors give all of the keywords to indicate that they know what they are doing.*
+
+>We agree that predicting smaller, transient shifts in richness may be more difficult than predicting large, sustained ones.
+Nevertheless, we believe that our results have implications for ecologists' longer-term predictions of declining richness, as discussed on lines [[XX]].
 

### 2017-12-08 manuscript/draft.Rmd

--- 2017-12-08 manuscript/draft.Rmd
@@
 decade at `r numbers$N_predict_sites` sites distributed throughout the
-continental United States. While each model could explain most of the variance
-in richness, none of them consistently outperformed a baseline model that
-predicted constant richness at each site. In particular, we found no evidence
-that current methods (such as species distribution models) can successfully turn
-spatial data into useful temporal predictions about biodiversity at decadal
-time-scales. The best practices implemented in this study directly influence the
-forecasts, the relative performance of different modeling approaches, and the
-conclusions about the current state of biodiversity forecasting . To facilitate
-the rapid improvement of biodiversity forecasts, we emphasize the value of
-specific best practices in making forecasts and evaluating forecasting methods.
+continental United States. All models explained more than 50% of the variance in
+richness, but none of them consistently outperformed a baseline model that
+predicted constant richness at each site. The best practices implemented in this
+study directly influenced the forecasts and evaluations. Stacked species
+distribution models and "naive" forecasts produced poor estimates of uncertainty
+and accounting for this resulted in these models dropping in the relative
+performance compared to other models. Accounting for observer effects improved
+model performance overall, but also changed the rank ordering of models because
+it did not improve the accuracy of the "naive" model. Considering the forecast
+horizon revealed that the prediction accuracy decreased across all models as the
+time horizon of the forecast increased. To facilitate the rapid improvement of
+biodiversity forecasts, we emphasize the value of specific best practices in
+making forecasts and evaluating forecasting methods.
 

### 2017-12-09 manuscript/draft.Rmd
explain future predictions

--- 2017-12-08 manuscript/draft.Rmd
@@
 
-**Future environmental projections.** We made long term forecasts from 2014-2050
-using the CMIP5 multi-model ensemble dataset as the source for climate
-variables [@brekke2013]. Precipitation and temperature from 37 downscaled model
-runs [@brekke2013, see Table S1] using the RCP6.0 scenario were averaged
-together to create a single ensemble used to calculate the bioclimatic variables
-for North America. For NDVI we used the per-site average values from 2000-2013
-as a simple forecast. For observer effects (see below) each site was set to have
-zero observer bias.
+**Future environmental projections.** In addition to the analyses presented
+here, we have also generated and archived long term forecasts from 2014-2050.
+This will allow future researchers to assess the performance of our six models
+on longer time horizons as more years of BBS data become available.
+Precipitation and temperature were forecast using the CMIP5 multi-model ensemble
+dataset [@brekke2013]. 37 downscaled model runs [@brekke2013, see Table S1]
+using the RCP6.0 scenario were averaged together to create a single ensemble
+used to calculate the bioclimatic variables for North America. For NDVI, we used
+the per-site average values from 2000-2013 as a simple forecast. For observer
+effects (see below), each site was set to have zero observer bias. The
+predictions have been archived at [@zenodoPredict].
 

### 2017-12-09 manuscript/reviewer_reply.Rmd
explain future predictions
explain abstract changes

--- 2017-12-08 manuscript/reviewer_reply.Rmd
@@
 
+>We have added a summary of our results and findings.
+
 *57-58: this is not the definition of hindcasting.
@@
 *466: only other mention of future forecasts… not sure why this is included in this paper.*
+
+> Thank you again for pointing out that we were unclear here. The purpose of the future forecasts is now explained on lines [[xx]] and [[xx]].

### 2017-12-09 manuscript/reviewer_reply.Rmd
reply about deep time studies

--- 2017-12-08 manuscript/reviewer_reply.Rmd
@@
 
->See comment above on hindcasting.
+>See comment above on hindcasting. We agree that studies evaluating models using fossil records are an important aspect of this field of research and have included several references for them.
 

### 2017-12-09 manuscript/draft.Rmd
reply about deep time studies

--- 2017-12-08 manuscript/draft.Rmd
@@
 difficulty in assessing whether the predicted changes occurred; @dietze2016].
-What we do know comes from a small number of hindcasting studies, where models
-are built using data on species occurrence and richness from the past and
-evaluated on their ability to predict contemporary patterns [e.g., @algar2009;
-@distler2015]. These studies are a valuable first step, but lack several
-components that are important for developing forecasting models with high
-predictive accuracy, and for understanding how well different methods can
-predict the future. These "best practices" for effective forecasting and
-evaluation (Box 1) broadly involve: 1) expanding the use of data to include
-biological and environmental time-series [@tredennick2016]; 2) accounting for
-uncertainty in observations and processes, [@harris2015; @yu2010]; and 3)
-conducting meaningful evaluations of the forecasts by hindcasting, archiving
-short-term forecasts, and comparing forecasts to baselines to determine whether
-the forecasts are more accurate than assuming the system is basically static
-[@perretti2013].
+What we do know comes from a small number of biodiversity studies, where models
+are built using data on species occurrence, richness, or community composition
+from distinct timeperiods and evaluated on their ability to predict contemporary
+[@algar2009; @distler2015] or historic [@maguire2016; @blois2013] patterns.
+These studies are a valuable first step, but lack several components that are
+important for developing forecasting models with high predictive accuracy, and
+for understanding how well different methods can predict the future. These "best
+practices" for effective forecasting and evaluation (Box 1) broadly involve: 1)
+expanding the use of data to include biological and environmental time-series
+[@tredennick2016]; 2) accounting for uncertainty in observations and processes,
+[@harris2015; @yu2010]; and 3) conducting meaningful evaluations of the
+forecasts by hindcasting, archiving short-term forecasts, and comparing
+forecasts to baselines to determine whether the forecasts are more accurate than
+assuming the system is basically static [@perretti2013].
 

### 2017-12-09 manuscript/reviewer_reply.Rmd
answer about the long term forecasts

@@
 
->See comment above on hindcasting. We agree that studies evaluating models using fossil records are an important aspect of this field of research and have included several references for them.
+>See comment above on hindcasting.
 
@@
 *466: only other mention of future forecasts… not sure why this is included in this paper.*
+
+>As discussed throughout the paper we feel that forecast evaluation is beneficial to the field of ecology.
+While we do not perform any analysis on these long term forecasts, they will be archived for future evalation as a form of pre-registration.
+Since they are true forecasts (ie.
+predictions beyond the present day) evaluation of them will be a valuable contribution in the coming years.

### 2017-12-09 manuscript/reviewer_reply.Rmd

--- 2017-12-09 manuscript/reviewer_reply.Rmd
@@
 If the authors were to make splits based on expected periods of dynamism, they could better tell whether the drivers that are included in the models are accurate predictors.*
+
+>We agree that understanding the conditions under which forecasts are likely to perform well is an important step for ecological forecasting.
+We have added a sentence to the discussion making this point.
+That said, most ecological forecasts are intended to be applied to the future, which at least at the moment we would expect to remain "incredibly dynamic".
+As such the dynamics of the last decade are likely a reasonable representation of the kinds of dynamics we are trying to forecast.
 

### 2017-12-09 manuscript/draft.Rmd

@@
 difficulty in assessing whether the predicted changes occurred; @dietze2016].
-What we do know comes from a small number of biodiversity studies, where models
-are built using data on species occurrence, richness, or community composition
-from distinct timeperiods and evaluated on their ability to predict contemporary
-[@algar2009; @distler2015] or historic [@maguire2016; @blois2013] patterns.
-These studies are a valuable first step, but lack several components that are
-important for developing forecasting models with high predictive accuracy, and
-for understanding how well different methods can predict the future. These "best
-practices" for effective forecasting and evaluation (Box 1) broadly involve: 1)
-expanding the use of data to include biological and environmental time-series
-[@tredennick2016]; 2) accounting for uncertainty in observations and processes,
-[@harris2015; @yu2010]; and 3) conducting meaningful evaluations of the
-forecasts by hindcasting, archiving short-term forecasts, and comparing
-forecasts to baselines to determine whether the forecasts are more accurate than
-assuming the system is basically static [@perretti2013].
+What we do know comes from a small number of hindcasting studies, where models
+are built using data on species occurrence and richness from the past and
+evaluated on their ability to predict contemporary patterns [e.g., @algar2009;
+@distler2015]. These studies are a valuable first step, but lack several
+components that are important for developing forecasting models with high
+predictive accuracy, and for understanding how well different methods can
+predict the future. These "best practices" for effective forecasting and
+evaluation (Box 1) broadly involve: 1) expanding the use of data to include
+biological and environmental time-series [@tredennick2016]; 2) accounting for
+uncertainty in observations and processes, [@harris2015; @yu2010]; and 3)
+conducting meaningful evaluations of the forecasts by hindcasting, archiving
+short-term forecasts, and comparing forecasts to baselines to determine whether
+the forecasts are more accurate than assuming the system is basically static
+[@perretti2013].
 

### 2017-12-09 manuscript/draft.Rmd
changes suggested by ethan

--- 2017-12-09 manuscript/draft.Rmd
@@
 difficulty in assessing whether the predicted changes occurred; @dietze2016].
-What we do know comes from a small number of biodiversity studies, where models
-are built using data on species occurrence, richness, or community composition
-from distinct timeperiods and evaluated on their ability to predict contemporary
-[@algar2009; @distler2015] or historic [@maguire2016; @blois2013] patterns.
-These studies are a valuable first step, but lack several components that are
-important for developing forecasting models with high predictive accuracy, and
-for understanding how well different methods can predict the future. These "best
+What we do know comes from a small number of hindcasting studies, where models
+are built from different time periods and evaluated on their ability to predict
+biodiversity patterns in contemporary [@algar2009; @distler2015] or historic
+[@maguire2016; @blois2013] periods not used for model fitting. These studies are
+a valuable first step, but lack several components that are important for
+developing forecasting models with high predictive accuracy, and for
+understanding how well different methods can predict the future. These "best
 practices" for effective forecasting and evaluation (Box 1) broadly involve: 1)

### 2017-12-09 manuscript/draft.Rmd

--- 2017-12-09 manuscript/draft.Rmd
@@
 What we do know comes from a small number of hindcasting studies, where models
-are built using data on species occurrence and richness from the past and
-evaluated on their ability to predict contemporary patterns [e.g., @algar2009;
-@distler2015]. These studies are a valuable first step, but lack several
-components that are important for developing forecasting models with high
-predictive accuracy, and for understanding how well different methods can
-predict the future. These "best practices" for effective forecasting and
-evaluation (Box 1) broadly involve: 1) expanding the use of data to include
-biological and environmental time-series [@tredennick2016]; 2) accounting for
-uncertainty in observations and processes, [@harris2015; @yu2010]; and 3)
-conducting meaningful evaluations of the forecasts by hindcasting, archiving
-short-term forecasts, and comparing forecasts to baselines to determine whether
-the forecasts are more accurate than assuming the system is basically static
-[@perretti2013].
+are built from different time periods and evaluated on their ability to predict
+biodiversity patterns in contemporary [@algar2009; @distler2015] or historic
+[@maguire2016; @blois2013] periods not used for model fitting. These studies are
+a valuable first step, but lack several components that are important for
+developing forecasting models with high predictive accuracy, and for
+understanding how well different methods can predict the future. These "best
+practices" for effective forecasting and evaluation (Box 1) broadly involve: 1)
+expanding the use of data to include biological and environmental time-series
+[@tredennick2016]; 2) accounting for uncertainty in observations and processes,
+[@harris2015; @yu2010]; and 3) conducting meaningful evaluations of the
+forecasts by hindcasting, archiving short-term forecasts, and comparing
+forecasts to baselines to determine whether the forecasts are more accurate than
+assuming the system is basically static [@perretti2013].
 

### 2017-12-09 manuscript/reviewer_reply.Rmd

@@
 
+>Thanks!
+We tried to handle this using binder (https://mybinder.org/) to allow anyone to quickly engage with the full environment, but unfortunately the R support isn't quite ready yet.
+Fortunately it is under active development and as soon as it is available we plan to set it up and link to it from the repository to support this use case.
+
 *Line 269: See my above comments about spatially correlated random effects / GMRF models.

### 2017-12-09 manuscript/draft.Rmd
Reword sentence that used "forecast" three times in 13 words
equations for baselines

--- 2017-12-09 manuscript/draft.Rmd
@@
 
-**Future environmental projections.** In addition to the analyses presented
-here, we have also generated and archived long term forecasts from 2014-2050.
-This will allow future researchers to assess the performance of our six models
-on longer time horizons as more years of BBS data become available.
-Precipitation and temperature were forecast using the CMIP5 multi-model ensemble
-dataset [@brekke2013]. 37 downscaled model runs [@brekke2013, see Table S1]
-using the RCP6.0 scenario were averaged together to create a single ensemble
-used to calculate the bioclimatic variables for North America. For NDVI, we used
-the per-site average values from 2000-2013 as a simple forecast. For observer
-effects (see below), each site was set to have zero observer bias. The
-predictions have been archived at [@zenodoPredict].
+**Future environmental projections.** We made long term forecasts from 2014-2050
+using the CMIP5 multi-model ensemble dataset as the source for climate
+variables [@brekke2013]. Precipitation and temperature from 37 downscaled model
+runs [@brekke2013, see Table S1] using the RCP6.0 scenario were averaged
+together to create a single ensemble used to calculate the bioclimatic variables
+for North America. For NDVI we used the per-site average values from 2000-2013
+as a simple forecast. For observer effects (see below) each site was set to have
+zero observer bias.
 
@@
 **Baseline models.** We used two simple baseline models as a basis for
-comparison with the more complex models (Figure 2A). These baselines treated
-site-level richness observations either as uncorrelated noise around a
-site-level constant (the "average" model) or as an autoregressive model with a
-single year of history [the "naive" model, @hyndman2014]. Predictions from the
-"average" model are centered on the average richness observed during training,
-and the confidence intervals are narrow and constant-width. The "naive" model,
-in contrast, predicts that future observations will be similar to the final
-observed value (e.g., in our hindcasts the value observed in `r
-timeframe$last_train_year`), and the confidence intervals expand rapidly as the
-predictions extend farther into the future. Both models' richness predictions
-are centered on a constant value, so neither model can anticipate any trends in
-richness or any responses to future environmental changes.
+comparison with the more complex models (Figure 2A). The first baseline, called
+the "average" model, treated site-level richness observations as uncorrelated
+noise around a site-level constant:
+
+$$y_t = \mu + \epsilon_t.$$
+
+Predictions from the "average" model are centered on the average richness
+observed during training, and the confidence intervals are narrow and
+constant-width.
+
+The second baseline, called the "naive" model [@hyndman2014], was a simple
+autoregressive process with a single year of history:
+
+$$y_t = y_{t-1} + \epsilon_t$$
+
+In contrast to the "average" model, whose predictions were based on the average
+richness across the whole time series, the "naive" model predicts that future
+observations will be similar to the *final* observed value (e.g., in our
+hindcasts the value observed in `r timeframe$last_train_year`), and the
+confidence intervals expand rapidly as the predictions extend farther into the
+future. Both models' richness predictions are centered on a constant value, so
+neither model can anticipate any trends in richness or any responses to future
+environmental changes.
 

### 2017-12-09 manuscript/draft.Rmd
update details of environmental methods
Saying we did not use the calendar year is an important detail.
Updated months used in the NDVI calcs as per the current code.

--- 2017-12-09 manuscript/draft.Rmd
@@
 quarter, precipitation seasonality, precipitation of the wettest quarter, and
-precipitation of the warmest quarter. Satellite-derived NDVI, a primary
-correlate of richness in BBS data [@hurlbert2002], was obtained from the NDIV3g
-dataset with an 8 km resolution [@pinzon2014] and was available from 1981-2013.
-Average summer (May, June, July) and winter (December, January, Feburary) NDVI
-values were used as predictors. Elevation was from the SRTM 90m elevation
-dataset [@jarvis2008] obtained using the R package raster [@raster].
-Because BBS routes are 40-km transects rather than point counts, we used the
-average value of each environmental variable within a 40 km radius of each BBS
-route's starting point.
+precipitation of the warmest quarter. These variables were calculated for the 12
+months leading up to the annual survey (July-June) as opposed to the calendar
+year. Satellite-derived NDVI, a primary correlate of richness in BBS data
+[@hurlbert2002], was obtained from the NDIV3g dataset with an 8 km resolution
+[@pinzon2014] and was available from 1981-2013. Average summer (April, May, June)
+and winter (December, January, Feburary) NDVI values were used as predictors.
+Elevation was from the SRTM 90m elevation dataset [@jarvis2008] obtained using
+the R package raster [@raster]. Because BBS routes are 40-km transects rather
+than point counts, we used the average value of each environmental variable
+within a 40 km radius of each BBS route's starting point.
 

### 2017-12-09 manuscript/reviewer_reply.Rmd
Reword sentence that used "forecast" three times in 13 words
rough draft of letter

--- 2017-12-09 manuscript/reviewer_reply.Rmd
@@
+Dear XXX,
+We would like to thank you and the two reviewers for the productive feedback on our manuscript.
+We have made minor revisions in clarifying our results, being more descriptive in our data and methods, and including more language on future directions of this research.
+More detailed responses to individual comments are described below.
+
+Regards,
+
 ## Editors Comments

### 2017-12-09 manuscript/reviewer_reply.Rmd
minor formatting

@@
 More minor comments in General Comments section.*
+
 #### Experimental design
@@
 More minor comments in General Comments section.*
+
 #### Validity of the findings
@@
 
-*Line 138: Maybe also cite any of the interfaces you used, like RStan: http://mc-stan.org/users/citations/*
+*Line 138: Maybe also cite any of the interfaces you used, like RStan*: http://mc-stan.org/users/citations/
 

### 2017-12-09 manuscript/reviewer_reply.Rmd
more baseline eqn stuf

--- 2017-12-09 manuscript/reviewer_reply.Rmd
@@
 
+>We now clarify this, using an equation for each baseline model. In particular, we identify
+which parameters are estimated on a site-by-site basis and which ones are constant across
+all sites, as suggested.
+
 *Line 180 – Maybe also mention this is in the forecast package – as you do with gbm below?*
@@
 This isn’t your fault of course, but much harder for me to dig through the repo.*
-
->Thanks!
-We tried to handle this using binder (https://mybinder.org/) to allow anyone to quickly engage with the full environment, but unfortunately the R support isn't quite ready yet.
-Fortunately it is under active development and as soon as it is available we plan to set it up and link to it from the repository to support this use case.
 
@@
 
->We have added a summary of our results and findings.
-
 *57-58: this is not the definition of hindcasting.
@@
 *466: only other mention of future forecasts… not sure why this is included in this paper.*
-
-> Thank you again for pointing out that we were unclear here. The purpose of the future forecasts is now explained on lines [[xx]] and [[xx]].

### 2017-12-09 manuscript/draft.Rmd
more baseline eqn stuf

--- 2017-12-09 manuscript/draft.Rmd
@@
 
-Predictions from the "average" model are centered on the average richness
-observed during training, and the confidence intervals are narrow and
-constant-width.
+Predictions from the "average" model are thus centered on $\mu$, which could
+either be the mean of the raw training richness values, or an output from the
+observer model.  This model's confidence intervals have a constant width that
+depends on the standard deviation of $\epsilon$, which can either be the
+standard deviation of the raw training richness values, or
+$\sigma^\mathrm{residual}$ from the observer model; see supplement).
 
 The second baseline, called the "naive" model [@hyndman2014], was a simple
-autoregressive process with a single year of history:
+autoregressive process with a single year of history, i.e. an ARIMA(0,1,0)
+model:
 
-$$y_t = y_{t-1} + \epsilon_t$$
+$$y_t = y_{t-1} + \epsilon_t,$$
 
-In contrast to the "average" model, whose predictions were based on the average
+where the standard deviation of $\epsilon$ is a free parameter for each site. In
+contrast to the "average" model, whose predictions were based on the average
 richness across the whole time series, the "naive" model predicts that future
 observations will be similar to the *final* observed value (e.g., in our
-hindcasts the value observed in `r timeframe$last_train_year`), and the
-confidence intervals expand rapidly as the predictions extend farther into the
-future. Both models' richness predictions are centered on a constant value, so
-neither model can anticipate any trends in richness or any responses to future
-environmental changes.
+hindcasts the value observed in `r timeframe$last_train_year`). Moreover,
+because the $\epsilon$ values accumulate over time, the confidence intervals
+expand rapidly as the predictions extend farther into the future. Despite these
+differences, both models' richness predictions are centered on a constant value,
+so neither model can anticipate any trends in richness or any responses to
+future environmental changes.
 

### 2017-12-10 manuscript/draft.Rmd
were -> are

@@
 where the standard deviation of $\epsilon$ is a free parameter for each site. In
-contrast to the "average" model, whose predictions were based on the average
+contrast to the "average" model, whose predictions are based on the average
 richness across the whole time series, the "naive" model predicts that future

### 2017-12-10 manuscript/reviewer_reply.Rmd

@@
 *Line 329 – maybe reword, you use forecast 3x in this sentence*
+
+>Thanks for pointing this out. We have reduced this to 1x.
 

### 2017-12-10 manuscript/draft.Rmd

@@
 species richness. The results of this effort are both promising and humbling.
-When comparing forecasts across sites, many different approaches to forecasting
-produce reasonable forecasts (Figure 3). If a site is predicted to have a high
-number of species in the future, relative to other sites, it generally does.
-However, none of the methods evaluated reliably determined how site-level
-richness changes over time (Figure 6), which is generally the stated purpose of
-these forecasts. As a result, baseline models, which did not attempt to
-anticipate changes in richness over time, generally provided the best forecasts
-for future biodiversity. While this study is restricted to breeding birds in
-North America, its results are consistent with a growing literature on the
-limits of ecological forecasting, as discussed below.
+When comparing predictions across sites, many different approaches produce
+reasonable forecasts (Figure 3). If a site is predicted to have a high number of
+species in the future, relative to other sites, it generally does.  However,
+none of the methods evaluated reliably determined how site-level richness
+changes over time (Figure 6), which is generally the stated purpose of these
+forecasts. As a result, baseline models, which did not attempt to anticipate
+changes in richness over time, generally provided the best forecasts for future
+biodiversity. While this study is restricted to breeding birds in North America,
+its results are consistent with a growing literature on the limits of ecological
+forecasting, as discussed below.
 

### 2017-12-11 manuscript/reviewer_reply.Rmd
clarify deviance
input from ethan

--- 2017-12-09 manuscript/reviewer_reply.Rmd
@@
-Dear XXX,
-We would like to thank you and the two reviewers for the productive feedback on our manuscript.
-We have made minor revisions in clarifying our results, being more descriptive in our data and methods, and including more language on future directions of this research.
+We would like to thank the editor and the two reviewers for the productive feedback on our manuscript.
+We have addressed the comments related to our choice of data, clarified the results, added more description to our data and methods, and included more language on future directions of this research.
 More detailed responses to individual comments are described below.
-
-Regards,
 

### 2017-12-11 manuscript/reviewer_reply.Rmd
update "annual data" answer with ethans feedback.

--- 2017-12-10 manuscript/reviewer_reply.Rmd
@@
 
->BBS data are generally viewed as annual data, even though the observations happen on a single day.
-The presence of an individual species is primarily a function of suitable habitat and, for many species, timing of migration.
-Environmental variables with a shorter temporal scale could likely improve models, such as incorporating cool weather days into the observation model, and would be worth pursuing in a different study.
+>Bird data is collected once each year during the breeding season.
+Climate data is available monthly.
+We want to use the chunk of the climate data that is mostly likely to predict the the bird response.
+We used climate values from the 12 months leading up to the bird sampling since this is assumed to influence the habitat and other resources the birds are responding to (We have also clarified this in the methods).
+Other choices are possible, but rarely explored b/c what is most commonly used is long-term averages.
+Future work could explore shorter time scale environmental data such as monthly NDVI trends (instead of just means), migration indicators for migratory species, or more localized weather around each sampling site that could be incorporated into an observer model.
 
@@
 Var(deviance) should incorporate precision – but I’ve only seen this extracted from Bayesian models*
+
+>We were unclear here.
+We had only intended to convey that mean deviance was lowest for predictions that were made with appropriate uncertainty.
+We did not use Var(deviance).
+The new sentence should be clearer on this point.
 

### 2017-12-11 manuscript/reviewer_reply.Rmd
a

--- 2017-12-11 manuscript/reviewer_reply.Rmd
@@
 We have addressed the comments related to our choice of data, clarified the results, added more description to our data and methods, and included more language on future directions of this research.
-More detailed responses to individual comments are described below.
+More detailed responses to individual comments are included below.
 
@@
 
+>We were unclear here.
+We had only intended to convey that mean deviance was lowest for predictions that were made with appropriate uncertainty.
+We did not use Var(deviance).
+The new sentence should be clearer on this point.
+
 *Line 266: I think you guys have done a great job putting together this public repository + made the analysis totally transparent.
@@
 This isn’t your fault of course, but much harder for me to dig through the repo.*
+
+>Thanks!
+We tried to handle this using binder (https://mybinder.org/) to allow anyone to quickly engage with the full environment, but unfortunately the R support isn't quite ready yet.
+Fortunately it is under active development and as soon as it is available we plan to set it up and link to it from the repository to support this use case.
 

### 2017-12-11 manuscript/reviewer_reply.Rmd
ethan feedback

--- 2017-12-11 manuscript/reviewer_reply.Rmd
@@
 Climate data is available monthly.
-We want to use the chunk of the climate data that is mostly likely to predict the the bird response.
+We want to use the chunk of the climate data that is mostly likely to predict the bird response.
 We used climate values from the 12 months leading up to the bird sampling since this is assumed to influence the habitat and other resources the birds are responding to (We have also clarified this in the methods).
-Other choices are possible, but rarely explored b/c what is most commonly used is long-term averages.
+Other choices are possible, but rarely explored b/c what is most commonly used is long-term averages (e.g., 30 year climate means).
 Future work could explore shorter time scale environmental data such as monthly NDVI trends (instead of just means), migration indicators for migratory species, or more localized weather around each sampling site that could be incorporated into an observer model.

### 2017-12-11 manuscript/reviewer_reply.Rmd

@@
+We would like to thank the editor and the two reviewers for the productive feedback on our manuscript.
+We have addressed the comments related to our choice of data, clarified the results, added more description to our data and methods, and included more language on future directions of this research.
+More detailed responses to individual comments are included below.
+
 ## Editors Comments

### 2017-12-11 manuscript/reviewer_reply.Rmd
add line numbers

--- 2017-12-11 manuscript/reviewer_reply.Rmd
@@
-We would like to thank the editor and the two reviewers for the productive feedback on our manuscript.
-We have addressed the comments related to our choice of data, clarified the results, added more description to our data and methods, and included more language on future directions of this research.
-More detailed responses to individual comments are included below.
-
 ## Editors Comments
@@
 More minor comments in General Comments section.*
-
 #### Experimental design
@@
 More minor comments in General Comments section.*
-
 #### Validity of the findings
@@
 
-*Line 138: Maybe also cite any of the interfaces you used, like RStan*: http://mc-stan.org/users/citations/
+*Line 138: Maybe also cite any of the interfaces you used, like RStan: http://mc-stan.org/users/citations/*
 
@@
 
->We agree that this is likely. We make this point on [[lines]]
+>We agree that this is likely. We make this point on lines 423-429 of the revised manuscript.
 
@@
 
->Thank you for pointing out that we were unclear here.  The methods and purpose for the future predictions are now explained in more detail on lines [[XX]].
+>Thank you for pointing out that we were unclear here.  The methods and purpose for the future predictions are now explained in more detail on lines 136-139 and 502-505 of the revised manuscript.
 
@@
 
-> Thank you again for pointing out that we were unclear here. The purpose of the future forecasts is now explained on lines [[xx]] and [[xx]].
+> Thank you again for pointing out that we were unclear here. The purpose of the future forecasts is now explained on lines 136-139 and 502-505.

### 2017-12-11 manuscript/reviewer_reply.Rmd
reviewer comments PDF

@@
+---
+output:
+  pdf_document: default
+geometry: left=2.5cm, right=2.5cm, top=2.25cm, bottom=2.25cm, headheight=12pt, letterpaper
+fontsize: 12pt
+---
+
+\raggedright
+
 We would like to thank the editor and the two reviewers for the productive feedback on our manuscript.
@@
 ## Reviewer 1: Eric Ward
+
 #### Basic reporting
@@
 #### Introduction:
+
 *Line 41: I think this is generally more common in terrestrial ecology – in fisheries these SDM models with more of a theoretical background don’t seem to be used as much*

### 2017-12-11 manuscript/draft.Rmd
Add new retriever reference

@@
 Breeding Bird Survey (BBS) [@pardieck2017] using the Data Retriever Python
-package [@morris2013] and rdataretriever R package [@rdataretriever]. BBS
-observations are three-minute point counts made at 50 fixed locations along a
-40km route. Here we denote each route as a site and summarize richness as the
-total species observed at all 50 locations in each surveyed year. Prior to
-summarizing the data was filtered to exclude all nocturnal, cepuscular, and
-aquatic species [since these species are not well sampled by BBS methods;
-@hurlbert2005], as well as unidentified species, and hybrids. All data from
-surveys that did not meet BBS quality criteria were also excluded.
+package [@morris2013; @senyondo2017] and rdataretriever R package
+[@rdataretriever]. BBS observations are three-minute point counts made at 50
+fixed locations along a 40km route. Here we denote each route as a site and
+summarize richness as the total species observed at all 50 locations in each
+surveyed year. Prior to summarizing the data was filtered to exclude all
+nocturnal, cepuscular, and aquatic species [since these species are not well
+sampled by BBS methods; @hurlbert2005], as well as unidentified species, and
+hybrids. All data from surveys that did not meet BBS quality criteria were also
+excluded.
 
@@
 range of possible outcomes; and 3) deviance (i.e. negative 2 times the Gaussian
-log-likelihood) as an integrative measure of fit incorporating good point
-estimates, precision, and coverage. In addition to evaluating forecast
-performance in general, we evaluated how performance changed as the time horizon
-of forecasting increased by plotting performance metrics against year. Finally,
-we decomposed each model's squared error into two components: the squared error
-associated with site-level means and the squared error associated with annual
-fluctuations in richness within a site. This decomposition describes the extent
-to which each model's error depends on consistent differences among sites versus
-changes in site-level richness from year to year.
+log-likelihood) as an integrative measure of fit that incorporates both accuracy
+and uncertainty. In addition to evaluating forecast performance in general, we
+evaluated how performance changed as the time horizon of forecasting increased
+by plotting performance metrics against year. Finally, we decomposed each
+model's squared error into two components: the squared error associated with
+site-level means and the squared error associated with annual fluctuations in
+richness within a site. This decomposition describes the extent to which each
+model's error depends on consistent differences among sites versus changes in
+site-level richness from year to year.
 

### 2017-12-11 manuscript/draft.Rmd

@@
 the forecast relative to a simple baseline [@jolliffe2003]. We recommend
-following the example of @ye2015 and adopting this approach in future ecological
-forecasting research.
+following the example of @perretti2013 and adopting this approach in future
+ecological forecasting research.
 
@@
 the modeled forecasts are more accurate than the naive assumption that the world
-is static [@jolliffe2003; @ye2015].
+is static [@jolliffe2003; @perretti2013].
 

### 2018-01-05 manuscript/draft.Rmd
Editor comments: pp 1-9

@@
 @diaz2015; @tilman2017]. High-profile studies forecasting large biodiversity
-declines over the coming decades have played a large role in shaping ecologists'
-priorities [as well as those of policymakers; e.g. @ipcc2014], but it is
-inherently difficult to evaluate such long-term predictions before the projected
-biodiversity declines have occurred.
+declines over the coming decades have played a large role in shaping the
+priorities of ecologists and policymakers [e.g. @ipcc2014], but it is inherently
+difficult to evaluate such long-term predictions prior to the predicted decline
+in diversity.
 
@@
 richness [e.g. @calabrese2014]. Alternatively, models that directly relate
-spatial patterns of species richness to environment conditions have been
+spatial patterns of species richness to environment conditions are available
 developed and generally perform equivalently to stacked SDMs [@algar2009;
@@
 surveyed year. Prior to summarizing the data was filtered to exclude all
-nocturnal, cepuscular, and aquatic species [since these species are not well
+nocturnal, crepuscular, and aquatic species [since these species are not well
 sampled by BBS methods; @hurlbert2005], as well as unidentified species, and
@@
 **Future environmental projections.** In addition to the analyses presented
-here, we have also generated and archived long term forecasts from 2014-2050.
-This will allow future researchers to assess the performance of our six models
-on longer time horizons as more years of BBS data become available.
-Precipitation and temperature were forecast using the CMIP5 multi-model ensemble
-dataset [@brekke2013]. 37 downscaled model runs [@brekke2013, see Table S1]
-using the RCP6.0 scenario were averaged together to create a single ensemble
-used to calculate the bioclimatic variables for North America. For NDVI, we used
-the per-site average values from 2000-2013 as a simple forecast. For observer
-effects (see below), each site was set to have zero observer bias. The
-predictions have been archived at [@zenodoPredict].
+here, we also generated and archived long term forecasts from 2014-2050. This
+will allow future researchers to assess the performance of our six models on
+longer time horizons as more years of BBS data become available. Precipitation
+and temperature were forecast using the CMIP5 multi-model ensemble dataset
+[@brekke2013]. We averaged 37 downscaled model runs [@brekke2013, see Table S1]
+using the RCP6.0 scenario to create a single ensemble used to calculate the
+bioclimatic variables for North America. For NDVI, we used the per-site average
+values from 2000-2013 as a simple forecast. For observer effects (see below),
+each site was set to zero observer bias. The predictions were archived at
+[@zenodoPredict].
 
@@
 
-This research was supported by the Gordon and Betty Moore Foundation's
-Data-Driven Discovery Initiative through Grant GBMF4563 to E.P. White. We thank
-the developers and providers of the data and software that made this research
-possible including: the PRISM Climate Group at Oregon State University, the
-staff at USGS and volunteer citizen scientists associated with the North
+We thank the developers and providers of the data and software that made this
+research possible including: the PRISM Climate Group at Oregon State University,
+the staff at USGS and volunteer citizen scientists associated with the North
 American Breeding Bird Survey, NASA, the World Climate Research Programme's

### 2018-01-05 manuscript/draft.Rmd

@@
 either be the mean of the raw training richness values, or an output from the
-observer model.  This model's confidence intervals have a constant width that
-depends on the standard deviation of $\epsilon$, which can either be the
+observer model. The confidence iterval width in this model is constant and a 
+consequence of the standard deviation $\epsilon$, which can either be the
 standard deviation of the raw training richness values, or
@@
 longer-term dependencies in the underlying process), a moving average component
-(where the noise can have serial autocorrelation) and an
+(where the noise may be serially autocorrelation) and an
 integration/differencing component (so that the analysis could be performed on
@@
 implicitly assumes that two sites with identical Bioclim, elevation, and NDVI
-values should have identical richness distributions. As above, we included
+values will have identical richness distributions. As above, we included
 observer effects and the associated uncertainty by running these models 500
@@
 falling inside its 95% confidence intervals. This stacked SDM's narrow
-predictive distribution caused it to have notably higher deviance (Figure 5B)
+predictive distribution caused notably greater deviance (Figure 5B)
 than the next-worst model, even though its point estimates were not unusually
@@
 develop and follow best practices for conducting and evaluating ecological
-forecasts [@clark2001]. We have used a number of these practices (Box 1) in a
+forecasts [@clark2001]. We used a number of these practices (Box 1) in a
 single study that builds and evaluates forecasts of biodiversity in the form of
@@
 When comparing predictions across sites, many different approaches produce
-reasonable forecasts (Figure 3). If a site is predicted to have a high number of
-species in the future, relative to other sites, it generally does.  However,
+reasonable forecasts (Figure 3). If a site is predicted to have high species
+richness in the future, relative to other sites, it generally does.  However,
 none of the methods evaluated reliably determined how site-level richness
@@
 biodiversity change should be viewed with skepticism unless the underlying
-models have been validated temporally, via hindcasting and comparison with
+models are temporally validated, via hindcasting and comparison with
 simple baselines. Since site-level richness is relatively stable, spatial
-validation is not enough: a model can have high accuracy across spatial
-gradients without being able to predict changes over time. This gap between
-spatial and temporal accuracy is known to be important for species-level
+validation is not enough: a model be accurate across spatial gradients while
+being unable to predict changes over time. This gap between
+spatial and temporal accuracy is important for species-level
 predictions [@rapacciuolo2012; @oedekoven2017]; our results indicate that it is
-substantial for higher-level patterns like richness as well. SDMs' poor temporal
-predictions are particularly sobering, as these models have been one of the main
-foundations for estimates of the predicted loss of biodiversity to climate
-change over the past decade or so [@thomas2004; @thuiller2011; @urban2015]. Our
+substantial for higher-level patterns like richness as well. The poor quality 
+temporal predictions of SDM's are particularly sobering because of their importance
+over the last decade or so for predicting biodiversity loss as a response to 
+climate change [@thomas2004; @thuiller2011; @urban2015]. Our
 results also highlight the importance of comparing multiple modeling approaches
@@
 results to simple baselines to avoid over-interpreting the information present
-in these forecasts [Box 1]. Disciplines that have more mature forecasting
+in these forecasts [Box 1]. Disciplines with more mature forecasting
 cultures often do this by reporting "forecast skill", i.e., the improvement in
@@
 
-We have only evaluated annual forecasts up to a decade into the future, but
+We only evaluated annual forecasts up to a decade into the future, but
 forecasts are often made with a lead time of 50 years or more. These long-term
@@
 datasets, but are important for understanding changes in biodiversity at some of
-the lead times relevant for conservation and management. Two studies have
+the lead times relevant for conservation and management. Two studies
 assessed models of species richness at longer lead times [@algar2009;
@@
 been observed. However, in real forecasts the environmental conditions
-themselves have to be predicted, and environmental forecasts will also have
+themselves must be predicted, and environmental forecasts will also have
 uncertainty and bias. Ultimately, ecological forecasts that use environmental
@@
 determine how far into the future we can successfully forecast [@dietze2016;
-@tredennick2016]. We have openly archived forecasts from all six models through
+@tredennick2016]. We openly archived forecasts from all six models through
 the year 2050 [@zenodoPredict], so that we and others can assess how well they
